### 2015-09-20 control.tex
split into 3 papers

\documentclass[letterpaper,12pt,fullpage]{article}

\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{cite}
\usepackage{graphicx}
% \usepackage[dvips]{graphicx}
% \usepackage{epsfig} % for postscript graphics files
  % \graphicspath{{../eps/}}
% \DeclareGraphicsExtensions{.eps}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[cmex10]{amsmath}
%\usepackage{array}
%\usepackage{mdwmath}
%\usepackage{mdwtab}
%\usepackage{eqparbox}
\usepackage[tight,footnotesize]{subfigure}
%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
%\usepackage{fixltx2e}
%\usepackage{stfloats}
\usepackage{hyperref}

% correct bad hyphenation here
%\hyphenation{op-tical net-works semi-conduc-tor}

\input latex-commands

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}

\newcommand{\shrinkfig}{\def\baselinestretch{1.0}\small} % 0.9 okay
\newcommand{\shrink}{\def\baselinestretch{1.1}\small} % 0.97 % 0.95 okay

\begin{document}

\title{A Survey of Possible Exoskeleton Control Architectures and
Algorithms\\
(Draft 2.0)}

\author{Alex Ansari, Christopher G. Atkeson, Howie Choset, and Matthew Travers\\
Carnegie Mellon University}

\maketitle

\begin{abstract}
Abstract to be written.
\end{abstract}

\section{Executive Summary}

Executive Summary to be written.

\section{Scope: What is this paper about?}

This paper surveys possible exoskeleton control architectures and
algorithms.
One goal is to address the question:
{\it How do exoskeleton control approaches compare with 
approaches to humanoid robot control?}
A companion paper surveys implemented exoskeleton control architectures
and algorithms~\cite{}.

The focus is on exoskeleton control that allows a
highly trained and top percentile athletic 
operator to carry a payload that weighs approximately the same amount
as the operator. We envisage these types of exoskeletons to be useful
in carrying protective and safety equipment for SWAT teams, police,
firefighters, and soldiers.

We expect each exoskeleton controller
to be used by and optimized for a single operator.
A substantial investment in capturing the operators normal behavior,
operator training and learning, and controller customization can be made.

We focus this survey on exoskeleton control for lower body tasks (standing, walking,
running, jumping, kicking, dodging, ...).
We do not survey exoskeleton control for human arms or manipulation.

We treat the torso and helmet of the exoskeleton as the payload,
and focus on a lower body exoskeleton to support any payloads that
are on the torso or head.

%\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
A later white paper will discuss
possible combinations of response to user and exoskeleton
autonomy including balance control; response to trips, slips,
stumbles, and fumbles; task assistance and guidance (guide operator to
doorknob, button, or light switch); superhuman response to external
perturbations such as projectiles and explosions; and autonomous
execution (operator/exoskeleton symbiosis with multi-tasking)?

\subsection{Let's keep it simple}

Terms like impedance and admittance control are used, but are often confusing,
as in controller design one can choose from several possible
inputs into the exoskeleton (exoskeleton positions,
velocities, accelerations, operator-exo contact forces and
exo-world contact forces), and choose from several possible exoskeleton
outputs: exoskeleton actuator forces and torques, exoskeleton
motion (position, velocity, and acceleration), as well as operator-exo
contact forces or exo-world contact forces.

A useful background paper: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}

Variants of nonlinear feedback control such as feedback linearization or
sliding mode control are largely ignored in this paper.
Once the decision to use feedback control
based on a set of observable quantities and with particular outputs is made,
one can try out the various linear and nonlinear feedback control paradigms
to see what works well.

Variants of function approximation methods such as lookup tables, fuzzy logic,
sigmoidal neural networks, radial basis functions, and locally weighted regression
are also largely ignored. Once the decision to use a function approximator
and what the inputs and outputs are has been made, one can try out the various
approaches to see what works well.

Same for optimization methods.

Same for constraint enforcement (such as avoiding self-collisions)
in either optimization or feedback control (barrier
Lyapunov functions (BLF)~\cite{IEEE06911561}).

Stability proofs of any of these methods should be viewed skeptically due
to the unmodeled operator-exoskeleton and exoskeleton-world contact dynamics, 
actuator unmodeled dynamics, joint play and exoskeleton
structural deformation, and controller
time delays which are typically ignored, especially in proofs involving
passivity arguments or Lyapunov functions.
We ignore stability proofs as well.

\section{What are some design philosophy alternatives?}

\subsection{``Invisible'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally
and exerts little or no force on the operator?

This could be achieved through active control using feedback of the
operator-exoskeleton forces. Additional gravity, friction, actuator, and
inertial compensation (inverse dynamics) can improve performance.

\subsection{``Natural'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally?
The operator can feel the exoskeleton, but can overpower it when necessary.

Bypass valves and clutches could allow the operator to physically ``take over''
and push the relatively lightweight leg mounted exoskeleton around.

``Passive'' degrees of freedom can have physical or ``virtual'' springs
applied to assist in support (all abduct/adduct DOF, rotational DOF).
These aids can be clutched out or turned off when necessary.
``Virtual'' springs (springs implemented actively through exoskeleton
actuators) can be modulated.

\subsection{``Symbiotic'' exoskeleton}

Can we build an exoskeleton that allows the operator/exoskeleton team
to be effective, but not necessarily behave naturally? 
The operator needs to learn to ``fly'' the suit,
just as operators learn to operate parachutes,
wing suits, diving equipment, high and low temperature protective
suits, firefighting equipment, and high altitude flight
and parachute suits.

The exoskeleton can adapt its behavior to the task.
For example, for jumping, the exoskeleton can operate like a pogo 
stick \url{https://www.youtube.com/watch?v=lbp41vWP4o4}, 
and for running it could act like jumping 
stilts \url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},

\section{Effects of underlying exoskeleton actuator technology}

One basic distinction is whether the actuators are thought of as position
or force sources. Often this is revealed by the design of the ``low level''
controller. Note that this is often somewhat confusing. Electric motors
are torque sources (low impedance), but when a high gear ratio transmission
is added (such as a harmonic drive) the whole system becomes high impedance
and thus is best thought of as a position (or velocity) source. Therefore,
position control is performed by the low level controller. Hydraulic actuators
are high impedance, but when a force or load sensor is added (or piston differential
oil pressure is used as a force sensor), the whole system becomes low impedance
and is best thought of as a force source. Force (or joint torque) control is
performed by the low level controller. Force or torque can be added to an
electromechanical drive train (electric motor plus gears or motor plus ballscrew)
to make it lower impedance and more like a force source as well.

\section{Recommended Control Architecture}

We recommend the following hierarchical control architecture~\cite{IEEE06907051}:

{\bf Low Level Control:} Present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing active stiffness
or damping.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation are implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
or distibuted limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
Often, this controller is provided by the manufacturer of the robot
or exoskeleton and runs on a separate computer system from the
``user's'' or task/application controller~\cite{Sarcos,Atlas}.

{\bf ``Phoneme'' Control:} Mid-level controllers
for relatively constant phases of behavior.
Often involves position or velocity targets (goto or
stay-at X), trajectory references, ...
Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
Finer resolution may be needed. For example a walking stance period may
be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
The first and second halves of stance are divided by when the center of mass
passes over the ankle.
Local inverse kinematics (finding small joint motions for small target changes)
is performed at this level.
This level often provides the behavioral primitives or ``verbs'' for robot programming.

{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
units.
Often involves finite state machines, timers, if-then condition testing, gain
scheduling, ...
Examples: walk-to-location( x=7.8m, y=2.3m ), run-to-location( speed=3.7m/s ),
horizontal-jump( height= 2m, length= 5m ).
Global inverse kinematics (selection of solution branch) is performed at this level.
This level often determines the control flow of a behavior.

{\bf Highest Level Control} aka {\bf Behavior Selection:} 
Answer the question: ``What do I do next?''
Select behaviors and behavioral parameters (targets,
speeds, durations).
Often involves finite state machines. timers, if-then condition testing, ...
This level often involves a great deal of human operator online interaction,
or is completely done by the operator.

{\bf Error Handling:} Errors must be detected and handled by parameter (target)
adjustment and/or behavior switching
at all of these levels.

This control architecture is intuitive, abstracts the hardware 
and lower level control in a straightforward way, is easy to program, and is
commonly used. Variants of this architecture were used by almost all teams
in the DARPA Robotics Challenge~\cite{}.

Closely related control architectures combine Phoneme and Word control,
or combine Phoneme, Word, and Behavior Selection.
This is not a major design change.

An alternative behavioral architecture might not have explicit levels but instead
has a ``soup'' of behaviors that compete or cooperate for the ability to
control the robot or exoskeleton. Subsumption architecture and other behavior-based
architectures are closer to the ``soup'' approach and to some extent
use inhibition and excitation
among behaviors to select active behaviors.

Another alternative behavioral architecture does not have explicit separate behaviors,
but one or a small number of functions or policies. 
We are seeing more of these architectures
due to the success and popularity of function approximation such as
deep learning.

\section{Exoskeleton dynamics}

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskelton parts which they are attached to in terms
of link inertias, location of center of mass, and moments of inertia,
and just think of the whole system as one robot.
We can also separate out the operator dynamics if this is more convenient
(we separate them out and ignore them in what follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles and position of the ``root'' of
the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques.
$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
that correspond to root velocities, since those dimensions are not actuated.

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
and so is the world ($\vtau_w$) (very unrealistic assumption). In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

Note that we are ignoring the operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. We will add these in later (or maybe
we won't ???).

We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector.

\section{Passive control with no active control}

One control option is to use only passive controls, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
An operator would rely on mechanical backdrivability to drive
exokeleton to do tasks.
In this case, 
$\vtau$ is generated by passive devices such as springs or dampers.

\section{Open loop control}

Another control option is open loop control, in which actuator commands are
constant, or a function of time or phase of a behavior or task
(at least on the small time scale).
Commands or behaviors may be selected at a relatively coarse time interval
by the operator or by the exoskeleton controller.
This approach typically 
uses position and velocity control in either joint or
Cartesian coordinates
with either pre-generated references, selected targets,
or online generated trajectories to move to targets or perform tasks.
The operator relies on mechanical backdrivability 
to correct exoskeleton behavior.
In this case, it is common to generate the exoskeleton actuator commands using
a time or phase index:
$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
dependent feedforward torque vector.

Recognizing/estimating phase goes here.
Phase Triggered References could go here.
\begin{verbatim}
H.Kawamoto and Y.Sankai, “Power Assist Method Based on Phase
Sequence Driven by Interaction between Human and Robot Suit,”
Proceedings of the IEEE International Workshop on Robot and Human
Interactive Communication, pp. 491–496, 2004.
\end{verbatim}

Motion capture, motion libraries goes here.

Often finite state machines whose states correspond to left stance, double support 1,
right stance, and double support 2 are used to select behaviors. Gait transitions
can be triggered by foot touchdown and liftoff events, or
displacement of the Center of Mass (CoM).

Central pattern generator

Iterative learning control to acquire $\vtau_{ff}(t)$ for desired trajectories.

\section{Active gravity compensation}

Active gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia).
\begin{equation}
\vtau = \hat{\mG}(\hat{\vq})
\end{equation}
where $\vtau$ are the exoskeleton joint torques, 
and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.

Sarcos Primus Humanoid (ATR) video of active gravity compenstation applied
to a humanoid robot:\\
\url{https://www.youtube.com/watch?v=KNxxLm4sPys}

A simple approximation for a lower body exoskeleton with a heavy payload at the
torso is:
\begin{equation}
\vtau \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
\end{equation}
where $\hat{\mJ}$ is the estimated Jacobian matrix,
$\hat{m}$ is the estimated
weight of the exoskeleton (which could include the operator),
and $\hat{\vg}$ is the gravity vector (it is estimated due to possible orientation
errors in the controller's estimate of vertical).

\section{Friction compensation}

Friction can also be compensated.
\begin{equation}
\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

\section{Online inverse dynamics control}

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

These force/torque estimates rely on estimated dynamic models and their parameters.
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

Simplifying equation~\ref{eq:cancel} by 
assuming the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
The operator sees the true inertia of the exoskeleton, but no gravitational
loads and no other dynamics (Coriolis, centripetal, or frictional).

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
the state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}

Online inverse dynamics control is a generalization of 
active gravity compensation to include
inertial, Coriolis, and centripetal forces (forces due to acceleration and
velocity), and potentially frictional forces.
Online inverse dynamics can be added to open loop behavior execution
and many other types of control such as operator force
feedback control, impedance control, and admittance control.
Computed torque control and feedback linearization are forms of online
inverse dynamics.

A key question is where does desired acceleration come from?
\begin{verbatim}
  Operator force control.
  Receding Horizon Control of optimal trajectory of 
    idealized LIPM+flywheel model (Atkeson)
  intent prediction (see above)
  neuromuscular model (Geyer, Wang, Van de Panne)
  hybrid: Optimal control + neuromuscular model (Atkeson NSF Proposal)
    - NMM provides bias through changes to optimization criterion
    - NMM provides hard constraints
    - NMM replaces components of model-based optimization scheme
\end{verbatim}

\section{Operator-Exoskeleton Impedance Control}

What is impedance?
\url{https://en.wikipedia.org/wiki/Mechanical_impedance}
Stiffness, damping, and mass are components of an impedance, as
they map kinematic variables (position, velocity, and acceleration)
to forces and torques.

The goal here is to make the exoskeleton imitate a desired linear dynamic system
from the point of view of the operator.
A future white paper will discuss how to make the exoskeleton 
imitate a desired linear dynamic system 
from the point of view of external perturbations.
In the case where there is no force sensing between the operator and the
exoskeleton, the exoskeleton controller maps from exoskeleton configuration
and velocity to desired joint/actuator/synergy forces.
In this case the actuators need to generate forces and torques rather than
positions or angles, or linear or angular velocities.

Here we make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator.
We can also make the exoskeleton appear as a locally linear impedance to the
operator in some other coordinate system, such as Cartesian coordinates.
A different goal is to make the exoskeleton appear as a locally linear impedance to
the outside world in some coordinate system.

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
\end{equation}
and the operator sees the desirecd impedance.

Although impedance control is extensively utilized in rehabilitation
robotics [18], limited number of studies has been focused on
this method for power augmentation [19, 20].
\begin{verbatim}
[18] R. Riener, L. Lunenburger, S. Jezernik, M. Anderschitz, G. Colombo,
and V. Dietz, "Patient-cooperative strategies for robot-aided treadmill
training: first experimental results," Neural Systems and Rehabilitation
Engineering, IEEE Transactions on, vol. 13, pp. 380-394, 2005.
[19] B.-K. Lee, H.-D. Lee, J.-y. Lee, K. Shin, J.-S. Han, and C.-S. Han,
"Development of dynamic model-based controller for upper limb
exoskeleton robot," in Robotics and Automation (ICRA), 2012 IEEE
International Conference on, 2012, pp. 3173-3178.
[20] W. Yu, J. Rosen, and X. Li, "PID admittance control for an upper limb
exoskeleton," in American Control Conference (ACC), 2011, 2011, pp.
1124-1129.
\end{verbatim}

\section{Force feedback and get out of the way control (inverse dynamics version)}

The goal here is to use force sensing between the operator and the
exoskeleton to ultimately generate exoskeleton velocities or angular velocities,
imitating a desired linear dynamic system.
This type of control is often referred to as admittance control as the
exoskeleton maps contact forces into joint motion.
Inverse dynamics control
can be used to improve this type of control performance.

What is admittance?
\url{https://en.wikipedia.org/wiki/Admittance}
Compliance, inverse damping, inverse mass are components of an admittance,
as they map force and torques to kinematic variables (position, velocity, and
acceleration).

We can implement force control 
if measurements of the contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force as possible 
assuming the actuators are torque sources.
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vf} \mJ^{\tr} ( \hat{\vf} - \vf_d )
\end{equation}
$\mK_{\vf}$ is a gain matrix, $\mJ$ is the appropriate Jacobian matrix for
force $\vf$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

Note that impedance control can also be implemented using force control
by setting:
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
in addition to directly including the stiffness and damping terms in
equation~\ref{eq:impedance}.

\section{Mapping External Forces To Operator Forces Using Force Control}

So far we have ignored or canceled the effect of 
the exoskeleton contact with the outside world.
A different control objective could be to map the external forces to 
desired operator forces:
\begin{equation}
{\vf_o}_d = F_{ow} ( \vf_w )
\end{equation}
and cancel the external forces not transferred in the inverse dynamics.
Everything in the above equation is estimated, so we are dropping the hats 
$\hat{\mbox{}}$.

For example, we could want the operator to feel a scaled (scale factor $\alpha$) 
version of the external
forces, as if they were contacting the operator directly.
\begin{equation}
{\vf_o}_d = \alpha J_o^{-\tr}(\vq) J_w^{\tr}(\vq) \vf_w
\end{equation}
and we scale the $\mJ_w^{\tr}(\vq) \vf_w$ term in the inverse dynamics
by $(1-\alpha)$
We are assuming we have any necessary force sensing between the exoskeleton and
the world.

If the operator applies joint torques, this simplifies to
\begin{equation}
{\vtau_o}_d = \alpha J_w^{\tr}(\vq) ( \vf_w )
\end{equation}
and the inversion of a Jacobian matrix is unnecessary.

This would provide a way of enabling the operator to perceive and control
external contact forces.

\section{Autonomous exoskeleton-world control}

\subsection{Autonomous exoskeleton-world position/velocity control}

A control objective could be to have the exoskeleton to move autonomously.

To be written.

\subsection{Autonomous exoskeleton-world force control}

A control objective could be to have the exoskeleton to apply a desired force.

To be written.

\subsection{Autonomous exoskeleton-world impedance control}

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.

To be written.

\section{Using Online Optimization: Quadratic Programming}

The dominant paradigm in force controlled (typically
hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
such as torque and center of pressure limits using online quadratic programming
to solve the inverse dynamics equations. This approach can also be used
to implement and trade off among the many control objectives described so far.

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
torques $\vtau$, and contact forces $\vf$.
For robots the discrepancy between actual and desired accelerations, joint
torques, and contact forces are all minimized:
\begin{equation}
C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
+ \vf^{\tr} \mW_{\vf} \vf
\end{equation}
The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
Torque limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
Variants of this optimization
were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.

This approach can be used to achieve desired joint or synergy (such as center
of mass) accelerations, center of mass dynamic behavior, center of pressure location,
weight distribution, maintenance of a desired body part position, orientation,
velocity, or trajectory. and joint/actuator position and velocity limits (indirectly).

In the case of exoskeletons, part of the optimization criterion could be to
minimize contact forces on the operator:
\begin{equation}
C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
\end{equation}
Operator force feedback and impedance control could be implemented by specifying a
desired operator force ${\vf_o}_d$:
\begin{equation}
C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
\end{equation}

To implement operator impedance control we set the desired operator contact force to:
\begin{equation}
{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
\end{equation}
Assuming that the operator directly applies exoskeleton joint torques,
\begin{equation}
C_e(\vx) = C_r(\vx) 
+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
\end{equation}
and
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}

\section{Operator-Exoskeleton Force Control using position controlled actuation}

At this point our dynamics formulation is no longer appropriate, since we
are assuming the actuators are now position rather than torque sources.

\subsection{Get out of the way control with position sources}

The exoskeleton moves as to create as little force as possible between the
operator and the exoskeleton. The actuators in this case may be position sources or
force sources with high servo gains,
but ultimately operator-exoskeleton forces are mapped
to exoskeleton velocities or angular velocities.

\begin{equation}
\dot{\vq}_d = \mJ^{-1} (\mK_1 \vf)
\label{eq:gootwcps}
\end{equation}
where $\dot{\vq}_d$ are the commanded exoskeleton joint velocities,
$\mJ$ is an appropriate Jacobian matrix, $\mK_1$ is a gain matrix,
and $\vf$ is the force vector to be controlled~\cite{IEEE06990981}.

Inverting a Jacobian matrix is problematic when the matrix is
nearly singular.
Using a fixed or gain scheduled gain matrix may make more sense.
\begin{equation}
\dot{\vq}_d = \mK_3 \vf
\end{equation}
Simplifying equation~\ref{eq:gootwcps}
by assuming the forces to be controlled
are directly applied to the exoskeleton joints, and thus the Jacobian matrix
is the identity matrix, gives us a similar equation:
\begin{equation}
\dot{\vq}_d = \mK_1 \vtau_o
\end{equation}
and we avoid inverting a Jacobian and problems with singularities.

\subsection{Operator force feedback with position sources}

We introduce a desired force to allow for more complex force control.
\begin{equation}
\dot{\vq}_d = \mJ^{-1} (\mK_1 (\vf - \vf_d))
\end{equation}

The exoskeleton could move as to provide a scaled version of exoskeleton-world
contact forces, or some other (usually simple) mapping. 
\begin{equation}
\vf_d = \alpha \mJ_{ow} \vf_w
\end{equation}

\section{Specific Proposed and Implmented Exoskeleton Control Schemes}

\subsection{Sensitivity Amplification Control}

Sensitivity Amplification Control (SAC) used to control BLEEX
is a dynamic cancellation technique (a.k.a.
inverse dynamics, computed torque, or feedback linearization) that also tries
to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
using acceleration feedback. It is not clearly stated but it appears that the
acceleration is the result of double differentiating position, as there do not
appear to be velocity sensors on BLEEX. Low pass filtering is applied to reduce
the high frequency noise amplified by the double differentiation process.

\subsection{Integral admittance control}

{\it [Exoskeletons] have the implicit property
of causing a virtual modification of the dynamic response of
the human limb. We use this property of the exoskeletons
action to formulate a unified control design framework called
Integral Admittance [torque to angle] Shaping, which designs exoskeleton con-
trollers capable of producing the desired dynamic response
for the assisted limb. In this framework, a virtual increase
in the admittance of the limb is produced by coupling it
to an exoskeleton that exhibits active behavior. Specifically,
our framework shapes the magnitude profile of the integral
admittance (i.e. torque-to-angle relationship) of the coupled
human-exoskeleton system, such that the desired assistance is
achieved. This framework also ensures that the coupled stability
and passivity are guaranteed.}~\cite{Nagarajan_etal_2015}

{\it ... the impedance of the coupled human-exoskeleton
system needs to be reduced below that of the unassisted
human limb. This implies that the exoskeleton needs to
cancel its own impedance first and then compensate for
at least a part of the human limb’s impedance. Therefore,
the desired exoskeleton behavior must be that of a negative
impedance ... Consequently,
the feedback gains will all be
positive. In other words, the exoskeleton controller uses
positive feedback, and hence the exoskeleton exhibits active
behavior, which is capable of performing net positive work
on the limb. ...
However, positive feedback naturally raises the question of
stability, and so we now explain how coupled stability can be
achieved. Although the exoskeleton exhibits active behavior,
which can be potentially destabilizing, the controller can be
designed such that the coupled human-exoskeleton system
is stable and passive. ... 
}~\cite{Nagarajan_etal_2015}

{\it
... Inertia compensation is more complex ...
It can be shown that using only positive acceleration
feedback ..., the gain margin of the coupled system
reduces to the moment of inertia of the exoskeleton, which
implies that the exoskeleton controller ... can at the most
compensate for the exoskeleton’s own moment of inertia
before going unstable. This implies that the moment of
inertia of the coupled human-exoskeleton system cannot be
reduced below that of the unassisted human limb, without
compromising coupled stability. However, using low-pass
filtered acceleration feedback, it can be shown that inertia 
reduction can be achieved. ...
This work uses filtered acceleration
feedback with a second-order low-pass Butterworth filter. ...
}~\cite{Nagarajan_etal_2015}

This is a form of frequency domain virtual model control.

\begin{verbatim}
Kalman filter:
P. Canet, “Kalman filter estimation of angular velocity and accelera-
tion: On-line implementation,” McGill University, Montr ́eal, Canada,
Tech. Rep. TR-CIM-94-15, Nov. 1994.
\end{verbatim}

\subsection{Dual Control Appproach}

This approach implements a passive swing leg in addition to Sensitivity 
Amplification Control for the stance leg.

{\it The robot utilized the dual-mode control scheme, which is
comprised of the active control for the stance phase and the
passive control (using bypass valves) for the swing phase, to achieve high walking
speed in the swing phase while supporting heavy loads in
the stance phase. To reduce the sudden change of the torque
command at the transition from the swing phase to the stance
phase, a smoothing method is adopted. We also implemented
a pre-transition method to take a foot off quickly for fast
walking by predicting the change from the swing to the stance
in advance.}~\cite{IEEE07222598}

Bypass valves implement what Sarcos calls ``dangle''.

Stance: virtual joint torque control method Kazerooni et al.(2005)
When contact location between the wearer and the
exoskeleton is not fixed and difficult to estimate, this method
has been shown to be an effective method to generate the
locomotion for an exoskeleton robot.[3],[4],[21]
\begin{verbatim}
H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
Conf. Robot. Autom., pp. 4364-4371, 2005.
Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
for Human Performance Amplification,” University of California,2003.
582
Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
Control,” Journal of computers, vol.9, No.1,202-208. 2014.
\end{verbatim}

To reduce
sudden changes(command jump) at the phase transitions, a
smoothing method [5],[19] is introduced and a pre-transition
method is used to solve the swing delay due to the internal
pressure(approx 5 bar).
\begin{verbatim}
H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
Robotics Research, pp. 561-573, 2006.
oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
Based Control for a Rotary Series Elastic Actuator Assisting the knee
\end{verbatim}

Transition Control

1) Smoothing Method: During transitions of the gait
phase, discontinuity of the control command torque are
occurred by the different condition for the fixed coordinate
which is on the backpack at the swing phase or the foot at
the stance phase. In this paper, to reduce this sudden change
due to gait phase changes, a smoothing method is proposed
as shown in Fig. 7. An exponential function is considered
as the weighting function for smoothing as shown in (6).
In this case, the weighting is small at the initial stage but
it exponentially converges to one for supporting the load
quickly.

2) Pre-transition Method: The pre-transition method is
that the passive mode is executed in the pre-swing phase
prior to toe off. This dramatically reduces the moving

\subsection{Ground Reaction Force Control}

{\it
The ground reaction
forces (GRF) magnitude and direction is used to command the
actuators. In some researches the GRF sensors are used
together with other sensors in the control architectures [17],
while in other exoskeleton the control system is based on the
GRF merely [RoboKnee, Honda].
}
\begin{verbatim}
17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
"Intention-based walking support for paraplegia patients with Robot
Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
\end{verbatim}

\subsection{Virtual model control}

The goal here is to make the exoskeleton imitate a desired (usually nonlinear)
dynamic system,
For running it could be a pogo stick or trampoline, for example.

There are several possible versions of this type of control:
\begin{enumerate}
\item
Map from operator-exoskeleton contact forces and system state
to exoskeleton actuator or internal forces.
\item
Map from operator-exoskeleton contact forces and system state
to exoskeleton acceleration.
\item
Map from operator-exoskeleton contact forces and system state
to exoskeleton-world contact forces.
\item
Map from exoskeleton-world contact forces and system state to exoskeleton actuator or internal forces.
\item
Map from exoskeleton-world contact forces and system state to exoskeleton acceleration.
\item
Some combination of the above.
\end{enumerate}

\begin{verbatim}
virtual model approach: hopper, compass gait (J. Pratt)
estimate trajectory (jumping) vs. program behavior (trampoline, hopper,
   compass gait)
- J' control + behavior
- Inverse dynamics + behavior
\end{verbatim}

\subsection{Task Specific Control}

Task specific control can involve switching low level control modes, or abstracting
the behavior of the exoskeleton with a hierarchy:
\begin{enumerate}
\item
Task specific control: Generate desired motions and contact forces between
the operator and the exoskeleton, and the exoskeleton and the world.
\item
Sometimes there are intermediate levels of control.
\item
Exoskeleton control: Control the exoskeleton to generate the desired motions and
desired contact forces, with the desired impedance or admittance.
\end{enumerate}

\section{State Estimation}

We are surprised that acceleration feedback based on double differentiation
works well on real systems (even low pass filtered acceleration
estimated by a Kalman filter).
Impacts, shock waves, and the fact that neither
the operator's body parts or the exoskeletons parts are rigid bodies and the
joints are not well defined suggest this approach would be fragile to modeling
error and unmodeled dynamics.
It is not possible to distribute MEMs IMUs throughout the exoskeleton to
improve both velocity and 
acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
Sensitivity Amplification Control.

Role of phase estimation, behavioral clocks, and phase reset events
such as impacts.

\section{Discussion}

Discussion to be written.

\section{Conclusions and Recommendations}

Conclusions and Recommendations to be written.

\bibliographystyle{plain}
\bibliography{exo}

\end{document}

### 2015-09-20 control.tex
more updates

@@
 Algorithms\\
-(Draft 2.0)}
+(Draft 2.1)}
 
@@
 
-\section{Autonomous exoskeleton-world control}
-
-\subsection{Autonomous exoskeleton-world position/velocity control}
-
-A control objective could be to have the exoskeleton to move autonomously.
+\section{Autonomous exoskeleton control}
+
+\subsection{Autonomous exoskeleton position/velocity control}
+
+A control objective could be to have the exoskeleton to move autonomously,
+which is useful if the operator would like to rest, is attending to some other
+task, or is disabled.
+
+This is usually done by using stored trajectories or trajectories generated
+online that specify desired position $\vq_d$, 
+desired velocity $\dot{\vq}_d$, and
+desired acceleration $\ddot{\vq}_d$.
+The feedforward method is to compute the needed torques (potentially offline)
+using inverse
+dynamics (including the operator link inertias, link centers of mass, and link
+moments of inertia in the dynamic model) based on only the desired postion, 
+desired velocity, and desired acceleration. Trajectory tracking errors are compensated
+for by feedback control designed using some other method. In most robotics
+fixed gain independent joint linear control is used.
+This method is safest in the sense that the feedback controller can be tested
+and verified independently of the desired trajectories and dynamic models.
+
+Computed torque approaches use online inverse dynamics computed from the estimated
+position, estimated velocity, and desired acceleration.
+In this case the feedback control to compensate for trajectory tracking errors can
+be independent of the inverse dynamics and its modeling errors, or the desired
+acceleration can be modified with feedback terms.
+\begin{equation}
+\ddot{\vq}_d^* = \ddot{\vq}_{d} + \mK_{\vq}^{\ddot{\vq}} ( \hat{\vq} - \vq_d ) 
++ \mK_{\dot{\vq}}^{\ddot{\vq}} ( \hat{\dot{\vq}} - \dot{\vq_d} )
+\end{equation}
+The problem with putting the inverse dynamics in the feedback loop is that the
+effective gains of this control vary widely with the configuration $\vq$.
+Unmodeled dynamics (that also vary with $\vq$) limit the size of the gain matrices
+$\mK_{\vq}^{\ddot{\vq}}$ and $\mK_{\dot{\vq}}^{\ddot{\vq}}$. Because so many quantities
+are varying quite a bit, constant gain matrices may be quite low for a worst case
+design. It is also very difficult to verify this type of nonlinear feedback controller.
+For this reason most current approaches to humanoid
+robots design the feedback controller
+separately and use constant gain independent joint linear feedback control.
+
+\subsection{Autonomous exoskeleton-world force control}
+
+A control objective could be to have the exoskeleton to apply a desired force.
 
@@
 
-\subsection{Autonomous exoskeleton-world force control}
-
-A control objective could be to have the exoskeleton to apply a desired force.
-
-To be written.
+\begin{verbatim}
+include J' term
+\end{verbatim}
+
+Why isn't this identical to operator force control?
 
@@
 To be written.
+
+\begin{verbatim}
+Cartesis version - j-1
+joint version, w jKj =k
+\end{verbatim}
+
+Why isn't this identical to operator impedance control?
 

### 2015-09-20 control.tex
more updates

@@
 \section{Recommended Control Architecture}
+\label{sec:architecture}
 
@@
 
+[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
+
 \section{Passive control with no active control}
@@
 
+1) We recommend a particular control 
+overall architecture in Section~\ref{sec:architecture}.
+
+2) We recommend the use of dynamic models of the exoskeleton, as has been done
+for the last decade starting with the BLEEX exoskeleton.
+
+3) We recommend using online optimization as part of the Lowest Level Controller
+in the form of Quadratic Programming. This was tested by the top biped robot
+teams in the DARPA Robotics Challenge, and worked well.
+
+4) We recommend applying as many sensors as possible, and asking the performers
+to make it easy to add more by making the sensor network available in the design.
+This maximizes the probability of success
+with respect to control architectures and algorithms by enabling multiple control
+approaches to be implemented, refined, and support each other.
+In particular, we would like to see force sensing between the operator and the
+exoskeleton, including at the feet, as much force sensing between the exoskeleton 
+and the world as possible, but minimally full six-axis force/torque sensing at
+the exoskeleton feet (where they touch the ground). We would like to see multiple MEMs
+IMUs (measuring linear acceleration and angular velocity) installed across 
+the exoskeleton. We would like to see high quality direct velocity sensing
+such as high count (100,000 counts/revolution) rotary encoders and 
+analog linear tachometers.
+
+5) {\bf Controversial:} We recommend a ``symbiotic'' control system design,
+which expects the operator to adapt to the exoskeleton and its control,
+and which has the control customized for the operator.
+
 \bibliographystyle{plain}

### 2015-09-21 control.tex
more updates

--- 2015-09-20 control.tex
@@
 inputs into the exoskeleton (exoskeleton positions,
-velocities, accelerations, operator-exo contact forces and
-exo-world contact forces), and choose from several possible exoskeleton
+velocities, accelerations, operator-exoskeleton contact forces and
+exoskeleton-world contact forces), and choose from several possible exoskeleton
 outputs: exoskeleton actuator forces and torques, exoskeleton
@@
 and acceleration.
+\item
+Online optimization of accelerations, actuator forces and torques, and contact forces
+using quadratic programming.
 \end{itemize} 
@@
 scheduling, ...
-Examples: walk-to-location( x=7.8m, y=2.3m ), run-to-location( speed=3.7m/s ),
-horizontal-jump( height= 2m, length= 5m ).
+Examples: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
+horizontal-jump(height= 2m, length= 5m).
 Global inverse kinematics (selection of solution branch) is performed at this level.
@@
 body parts with the exoskelton parts which they are attached to in terms
-of link inertias, location of center of mass, and moments of inertia,
+of link inertias, locations of center of mass, and moments of inertia,
 and just think of the whole system as one robot.
@@
 at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
-and so is the world ($\vtau_w$) (very unrealistic assumption). In this case:
+and so is the world ($\vtau_w$) (a very unrealistic assumption). In this case:
 \begin{equation}
@@
 the state vector.
+
+[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
 
@@
 
-Sarcos Primus Humanoid (ATR) video of active gravity compenstation applied
+Sarcos Primus Humanoid videos of active gravity compensation applied
 to a humanoid robot:\\
-\url{https://www.youtube.com/watch?v=KNxxLm4sPys}
+CMU: cancel gravity and friction: \url{https://www.youtube.com/watch?v=Ac5cowTwOPw}\\
+ATR: cancel gravity only: \url{https://www.youtube.com/watch?v=KNxxLm4sPys}
 
@@
 correctly handle stiction and hysteresis effects.
+
+Sarcos Primus Humanoid video of friction compensation applied
+to a humanoid robot:\\
+CMU: cancel gravity and friction: \url{https://www.youtube.com/watch?v=Ac5cowTwOPw}\\
 
@@
 + \mK_{\dot{\vq}}^{\ddot{\vq}} ( \hat{\dot{\vq}} - \dot{\vq_d} )
-\end{equation}
-The problem with putting the inverse dynamics in the feedback loop is that the
+\label{eq:computed-torque-control}
+\end{equation}
+The problem with putting the inverse dynamics in the feedback loop 
+(equation~\ref{eq:computed-torque-control})
+is that the
 effective gains of this control vary widely with the configuration $\vq$.
@@
 $\mK_{\vq}^{\ddot{\vq}}$ and $\mK_{\dot{\vq}}^{\ddot{\vq}}$. Because so many quantities
-are varying quite a bit, constant gain matrices may be quite low for a worst case
-design. It is also very difficult to verify this type of nonlinear feedback controller.
-For this reason most current approaches to humanoid
+are varying quite a bit, constant gain matrices may be quite small for a worst case
+design. It is also very difficult to evaluate and 
+verify this type of nonlinear feedback controller.
+For these reasons most current approaches to humanoid
 robots design the feedback controller
-separately and use constant gain independent joint linear feedback control.
+separately (and manually) 
+and use constant gain independent joint linear feedback control.
 
@@
 Why isn't this identical to operator impedance control?
+
+\subsection{How can the operator share control with autonomous control?}
+
+To be written.
 
@@
 \end{equation}
-Assuming that the operator directly applies exoskeleton joint torques,
+Assuming that the operator directly applies exoskeleton joint torques, this becomes:
 \begin{equation}
@@
 forces (GRF) magnitude and direction is used to command the
-actuators. In some researches the GRF sensors are used
+actuators. In some research the GRF sensors are used
 together with other sensors in the control architectures [17],
@@
 
-We are surprised that acceleration feedback based on double differentiation
-works well on real systems (even low pass filtered acceleration
-estimated by a Kalman filter).
-Impacts, shock waves, and the fact that neither
+Estimating acceleration using double differentiation with low pass filtering
+currently works well on real exoskeletons.
+However, impacts, shock waves, and the fact that neither
 the operator's body parts or the exoskeletons parts are rigid bodies and the
-joints are not well defined suggest this approach would be fragile to modeling
-error and unmodeled dynamics.
-It is not possible to distribute MEMs IMUs throughout the exoskeleton to
+joints are not well defined suggest this approach can be improved with actual
+acceleration measurements.
+It is now possible to distribute MEMs IMUs throughout the exoskeleton to
 improve both velocity and 
@@
 
+Disturbance observers~\cite{IEEE06197032}.
+
 \section{Discussion}
@@
 Conclusions and Recommendations to be written.
+
+1) We recommend a particular control 
+overall architecture in Section~\ref{sec:architecture}.
+We do not expect this to be controversial.
+
+2) We recommend actuation that can be treated as force or torque sources.
+This is a potentual problem for revolute electric actuation, as it
+is difficult to get necessary performance at low gear ratios (20-50)
+and fit the necessary torque sensing,
+
+3) We recommend the use of dynamic models of the exoskeleton, as has been done
+for the last decade starting with the BLEEX exoskeleton.
+We do not expect this to be controversial.
+
+4) We recommend using online optimization as part of the Lowest Level Controller
+in the form of quadratic programming. Online optimization
+was tested by the top biped robot
+teams in the DARPA Robotics Challenge, and worked well.
+A possible objection to online optimization
+is that it requires more substantial computing
+resources than current exoskeleton control methods. We feel the feasibility
+of this approach has been demonstrated and is easily achievable. Perceptual
+computing costs will dwarf computing costs for control, 
+so the computation cost is a non-issue.
+
+5) We recommend applying as many sensors as possible, and asking the performers
+to make it easy to add more by making the sensor network available in the design.
+This maximizes the probability of success
+with respect to control architectures and algorithms by enabling multiple control
+approaches to be implemented, refined, and support each other.
+In particular, we would like to see force sensing between the operator and the
+exoskeleton, including at the feet, as much force sensing between the exoskeleton 
+and the world as possible, but minimally full six-axis force/torque sensing at
+the exoskeleton feet (where they touch the ground). We would like to see multiple MEMs
+IMUs (measuring linear acceleration and angular velocity) installed across 
+the exoskeleton and at least one high quality fiber optic gyro IMU. 
+We would like to see high quality direct velocity sensing
+such as high count (100,000 counts/revolution) encoders and/or 
+analog rotary or linear tachometers on actuators and joints. 
+We would like to see actuator force or torque sensors
+(load cells or equivalents) with the measurement on the link side (rather than
+the actuator side) of any transmission. Electric current in motors and oil
+pressure in hydraulic pistons can also be used for actuator force estimation,
+but because these measurements are on the actuator side before the transmission
+and in the case of hydraulics before the oil seals on the piston, these measurements
+are greatly contaminated by friction. On the Atlas humanoid we typically saw 10Nm
+joint torque estimation errors for a system that estimated actuator output using
+oil pressure on each side of the piston head.
+A possible objection to this is additional cost. We feel it is a false
+economy to skimp on sensing. Leaving practical sensing out greatly increases
+the risk of poor performance.
+
+6) We recommend a ``symbiotic'' control system design should be used as a backup,
+in case more aggressive design philosophies such as ``invisible'' and ``natural''
+control system designs are not achievable.,
+The ``symbiotic'' control system design
+expects the operator to adapt to the exoskeleton and its control,
+and has the control customized for the operator.
+
+7) We recommend making use of virtual model control specialize
+for the various tasks. This can greatly increase operator/exoskeleton system
+performance over an exoskeleton that just carries a load and is otherwise
+``invisible''.
 

### 2015-09-21 exos/bleex.tex
added half of BLEEX section.

% \documentclass[letterpaper,12pt,fullpage]{article}

% \usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
% \usepackage{cite}
% \usepackage{graphicx}
% % \usepackage[dvips]{graphicx}
% % \usepackage{epsfig} % for postscript graphics files
%   % \graphicspath{{../eps/}}
% % \DeclareGraphicsExtensions{.eps}
% \usepackage{amsmath}
% \usepackage{amssymb}
% %\usepackage[cmex10]{amsmath}
% %\usepackage{array}
% %\usepackage{mdwmath}
% %\usepackage{mdwtab}
% %\usepackage{eqparbox}
% \usepackage[tight,footnotesize]{subfigure}
% %\usepackage[caption=false]{caption}
% %\usepackage[font=footnotesize]{subfig}
% %\usepackage{fixltx2e}
% %\usepackage{stfloats}
% \usepackage{hyperref}

% % correct bad hyphenation here
% %\hyphenation{op-tical net-works semi-conduc-tor}

% % \input latex-commands

% \newcommand{\T}{{\mbox {T}}}

% \newcommand{\gc}{{\mbox {GravityCompensation}}}

% \newcommand{\shrinkfig}{\def\baselinestretch{1.0}\small} % 0.9 okay
% \newcommand{\shrink}{\def\baselinestretch{1.1}\small} % 0.97 % 0.95 okay

% \begin{document}

\subsection{BLEEX}

keywords: sensitivity amplification; strength augmentation;\\

The Berkeley lower extremity exoskeleton (BLEEX) is an anthropomorphic, powered exoskeleton designed for human strength augmentation.  It is described as the first field-operational robotic system to allow its operator to carry significant loads over unstructured terrain without external power \cite{bleex_design_2006}.

The BLEEX system includes two 7 DOF, three-segment legs with thigh, shank, and foot links, on-board power supply, and a backpack-like frame.  The human wearer is rigidly connected at the feet and torso such that the frame shelters the user by transferring load forces to the ground.  The leg segments are connected by rotational joints including 3 DOF (2 actuated) at the hip, 1 DOF (actuated) at each knee, and 3 DOF (1 actuated f/e in sagittal 
plane; 2 passive) at the ankles.  Joint angles, torque, and power requirements are determined from human motion analysis based on a 75-kg human walking on flat ground at roughly 1.3 m/s (the average military male's maximum reported joint limits are also used to derive joint range of motion targets).  During design, joint motion was intended to be slightly less than the maximum human range of motion for safety; however, some joint ranges had to be reduced to avoid singularities.
% Ideally, joint motion would provide the maximum human range of motion for safety

Due to its high power to weight ratio (twice that of electric motors), BLEEX uses a hydraulic actuation system.  An on-board internal combustion engine provides both electric and hydraulic power.  The joints are driven by commercial small bore (2cm) dual action hydraulic actuators operating at 6.9 MPa. Though the operating pressure is relatively low, the hydraulic actuation system exhibits significant pressures losses across servo valves when less pressure is required than this system pressure.  Table~\ref{tab:bleex_joints} provides details regarding the range of motion and torque capabilities of BLEEX's joints.  As reported in \cite{bleex_design_2006}, BLEEX requires 1,143 W for walking relative to 165 W for human walking (14\% efficient compared to a human of the same size).  Altogether the suit needs 2.27 kW of hydraulic power and 220 W of electric power to accommodate climbing (540 W) and remaining electrical loads including 240 W to power the second stages on servo vales.

%
\begin{table}
\centering
\begin{tabular}{|l|*{3}{c|}}  % repeats {c|} 6 times
\hline
& BLEEX & human max & BLEEX \\
& ROM & torque \& power & max torque \\ \hline
Ankle flexion / extension & $\pm 45^\circ$ & $-120$ N-m; 250 W & $-200 / 155$ N-m\\ \hline
Ankle abduction / adduction & $\pm 20^\circ$ & N/A & N/A \\ \hline
Knee flexion & $121^\circ$ & $-35 / 60$ N-m; $-150 / 50$ W & $-100 / 140$ N-m \\ \hline
Hip flexion / extension & $\pm 121^\circ / 10^\circ$ & $-80 / 60$ N-m; $-60 / 115$ W & $-150 / 130$ N-m \\ \hline
Hip abduction / adduction & $\pm 16^\circ$ & N/A & N/A\\ \hline
total rotation external & $35^\circ$ & N/A & N/A \\ \hline
total rotation internal & $35^\circ$ & N/A & N/A \\ \hline
\end{tabular}
\caption{BLEEX joint range of motion (ROM) is near anthropomorphic.  The max torques are designed to meet the torque / power requirements of similarly sized human walking at 1.3 m/s \cite{bleex_design_2006}.}\label{tab:bleex_joints}
\end{table}
%

\subsubsection{Control}

\bibliographystyle{plain}
\bibliography{exos/bleex}

% \end{document}

### 2015-09-21 control.tex
added half of BLEEX section.
more updates

@@
 
+A later white paper will discuss how to handle malfunctions and damage to the
+system. Basically this boils down to fault detection, and switchover to
+an approapriate safe mode of operation.
+
 \subsection{Let's keep it simple}
@@
 
-A useful background paper: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}
+A useful background paper on impedance and admittance: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}
 
@@
 and thus is best thought of as a position (or velocity) source. Therefore,
-position control is performed by the low level controller. Hydraulic actuators
+position or velocity 
+control is performed by the low level controller. Hydraulic actuators
 are high impedance, but when a force or load sensor is added (or piston differential
@@
 and is best thought of as a force source. Force (or joint torque) control is
-performed by the low level controller. Force or torque can be added to an
-electromechanical drive train (electric motor plus gears or motor plus ballscrew)
+performed by the low level controller. Force or torque sensing can be added to an
+electromechanical drive train (electric motor plus gears or electric
+motor plus ballscrew)
 to make it lower impedance and more like a force source as well.
@@
 
+\section{Test case: Atlas humanoid robot}
+
+Gazebo and SDFast models.
+
 \section{Passive control with no active control}
@@
 \end{equation}
+
+\subsection{Full Atlas robot implementation.}
 
@@
 
+DISCUSS HOW TO HANDLE SENSOR NOISE
+
 Estimating acceleration using double differentiation with low pass filtering
@@
 Disturbance observers~\cite{IEEE06197032}.
+
+\subsection{Full Atlas robot.}
+
+\section{Robustness to parametric modeling error}
+
+\section{Robustness to unmodeled dynamics}
+
+\subsection{Example unmodeled dynamics}
+
+Examples:
+\begin{itemize}
+\item
+A) hydraulic actuator dynamics.
+\item
+B) delay. 
+\item
+C) elastic straps sinusoidal oscillation.
+\item
+D) loose straps (bang coast bang oscillation)
+\end{itemize}
+
+\subsection{1 joint rotary example.}
+
+\subsection{2 joint rotary example: stance.}
+
+\subsection{2 joint rotary example: swing.}
+
+\subsection{4 link body example.}
+
+\subsection{Full Atlas robot.}
+
+\section{Underactuated control}.
+
+Not talking about floating body dynamics, where global (``root'') position
+and velocity is unactuated.
+
+What about only sagittal actuation?
+
+\subsection{No actuation on lateral and twist}
+
+\subsection{Passive actuation on lateral and twist}
+
+Address what is the complexity penalty, weight penalty, 
+and ``drag'' of actuator on individual joint motion.
+
+\subsection{Active actuation on lateral and twist}
+
+Address what is the complexity penalty, weight penalty, 
+and ``drag'' of actuator on individual joint motion.
 
@@
 \end{document}
+
+sensitivity to
+ modeling error
+ sensor error
+ unmodeled dynamics
+
+underactuated control problem
+ - not fully actuated sagittal only.
+ - how necessary is lateral actuation
+ - what is the penalty
+  weight
+  drag of actuator
+
+how handle failure and damage after Oct 1
+
+what about a full model of the system
+
+Oct 5-9

### 2015-09-22 control.tex
more updates

@@
 Algorithms\\
-(Draft 2.1)}
+(Draft 2.2)}
 
@@
 
-A useful background paper on impedance and admittance: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}
+A useful background paper on impedance and admittance:\\
+\url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}
 
@@
 For example, for jumping, the exoskeleton can operate like a pogo 
-stick \url{https://www.youtube.com/watch?v=lbp41vWP4o4}, 
-and for running it could act like jumping 
-stilts \url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},
+stick:\\
+\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
+and for running it could act like jumping stilts:\\
+\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},
 
@@
 Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
-or distibuted limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
+or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
 Often, this controller is provided by the manufacturer of the robot
@@
 
-This control architecture is intuitive, abstracts the hardware 
+{\bf State Estimation:}
+In this architecture, state estimation to deal with sensor noise and incomplete or
+redundant sensing is a separate process and is as independent as possible
+from the control method~\cite{certainty-eq,separation-prin}.
+This makes evaluating different control methods much easier.
+
+The recommended control architecture is intuitive, abstracts the hardware 
 and lower level control in a straightforward way, is easy to program, and is
@@
 deep learning.
+
+An alternative approach to state estimation is to design state estimation
+and control laws simultaneously~\cite{LQG-LTR}. Another alternative is
+to have a controller with internal state, and design or learn a policy
+that works well, but does not have an explicit state estimation 
+part~\cite{lead-lag}.
 
@@
 where $\vtau$ are the exoskeleton joint torques, 
+$\hat{\mG}()$ are the gravitational forces calculated using an imperfect model, 
 and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
@@
 errors in the controller's estimate of vertical).
+All weight is assumed to be concentrated at the center of mass of the full system.
 
@@
 \section{Online inverse dynamics control}
+\label{sec:invdyn}
 
@@
 
-We can implement force control 
-if measurements of the contact forces with
+We can implement operator force control 
+if measurements of the operator contact forces with
 the exoskeleton are available.
-The exoskeleton moves as to create as little force as possible 
-assuming the actuators are torque sources.
-\begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vf} \mJ^{\tr} ( \hat{\vf} - \vf_d )
-\end{equation}
-$\mK_{\vf}$ is a gain matrix, $\mJ$ is the appropriate Jacobian matrix for
-force $\vf$
+The exoskeleton moves as to create as little force on the operator as possible 
+assuming the actuators are torque sources:
+\begin{equation}
+\vtau = \vtau_{\invdyn} + \mJ_o^{\tr} {\vf_o}_d
++ \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
+\end{equation}
+$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
+force $\vf_o$
 Note that this does not require inverting the Jacobian matrix.
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 \end{equation}
@@
 \begin{equation}
-\ddot{\vq} = - \mM^{-1} (\vq) \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 - \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
@@
 dynamics (including the operator link inertias, link centers of mass, and link
-moments of inertia in the dynamic model) based on only the desired postion, 
+moments of inertia in the dynamic model) based on the desired postion, 
 desired velocity, and desired acceleration. Trajectory tracking errors are compensated
@@
 
-A control objective could be to have the exoskeleton to apply a desired force.
-
-To be written.
-
-\begin{verbatim}
-include J' term
-\end{verbatim}
-
-Why isn't this identical to operator force control?
+A control objective could be to have the exoskeleton to autonomously
+apply a desired force.
+One could add
+\begin{equation}
+\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
+\end{equation}
+to a gravity compensated control or to control including inverse dynamics.
+
+Active force control could be added to improve the performance:
+\begin{equation}
+\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
+\end{equation}
 
@@
 be compliant or have an assigned impedance.
-
-To be written.
-
-\begin{verbatim}
-Cartesis version - j-1
-joint version, w jKj =k
-\end{verbatim}
-
-Why isn't this identical to operator impedance control?
-
-\subsection{How can the operator share control with autonomous control?}
-
-To be written.
+Let's describe a small position offset of the exoskeleton-world
+contact points as $\vq_w$. We know that:
+\begin{equation}
+\dot{\vq}_w = \mJ_w \dot{\vq}
+\end{equation}
+and that
+\begin{equation}
+\vtau = \mJ_w^{\tr} \vf_w 
+\end{equation}
+This allows us to relate stiffness and damping in the external world coordinates
+with the stiffness and damping in exoskeleton joint coordinates.
+Here is the relationship for damping:
+\begin{equation}
+\vtau = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w \dot{\vq}
+\end{equation}
+So the damping in joint coordinates is $\vtau = \mK_{\dot{\vq}} \dot{\vq}$ and
+therefore
+\begin{equation}
+\mK_{\dot{\vq}} = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w
+\end{equation}
+The mapping is the same for stiffnes:
+\begin{equation}
+\mK_{\vq} = \mJ_w^{\tr} \mK_{w} \mJ_w
+\end{equation}
+So we should assign the exoskeleton joint position gains to be $\mK_{\vq}$ and
+the joint velocity gains to be $\mK_{\dot{\vq}}$ in equation~\ref{eq:impedance}.
+
+Similarly, we can use exoskeleton-world force control to improve performance,
+setting
+\begin{equation}
+{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
+\end{equation}
+and add the force error to equation~\ref{eq:impedance}: 
+\begin{equation}
+\mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
+\end{equation}
 
 \section{Using Online Optimization: Quadratic Programming}
+\label{sec:qp}
 
@@
 \end{equation}
+$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
+different errors and error directions.
 The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
@@
 
-In the case of exoskeletons, part of the optimization criterion could be to
+In the case of exoskeletons, 
+part of the optimization criterion could be to
 minimize contact forces on the operator:
@@
 Operator force feedback and impedance control could be implemented by specifying a
-desired operator force ${\vf_o}_d$:
+desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
+\begin{equation}
+\vf_o = {\vf_o}_d
+\end{equation}
+or by adding a ``soft'' constraint in the form of a new term in the optimization
+criterion:
 \begin{equation}
@@
 \end{equation}
-Assuming that the operator directly applies exoskeleton joint torques, this becomes:
+Assuming that the operator directly applies exoskeleton joint torques, this becomes
+the equality constraint
+\begin{equation}
+\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+or is added to the optimization criterion:
 \begin{equation}
@@
 \end{equation}
+with
+\begin{equation}
+{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+
+\subsection{Full Atlas robot implementation.}
+
+\section{Underactuated and overactuated control}
+
+There are two kinds of underactuation we want to consider. The first
+is ``floating body dynamics'', where the robot is not bolted to the floor.
+Humanoid robots have floating body dynamics. A ``root'' location (usually somewhere
+on the pelvis) is chosen, and the position, velocity, orientation, and angular
+velocity of that point and the rigid body it is on
+are included in the state vector $\vq$. The corresponding dimensions of the
+velocity vector $\dot{\vq}$ are not actuated. We have been sweeping this
+issue under the rug so far. It is now useful to be more explicit. A selection matrix
+$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
+\begin{equation}
+\vtau = \mS \vu
+\end{equation}
+The rows in $\mS$ corresponding to the root linear and angular velocities
+are all zero.
+
+The second kind of actuation appears in exoskeletons where we rely on the
+operator to actuate a particular joint. For example, many lower body
+exoskeltons only actuate the joints in the sagittal plane, leaving lateral
+and twist (yaw) joints unactuated. We can actually handle this by using
+the matrix $\mS$ and reducing the size of the $\vu$ vector.
+
+We can also have passive actuation. This we can include by augmenting the
+dynamics equations to include the force equation of the passive element,
+such as a nonlinear spring.
+
+We can have multijoint actuation, which can be handled in $\mS$ as well.
+
+We can have redundant actuation (more than one actuator across any particular
+joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
+
+The modified dynamics equations are
+\begin{equation}
+\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
++ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_o^{\tr}(\vq) \vf_o
+\label{eq:dynamics2}
+\end{equation}
+Note that $\ddot{\vq}$ and the new actuation vector $\vu$
+do not have the same dimensionality. What difference does that make?
+
+Underactuated joints reduce our ability to control behaviors directly
+through the actuators. In some cases we can control unactuated degrees of freedom
+through indirect effects over time, and in some cases those degrees of freedom
+are not ``controllable'' in the technical sense. Remember that we have a human
+operator that fully actuates the exoskeleton, so having some degrees of freedom
+not controllable by the actuators is not fatal to a design.
+
+The {\bf no control} case is unchanged with underactuation.
+
+{\bf Open loop control} continues to work with underactuation
+as long as a drive function can be found
+to generate the desired behavior.
+
+{\bf Gravity compensation} and {\bf friction compensation}
+still work on the actuated degrees
+of freedom. The operator needs to carry the load on the unactuated degrees of freedom.
+
+Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
+no longer works if $\mS$ is not invertable (not square or not full rank).
+In the underactuated case $\mS$ is not square. In this case {\bf online optimization}
+as described in Section~\ref{sec:qp} is used to ``perform inverse dynamics'' by
+finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
+and actuator commands $\vu$.
+
+Another approach to solving this problem is prioritized control, which does
+not include optimization~\cite{}. For reasons we will describe, we believe
+online optimization is much more flexible and can handle programming exoskeleton
+behavior much better, so we will not discuss prioritized control further.
+
+Operator {\bf force feedback} is implemented using online optimization by
+add the following ``hard'' constraint
+to the equality constraints in the quadratic program:
+\begin{equation}
+\vf_o = {\vf_o}_d 
+\end{equation}
+or adding a ``soft'' constraint as a term
+\begin{equation}
+(\vf_o - {\vf_o}_d)^{\tr} W_{\vf_o} (\vf_o - {\vf_o}_d)
+\end{equation}
+in the optimization criterion. $W_{\vf_o}$ is a weight to indicate how much operator
+force is acceptable (if ${\vf_o}_d = 0$) or how much operator force error is
+acceptable.
+Force damping can be implemented by adding a similar ``soft'' constraint:
+\begin{equation}
+\dot{\vf}_o^{\tr} W_{\dot{\vf}_o} \dot{\vf}_o
+\end{equation}
+to the optimization criterion.
+
+{\bf Impedance control} is implemented using online optimization by adding
+the impedance equations as ``hard'' equality constraints (to achieve them exactly)
+\begin{equation}
+\mJ_o^{\tr} {\vf_o} = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+or by adding the ``soft'' constraint ``equation error'' to the optimization criterion
+weighted by a weight factor $W_{imp}$ to indicate
+how large an error is tolerated:
+\begin{equation}
+\ve = \mJ_o^{\tr} {\vf_o} - \mK_{\vq}( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
 and
 \begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-
-\subsection{Full Atlas robot implementation.}
+\ve^{\tr} W_{imp} \ve
+\end{equation}
+
+{\bf Mapping external forces to operator forces} can be done either with the
+``hard'' equality constraint:
+\begin{equation}
+\vf_o = F_{ow} ( \vf_w )
+\end{equation}
+or using the ``soft'' constraint term:
+\begin{equation}
+(\vf_o - F_{ow} ( \vf_w ))^\tr \mW_{fmap}  (\vf_o - F_{ow} ( \vf_w ))
+\end{equation}
+
+Similar techniques are used to handle {\bf autonomous exoskeleton position, velocity,
+force, and impedance control.}
+
+\subsection{How can the operator share control with autonomous control?}
+
+The use of online optimization makes sharing control with an operator easy.
+Dimensions (even powered dimensions) that are under the operator's control have
+low weights in the online optimization. It is also easy to map operator controls
+such as joysticks, sliders, and alternative interfaces to almost any desired
+control outcome by including these measurements in the online optimization.
+The operator and the controller can ``share'' or ``blend''
+control by making their respective
+weight matrices similar.
 
@@
 
+We will now consider position controlled actuators.
 At this point our dynamics formulation is no longer appropriate, since we
@@
 \subsection{Full Atlas robot.}
-
-\section{Underactuated control}.
-
-Not talking about floating body dynamics, where global (``root'') position
-and velocity is unactuated.
-
-What about only sagittal actuation?
-
-\subsection{No actuation on lateral and twist}
-
-\subsection{Passive actuation on lateral and twist}
-
-Address what is the complexity penalty, weight penalty, 
-and ``drag'' of actuator on individual joint motion.
-
-\subsection{Active actuation on lateral and twist}
-
-Address what is the complexity penalty, weight penalty, 
-and ``drag'' of actuator on individual joint motion.
 

### 2015-09-22 exos/bleex.tex

@@
-% \documentclass[letterpaper,12pt,fullpage]{article}
-
-% \usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
-% \usepackage{cite}
-% \usepackage{graphicx}
-% % \usepackage[dvips]{graphicx}
-% % \usepackage{epsfig} % for postscript graphics files
-%   % \graphicspath{{../eps/}}
-% % \DeclareGraphicsExtensions{.eps}
-% \usepackage{amsmath}
-% \usepackage{amssymb}
-% %\usepackage[cmex10]{amsmath}
-% %\usepackage{array}
-% %\usepackage{mdwmath}
-% %\usepackage{mdwtab}
-% %\usepackage{eqparbox}
-% \usepackage[tight,footnotesize]{subfigure}
-% %\usepackage[caption=false]{caption}
-% %\usepackage[font=footnotesize]{subfig}
-% %\usepackage{fixltx2e}
-% %\usepackage{stfloats}
-% \usepackage{hyperref}
-
-% % correct bad hyphenation here
-% %\hyphenation{op-tical net-works semi-conduc-tor}
-
-% % \input latex-commands
-
-% \newcommand{\T}{{\mbox {T}}}
-
-% \newcommand{\gc}{{\mbox {GravityCompensation}}}
-
-% \newcommand{\shrinkfig}{\def\baselinestretch{1.0}\small} % 0.9 okay
-% \newcommand{\shrink}{\def\baselinestretch{1.1}\small} % 0.97 % 0.95 okay
-
-% \begin{document}
-
 \subsection{BLEEX}
+\label{exo:bleex}
 
@@
 It is described as the first field-operational robotic system to allow its operator to carry significant loads over unstructured terrain without external power \cite{bleex_design_2006}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/bleex_exo.png}
+\end{figure}
 
@@
 
+The BLEEX team has successfully implemented both sensitivity amplification \cite{sesitivityAmpPaper2005} and a hybrid assitive control scheme \cite{bleex_hybrid_control_2006} that switches (based on gait phase) between sensitivity amplification and a position control regulating desired torque.
+This section focuses on sensitivity amplification, as the hybrid assitive strategy did not perform as well in walking trials.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/bleex_support_phases.png}
+\end{figure}
+
+In its sensitivity amplification implementation, BLEEX controllers attempt to minimize interaction forces between the human pilot and the exoskeleton.
+However, the BLEEX team makes a point to avoid direct measurements of the interaction force between the human and the exoskeleton suit.
+They highlight difficulties in properly outfitting humans with necessary sensing equipment and challenges in modeling the interaction between the human and exoskeleton (e.g., non-rigid, non-fixed contact points).
+Instead, the team develops controllers based on an inverse dynamic model (in the saggital plane) of only the exoskeleton.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/bleex_1dof_ex.png}
+\end{figure}
+
+Assuming no outside disturbances, the BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as $d$ (assuming no outside disturbances).
+Neglecting gravity, the exokeleton angular velocity is modeled as
+\[v = G r + S d ,\] 
+where $v$ represents the angular velocity of the exokeleton, $G$ is the transfer function from actuator inputs ($G$ is the exoskeleton's dynamics), $r$ is the actuator input, and $S$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=4.0in]{exos/figs/bleex_control_diag_1.png}
+\end{figure}
+
+The goal is to maximize sensitivity to $d$ \emph{without direct measurement.}  Sensitivity amplification accomplishes this by creating a feedback loop from a controller, $C$, acting only on exoskeleton variables.
+A new sensitivity equation,
+\[S_{new} = \frac{v}{d} = \frac{S}{1 + G C} ,\]
+is maximized by applying positive feedback.
+To achieve a large sensitivity, BLEEX uses $C = (1-\alpha^{-1})G^{-1}$ so that $\alpha$ provides a direct (scalar) amplification factor.
+A low pass filter is often added to the $C$ term in order to damp out high frequency dynamics of the exoskeleton, which are not captured in these models.
+Note that the controller depends on an inverse dynamic model of the exoskeleton, $G^{-1}$.
+Since the model is hybrid, these dynamics switch according to gait phases (single support, double support, stance).
+BLEEX detects these transitions using foot sensors.
+Assuming the single leg support dynamics are in the form,
+\begin{equation}
+\mM({\bf\theta})\ddot{{\bf \theta}} + \mC({\bf\theta},\dot{{\bf\theta}}) \dot{{\bf \theta}} + \mV({\bf\theta}) = {\bf T} + {\bf d} ,
+\end{equation}
+where $T$ is a vector of actuator torques, the BLEEX control torque would be
+\begin{equation}
+{\bf T} = \mV({\bf\theta}) + (1 - \alpha^{-1}) \big [ \mM({\bf\theta})\ddot{{\bf \theta}} + \mC({\bf\theta},\dot{{\bf\theta}}) \dot{{\bf \theta}} \big ] .
+\end{equation}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=4.0in]{exos/figs/bleex_control_diag_2.png}
+\end{figure}
+
+As a note, positive feedback is normally avoided in control design because it amplifies disturbances.
+In the case of BLEEX, designers sacrifice disturbance rejection to maximize the response of the suit to its wearer.
+Users must therefore take action to stabilize and balance out disturbances.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=4.0in]{exos/figs/bleex_control_diag_3.png}
+\end{figure}
+
+For sensing, BLEEX uses information from \textbf{8 encoders} and \textbf{16 accelerometers} to determine angle, angular velocity, and angular acceleration of eight actuated joints.
+It includes a \textbf{foot switch and load distribution sensor} at each foot.
+Eight single axis \textbf{force sensors} provide measurements required to perform low level force control at each actuator.
+An \textbf{inclinometer} indicates the orientation of the backpack relative to gravity.
+Using this sensitivity amplification control scheme, BLEEX has achieved successful walking at 1.3 m/s with a 34kg payload \cite{sesitivityAmpPaper2005}.
+
+\subsection{Assessment and Recommendations}
+
+Sensitivity amplification is a current best-in-class control strategy.
+The exoskeleton shadows the user and uses exoskeleton data for joint angle, velocity, and acceleration and the exoskeleton model to minimize torques experienced by the human torque.
+The process and hardware are relatively simple yet effective.
+However, the approach has several significant disadvantages including heavy reliance on an accurate exoskeleton model and the fact that sensitivity amplification will amplify disturbances.
+Note that in combat situations this latter point is critical as a sensitivity amplification will amplify forces acting on the suit and the operator will have to re-act to compensate.
+The only way to compensate for or resist external forces in this setting is to filter them out, which would be extremely challenging unless additional sensory equipment is provided.
+As one possibility, a sensitivity optimization approach may be paired with sEMG data to estimate user intent and provide such a filter.
+
 \bibliographystyle{plain}
@@
 
+The figures in this section were obtained from \cite{sesitivityAmpPaper2005}. Materials presented are based on the references above.
+
 % \end{document}

### 2015-09-22 control.tex
cga updates

@@
+% SIGN ISSUE ON FORCE FEEDBACK o and w: sort it out
+
 \documentclass[letterpaper,12pt,fullpage]{article}
@@
 \newcommand{\invdyn}{{\mbox {\tiny ID}}}
+\newcommand{\invdynw}{{\mbox {\tiny ID+Fw}}}
 
@@
 A companion paper surveys implemented exoskeleton control architectures
-and algorithms~\cite{}.
+and algorithms [survey.pdf].
 
@@
 system. Basically this boils down to fault detection, and switchover to
-an approapriate safe mode of operation.
+an appropriate safe mode of operation.
+
+A later white paper will address robustness to modeling error and unmodeled
+dynamics using simulation.
 
@@
 
-\section{What are some design philosophy alternatives?}
-
-\subsection{``Invisible'' exoskeleton}
-
-Can we build an exoskeleton allows the operator to behave naturally
-and exerts little or no force on the operator?
-
-This could be achieved through active control using feedback of the
-operator-exoskeleton forces. Additional gravity, friction, actuator, and
-inertial compensation (inverse dynamics) can improve performance.
-
-\subsection{``Natural'' exoskeleton}
-
-Can we build an exoskeleton allows the operator to behave naturally?
-The operator can feel the exoskeleton, but can overpower it when necessary.
-
-Bypass valves and clutches could allow the operator to physically ``take over''
-and push the relatively lightweight leg mounted exoskeleton around.
-
-``Passive'' degrees of freedom can have physical or ``virtual'' springs
-applied to assist in support (all abduct/adduct DOF, rotational DOF).
-These aids can be clutched out or turned off when necessary.
-``Virtual'' springs (springs implemented actively through exoskeleton
-actuators) can be modulated.
-
-\subsection{``Symbiotic'' exoskeleton}
-
-Can we build an exoskeleton that allows the operator/exoskeleton team
-to be effective, but not necessarily behave naturally? 
-The operator needs to learn to ``fly'' the suit,
-just as operators learn to operate parachutes,
-wing suits, diving equipment, high and low temperature protective
-suits, firefighting equipment, and high altitude flight
-and parachute suits.
-
-The exoskeleton can adapt its behavior to the task.
-For example, for jumping, the exoskeleton can operate like a pogo 
-stick:\\
-\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
-and for running it could act like jumping stilts:\\
-\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\input{design-goals.tex}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-{\bf Low Level Control:} Present interface to actuators and
-high bandwidth feedback control, degrees of freedom,
-or synergies of either:
+\subsection{Low Level Control}
+
+Low level control presents an API (Application Programmer Interface)
+for higher control levels
+to access the actuators, sensors, and other robot hardware.
+It implements high servo rate processing including feedback control and online
+optimization.
+This level provides an idealization of the exoskeleton in two ways: 
+What degrees of freedom (DOF) are controlled, and how is a degree of
+freedom controlled?
+
+\subsubsection{What degress of freedom are controlled?}
+
+The low level control API may directly map on to the exoskeleton hardware
+(physical joints,
+raw actuators, and raw sensors), or
+it may present an ``abstracted'' view or ``virtual'' hardware.
+For example, it is often useful to create virtual degrees of freedom that are
+actually combinations of joints, or to control points on the robot in Cartesian
+or other coordinates. We refer to these virtual degress of freedom as {\bf synergies.}
+A complex synergy is the center of mass. Higher level control is made easier by
+providing a virtual sensor that measures center of mass location and velocity,
+as well as providing a virtual actuator that acts at the center of mass in
+a Cartesian fashion (push horizontally and vertically in world coordinates).
+On a smaller scale, it may simplify higher level control 
+to create virtual actuators that cross
+multiple joints. These multi-joint actuators are actually implemented by commands
+to individual physical single-joint actuators.
+
+\subsubsection{How is a degree of freedom controlled?}
+
+It is useful for human programmers
+to provide them with idealized abstractions of control of a single degree of
+freedom. Here are some common alternatives:
 \begin{itemize}
 \item
-Position control.
+Position control. The joint moves to a specified position and stays there independent
+of load.
 \item
-Pure force or torque control.
+Velocity control. The joint moves at a specified velocity independent of load.
 \item
-Impedance control: specify desired position (for spring), and generate
+Force or torque control. The joint provides a specified force or torque independent
+of its motion.
+\item
+Impedance control: The user specifies a desired position (for a spring), and commands
+a desired
 stiffness, damping, and potentially a modified inertia or moment of inertia.
-Modifying inertia is more dangerous than providing active stiffness
+Modifying inertia is more dangerous than providing variable active stiffness
 or damping.
 Or (almost equivalently),
-specify a desired force and change the force according to position, velocity,
-and acceleration.
+the user specifies a desired force or torque
+which changes according to the position, velocity,
+and acceleration of the joint.
 \item
 Online optimization of accelerations, actuator forces and torques, and contact forces
-using quadratic programming.
+using quadratic programming. See Section \ref{sec:qp} for further explanation.
 \end{itemize}
-One important synergy is coordinated control of the center of mass location and
-velocity for balance.
-PID or fancier feedback control laws are implemented at this level.
+
+\subsubsection{Feedback control}
+
+PID or more complex feedback control laws are implemented at this level.
+
+\subsubsection{Feedforward control}
+
 Gravity, friction, actuator dynamics, and other feedforward disturbance
@@
 or other types of feedforward reference tracking control.
+
+\subsubsection{State estimation}
+
+State estimation is done at this level to compensate for sensor noise.
 Disturbance observers for model errors and contact/wind forces
 are implemented at this level.
-Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
+
+\subsubsection{Existing examples}
+
+Typically low level control is implemented
+by a high sampling rate global servo~\cite{Atlas-robot}
 or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
@@
 
-{\bf ``Phoneme'' Control:} Mid-level controllers
+\subsection{``Phoneme'' Control}
+
+This level often provides the behavioral primitives or ``verbs'' for robot programming:
+providing mid-level controllers
 for relatively constant phases of behavior.
-Often involves position or velocity targets (goto or
-stay-at X), trajectory references, ...
-Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
-Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
-Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
-Finer resolution may be needed. For example a walking stance period may
+These behaviors often involves position or velocity targets (goto or
+stay-at X), and trajectory references.
+Example phoneme behaviors from walking: left-stance, double-support-1, right-stance, double-support-2, ....
+Example behaviors from running: left-foot-down, flight-1, right-foot-down, flight-2, ....
+Example behaviors from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ....
+Finer or multiple behavioral resolutions may be needed.
+For example a walking stance period may
 be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
@@
 is performed at this level.
-This level often provides the behavioral primitives or ``verbs'' for robot programming.
-
-{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
+
+\subsection{``Word'' Control}
+
+This level determines the control flow of a behavior: what happens next?
+Mid-level controllers are provided for sequences of ``phoneme'' behavioral
 units.
-Often involves finite state machines, timers, if-then condition testing, gain
-scheduling, ...
-Examples: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
-horizontal-jump(height= 2m, length= 5m).
+This level often involves finite state machines, timers, if-then condition testing,
+and gain or behavior
+scheduling.
+Example word behaviors: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
+and horizontal-jump(height=2m, length=5m).
 Global inverse kinematics (selection of solution branch) is performed at this level.
-This level often determines the control flow of a behavior.
-
-{\bf Highest Level Control} aka {\bf Behavior Selection:} 
-Answer the question: ``What do I do next?''
-Select behaviors and behavioral parameters (targets,
-speeds, durations).
-Often involves finite state machines. timers, if-then condition testing, ...
+
+\subsection{Behavior Selection}
+
+The highest level of control in the recommended architecture is behavior selection,
+which answers the question:
+``What should I do next?''
+This level selects behaviors and behavioral parameters (targets,
+speeds, durations, etc.).
+This level often also 
+includes decision tables, finite state machines. timers, if-then condition testing, ...
 This level often involves a great deal of human operator online interaction,
@@
 
-{\bf Error Handling:} Errors must be detected and handled by parameter (target)
-adjustment and/or behavior switching
-at all of these levels.
-
-{\bf State Estimation:}
+\subsection{State Estimation}
+
 In this architecture, state estimation to deal with sensor noise and incomplete or
@@
 
+\subsection{Error Handling}
+
+Errors must be detected and handled by parameter (target)
+adjustment and/or behavior switching
+at all of these levels.
+
+\subsection{Why do we recommend this architecture?}
+
 The recommended control architecture is intuitive, abstracts the hardware 
@@
 
+\subsection{What are the alternatives?}
+
 Closely related control architectures combine Phoneme and Word control,
 or combine Phoneme, Word, and Behavior Selection.
-This is not a major design change.
+This is not a major change.
 
@@
 operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
+We do not address the loosely connected case in this paper.
 
@@
 \end{equation}
-where $\vq$ are the joint angles and position of the ``root'' of
+where $\vq$ are the joint angles of
 the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
@@
 as exoskeleton joint torques.
-$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
-that correspond to root velocities, since those dimensions are not actuated.
+
+We do not describe how to derive this equation or identify appropriate parameters
+for an exoskeleton. We can provide this information if requested.
+We are assuming that the equations have been derived taking into account constraints
+such as one or two feet on the ground. Until we discuss ``floating body'' dynamics
+in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
+We are assuming all degrees of freedom (joints) are actuated until 
+Section~\ref{sec:float}.
 
@@
 
+[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
 Note that we are ignoring the operator dynamics and operator muscle and tissue
@@
 We are also ignoring actuator dynamics, which we can handle by augmenting 
-the state vector.
-
-[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
+the state vector and adding additional equations.
 
@@
 attempt to match the torques caused by inertial, Coriolis, centripetal,
-gravitational, frictional, and exoskeleton-world contact forces.
+gravitational, and frictional forces.
 \begin{equation}
@@
 + \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+\end{equation}
+If in addition we want to cancel exoskeleton-world contact forces,
+we add another term:
+\begin{equation}
+\vtau = \vtau_{\invdynw} = \vtau_{\invdyn} 
 + \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
+\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
 \label{eq:change-inertia1}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
+\vtau = \vtau_{\invdynw} + \mK \hat{\vf}_o
 \end{equation}
@@
 from the point of view of the operator.
-A future white paper will discuss how to make the exoskeleton 
+Section~\ref{sec:impw} discusses how to make the exoskeleton 
 imitate a desired linear dynamic system 
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\vtau = \vtau_{\invdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \label{eq:impedance}
@@
 \begin{equation}
-\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
-\end{equation}
-and the operator sees the desirecd impedance.
-
-Although impedance control is extensively utilized in rehabilitation
-robotics [18], limited number of studies has been focused on
-this method for power augmentation [19, 20].
+\mM \ddot{\vq} - \mK_{\dot{\vq}} \dot{\vq} - \mK_{\vq} ( \vq - \vq_d ) = \vtau_o
+\end{equation}
+and the operator sees the desired impedance.
+
+{\it Although impedance control is extensively utilized in rehabilitation
+robotics [18], only a limited number of studies has been focused on
+this method for power augmentation [19, 20].}
 \begin{verbatim}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_o^{\tr} {\vf_o}_d
+\vtau = \vtau_{\invdynw} + \mJ_o^{\tr} {\vf_o}_d
 + \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 - \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
@@
 \subsection{Autonomous exoskeleton-world impedance control}
+\label{sec:impw}
 
@@
 \end{equation}
-and add the force error to equation~\ref{eq:impedance}: 
-\begin{equation}
-\mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
-\end{equation}
+and add the force error: 
+\begin{equation}
+\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
++ \mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
+\end{equation}
+
+\section{Underactuated and overactuated control and floating body dynamics}
+\label{sec:float}
+
+This section describes how to handle underactuation and overactuation using
+``floating body dynamics.''
+We have been sweeping this
+issue under the rug so far. It is now useful to be more explicit.
+There are two kinds of underactuation we want to consider. The first
+is ``floating body dynamics'', where the robot is not bolted to the floor.
+Humanoid robots (and humans) have floating body dynamics.
+A ``root'' location (usually somewhere
+on the pelvis) is chosen, and the position, velocity, orientation, and angular
+velocity of that point and the rigid body it is on
+are included in the state vector $\vq$. The corresponding dimensions of the
+velocity vector $\dot{\vq}$ are not actuated. A selection matrix
+$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
+\begin{equation}
+\vtau = \mS \vu
+\end{equation}
+The rows in $\mS$ corresponding to the root linear and angular velocities
+are all zero.
+
+The second kind of underactuation appears in exoskeletons where we rely on the
+operator to actuate a particular joint. For example, many lower body
+exoskeltons only actuate the joints in the sagittal plane, leaving lateral
+and twist (yaw) joints unactuated or passively actuated.
+
+Passive actuation can be handled by augmenting the
+dynamics equations to include the force equation of the passive element,
+such as a nonlinear spring.
+
+The underactuation can be handled by reducing the size of the $\vu$ vector
+and changing the matrix $\mS$ to reflect the missing actuation.
+We can have redundant actuation (more than one actuator across any particular
+joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
+We can have multijoint actuation, which can be handled in $\vu$ and $\mS$ as well.
+
+The modified dynamics equations are
+\begin{equation}
+\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
++ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_o^{\tr}(\vq) \vf_o
+\label{eq:dynamics2}
+\end{equation}
+Note that $\ddot{\vq}$ and the new actuation vector $\vu$
+do not have the same dimensionality. What difference does that make?
+
+Underactuated joints reduce our ability to control behaviors directly
+through the actuators. In some cases we can control unactuated degrees of freedom
+through indirect effects over time, and in some cases those degrees of freedom
+are not ``controllable'' in the technical sense. Remember that we have a human
+operator that fully actuates the exoskeleton, so having some degrees of freedom
+that are not controllable by the actuators is not fatal to a design.
+
+The {\bf no control} case is unchanged with underactuation.
+
+{\bf Open loop control} continues to work with underactuation
+as long as a drive function can be found
+to generate the desired behavior.
+
+{\bf Gravity compensation} and {\bf friction compensation}
+still work on the actuated degrees
+of freedom. Passive actuation or 
+the operator needs to carry the load on the unactuated degrees of freedom.
+
+Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
+no longer works if $\mS$ is not invertable (not square or not full rank).
+In the underactuated and overactuated cases $\mS$ is not square. 
+In these cases {\bf online optimization}
+as described in the next Section is used to ``perform inverse dynamics'' by
+finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
+and actuator commands $\vu$.
+
+Another approach to solving this problem is prioritized control, which does
+not include optimization~\cite{}. For reasons we will describe, we believe
+online optimization is much more flexible and can handle programming exoskeleton
+behavior much better, so we will not discuss prioritized control further.
 
@@
 The dominant paradigm in force controlled (typically
-hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
-such as torque and center of pressure limits using online quadratic programming
-to solve the inverse dynamics equations. This approach can also be used
+hydraulic) full size humanoid robots is to use online optimization
+to handle under and overactuation
+and to handle constraints in inverse dynamics
+such as torque and center of pressure limits.
+Online optimization typically solves the constrained inverse dynamics problem
+using quadratic programming.
+This approach is used
 to implement and trade off among the many control objectives described so far.
@@
 \end{equation}
-The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
-torques $\vtau$, and contact forces $\vf$.
-For robots the discrepancy between actual and desired accelerations, joint
-torques, and contact forces are all minimized:
+The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, actuator
+commands $\vu$, and contact forces $\vf$.
+For robots the discrepancy between actual and desired accelerations, actuator command
+magnitude, and contact force magnitude are all minimized:
 \begin{equation}
@@
 $\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
-different errors and error directions.
-The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
-Torque limits, friction limits, center of pressure location constraints,
+different errors/magnitudes and error/magnitude directions.
+The dynamics (equation~\ref{eq:dynamics2}) are enforced as an equality constraint.
+Actuator limits, friction limits, center of pressure location constraints,
 and unidirectional contact constraints are all
@@
 
+\subsection{Controlling operator forces and perceived exoskeleton impedance}
+
 In the case of exoskeletons, 
@@
 \end{equation}
-
-To implement operator impedance control we set the desired operator contact force to:
-\begin{equation}
-{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
-\end{equation}
-Assuming that the operator directly applies exoskeleton joint torques, this becomes
-the equality constraint
-\begin{equation}
-\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-or is added to the optimization criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) 
-+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
-\end{equation}
-with
-\begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-
-\subsection{Full Atlas robot implementation.}
-
-\section{Underactuated and overactuated control}
-
-There are two kinds of underactuation we want to consider. The first
-is ``floating body dynamics'', where the robot is not bolted to the floor.
-Humanoid robots have floating body dynamics. A ``root'' location (usually somewhere
-on the pelvis) is chosen, and the position, velocity, orientation, and angular
-velocity of that point and the rigid body it is on
-are included in the state vector $\vq$. The corresponding dimensions of the
-velocity vector $\dot{\vq}$ are not actuated. We have been sweeping this
-issue under the rug so far. It is now useful to be more explicit. A selection matrix
-$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
-\begin{equation}
-\vtau = \mS \vu
-\end{equation}
-The rows in $\mS$ corresponding to the root linear and angular velocities
-are all zero.
-
-The second kind of actuation appears in exoskeletons where we rely on the
-operator to actuate a particular joint. For example, many lower body
-exoskeltons only actuate the joints in the sagittal plane, leaving lateral
-and twist (yaw) joints unactuated. We can actually handle this by using
-the matrix $\mS$ and reducing the size of the $\vu$ vector.
-
-We can also have passive actuation. This we can include by augmenting the
-dynamics equations to include the force equation of the passive element,
-such as a nonlinear spring.
-
-We can have multijoint actuation, which can be handled in $\mS$ as well.
-
-We can have redundant actuation (more than one actuator across any particular
-joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
-
-The modified dynamics equations are
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_o^{\tr}(\vq) \vf_o
-\label{eq:dynamics2}
-\end{equation}
-Note that $\ddot{\vq}$ and the new actuation vector $\vu$
-do not have the same dimensionality. What difference does that make?
-
-Underactuated joints reduce our ability to control behaviors directly
-through the actuators. In some cases we can control unactuated degrees of freedom
-through indirect effects over time, and in some cases those degrees of freedom
-are not ``controllable'' in the technical sense. Remember that we have a human
-operator that fully actuates the exoskeleton, so having some degrees of freedom
-not controllable by the actuators is not fatal to a design.
-
-The {\bf no control} case is unchanged with underactuation.
-
-{\bf Open loop control} continues to work with underactuation
-as long as a drive function can be found
-to generate the desired behavior.
-
-{\bf Gravity compensation} and {\bf friction compensation}
-still work on the actuated degrees
-of freedom. The operator needs to carry the load on the unactuated degrees of freedom.
-
-Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
-no longer works if $\mS$ is not invertable (not square or not full rank).
-In the underactuated case $\mS$ is not square. In this case {\bf online optimization}
-as described in Section~\ref{sec:qp} is used to ``perform inverse dynamics'' by
-finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
-and actuator commands $\vu$.
-
-Another approach to solving this problem is prioritized control, which does
-not include optimization~\cite{}. For reasons we will describe, we believe
-online optimization is much more flexible and can handle programming exoskeleton
-behavior much better, so we will not discuss prioritized control further.
-
-Operator {\bf force feedback} is implemented using online optimization by
-add the following ``hard'' constraint
-to the equality constraints in the quadratic program:
-\begin{equation}
-\vf_o = {\vf_o}_d 
-\end{equation}
-or adding a ``soft'' constraint as a term
-\begin{equation}
-(\vf_o - {\vf_o}_d)^{\tr} W_{\vf_o} (\vf_o - {\vf_o}_d)
-\end{equation}
-in the optimization criterion. $W_{\vf_o}$ is a weight to indicate how much operator
-force is acceptable (if ${\vf_o}_d = 0$) or how much operator force error is
-acceptable.
-Force damping can be implemented by adding a similar ``soft'' constraint:
+Force damping can be implemented by adding the following ``soft'' constraint term
+to the optimization criterion:
 \begin{equation}
@@
 \end{equation}
-to the optimization criterion.
-
-{\bf Impedance control} is implemented using online optimization by adding
+
+We can implement operator {\bf impedance control} in several ways:
+We can add 
 the impedance equations as ``hard'' equality constraints (to achieve them exactly)
@@
 
+We can alternatively just set the desired operator contact force to:
+\begin{equation}
+{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
+\end{equation}
+Unfortunately, this involves inverting the Jacobian matrix, which may introduce
+problems.
+Assuming that the operator directly applies exoskeleton joint torques, 
+we can get rid of the Jacobian matrix, and this becomes
+the equality constraint
+\begin{equation}
+\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+or is added to the optimization criterion:
+\begin{equation}
+C_e(\vx) = C_r(\vx) 
++ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
+\end{equation}
+with
+\begin{equation}
+{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+
 {\bf Mapping external forces to operator forces} can be done either with the
@@
 The use of online optimization makes sharing control with an operator easy.
-Dimensions (even powered dimensions) that are under the operator's control have
+Dimensions (even actuated dimensions) that are under the operator's control have
 low weights in the online optimization. It is also easy to map operator controls
@@
 weight matrices similar.
+
+\subsection{Full Atlas robot implementation.}
+
+Discuss or point to DRC papers by multiple groups.
 
@@
 
+We have not seen a lot of discussion of state estimation of exoskeletons in
+the literature.
+We will discuss our approach to state estimation on humanoid robots here.
+
+We implemented two kinds of Kalman filters for state estimation.
+Low level state estimation is done by an extended Kalman filter that takes into
+account joint position, actuator load measurements, actuator commands,
+and the dynamic model of the robot,
+and measured contact forces (feet and wrists)~\cite{}.
+For higher level (overall) state estimation,
+we developed a center of mass location and velocity estimator, and its
+application in full body inverse dynamics control, fall detection
+and fall prevention for humanoid robots and exoskeletons~\cite{}.
+
+We use Linear
+Inverted Pendulum Model (LIPM) dynamics with an offset to predict
+the center of mass motion. This offset can be interpreted
+as a modelling error on the center of mass position, or an
+external force exerted on the center of mass of the robot,
+or a combination of both. The center of mass estimator was
+implemented on our Atlas humanoid robot. During the DARPA
+Robotics Challenge Finals, it successfully saved our robot from
+falling on two occasions, and made us the only competitive team
+without a fall or need for a human rescue among all teams.
+
+Unscented,
+Particle filters
+
 DISCUSS HOW TO HANDLE SENSOR NOISE
@@
 
-\section{Robustness to parametric modeling error}
-
-\section{Robustness to unmodeled dynamics}
-
-\subsection{Example unmodeled dynamics}
-
-Examples:
-\begin{itemize}
-\item
-A) hydraulic actuator dynamics.
-\item
-B) delay. 
-\item
-C) elastic straps sinusoidal oscillation.
-\item
-D) loose straps (bang coast bang oscillation)
-\end{itemize}
-
-\subsection{1 joint rotary example.}
-
-\subsection{2 joint rotary example: stance.}
-
-\subsection{2 joint rotary example: swing.}
-
-\subsection{4 link body example.}
-
-\subsection{Full Atlas robot.}
+\section{Robustness to parametric modeling error and unmodelled dynamics}
+
+We need to be robust to parameter modeling errors such as the wrong mass
+or moment of inertia for an exoskeleton part, variations in the operator
+parameters, and variations in attachments to the exoskeleton and payload.
+Examples of typically unmodeled dynamics (non-parametric modeling error) include:
+backlash, joint play, structural deformation, flexible wires and hoses,
+actuator dynamics, processing delay,
+elastic operator tissue and straps, and 
+loose or slack operator tissue and strap.
+
+There are two questions to focus on:
+
+\subsection{How robust are these control schemes to parametric and non-parametric
+modeling error?}
+
+This question needs to be addressed using simulation, and is deferred to a future
+white paper [robust.pdf].
+
+\subsection{How can we make control more robust?}
+
+One way to understand control systems is to separate 1) what happens when there
+are no errors and feedback control is not active, and 2) what happens in response
+to errors (feedback control). 
+What happens when there are no errors in model-based control is 
+feedforward control, which does not have stability issues (except for some ignorable
+technical issues). If your model is good, you get good performance. If your
+model is bad, you get larger errors and have to rely on feedback control.
+So feedforward control is robust, and we don't have to worry about it. We do have
+to worry about how to get good models, and what to do about stuff that varies:
+contacts, operator body, operator behavior, payloads, attachments, etc.
+We do have to worry about performance.
+
+On the other hand, feedback control can lead to steady state or growing
+oscillation, which we colloquially refer to as being unstable or instability.
+Since walking and running are also oscillations which we refer to as stable
+behaviors, the technical meaning of the word ``stable'' is more complex than
+just ``it keeps moving'' or ``it doesn't do what the operator wants''.
+In any event, we don't want oscillations.
+
+The best way to avoid oscillations is to manually design a conservative 
+independent joint feedback controller.
+This way it can be tested independently of any
+feedforward control.
+Unfortunately, inverse dynamics and online optimization-based control mix
+feedforward and feedback control so we cannot use the recommended, well established,
+and safe approach of manual conservative independent joint feedback control design.
+
+What to do about this to achieve robust online optimalization-based control is
+currently a research issue. In actual implementations, skilled engineers manually
+tune optimization weights and the various constraints that are active to
+make this control work reliably. The fact that this has been done on many actual
+humanoid robots in many situations is comforting. The fact that none of these
+implementations have moved with the speed and agility of a top percentile
+human athlete is disconcerting.
+
+We can ``verify'' designs using simulation, but making accurate simulations of
+the effects of contact and unmodelled dynamics is also a current research issue.
+There is no getting around the fact that our operators will be like the test
+pilots of early planes. The exoskeletons may crash, flail wildly, or oscillate.
+We need to plan for operator safety from day 1 of operation, not try to retrofit
+it later.
+
+If the robustness of online optimization is not adequate, an alternative is
+combining feedforward and feedback control into one function and using gradient
+descent in simulation to improve the control robustness.
+One technique to do this is multiple-model design~\cite{acc12}.
+
+Note that adaptive or learning
+control does not promise to improve the kind of robustness exoskeltons need. 
+These approaches promise
+to improve performance, and try to track (slowly) changes in the operator, exoskeleton,
+or contacts to maintain performance. If something changes quickly, we need robust
+control design.
 

### 2015-09-22 exos/exo-ul7.tex
finished hal

\subsection{EXO-UL7}
\label{exo:exo-ul7}

keywords: model-based control; strength augmentation;\\

\begin{figure}[ht]
  \centering
  \includegraphics[width=4.0in]{exos/figs/exo-ul7.png}
\end{figure}

Although this study focuses on lower-body exoskeltons, we include the EXO-UL7 7 DOF upper-extremity exoskeleton for its novel actuation scheme and principled design.

-------------------------------------------------

(Called CADEN - "cable driven dexterous exoskeleton for neurorehabilitation" in [[EXOul7design]])

Design:

7 DOF model; cable-actuated; proximal motor placement and distal cable-pulley reductions for low inertia, high stiffness links with backdrivable transmissions with zero backlash.  "Enables full glenohumeral, elbow, and wrist joint functionality." -- [[EXOul7design]]

% set the Human Machine Interface (HMI) at the neuromuscular level ... as one of the primary command signals... Use sEMG signals.  -- http://bionics.seas.ucla.edu/research/exoskeleton_device_3.html

[[EXOul7design]] contains data from a dynamic / kinematic study of human motion during everyday tasks in order to determine design requirements for exo.

3 lvls of safety mechanisms: 1) built in mechanical -- phys. stops, 2) electrical -- three wearer e-stops including an observer e-stop, an enable button which must be held, and a foot switch e-stop, and 3) software -- redundant position sensing (pots, Midori, Fullerton; shaft encoder, HP), one at either side of power train, monitor both joint motion and motor position. -- [[EXOul7design]]

Control bandwidth -- targeted 10Hz based on achievable frequency range of human arm which is btwn 2Hz and 5Hz.  They found shoulder joint had 1st resonant mode at 6 Hz. -- [[EXOul7design]]

Weight: 3.5 and 6kg from links 1 and links 2-7 (resp.). -- [[EXOul7design]]

Articulation: 7 single axis revolute joints.  Use three pulleys (90 and 180 deg configurations) at joints to keep cable length constant. Stack of pulleys required -- two pulleys representing agonist muscle groups and two for antagonist -- [[EXOul7design]]

Singularities: Design movable joint mechanisms that allow adjustment of singularity within 15deg based on user preferences. -- [[EXOul7design]]

Power transmission: shafts, cables, fluid lines, and gear trains.  Use pulleys for speed reductions and backdrivability.  2-stage pulleys in proximal joints 1-4 and single stage for more distal (5-7). -- [[EXOul7design]]

Motors for joints 1–4 were mounted on the stationary base, achieving a 60% reduction in overall weight of the moving parts. The remaining three motors, whose torque requirements are substantially less, were positioned on the forearm -- [[EXOul7design]]

Control Strategy:

Model-based control -- "proposed HMI takes advantage of the electro-chemical-mechanical delay, which inherently exists in the musculoskeletal system, between the time when the neural system activates the muscular system and the time when the muscles generate moments around the joints. The myoprocessor is a model of the human muscle running in real-time and in parallel to the physiological muscle.   During the electro-chemical-mechanical time delay, the system will gather information regarding the physiological muscle’s neural activation level based on processed sEMG signals, the joint position, and angular velocity, and will predict using the myoprocessor the force that will be generated by the muscle before physiological contraction occurs. By the time the human muscles contract, the exoskeleton will move with the human in a synergistic fashion, allowing natural control of the exoskeleton as an extension of the operator's body." %-- http://bionics.seas.ucla.edu/research/exoskeleton_device_3.html

Say they use neuro control strategy to avoid having to trigger motion based on human movement / force (required in position, force-impedance control) -- [[EXOul7design]]

References:

Cavallaro E., J. Rosen, J. C. Perry, S. Burns, B. Hannaford, Hill Based Model as a Myoprocessor for a Neural Controlled Powered Exoskeleton Arm – Parameter Optimization, Proceedings of the 2005 IEEE international Conference on Robotics and Automation, ICRA 2005, pp. 4525 – 4530, Barcelona Spain, April 2005

Rosen J,, J. C. Perry, N. Manning, S. Burns, B. Hannaford, The Human Arm Kinematics and Dynamics During Daily Activities – Toward a 7 DOF Upper Limb Powered Exoskeleton, - ICAR 2005 – Seattle WA, July 2005. [ CP19]

Perry J.C., J. Rosen, Design of a 7 Degree-of-Freedom Upper-Limb Powered Exoskeleton Proceedings of the 2006 BioRob Conference, Pisa, Italy, February, 2006.

Cavallaro E., J. Rosen, J. C. Perry, S. Burns, Myoprocessor for Neural Controlled Powered Exoskeleton Arm, IEEE Transactions on Biomedical Engineering, pp. 2387-2396, Vol. 53, No. 11, November 2006

Perry J. C., J. Rosen, S. Burns, Upper-Limb Powered Exoskeleton Design, IEEE Transactions on Mechatronics, Volume 12, No. 4, pp. 408-417, August 2007 <<EXOul7design>>

Rosen J., and J.C. Perry, Upper Limb Powered Exoskeleton, Journal of Humanoid Robotics, Vol. 4, No. 3 (2007) 1–20

Miller, L.M. and J. Rosen, 2010." Comparison of multi-sensor admittance control in joint space and task space for a seven degree of freedom upper
limb exoskeleton," 3rd IEEE RAS and EMBS International Conference on Biomedical Robotics and Biomechatronics (BioRob). <<EXOul7admitJointTaskCmp>>

Wen, Y., J. Rosen, and L. Xiaoou, 2011." PID admittance control for an upper limb exoskeleton," American Control Conference (ACC).  <<EXOul7admit>>

-------------------------------------------------

\subsection{Assessment and Recommendations}

The EXO-UL7 is a well-designed upper body exosekelton that is experimenting with new sEMG based control mechanisms.  The control approach is not as refined and needs to be developed further before field applications due to reliance on sEMG data.  
% As mentioned in Sections~\ref{exo:bleex} and \ref{exo:hal}, there are advantages to .

The figures in this section were obtained from \cite{EXOul7design2007,EXOul7pidAdmit2011}.  Materials presented are based on the references above.

### 2015-09-22 control.tex

@@
 sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
-the state estimation produces not only perfect positions and velocities but
+state estimation produces not only perfect positions and velocities but
 also perfect acceleration estimates. In the case 
@@
 
-We implemented two kinds of Kalman filters for state estimation.
-Low level state estimation is done by an extended Kalman filter that takes into
-account joint position, actuator load measurements, actuator commands,
-and the dynamic model of the robot,
-and measured contact forces (feet and wrists)~\cite{}.
+We use extended Kalman Filtering (EKF) to perform state estimation.
+Currently, using the full body dynamics (equation~\ref{eq:dynamics2})
+and any addtional constraint equations for ground contacts (feet on the ground):
+\begin{equation}
+\mJ_w \ddot{\vq} + \dot{\mJ}_w \dot{\vq} = 0
+\end{equation}
+is too expensive to produce both a forward prediction and a linearized model
+about the current state in the desired 1ms time step.
+Therefore we approximate the dynamics by separating the center of mass dynamics
+from the joint dynamics and use two separate extended Kalman filters:
+a filter for the center of mass dynamics, and a filter for the joint dynamics.
+We use an orthogonal (QR) decomposition to project motion into the orthogonal
+complement of the contact Jacobian where the ``external'' dynamics (the dynamics
+of the center of mass) are separate from the ``internal'' dynamics (the motion
+of the joints).
+
+Joint velocities are estimated using actuator position sensors, actuator load
+sensors, actuator commands, a single high
+quality IMU at the pelvis, multiple low quality (but cheap) MEMs IMUs
+spread throughout the robot. We also take into account feet and hand
+contact forces, which
+are measured. We use the full internal dynamics as the process model,
+assuming the root dynamics are known. See Chapter 3 of~\cite{Xinjelefu-thesis}.
+The internal dynamics are linearized and covariances updated
+by a separate process at a 100Hz rate.
+The joint positions are predicted by integrating the estimated velocities.
+
 For higher level (overall) state estimation,
@@
 and fall prevention for humanoid robots and exoskeletons~\cite{}.
-
+The center of mass filter is modeled as a multiple model EKF with contact switching
+to handle changes in support: single support, double support, and off the ground.
+We use the robot joint sensing (kinematics) to create a virtual sensor that measures
+center of mass position and velocity relative to a stance foot.
 We use Linear
@@
 or a combination of both. The center of mass estimator was
-implemented on our Atlas humanoid robot. During the DARPA
-Robotics Challenge Finals, it successfully saved our robot from
+implemented on our Atlas humanoid robot. 
+This estimator is especially helpful when compensating
+for unplanned slow changing external forces applied at
+unknown locations on the robot, which is quite likely when
+operating in tight spaces. It also handles relatively small
+dynamic forces well when walking, e.g. dragging a tether
+or pushing through a spring loaded door. Thanks to
+the estimator, very little tuning is done for our mass model.
+During the DARPA
+Robotics Challenge Finals
+where no safety belay was
+allowed, our early warning system
+successfully saved our robot from
 falling on two occasions, and made us the only competitive team
@@
 
-Unscented,
-Particle filters
-
-DISCUSS HOW TO HANDLE SENSOR NOISE
-
+{\bf Safety Code: Fall Prediction:}
+The most significant contribution of the external force estimator is that it
+can detect when a large external force is being applied that might
+push the robot over.
+We compute a ``corrected capture point'' (CCP)~\cite{capture},
+which is an offset to the current capture point.
+The offset 
+takes into account the estimated external force, represented as an offset
+to the center of mass.
+The corrected capture point getting close to the boundary of the polygon of
+support warns the controller that the robot might fall if the external
+force is maintained.
+We can also compute the corrected capture point assuming that the
+external force follows a known time course plus an estimated constant offset,
+or steadily increases or decreases for a fixed time interval
+based on an estimated derivative.
+We assume the external force is due to the robot's actions, and not
+due to external disturbances such as wind, a moving support platform
+(although earthquakes do happen at the DRC location in California),
+or external agents pushing on the robot.
+The current
+behavior is stopped and the robot ``frozen'' in place.
+
+A derivation of the corrected capture point starts with LIPM dynamics
+augmented with a true center of mass offset and a true external force:
+\begin{equation}
+\ddot{c} = (c + c_{offset} + f_{ext}*z/mg - cop)*g/z
+= (c + \Delta - cop)*g/z
+\end{equation}
+where $c$ is the location of the center of mass projected on the ground
+plane,
+$cop$ is the center of pressure
+location in that ground plane, and $\Delta$ is the sum of the true center of mass
+offset from the modeled center of mass and any external horizontal force.
+Our extended Kalman filter estimates $\hat{c}$, $\hat{\dot{c}}$, and
+$\hat{\Delta}$,
+taking into account the
+current center of mass height $z$. We assume a constant center of mass height
+in estimating the corrected capture point based on the estimated capture point
+as described in~\cite{capture}:
+\begin{equation}
+\widehat{CCP} = \widehat{CP} + \hat{\Delta} 
+= \hat{c} + \hat{\dot{c}}\hat{z}/g + \hat{\Delta}
+\end{equation}
+
+For dynamic walking, the corrected capture point goes beyond the
+current single foot support polygon during swing phase and is captured by
+the touchdown foot. One way to predict a fall is to use the
+swing foot time to touchdown and predict if the current
+capture point is within the support polygon of possible or desired
+footholds. This approach is complicated because it depends
+on the controller. Our simpler solution is to use a heuristic
+that detects a fall only if the corrected capture point
+is outside of the current support
+polygon for a continuous period of time. The time threshold was set to
+0.6 seconds after extensive testing. As soon as a fall is
+predicted, there are several things the humanoid can do, such
+as using angular momentum to maintain balance, or changing the
+current or future foot placements.
+We have implemented a simple
+step recovery controller that works in single support where
+the robot corrects the next foot placement and timing
+using a simple heuristic to avoid self collision.
+We would only step left if the robot is falling left, and similarly
+for right steps. We avoid crossing the legs in using foot placement
+to balance laterally.
+
+\begin{figure}[t]
+  \centering
+  \hbox{}
+      \includegraphics[height=5cm]{figs/door1-5.png}
+%      \includegraphics[height=5cm]{figs/door2-1c.png}
+      \hbox{}
+      \includegraphics[height=5cm]{figs/door1-3.png}
+      \hbox{}
+      \includegraphics[height=5cm]{figs/door-stuckc2.png}
+      \hbox{}
+      \caption{Successful sidestepping through the door (left, middle)
+        and the failure in the DRC rehearsal (right) in which the protective
+        cage (black padding) for the head is against the white door frame.}
+      \label{figDoor}
+\end{figure}
+
+\begin{figure}[t]
+      \centering
+      % \includegraphics[width=0.48\textwidth]{figs/door-traces.png}
+\subfigure
+{\includegraphics[width=0.48\textwidth]{figs/cop_x1.pdf} \label{subfig:ext_cop_x}}
+%\subfigure
+%{\includegraphics[width=0.48\textwidth]{figs/cop_y.pdf} \label{subfig:ext_cop_y}}
+\subfigure
+{\includegraphics[width=0.48\textwidth]{figs/ext_f1.pdf} \label{subfig:ext_f}}
+\caption{
+Atlas was caught on the door frame when sidestepping through it during
+the DRC rehearsal.
+The walking controller delayed liftoff and remained in double
+support when the external force estimator detected a large change
+in the estimated external force in the robot's sideways
+direction ($F_x$, through the door).
+The single support phase is shown by the shaded area, and the black dashed
+lines indicates the planned liftoff time.
+The estimated CoM is the sum of the model CoM and the estimated CoM offset.
+}
+\label{figDoorError}
+\end{figure}
+
+{\bf Results of Safety Code: Robot caught on door frame:}
+In the DRC rehearsal, the robot was caught on
+the door frame when sidestepping though (Figure~\ref{figDoor}).
+The walking
+controller detected an anomaly in the estimated external force in the
+sideways direction ($F_x$),
+delayed liftoff and remained in double support, and stopped the
+current behavior
+to allow for manual recovery (Figure~\ref{figDoorError}).
+
+{\bf Results of Safety Code: Manipulation Error:}
+For the manipulation controller, the robot is always assumed
+to be in double support, and the support polygon
+is computed by finding the convex hull of the foot corner
+points (light green in Figure~\ref{figDrillCOP}),
+computed using forward kinematics.
+To prevent the robot from falling during
+manipulation, we require the corrected capture point to be within a
+subset of the support polygon called the safe region, 
+(dark green in Figure~\ref{figDrillCOP}),
+When the corrected capture point escapes the safe region,
+a freeze signal is sent to the manipulation controller, and it
+clears all currently executing joint trajectories and freezes the robot
+at the current pose, with the balance controller still running.
+
+During our second
+run in the Finals, our right electric forearm mechanically failed
+when the cutting motion was initiated for the drill task. The
+uncontrolled forearm wedged the drill into the wall and
+pushed the robot backwards.
+The controller stopped the behavior (a freeze),
+and saved the robot from falling (Figure~\ref{figDrillCOP}).
+The operator was then able to
+recover from an otherwise catastrophic scenario.
+
+The time plot in Figure~\ref{figDrillCOP} shows candidate fall predictors
+during this event. We can eliminate some candidate fall predictors
+easily. The center of mass (CoM) (and a ``corrected'' CoM (not shown))
+usually provide a fall warning too late, because the CoM velocity is not
+included. The capture point (CP) does not include information about
+external forces. The center of pressure (CoP) is 
+noisy and gives too many false alarms. It can warn of 
+foot tipping, but it is less reliable about warning about robot falling,
+which is not the same thing as foot tipping in a force controlled robot
+or if there are non-foot contacts and external
+forces. In this plot, we see that the CoP moves away from the safe region
+during recovery, predicting that the robot is falling again, while the
+corrected capture point (CCP) moves towards the interior of the safe region.
+
+\begin{figure}[t]
+      \centering
+      \includegraphics[height=30mm]{figs/near-fall2c.png}
+      \includegraphics[height=30mm]{figs/drill-cop1.png}
+      \includegraphics[height=30mm]{figs/drill-cop2.png}
+      \includegraphics[height=30mm]{figs/drill-cop3.png}\\
+\includegraphics[width=0.48\textwidth]{figs/cp_time_xes.pdf}
+      \caption{Top Left: Robot posture after 
+error detection.
+Top Right: Black trace: The corrected capture point (CCP)
+up to the error detection,
+Red trace: The CCP during the ``frozen'' period, and Blue trace: The CCP moves
+back to the center of the polygon of support during manual recovery.
+Bottom: Plots of candidate fall predictors in the fore/aft direction
+during this event. The black vertical dashed lines mark the freeze time and
+the start of manual recovery.}
+      \label{figDrillCOP}
+\end{figure}
+
+Unscented and particle filters could be used instead of extended
+Kalman filtering. It may well be the case that unscented filtering
+outperforms extended Kalman filtering.
+
+\subsection{Current state estimation approaches in exoskeletons}
+
+We see that sensitivity amplification control uses acceleration as an input.
 Estimating acceleration using double differentiation with low pass filtering
@@
 
-\subsection{Full Atlas robot.}
-
 \section{Robustness to parametric modeling error and unmodelled dynamics}
 
-We need to be robust to parameter modeling errors such as the wrong mass
+We need to be robust to parameter modeling errors such as the wrong mass,
+center of mass location, joint location,
 or moment of inertia for an exoskeleton part, variations in the operator

### 2015-09-22 design-goals.tex

\section{What are some design philosophy alternatives?}

\subsection{``Invisible'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally
and exerts little or no force on the operator?

This design goal
could be achieved through active control using feedback of the
operator-exoskeleton forces and/or high quality prediction of what
the operator will do next. Additional gravity, friction, actuator, and
inertial compensation (inverse dynamics) can improve performance.

\subsection{``Natural'' exoskeleton}

Can we build an exoskeleton allows the operator to behave naturally?
The operator can feel the exoskeleton, but can overpower it when necessary.

One way to achieve this design goal is to have extremely light attachments
to the limbs, and put all heavy exoskeleton components on the
torso and head. The limbs can be physically driven by the operators muscles
with little ``drag'' on the operator.
Bypass valves and clutches could allow the operator to physically ``take over''
and push the relatively lightweight limb mounted parts of the exoskeleton around.

A complementary way to achieve this goal is through active control. Low impedance
(torque source) actuation makes combining active control with operator backdriving
easier.

A ``natural'' exoskeleton is easier to build than 
an ``invisible'' exoskeleton.

\subsection{``Symbiotic'' exoskeleton}

Can we build an exoskeleton that allows the operator/exoskeleton team
to be effective, but not necessarily behave naturally? 
The operator needs to learn to ``fly'' the suit,
just as operators learn to operate parachutes,
wing suits, diving equipment, high and low temperature protective
suits, firefighting equipment, and high altitude flight
and parachute suits.

An advantage of this approach is that 
the exoskeleton can adapt its behavior to the task.
For example, for jumping, the exoskeleton can operate like a pogo 
stick:\\
\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
and for running it could act like jumping stilts:\\
\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},

A ``symbiotic'' exoskeleton is easier to build than an ``invisible'' or ``natural''
exoskeleton.

### 2015-09-23 control.tex
cga updates

@@
 Algorithms\\
-(Draft 2.2)}
+(Draft 2.3)}
 
@@
 
-A later white paper will address robustness to modeling error and unmodeled
-dynamics using simulation.
+A later white paper will evaluate using simulation
+the various control approaches' robustness to modeling error and unmodeled
+dynamics.
 
@@
 
-\subsection{Low Level Control}
+\subsection{Low Level Control (LLC)}
 
@@
 
-\subsubsection{What degress of freedom are controlled?}
+\subsubsection{LLC: What degrees of freedom are controlled?}
 
@@
 actually combinations of joints, or to control points on the robot in Cartesian
-or other coordinates. We refer to these virtual degress of freedom as {\bf synergies.}
+or other coordinates. We refer to these virtual degrees of freedom as {\bf synergies.}
 A complex synergy is the center of mass. Higher level control is made easier by
@@
 
-\subsubsection{How is a degree of freedom controlled?}
-
-It is useful for human programmers
-to provide them with idealized abstractions of control of a single degree of
-freedom. Here are some common alternatives:
+\subsubsection{LLC: How is a degree of freedom controlled?}
+
+It is useful to provide human programmers
+with idealized abstractions of control of a single degree of
+freedom. Here are some common alternatives that could be provided by low level control:
 \begin{itemize}
@@
 
-\subsubsection{Feedback control}
-
-PID or more complex feedback control laws are implemented at this level.
-
-\subsubsection{Feedforward control}
+\subsubsection{LLC: Feedback control}
+
+PID or more complex feedback control laws are implemented by low level control.
+
+\subsubsection{LLC: Feedforward control}
 
 Gravity, friction, actuator dynamics, and other feedforward disturbance
-compensation are implemented at this level, as well as any inverse dynamics
+compensation are implemented by low level control, as well as any inverse dynamics
 or other types of feedforward reference tracking control.
 
-\subsubsection{State estimation}
-
-State estimation is done at this level to compensate for sensor noise.
-Disturbance observers for model errors and contact/wind forces
-are implemented at this level.
-
-\subsubsection{Existing examples}
+\subsubsection{LLC: Existing examples}
 
@@
 
-In this architecture, state estimation to deal with sensor noise and incomplete or
+In this architecture, state estimation to deal with sensor noise,
+disturbance estimation, and incomplete or
 redundant sensing is a separate process and is as independent as possible
@@
 This makes evaluating different control methods much easier.
+Disturbance observers for disturbances such as
+moving support surfaces, contact or wind forces, and model errors
+are also implemented.
 
@@
 or combine Phoneme, Word, and Behavior Selection.
-This is not a major change.
+Neither of these is a major change.
 
@@
 
-The dynamics of a revolute exoskeleton are:
+The dynamics of a revolute exoskeleton standing on the ground are:
 \begin{equation}
@@
 
-[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
-Note that we are ignoring the operator dynamics and operator muscle and tissue
-forces resulting in operator joint torques. We will add these in later (or maybe
-we won't ???).
+Note that we are ignoring the human operator dynamics and operator muscle and tissue
+forces resulting in operator joint torques. Intent prediction will model these
+effects [intent.pdf].
 
@@
 the state vector and adding additional equations.
-
-\section{Test case: Atlas humanoid robot}
-
-Gazebo and SDFast models.
 
@@
 
-Recognizing/estimating phase goes here.
-Phase Triggered References could go here.
-\begin{verbatim}
-H.Kawamoto and Y.Sankai, “Power Assist Method Based on Phase
-Sequence Driven by Interaction between Human and Robot Suit,”
-Proceedings of the IEEE International Workshop on Robot and Human
-Interactive Communication, pp. 491–496, 2004.
-\end{verbatim}
-
-Motion capture, motion libraries goes here.
-
-Often finite state machines whose states correspond to left stance, double support 1,
-right stance, and double support 2 are used to select behaviors. Gait transitions
-can be triggered by foot touchdown and liftoff events, or
-displacement of the Center of Mass (CoM).
-
-Central pattern generator
-
-Iterative learning control to acquire $\vtau_{ff}(t)$ for desired trajectories.
-
 \section{Active gravity compensation}
@@
 
-A key question is where does desired acceleration come from?
-\begin{verbatim}
-  Operator force control.
-  Receding Horizon Control of optimal trajectory of 
-    idealized LIPM+flywheel model (Atkeson)
-  intent prediction (see above)
-  neuromuscular model (Geyer, Wang, Van de Panne)
-  hybrid: Optimal control + neuromuscular model (Atkeson NSF Proposal)
-    - NMM provides bias through changes to optimization criterion
-    - NMM provides hard constraints
-    - NMM replaces components of model-based optimization scheme
-\end{verbatim}
+A key question is where does the desired acceleration $\ddot{\vq}_d$ come from?
+This is discussed in the companion paper [intent.pdf].
 
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + \mJ_o^{\tr} {\vf_o}_d
-+ \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
+\vtau = \vtau_{\invdynw} + \mJ_o^{\tr}(\vq) {\vf_o}_d
++ \mJ_o^{\tr}(\vq) \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
 \end{equation}
@@
 \begin{equation}
-\dot{\vq}_w = \mJ_w \dot{\vq}
+\dot{\vq}_w = \mJ_w(\vq) \dot{\vq}
 \end{equation}
@@
 \begin{equation}
-\vtau = \mJ_w^{\tr} \vf_w 
+\vtau = \mJ_w^{\tr}(\vq) \vf_w 
 \end{equation}
@@
 \begin{equation}
-\vtau = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w \dot{\vq}
+\vtau = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq) \dot{\vq}
 \end{equation}
@@
 \begin{equation}
-\mK_{\dot{\vq}} = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w
+\mK_{\dot{\vq}}(\vq) = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq)
 \end{equation}
@@
 \begin{equation}
-\mK_{\vq} = \mJ_w^{\tr} \mK_{w} \mJ_w
-\end{equation}
-So we should assign the exoskeleton joint position gains to be $\mK_{\vq}$ and
-the joint velocity gains to be $\mK_{\dot{\vq}}$ in equation~\ref{eq:impedance}.
+\mK_{\vq}(\vq) = \mJ_w^{\tr}(\vq) \mK_{w} \mJ_w(\vq)
+\end{equation}
+So we should assign the exoskeleton joint position gains to be $\mK_{\vq}(\vq)$ and
+the joint velocity gains to be $\mK_{\dot{\vq}}(\vq)$ in equation~\ref{eq:impedance}.
 
@@
 \vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
-+ \mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
++ \mJ_w^{\tr}(\vq) \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
 \end{equation}
@@
 inverse dynamics, computed torque, or feedback linearization) that also tries
-to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
+to modify the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
 using acceleration feedback. It is not clearly stated but it appears that the
@@
 We will discuss our approach to state estimation on humanoid robots here.
-
-We use extended Kalman Filtering (EKF) to perform state estimation.
+State estimation to deal with sensor noise,
+disturbance estimation, and incomplete or
+redundant sensing is a separate process and is as independent as possible
+from the control method~\cite{certainty-eq,separation-prin}.
+This makes evaluating different control methods much easier.
+Disturbance observers for disturbances such as
+moving support surfaces, contact or wind forces, and model errors
+are also implemented.
+
+For humanoid robot state estimation (for proposed exoskeleton state estimation)
+we use extended Kalman Filtering (EKF) to perform state estimation.
 Currently, using the full body dynamics (equation~\ref{eq:dynamics2})
-and any addtional constraint equations for ground contacts (feet on the ground):
+and any additional constraint equations for ground contacts (feet on the ground):
 \begin{equation}
@@
 sensors, actuator commands, a single high
-quality IMU at the pelvis, multiple low quality (but cheap) MEMs IMUs
-spread throughout the robot. We also take into account feet and hand
+quality IMU at the pelvis, and multiple low quality (but cheap) MEMs IMUs
+spread throughout the robot or exoskeleton. We also take into account feet and hand
 contact forces, which
@@
 to handle changes in support: single support, double support, and off the ground.
-We use the robot joint sensing (kinematics) to create a virtual sensor that measures
+We use the robot or exoskeleton
+joint sensing (kinematics) to create a virtual sensor that measures
 center of mass position and velocity relative to a stance foot.
@@
 as a modelling error on the center of mass position, or an
-external force exerted on the center of mass of the robot,
+external force exerted on the center of mass of the robot/exoskeleton,
 or a combination of both. The center of mass estimator was
@@
 for unplanned slow changing external forces applied at
-unknown locations on the robot, which is quite likely when
+unknown locations on the robot or exoskeleton, which is quite likely when
 operating in tight spaces. It also handles relatively small
@@
 where no safety belay was
-allowed, our early warning system
+allowed, our fall early warning system
 successfully saved our robot from
@@
 can detect when a large external force is being applied that might
-push the robot over.
+push the robot or exoskeleton over.
 We compute a ``corrected capture point'' (CCP)~\cite{capture},
@@
 The corrected capture point getting close to the boundary of the polygon of
-support warns the controller that the robot might fall if the external
+support warns the controller that the robot or exoskeleton might fall if the external
 force is maintained.
@@
 based on an estimated derivative.
-We assume the external force is due to the robot's actions, and not
+We assume the external force is due to the robot's or exoskeleton's
+movements, and not
 due to external disturbances such as wind, a moving support platform
-(although earthquakes do happen at the DRC location in California),
 or external agents pushing on the robot.
 The current
-behavior is stopped and the robot ``frozen'' in place.
+behavior is stopped and the robot or exoskeleton can safely be ``frozen'' in place.
 
@@
 0.6 seconds after extensive testing. As soon as a fall is
-predicted, there are several things the humanoid can do, such
+predicted, there are several things the humanoid or exoskeleton can do, such
 as using angular momentum to maintain balance, or changing the
@@
 step recovery controller that works in single support where
-the robot corrects the next foot placement and timing
+the robot or exoskeleton corrects the next foot placement and timing
 using a simple heuristic to avoid self collision.
-We would only step left if the robot is falling left, and similarly
+We would only step left if the robot or exoskeleton is falling left, and similarly
 for right steps. We avoid crossing the legs in using foot placement
@@
 
-We see that sensitivity amplification control uses acceleration as an input.
+We see that Sensitivity Amplification Control uses acceleration as an input.
 Estimating acceleration using double differentiation with low pass filtering
@@
 
-Role of phase estimation, behavioral clocks, and phase reset events
-such as impacts.
-
 Disturbance observers~\cite{IEEE06197032}.
@@
 parameters, and variations in attachments to the exoskeleton and payload.
-Examples of typically unmodeled dynamics (non-parametric modeling error) include:
+Examples of unmodeled dynamics (non-parametric modeling error) include:
 backlash, joint play, structural deformation, flexible wires and hoses,
@@
 elastic operator tissue and straps, and 
-loose or slack operator tissue and strap.
+loose or slack operator tissue and straps.
 
@@
 model is bad, you get larger errors and have to rely on feedback control.
-So feedforward control is robust, and we don't have to worry about it. We do have
+You may crash the exoskeleton, or fall down, but you will not oscillate
+(except in some contrived examples) and errors do not grow exponentially in
+response to feedforward control errors.
+So feedforward control is technically robust,  
+and we don't have to worry about it blowing up (except for chaotic systems).
+We do have to worry about performance.
+We have
 to worry about how to get good models, and what to do about stuff that varies:
 contacts, operator body, operator behavior, payloads, attachments, etc.
-We do have to worry about performance.
 
@@
 tune optimization weights and the various constraints that are active to
-make this control work reliably. The fact that this has been done on many actual
+make this type of control work reliably. The fact that this has been done on many actual
 humanoid robots in many situations is comforting. The fact that none of these
@@
 2) We recommend actuation that can be treated as force or torque sources.
-This is a potentual problem for revolute electric actuation, as it
+This is a potential problem for revolute electric actuation, as it
 is difficult to get necessary performance at low gear ratios (20-50)
-and fit the necessary torque sensing,
+and fit the necessary torque sensing.
 
@@
 
-4) We recommend using online optimization as part of the Lowest Level Controller
+4) We recommend using online optimization as part of the Low Level Controller
 in the form of quadratic programming. Online optimization

### 2015-09-23 exos/bleex.tex

@@
 
-keywords: sensitivity amplification; strength augmentation;\\
+keywords: sensitivity amplification; strength augmentation.\\
 
@@
 \end{equation}
-where $T$ is a vector of actuator torques, the BLEEX control torque would be
+where ${\bf T}$ is a vector of actuator torques, the BLEEX control torque would be
 \begin{equation}
@@
 \end{equation}
+The user must provide the first torque component as the exoskeleton has no actuator acting between the foot and the ground (the system is underactuated).
 
@@
 As one possibility, a sensitivity optimization approach may be paired with sEMG data to estimate user intent and provide such a filter.
+Note this type of implementation would present its own challenges considering limitations in sensing and calibration of current EMG systems.
 
@@
 The figures in this section were obtained from \cite{sesitivityAmpPaper2005}. Materials presented are based on the references above.
-
-% \end{document}

### 2015-09-23 exos/hal.tex

\subsection{HAL}
\label{exo:hal}

keywords: model-based control; predefined gait trajectory; strength augmentation; rehabilitation.\\

The hybrid assistive limb (HAL) exoskeleton is developed by the University of Tsukuba and Cyberdyne for human strength augmentation and as an assistive gait device in rehabilitation.  Although several prototypes have been developed, this section focuses on the most recent HAL-5 model (specifically the non-clicial type, HAL-5 Type-B). Though HAL-5 is a full body exoskeleton, only hip, knee joints, and ankle are actuated in the sagittal plane.  The exoskeleton powers DC motors with harmonic drives at hip, knee, ankle joints (in some models the ankle joints act as passive springs).  HAL weighs 23kg, includes an on-board AC100V battery as the power source, which is designed to support the power requirements of maximum velocity human walking and standing torque requirements.  The battery allows for 160 min of continuous operation and enables the exoskeleton to lift up to 70kg \cite{HALassist2011}.

Human operators attach to HAL at the waist by a belt, and at the calf and thigh using harness. HAL's frame does not transfer load to the ground.  Instead, HAL adjusts hip, knee and ankle torques to amplify its operator's torque. The exoskelton's sensory system includes bioelectric sensing (including sEMG), angular sensors, acceleration sensors, and center of pressure / center of gravity (COP/COG) sensors.  COP/COG sensing is provided through shoes with ground reaction force sensors.  The joint measurements are provided by potentiometers.
In at least one version of HAL the bioelectric data comes from sEMG sensors installed below the operator's hip and above the knee (on both front and back).
An IMU installed in HAL's backpack is used to estimate torso pose.

\begin{figure}[ht]
  \centering
  \includegraphics[width=3.5in]{exos/figs/hal-5_B_diagram.png}
\end{figure}

\subsubsection{Control}

HAL uses two types of control systems depending on application domain.  For gait assistance and rehabilitation, HAL uses an autonomous control system that carries the user through predefined gait trajectories by controlling knee and hip joints (the ankles behave as passive springs).  Gait phase intention is estimated from COP/COG sensors.  The exoskeleton drives the wearer to follow pre-recorded desired joint patterns.

The second control strategy, a model-based approach for human strength augmentation, estimates human intention from sEMG activity and provides power to augment torque provided by the operator.  A relatively autonomous torque assist strategy in \cite{HALmodelControlKnee2010} recognizes user intention to take a step by thresholding sEMG data.  The approach provides a knee torque response including an assitive torque component, a viscous damping component that reduces high velocity motion for safety, and a gravity compensation torque.  
%In \cite{HALvTorqueImp2002}, HAL uses myoelectric data to estimate muscle torque based on the difference between flexor and extensor muscle activities.

In \cite{HALmuscleImped2005}, an impedance control strategy controls the viscoelastic properties of HAL's knee joint from a musculoskeletal model of operator's limb moving in concert with the exoskeleton.  The controller uses sEMG sensor data to estimate muscle torque based on the difference between flexor and extensor muscle activities.  The authors note the sEMG model requires significant calibration effort.  
%
\begin{figure}[ht]
  \centering
  \includegraphics[width=6in]{exos/figs/hal_viscoelastic_control.png}
\end{figure}
%
%
Viscoelastic torques are computed according to
\[\tau_{a,i} = \alpha_i(-D_i \theta_i - K_i \theta_i),\]  
based on a variable gain set by $\alpha_i$.  The net knee actuator torque is
\[\tau_i = \tau_{a,i} + \tau_\mu + \tau_c .\]
The torque includes a $\tau_c$ term compensating for mechanical (actuator) impedance and $\tau_\mu$, a scaled version of the human torque (as estimated  from sEMG data).  The $K_i$ and $D_i$ terms in $\tau_{a,i}$ are viscoelastic parameters (based on the operator's muscles) in the human-exoskeleton model.  These parameters are estimated on-line using a weighted least-squares method.  In this case, the angular velocity data required for the impedance controller is estimated from a state observer.

While experiments in \cite{HALmuscleImped2005} only consider leg swing-up and swing-down, a similar control approach in \cite{HALvTorqueImp2002} switches the dynamics based controller to compensate for swing and stance phases in walking gait.  In this case, the gait transitions are detected by thresholding foot sensors and required angular velocity (and acceleration) data is determined by numerically differentiating angular encoders.

\subsection{Assessment and Recommendations}

Compared to sensitivity amplification methods, HAL's use of EMG data allows it to estimate intention without amplifying external disturbances.  This is ideal for combat scenarios.  However, sEMG data is extremely difficult to work with due to high filtering and calibration requirements.  A hybrid strategy which uses sensitivity amplification and sEMG data (possibly thresholded) to filter user intention could prove effective.  As an alternative, research into nano-fabrication and MEMs technologies has produced new ``robotic skin'' sensing arrays. This technology may allow direct measurement of interaction forces (i.e. estimating pressure / force and exoskeleton contact locations), which has not been feasible in the past.  Such measurements could (potentially) filter external disturbances from user generated input to estimate intent.

%\bibliographystyle{plain}
%\bibliography{exos/hal}

The figures in this section were obtained from \cite{HALassist2011,HALmuscleImped2005}.  Materials presented are based on the references above.

### 2015-09-23 exos/xor.tex

\subsection{XoR}
\label{exo:XoR}

keywords: strength augmentation; hybrid actuation\\

\begin{figure}[ht]
  \centering
  \includegraphics[width=4.0in]{exos/figs/xor.png}
\end{figure}

muscular weakness

Purpose: prototype developed for "postural control of elderly people and persons with mobility disability" -- [[llimbCtrlReviw]]

Design:

"implements a hybrid driving concept combining pneumatic artificial muscles and electric motors together [[XoRdesign]]; [[XoRkinemExtraction]] ...motion intention is estimated based on EMG signals ... user posture based on joint angles and ground reaction forces."  -- [[llimbCtrlReviw]]

Control Strategy:

Model-based control -- "...Pneumatic artificial muscles provide gravity compensation and electric motors serve the dynamic compensator. The total generated torque tracks a desired value calculated on the basis of a bio-mechanical model... Only simulation results were available to prove the efficacy of this assistive strategy working with XoR." -- [[llimbCtrlReviw]]

References:

S. Hyon, J. Morimoto, T. Matsubara, T. Noda, M. Kawato, Xor: Hybrid drive exoskeleton robot that can balance, in: Intelligent Robots and Systems (IROS), 2011 IEEE/RSJ International Conference on, IEEE, 2011, pp. 3975–3981. <<XoRdesign>>

J. Morimoto, T. Noda, S. Hyon, Extraction of latent kinematic relationships between human users and assistive robots, in: Robotics and Automation (ICRA), 2012 IEEE International Conference on, IEEE, 2012, pp. 3909–3915. <<XoRkinemExtraction>>

***** Data from design paper [[XoRdesign]]

HMI: 
Brain-machine interface.

Acutation:  
Hybrid drive - pneumatic muscles with electrical motors arranged in "optimal ways."

Design criteria:

Precise torque control; high backdrivability; force/velocity profile for rehab and compensation; lightweight; quasi-autonomy for rehab; autonomous posture control.

Precise torque control -- they believe this is most important feature to mobility assist/rehab programs as softness, compliance, and safety are defined in force/toque space;

safety thru passive backdrivability;

air muscles -- use them because there is negligable stick-slip; better controllability but gnereated force decreases quadratically with contraction (vanish around 30% contraction); optimize arrangment of AM to generate desired forces in specific confgs.  Use air muscle (AM) as gravity compensator and electric motor as dynamic compensator.  AM has high power density; good for high-load torque but delayed due to air compressibility.  There is also steady-state error.  Electric motors (without gear reduction) have high response torque but can only generate for short periods.  Hybrid drive sums the two to produce desired torque profile.  Can combine AM with small electric servos or hydraulic servo actuators.  Small servos without high gear reduction; DC (brushed and brushless).

Use AM as a unilateral drive for gravity comp combined with bilateral electric motors (motion control); generates precise high torque.  The air compressor / hydraulic pump can be smaller and lighter electric batteries.

10 DOF; 6 active (all FE joints with hybrid drive) and 6 passive joints (hip ankle AA, hip rotation);

Transmission - AM connected to pulleys via tendons.  Rubber muscle (MDSP-40) with EP regulators mounted in backpack; testing constant pressure and cam to reduce weight required for backpack air valves.  Bruschless DC geared motors (4.7 A rating; 0.12Nm output; gear ratio 57.5 for backdrivability; for short operation motor can generate 34.5 Nm/joint) with belt and pulleys

Range of motion of 120 deg and load capacity 150 Nm.  Weight 33kg of which toro + cables ~15kg. Experiments show insufficient torque and need for bi-directional pneumatic acuation at hip joints.

Sensing -- Rotory encoders at each joint; IMU in backpack; load cells at feet; network connected with servers processing sensory info from EMG, NIRS and EEG.  Controller runs at 1ms on Linux PC.

Control:

PLANNED: Torque control; apply compliant control mechanisms designed for full humanoid to exoskeleton control

S. Hyon, “Compliant terrain adaptation for biped humanoids without
measuring ground surface and contact forces,” IEEE Trans. Robotics,
vol. 25, no. 1, pp. 171–178, 2009.

S. Hyon, “A motor control strategy with virtual musculoskeletal
systems for compliant anthropomorphic robots,” IEEE/ASME Transactions
on Mechatronics, vol. 14, issue 6, pp. 677–688, 2009.

In simulation:
Task-space controller; The paper describes CoM / CoP control strategy based on GRF.  Achieve desired GRF using PD feedback to track desired COM trajectory.  Full-body task space controller superimposed with a joint stiffness controller applying feedback tracking ref joint trajectories.  Simulations reveal max torque of 150 Nm (effective torque of 80Nm) for HFE and KFE joints.  Plan to include EMG sensors;

------------------------------------------------

\subsection{Assessment and Recommendations}

The XoR prototype uses a novel hybrid actuation scheme to achieve both fast peak torque response and high sustained torque.

% \bibliographystyle{plain}
% \bibliography{exos/xor}

The figures in this section were obtained from \cite{xorDesign2011}. Materials presented are based on the references above.

### 2015-09-23 survey.tex

--- 2015-09-22 control.tex
@@
-% SIGN ISSUE ON FORCE FEEDBACK o and w: sort it out
-
 \documentclass[letterpaper,12pt,fullpage]{article}
@@
 \newcommand{\invdyn}{{\mbox {\tiny ID}}}
-\newcommand{\invdynw}{{\mbox {\tiny ID+Fw}}}
 
@@
 A companion paper surveys implemented exoskeleton control architectures
-and algorithms [survey.pdf].
+and algorithms~\cite{}.
 
@@
 system. Basically this boils down to fault detection, and switchover to
-an appropriate safe mode of operation.
-
-A later white paper will address robustness to modeling error and unmodeled
-dynamics using simulation.
+an approapriate safe mode of operation.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\input{design-goals.tex}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{What are some design philosophy alternatives?}
+
+\subsection{``Invisible'' exoskeleton}
+
+Can we build an exoskeleton allows the operator to behave naturally
+and exerts little or no force on the operator?
+
+This could be achieved through active control using feedback of the
+operator-exoskeleton forces. Additional gravity, friction, actuator, and
+inertial compensation (inverse dynamics) can improve performance.
+
+\subsection{``Natural'' exoskeleton}
+
+Can we build an exoskeleton allows the operator to behave naturally?
+The operator can feel the exoskeleton, but can overpower it when necessary.
+
+Bypass valves and clutches could allow the operator to physically ``take over''
+and push the relatively lightweight leg mounted exoskeleton around.
+
+``Passive'' degrees of freedom can have physical or ``virtual'' springs
+applied to assist in support (all abduct/adduct DOF, rotational DOF).
+These aids can be clutched out or turned off when necessary.
+``Virtual'' springs (springs implemented actively through exoskeleton
+actuators) can be modulated.
+
+\subsection{``Symbiotic'' exoskeleton}
+
+Can we build an exoskeleton that allows the operator/exoskeleton team
+to be effective, but not necessarily behave naturally? 
+The operator needs to learn to ``fly'' the suit,
+just as operators learn to operate parachutes,
+wing suits, diving equipment, high and low temperature protective
+suits, firefighting equipment, and high altitude flight
+and parachute suits.
+
+The exoskeleton can adapt its behavior to the task.
+For example, for jumping, the exoskeleton can operate like a pogo 
+stick:\\
+\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
+and for running it could act like jumping stilts:\\
+\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},
 
@@
 
-\subsection{Low Level Control}
-
-Low level control presents an API (Application Programmer Interface)
-for higher control levels
-to access the actuators, sensors, and other robot hardware.
-It implements high servo rate processing including feedback control and online
-optimization.
-This level provides an idealization of the exoskeleton in two ways: 
-What degrees of freedom (DOF) are controlled, and how is a degree of
-freedom controlled?
-
-\subsubsection{What degress of freedom are controlled?}
-
-The low level control API may directly map on to the exoskeleton hardware
-(physical joints,
-raw actuators, and raw sensors), or
-it may present an ``abstracted'' view or ``virtual'' hardware.
-For example, it is often useful to create virtual degrees of freedom that are
-actually combinations of joints, or to control points on the robot in Cartesian
-or other coordinates. We refer to these virtual degress of freedom as {\bf synergies.}
-A complex synergy is the center of mass. Higher level control is made easier by
-providing a virtual sensor that measures center of mass location and velocity,
-as well as providing a virtual actuator that acts at the center of mass in
-a Cartesian fashion (push horizontally and vertically in world coordinates).
-On a smaller scale, it may simplify higher level control 
-to create virtual actuators that cross
-multiple joints. These multi-joint actuators are actually implemented by commands
-to individual physical single-joint actuators.
-
-\subsubsection{How is a degree of freedom controlled?}
-
-It is useful for human programmers
-to provide them with idealized abstractions of control of a single degree of
-freedom. Here are some common alternatives:
+{\bf Low Level Control:} Present interface to actuators and
+high bandwidth feedback control, degrees of freedom,
+or synergies of either:
 \begin{itemize}
 \item
-Position control. The joint moves to a specified position and stays there independent
-of load.
-\item
-Velocity control. The joint moves at a specified velocity independent of load.
-\item
-Force or torque control. The joint provides a specified force or torque independent
-of its motion.
-\item
-Impedance control: The user specifies a desired position (for a spring), and commands
-a desired
+Position control.
+\item
+Pure force or torque control.
+\item
+Impedance control: specify desired position (for spring), and generate
 stiffness, damping, and potentially a modified inertia or moment of inertia.
-Modifying inertia is more dangerous than providing variable active stiffness
+Modifying inertia is more dangerous than providing active stiffness
 or damping.
 Or (almost equivalently),
-the user specifies a desired force or torque
-which changes according to the position, velocity,
-and acceleration of the joint.
+specify a desired force and change the force according to position, velocity,
+and acceleration.
 \item
 Online optimization of accelerations, actuator forces and torques, and contact forces
-using quadratic programming. See Section \ref{sec:qp} for further explanation.
+using quadratic programming.
 \end{itemize} 
-
-\subsubsection{Feedback control}
-
-PID or more complex feedback control laws are implemented at this level.
-
-\subsubsection{Feedforward control}
-
+One important synergy is coordinated control of the center of mass location and
+velocity for balance.
+PID or fancier feedback control laws are implemented at this level.
 Gravity, friction, actuator dynamics, and other feedforward disturbance
@@
 or other types of feedforward reference tracking control.
-
-\subsubsection{State estimation}
-
-State estimation is done at this level to compensate for sensor noise.
 Disturbance observers for model errors and contact/wind forces
 are implemented at this level.
-
-\subsubsection{Existing examples}
-
-Typically low level control is implemented
-by a high sampling rate global servo~\cite{Atlas-robot}
+Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
 or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
@@
 
-\subsection{``Phoneme'' Control}
-
-This level often provides the behavioral primitives or ``verbs'' for robot programming:
-providing mid-level controllers
+{\bf ``Phoneme'' Control:} Mid-level controllers
 for relatively constant phases of behavior.
-These behaviors often involves position or velocity targets (goto or
-stay-at X), and trajectory references.
-Example phoneme behaviors from walking: left-stance, double-support-1, right-stance, double-support-2, ....
-Example behaviors from running: left-foot-down, flight-1, right-foot-down, flight-2, ....
-Example behaviors from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ....
-Finer or multiple behavioral resolutions may be needed.
-For example a walking stance period may
+Often involves position or velocity targets (goto or
+stay-at X), trajectory references, ...
+Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
+Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
+Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
+Finer resolution may be needed. For example a walking stance period may
 be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
@@
 is performed at this level.
-
-\subsection{``Word'' Control}
-
-This level determines the control flow of a behavior: what happens next?
-Mid-level controllers are provided for sequences of ``phoneme'' behavioral
+This level often provides the behavioral primitives or ``verbs'' for robot programming.
+
+{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
 units.
-This level often involves finite state machines, timers, if-then condition testing,
-and gain or behavior
-scheduling.
-Example word behaviors: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
-and horizontal-jump(height=2m, length=5m).
+Often involves finite state machines, timers, if-then condition testing, gain
+scheduling, ...
+Examples: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
+horizontal-jump(height= 2m, length= 5m).
 Global inverse kinematics (selection of solution branch) is performed at this level.
-
-\subsection{Behavior Selection}
-
-The highest level of control in the recommended architecture is behavior selection,
-which answers the question:
-``What should I do next?''
-This level selects behaviors and behavioral parameters (targets,
-speeds, durations, etc.).
-This level often also 
-includes decision tables, finite state machines. timers, if-then condition testing, ...
+This level often determines the control flow of a behavior.
+
+{\bf Highest Level Control} aka {\bf Behavior Selection:} 
+Answer the question: ``What do I do next?''
+Select behaviors and behavioral parameters (targets,
+speeds, durations).
+Often involves finite state machines. timers, if-then condition testing, ...
 This level often involves a great deal of human operator online interaction,
@@
 
-\subsection{State Estimation}
-
+{\bf Error Handling:} Errors must be detected and handled by parameter (target)
+adjustment and/or behavior switching
+at all of these levels.
+
+{\bf State Estimation:}
 In this architecture, state estimation to deal with sensor noise and incomplete or
@@
 
-\subsection{Error Handling}
-
-Errors must be detected and handled by parameter (target)
-adjustment and/or behavior switching
-at all of these levels.
-
-\subsection{Why do we recommend this architecture?}
-
 The recommended control architecture is intuitive, abstracts the hardware 
@@
 
-\subsection{What are the alternatives?}
-
 Closely related control architectures combine Phoneme and Word control,
 or combine Phoneme, Word, and Behavior Selection.
-This is not a major change.
+This is not a major design change.
 
@@
 operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
-We do not address the loosely connected case in this paper.
 
@@
 \end{equation}
-where $\vq$ are the joint angles of
+where $\vq$ are the joint angles and position of the ``root'' of
 the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
@@
 as exoskeleton joint torques.
-
-We do not describe how to derive this equation or identify appropriate parameters
-for an exoskeleton. We can provide this information if requested.
-We are assuming that the equations have been derived taking into account constraints
-such as one or two feet on the ground. Until we discuss ``floating body'' dynamics
-in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
-We are assuming all degrees of freedom (joints) are actuated until 
-Section~\ref{sec:float}.
+$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
+that correspond to root velocities, since those dimensions are not actuated.
 
@@
 
-[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
 Note that we are ignoring the operator dynamics and operator muscle and tissue
@@
 We are also ignoring actuator dynamics, which we can handle by augmenting 
-the state vector and adding additional equations.
+the state vector.
+
+[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
 
@@
 attempt to match the torques caused by inertial, Coriolis, centripetal,
-gravitational, and frictional forces.
+gravitational, frictional, and exoskeleton-world contact forces.
 \begin{equation}
@@
 + \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
-\end{equation}
-If in addition we want to cancel exoskeleton-world contact forces,
-we add another term:
-\begin{equation}
-\vtau = \vtau_{\invdynw} = \vtau_{\invdyn} 
 + \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
+\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
 \label{eq:change-inertia1}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + \mK \hat{\vf}_o
+\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
 \end{equation}
@@
 from the point of view of the operator.
-Section~\ref{sec:impw} discusses how to make the exoskeleton 
+A future white paper will discuss how to make the exoskeleton 
 imitate a desired linear dynamic system 
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \label{eq:impedance}
@@
 \begin{equation}
-\mM \ddot{\vq} - \mK_{\dot{\vq}} \dot{\vq} - \mK_{\vq} ( \vq - \vq_d ) = \vtau_o
-\end{equation}
-and the operator sees the desired impedance.
-
-{\it Although impedance control is extensively utilized in rehabilitation
-robotics [18], only a limited number of studies has been focused on
-this method for power augmentation [19, 20].}
+\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
+\end{equation}
+and the operator sees the desirecd impedance.
+
+Although impedance control is extensively utilized in rehabilitation
+robotics [18], limited number of studies has been focused on
+this method for power augmentation [19, 20].
 \begin{verbatim}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + \mJ_o^{\tr} {\vf_o}_d
+\vtau = \vtau_{\invdyn} + \mJ_o^{\tr} {\vf_o}_d
 + \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
 - \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
@@
 \subsection{Autonomous exoskeleton-world impedance control}
-\label{sec:impw}
 
@@
 \end{equation}
-and add the force error: 
-\begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
-+ \mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
-\end{equation}
-
-\section{Underactuated and overactuated control and floating body dynamics}
-\label{sec:float}
-
-This section describes how to handle underactuation and overactuation using
-``floating body dynamics.''
-We have been sweeping this
-issue under the rug so far. It is now useful to be more explicit.
+and add the force error to equation~\ref{eq:impedance}: 
+\begin{equation}
+\mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
+\end{equation}
+
+\section{Using Online Optimization: Quadratic Programming}
+\label{sec:qp}
+
+The dominant paradigm in force controlled (typically
+hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
+such as torque and center of pressure limits using online quadratic programming
+to solve the inverse dynamics equations. This approach can also be used
+to implement and trade off among the many control objectives described so far.
+
+Quadratic programming~\cite{Wikipedia} solves the following problem:
+Minimize
+\begin{equation}
+\vx^{\tr} \mQ \vx + c^{\tr} \vx
+\end{equation}
+with inequality constraints
+\begin{equation}
+\mA_1 \vx \le \vb_1
+\end{equation}
+and equality constraints
+\begin{equation}
+\mA_2 \vx = \vb_2
+\end{equation}
+The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
+torques $\vtau$, and contact forces $\vf$.
+For robots the discrepancy between actual and desired accelerations, joint
+torques, and contact forces are all minimized:
+\begin{equation}
+C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
++ \vf^{\tr} \mW_{\vf} \vf
+\end{equation}
+$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
+different errors and error directions.
+The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
+Torque limits, friction limits, center of pressure location constraints,
+and unidirectional contact constraints are all
+enforced as inequality constraints.
+Variants of this optimization
+were performed every 1-2ms by various teams to control the
+Atlas robots in the DARPA Robotics Challenge.
+
+This approach can be used to achieve desired joint or synergy (such as center
+of mass) accelerations, center of mass dynamic behavior, center of pressure location,
+weight distribution, maintenance of a desired body part position, orientation,
+velocity, or trajectory. and joint/actuator position and velocity limits (indirectly).
+
+In the case of exoskeletons, 
+part of the optimization criterion could be to
+minimize contact forces on the operator:
+\begin{equation}
+C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
+\end{equation}
+Operator force feedback and impedance control could be implemented by specifying a
+desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
+\begin{equation}
+\vf_o = {\vf_o}_d
+\end{equation}
+or by adding a ``soft'' constraint in the form of a new term in the optimization
+criterion:
+\begin{equation}
+C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
+\end{equation}
+
+To implement operator impedance control we set the desired operator contact force to:
+\begin{equation}
+{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
+\end{equation}
+Assuming that the operator directly applies exoskeleton joint torques, this becomes
+the equality constraint
+\begin{equation}
+\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+or is added to the optimization criterion:
+\begin{equation}
+C_e(\vx) = C_r(\vx) 
++ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
+\end{equation}
+with
+\begin{equation}
+{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\end{equation}
+
+\subsection{Full Atlas robot implementation.}
+
+\section{Underactuated and overactuated control}
+
 There are two kinds of underactuation we want to consider. The first
 is ``floating body dynamics'', where the robot is not bolted to the floor.
-Humanoid robots (and humans) have floating body dynamics.
-A ``root'' location (usually somewhere
+Humanoid robots have floating body dynamics. A ``root'' location (usually somewhere
 on the pelvis) is chosen, and the position, velocity, orientation, and angular
@@
 are included in the state vector $\vq$. The corresponding dimensions of the
-velocity vector $\dot{\vq}$ are not actuated. A selection matrix
+velocity vector $\dot{\vq}$ are not actuated. We have been sweeping this
+issue under the rug so far. It is now useful to be more explicit. A selection matrix
 $\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
@@
 
-The second kind of underactuation appears in exoskeletons where we rely on the
+The second kind of actuation appears in exoskeletons where we rely on the
 operator to actuate a particular joint. For example, many lower body
 exoskeltons only actuate the joints in the sagittal plane, leaving lateral
-and twist (yaw) joints unactuated or passively actuated.
-
-Passive actuation can be handled by augmenting the
+and twist (yaw) joints unactuated. We can actually handle this by using
+the matrix $\mS$ and reducing the size of the $\vu$ vector.
+
+We can also have passive actuation. This we can include by augmenting the
 dynamics equations to include the force equation of the passive element,
@@
 
-The underactuation can be handled by reducing the size of the $\vu$ vector
-and changing the matrix $\mS$ to reflect the missing actuation.
+We can have multijoint actuation, which can be handled in $\mS$ as well.
+
 We can have redundant actuation (more than one actuator across any particular
 joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
-We can have multijoint actuation, which can be handled in $\vu$ and $\mS$ as well.
 
@@
 operator that fully actuates the exoskeleton, so having some degrees of freedom
-that are not controllable by the actuators is not fatal to a design.
+not controllable by the actuators is not fatal to a design.
 
@@
 still work on the actuated degrees
-of freedom. Passive actuation or 
-the operator needs to carry the load on the unactuated degrees of freedom.
+of freedom. The operator needs to carry the load on the unactuated degrees of freedom.
 
@@
 no longer works if $\mS$ is not invertable (not square or not full rank).
-In the underactuated and overactuated cases $\mS$ is not square. 
-In these cases {\bf online optimization}
-as described in the next Section is used to ``perform inverse dynamics'' by
+In the underactuated case $\mS$ is not square. In this case {\bf online optimization}
+as described in Section~\ref{sec:qp} is used to ``perform inverse dynamics'' by
 finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
@@
 
-\section{Using Online Optimization: Quadratic Programming}
-\label{sec:qp}
-
-The dominant paradigm in force controlled (typically
-hydraulic) full size humanoid robots is to use online optimization
-to handle under and overactuation
-and to handle constraints in inverse dynamics
-such as torque and center of pressure limits.
-Online optimization typically solves the constrained inverse dynamics problem
-using quadratic programming.
-This approach is used
-to implement and trade off among the many control objectives described so far.
-
-Quadratic programming~\cite{Wikipedia} solves the following problem:
-Minimize
-\begin{equation}
-\vx^{\tr} \mQ \vx + c^{\tr} \vx
-\end{equation}
-with inequality constraints
-\begin{equation}
-\mA_1 \vx \le \vb_1
-\end{equation}
-and equality constraints
-\begin{equation}
-\mA_2 \vx = \vb_2
-\end{equation}
-The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, actuator
-commands $\vu$, and contact forces $\vf$.
-For robots the discrepancy between actual and desired accelerations, actuator command
-magnitude, and contact force magnitude are all minimized:
-\begin{equation}
-C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
-+ \vf^{\tr} \mW_{\vf} \vf
-\end{equation}
-$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
-different errors/magnitudes and error/magnitude directions.
-The dynamics (equation~\ref{eq:dynamics2}) are enforced as an equality constraint.
-Actuator limits, friction limits, center of pressure location constraints,
-and unidirectional contact constraints are all
-enforced as inequality constraints.
-Variants of this optimization
-were performed every 1-2ms by various teams to control the
-Atlas robots in the DARPA Robotics Challenge.
-
-This approach can be used to achieve desired joint or synergy (such as center
-of mass) accelerations, center of mass dynamic behavior, center of pressure location,
-weight distribution, maintenance of a desired body part position, orientation,
-velocity, or trajectory. and joint/actuator position and velocity limits (indirectly).
-
-\subsection{Controlling operator forces and perceived exoskeleton impedance}
-
-In the case of exoskeletons, 
-part of the optimization criterion could be to
-minimize contact forces on the operator:
-\begin{equation}
-C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
-\end{equation}
-Operator force feedback and impedance control could be implemented by specifying a
-desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
+Operator {\bf force feedback} is implemented using online optimization by
+add the following ``hard'' constraint
+to the equality constraints in the quadratic program:
 \begin{equation}
@@
 \end{equation}
-or by adding a ``soft'' constraint in the form of a new term in the optimization
-criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
-\end{equation}
-Force damping can be implemented by adding the following ``soft'' constraint term
-to the optimization criterion:
+or adding a ``soft'' constraint as a term
+\begin{equation}
+(\vf_o - {\vf_o}_d)^{\tr} W_{\vf_o} (\vf_o - {\vf_o}_d)
+\end{equation}
+in the optimization criterion. $W_{\vf_o}$ is a weight to indicate how much operator
+force is acceptable (if ${\vf_o}_d = 0$) or how much operator force error is
+acceptable.
+Force damping can be implemented by adding a similar ``soft'' constraint:
 \begin{equation}
@@
 \end{equation}
-
-We can implement operator {\bf impedance control} in several ways:
-We can add 
+to the optimization criterion.
+
+{\bf Impedance control} is implemented using online optimization by adding
 the impedance equations as ``hard'' equality constraints (to achieve them exactly)
@@
 
-We can alternatively just set the desired operator contact force to:
-\begin{equation}
-{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
-\end{equation}
-Unfortunately, this involves inverting the Jacobian matrix, which may introduce
-problems.
-Assuming that the operator directly applies exoskeleton joint torques, 
-we can get rid of the Jacobian matrix, and this becomes
-the equality constraint
-\begin{equation}
-\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-or is added to the optimization criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) 
-+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
-\end{equation}
-with
-\begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-
 {\bf Mapping external forces to operator forces} can be done either with the
@@
 The use of online optimization makes sharing control with an operator easy.
-Dimensions (even actuated dimensions) that are under the operator's control have
+Dimensions (even powered dimensions) that are under the operator's control have
 low weights in the online optimization. It is also easy to map operator controls
@@
 weight matrices similar.
-
-\subsection{Full Atlas robot implementation.}
-
-Discuss or point to DRC papers by multiple groups.
 
@@
 
-We have not seen a lot of discussion of state estimation of exoskeletons in
-the literature.
-We will discuss our approach to state estimation on humanoid robots here.
-
-We implemented two kinds of Kalman filters for state estimation.
-Low level state estimation is done by an extended Kalman filter that takes into
-account joint position, actuator load measurements, actuator commands,
-and the dynamic model of the robot,
-and measured contact forces (feet and wrists)~\cite{}.
-For higher level (overall) state estimation,
-we developed a center of mass location and velocity estimator, and its
-application in full body inverse dynamics control, fall detection
-and fall prevention for humanoid robots and exoskeletons~\cite{}.
-
-We use Linear
-Inverted Pendulum Model (LIPM) dynamics with an offset to predict
-the center of mass motion. This offset can be interpreted
-as a modelling error on the center of mass position, or an
-external force exerted on the center of mass of the robot,
-or a combination of both. The center of mass estimator was
-implemented on our Atlas humanoid robot. During the DARPA
-Robotics Challenge Finals, it successfully saved our robot from
-falling on two occasions, and made us the only competitive team
-without a fall or need for a human rescue among all teams.
-
-Unscented,
-Particle filters
-
 DISCUSS HOW TO HANDLE SENSOR NOISE
@@
 
-\section{Robustness to parametric modeling error and unmodelled dynamics}
-
-We need to be robust to parameter modeling errors such as the wrong mass
-or moment of inertia for an exoskeleton part, variations in the operator
-parameters, and variations in attachments to the exoskeleton and payload.
-Examples of typically unmodeled dynamics (non-parametric modeling error) include:
-backlash, joint play, structural deformation, flexible wires and hoses,
-actuator dynamics, processing delay,
-elastic operator tissue and straps, and 
-loose or slack operator tissue and strap.
-
-There are two questions to focus on:
-
-\subsection{How robust are these control schemes to parametric and non-parametric
-modeling error?}
-
-This question needs to be addressed using simulation, and is deferred to a future
-white paper [robust.pdf].
-
-\subsection{How can we make control more robust?}
-
-One way to understand control systems is to separate 1) what happens when there
-are no errors and feedback control is not active, and 2) what happens in response
-to errors (feedback control). 
-What happens when there are no errors in model-based control is 
-feedforward control, which does not have stability issues (except for some ignorable
-technical issues). If your model is good, you get good performance. If your
-model is bad, you get larger errors and have to rely on feedback control.
-So feedforward control is robust, and we don't have to worry about it. We do have
-to worry about how to get good models, and what to do about stuff that varies:
-contacts, operator body, operator behavior, payloads, attachments, etc.
-We do have to worry about performance.
-
-On the other hand, feedback control can lead to steady state or growing
-oscillation, which we colloquially refer to as being unstable or instability.
-Since walking and running are also oscillations which we refer to as stable
-behaviors, the technical meaning of the word ``stable'' is more complex than
-just ``it keeps moving'' or ``it doesn't do what the operator wants''.
-In any event, we don't want oscillations.
-
-The best way to avoid oscillations is to manually design a conservative 
-independent joint feedback controller.
-This way it can be tested independently of any
-feedforward control.
-Unfortunately, inverse dynamics and online optimization-based control mix
-feedforward and feedback control so we cannot use the recommended, well established,
-and safe approach of manual conservative independent joint feedback control design.
-
-What to do about this to achieve robust online optimalization-based control is
-currently a research issue. In actual implementations, skilled engineers manually
-tune optimization weights and the various constraints that are active to
-make this control work reliably. The fact that this has been done on many actual
-humanoid robots in many situations is comforting. The fact that none of these
-implementations have moved with the speed and agility of a top percentile
-human athlete is disconcerting.
-
-We can ``verify'' designs using simulation, but making accurate simulations of
-the effects of contact and unmodelled dynamics is also a current research issue.
-There is no getting around the fact that our operators will be like the test
-pilots of early planes. The exoskeletons may crash, flail wildly, or oscillate.
-We need to plan for operator safety from day 1 of operation, not try to retrofit
-it later.
-
-If the robustness of online optimization is not adequate, an alternative is
-combining feedforward and feedback control into one function and using gradient
-descent in simulation to improve the control robustness.
-One technique to do this is multiple-model design~\cite{acc12}.
-
-Note that adaptive or learning
-control does not promise to improve the kind of robustness exoskeltons need. 
-These approaches promise
-to improve performance, and try to track (slowly) changes in the operator, exoskeleton,
-or contacts to maintain performance. If something changes quickly, we need robust
-control design.
+\section{Robustness to parametric modeling error}
+
+\section{Robustness to unmodeled dynamics}
+
+\subsection{Example unmodeled dynamics}
+
+Examples:
+\begin{itemize}
+\item
+A) hydraulic actuator dynamics.
+\item
+B) delay. 
+\item
+C) elastic straps sinusoidal oscillation.
+\item
+D) loose straps (bang coast bang oscillation)
+\end{itemize}
+
+\subsection{1 joint rotary example.}
+
+\subsection{2 joint rotary example: stance.}
+
+\subsection{2 joint rotary example: swing.}
+
+\subsection{4 link body example.}
+
+\subsection{Full Atlas robot.}
+
+\section{Survey Existing Exoskeletons}
+
+\input{exos/bleex}
+
+\input{exos/hal}
+
+\input{exos/xor}
+
+\input{exos/exo-ul7}
 
@@
 Oct 5-9
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2015-09-23 exos/xor.tex
finished XoR

@@
 
-keywords: strength augmentation; hybrid actuation\\
+keywords: model-based control; strength augmentation; hybrid actuation; pneumatic actuation; electric motors;\\
 
@@
 
-muscular weakness
+The XoR is a prototype, light-weight lower-body exoskeleton that uses a hybrid pneumatic-electric drive system that \textbf{reduces weight} while providing \textbf{precise torque control}, \textbf{backdrivability}, and a \textbf{desirable force / velocity} profile.
+The exoskeleton is designed to serve in rehabilitation settings to augment operators' strength and assist with postural control for persons with disabilities.
 
-Purpose: prototype developed for "postural control of elderly people and persons with mobility disability" -- [[llimbCtrlReviw]]
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/xor_joint_rom.png}
+\end{figure}
 
-Design:
+The XoR weighs 30kg and includes 10 DOF with 6 active joints (flexion / extension of hip, knee, and ankles) and 6 passive (hip abduction/adduction joints, hip rotation, and ankle adduction/abduction).
+The active joints are powered by hybrid actuators comprised of an \emph{air muscle} and an electric motor.
+The hybrid actuation scheme uses a unilateral air muscle layout to compensate for gravity and bilateral electric motors with relatively small gear ratio (57.5) to serve as the dynamic compensator.
 
-"implements a hybrid driving concept combining pneumatic artificial muscles and electric motors together [[XoRdesign]]; [[XoRkinemExtraction]] ...motion intention is estimated based on EMG signals ...
-user posture based on joint angles and ground reaction forces."  -- [[llimbCtrlReviw]]
+The actuation scheme is complementary in that the electric motors have a quick response time and produce high peak torque for short amounts of time, while air muscles have a delayed response (true of pneumatic systems in general due to the compressibility of air) with better power density and sustained torque.
+The hybrid drive system sums the two to develop a desired torque profile that achieves the benefits of both and reduces weight and motor size.
+The system offers high torque control with negligable stick-slip and backdrivability.
 
-Control Strategy:
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/xor_hybrid_drive_torque_time.png}
+\end{figure}
 
-Model-based control -- "...Pneumatic artificial muscles provide gravity compensation and electric motors serve the dynamic compensator.
-The total generated torque tracks a desired value calculated on the basis of a bio-mechanical model...
-Only simulation results were available to prove the efficacy of this assistive strategy working with XoR." -- [[llimbCtrlReviw]]
+A challenge in using air muscles is that force reduces quadratically as a function of contraction.
+For instance, at 30\% contraction, the force produced by FESTO rubber air muscles vanishes.
+To address the issue, XoR is strategically places air muscles to generate forces in desired configurations.
 
-References:
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=3.5in]{exos/figs/xor_air_muscle_force_vs_contraction.png}
+\end{figure}
 
-S.
-Hyon, J.
-Morimoto, T.
-Matsubara, T.
-Noda, M.
-Kawato, Xor: Hybrid drive exoskeleton robot that can balance, in: Intelligent Robots and Systems (IROS), 2011 IEEE/RSJ International Conference on, IEEE, 2011, pp.
-3975–3981.
-<<XoRdesign>>
+In implementation, the XoR uses rubber air muscles (MDSP-40) connected to pulleys via tendons and to EP regulators in the backpack (a constant pressure and cam system is being tested to reduce weight required for backpack air valves).
+The electric actuator is a geared, 200W brushless DC motor (Maxon EC powermax 30) rated at 4.7 A, which transmits power to joints through a belt and pulley system with a 57.5 gear ratio.
+the motor's 0.12 Nm output torque provides up to 34.5 Nm (for short duration) at every joint and is backdrivable.
+The design yields a relatively high range of motion (up to 120 degrees) with a load capacity 150 Nm.
+However, experiments reveal the need for bi-directional pneumatic actuation at hip joints.
 
-J.
-Morimoto, T.
-Noda, S.
-Hyon, Extraction of latent kinematic relationships between human users and assistive robots, in: Robotics and Automation (ICRA), 2012 IEEE International Conference on, IEEE, 2012, pp.
-3909–3915.
-<<XoRkinemExtraction>>
+For sensing, XoR is equipped with rotary encoders at joints, an IMU in the backpack, load cells in the feet, and is networked to servers capable of providing EMG, NIRS, and EEG data.
+A control PC running at 1 KHz, pulse counter, amplifiers, Digital IO are external and so not included in the weight of the unit.
+Human operators are outfitted with a goniometer.
 
-***** Data from design paper [[XoRdesign]]
+\subsubsection{Control}
 
-HMI: 
-Brain-machine interface.
+The XoR control strategy applied in \cite{XoRkinemExtraction2012} has two main components.
+First, a proportional-derivative (PD) feedback controller tracks desired joint angles and angular velocities that correspond to the state of the exoskeleton required to assist the human operator.
+To obtain this state (specifying the desired joint angles / velocities), the controller simultaneously measures the joint angle trajectories of the human user (goniometer) and of the robot (encoders), and uses canonical correlation analysis (CCA) to extract latent variables in the kinematic relationship between the two.
 
-Acutation:  
-Hybrid drive - pneumatic muscles with electrical motors arranged in "optimal ways."
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=6.0in]{exos/figs/xor_joint_angles.png}
+\end{figure}
 
-Design criteria:
+To avoid relying on high gain feedback, XoR incorporates user intent through EMG data.
+The controller rectifies and low pass filters (10 Hz cut-off) data measured in the pilot's quadriceps femoris, tensor fasciae latate, gluteus medius, and tibialis anterior.
+It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta} \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1 \cdot EMG_n)$, using a linear prediction model, 
+\[{\bf x}(k+1) = \mA{\bf x}(k)+\mB{\bf u}(k),\].  
+The controller uses the angular measurements required to perform CCA and EMG data to derive the model's $\mA$ and $\mB$ terms.
+The predicted state from EMG model provides the desired input to yet another PD controller:
+\[\tau_i = K_p (\theta_i^d - \theta_i) + K_d (\dot\theta_i^d - \dot \theta_i),\] 
+with gains of $K_p = 1000$ and $K_d = 100$.
 
-Precise torque control; high backdrivability; force/velocity profile for rehab and compensation; lightweight; quasi-autonomy for rehab; autonomous posture control.
-
-Precise torque control -- they believe this is most important feature to mobility assist/rehab programs as softness, compliance, and safety are defined in force/toque space;
-
-safety thru passive backdrivability;
-
-air muscles -- use them because there is negligable stick-slip; better controllability but gnereated force decreases quadratically with contraction (vanish around 30% contraction); optimize arrangment of AM to generate desired forces in specific confgs.
-Use air muscle (AM) as gravity compensator and electric motor as dynamic compensator.
-AM has high power density; good for high-load torque but delayed due to air compressibility.
-There is also steady-state error.
-Electric motors (without gear reduction) have high response torque but can only generate for short periods.
-Hybrid drive sums the two to produce desired torque profile.
-Can combine AM with small electric servos or hydraulic servo actuators.
-Small servos without high gear reduction; DC (brushed and brushless).
-
-Use AM as a unilateral drive for gravity comp combined with bilateral electric motors (motion control); generates precise high torque.
-The air compressor / hydraulic pump can be smaller and lighter electric batteries.
-
-10 DOF; 6 active (all FE joints with hybrid drive) and 6 passive joints (hip ankle AA, hip rotation);
-
-Transmission - AM connected to pulleys via tendons.
-Rubber muscle (MDSP-40) with EP regulators mounted in backpack; testing constant pressure and cam to reduce weight required for backpack air valves.
-Bruschless DC geared motors (4.7 A rating; 0.12Nm output; gear ratio 57.5 for backdrivability; for short operation motor can generate 34.5 Nm/joint) with belt and pulleys
-
-Range of motion of 120 deg and load capacity 150 Nm.  Weight 33kg of which toro + cables ~15kg. Experiments show insufficient torque and need for bi-directional pneumatic acuation at hip joints.
-
-Sensing -- Rotory encoders at each joint; IMU in backpack; load cells at feet; network connected with servers processing sensory info from EMG, NIRS and EEG.  Controller runs at 1ms on Linux PC.
-
-Control:
-
-PLANNED: Torque control; apply compliant control mechanisms designed for full humanoid to exoskeleton control
-
-S. Hyon, “Compliant terrain adaptation for biped humanoids without
-measuring ground surface and contact forces,” IEEE Trans. Robotics,
-vol. 25, no. 1, pp. 171–178, 2009.
-
-S. Hyon, “A motor control strategy with virtual musculoskeletal
-systems for compliant anthropomorphic robots,” IEEE/ASME Transactions
-on Mechatronics, vol. 14, issue 6, pp. 677–688, 2009.
-
-In simulation:
-Task-space controller; The paper describes CoM / CoP control strategy based on GRF.
-Achieve desired GRF using PD feedback to track desired COM trajectory.
-Full-body task space controller superimposed with a joint stiffness controller applying feedback tracking ref joint trajectories.
-Simulations reveal max torque of 150 Nm (effective torque of 80Nm) for HFE and KFE joints.
-Plan to include EMG sensors;
-
-------------------------------------------------
+In hip tracking experiments, the team found EMG data to be helpful.
+When using EMG signals, XoR achieved a mean squared error of $1.8E^{-3}$.
+Without the EMG data, the mean squared error increased to $2.1$.
+Additional experiments confirm the air muscles can effectively perform gravity compensation when used in conjunction with electric actuators (to correct for torque errors induced by inaccuracies in the model mapping position to toque in air muscles).
 
@@
 
-The XoR prototype uses a novel hybrid actuation scheme to achieve both fast peak torque response and high sustained torque.
+The XoR prototype uses a promising hybrid actuation scheme to achieve both fast peak torque response and higher sustained torque while remaining lightweight.
+They track human motion with feedback from a goniometer and feedforward control provided by EMG data.
+These human intention / sensing modalities are both difficult to implement due to noise and calibration issues.
+However, capturing the user intent is important, especially in the presence of external disturbances.
+If the sensory system were improved (e.g, wearable ``robot skin'' sensing arrays), this exoskeleton design could be highly effective.
 
@@
 
-The figures in this section were obtained from \cite{xorDesign2011}. Materials presented are based on the references above.
+The figures in this section were obtained from \cite{xorDesign2011,XoRkinemExtraction2012}. Materials presented are based on the references above.

### 2015-09-24 control.tex
cga-fixes

@@
 for higher control levels
-to access the actuators, sensors, and other robot hardware.
+to access the actuators, sensors, and other robot hardware~\cite{IEEE07271219}.
 It implements high servo rate processing including feedback control and online
@@
 weight matrices similar.
+
+Another approach to shared control is allocating time slices or providing assistance
+on an as-needed basis~\cite{IEEE07140030}.
 

### 2015-09-24 survey.tex
finished XoR
cga-fixes

@@
 %\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input latex-commands
-
-\newcommand{\gc}{{\mbox {GravityCompensation}}}
-\newcommand{\invdyn}{{\mbox {\tiny ID}}}
-
-\newcommand{\shrinkfig}{\def\baselinestretch{1.0}\small} % 0.9 okay
-\newcommand{\shrink}{\def\baselinestretch{1.1}\small} % 0.97 % 0.95 okay
+\input{latex-commands}
 
@@
 
-\title{A Survey of Possible Exoskeleton Control Architectures and
+\title{A Survey of Exoskeleton Control Architectures and
 Algorithms\\
-(Draft 2.2)}
+(Draft 3.0)}
 
@@
 
-This paper surveys possible exoskeleton control architectures and
+This paper surveys exoskeleton control architectures and
 algorithms.
-One goal is to address the question:
-{\it How do exoskeleton control approaches compare with 
-approaches to humanoid robot control?}
-A companion paper surveys implemented exoskeleton control architectures
-and algorithms~\cite{}.
-
 The focus is on exoskeleton control that allows a
@@
 firefighters, and soldiers. 
-
-We expect each exoskeleton controller
-to be used by and optimized for a single operator.
-A substantial investment in capturing the operators normal behavior,
-operator training and learning, and controller customization can be made.
-
-We focus this survey on exoskeleton control for lower body tasks (standing, walking,
+We also 
+focus this survey on exoskeleton control for lower body tasks (standing, walking,
 running, jumping, kicking, dodging, ...).
-We do not survey exoskeleton control for human arms or manipulation.
-
-We treat the torso and helmet of the exoskeleton as the payload,
-and focus on a lower body exoskeleton to support any payloads that
-are on the torso or head.
-
-%\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
-A later white paper will discuss
-possible combinations of response to user and exoskeleton
-autonomy including balance control; response to trips, slips,
-stumbles, and fumbles; task assistance and guidance (guide operator to
-doorknob, button, or light switch); superhuman response to external
-perturbations such as projectiles and explosions; and autonomous
-execution (operator/exoskeleton symbiosis with multi-tasking)?
-
-A later white paper will discuss how to handle malfunctions and damage to the
-system. Basically this boils down to fault detection, and switchover to
-an approapriate safe mode of operation.
-
-\subsection{Let's keep it simple}
-
-Terms like impedance and admittance control are used, but are often confusing,
-as in controller design one can choose from several possible
-inputs into the exoskeleton (exoskeleton positions,
-velocities, accelerations, operator-exoskeleton contact forces and
-exoskeleton-world contact forces), and choose from several possible exoskeleton
-outputs: exoskeleton actuator forces and torques, exoskeleton
-motion (position, velocity, and acceleration), as well as operator-exo
-contact forces or exo-world contact forces.
-
-A useful background paper on impedance and admittance:\\
-\url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}
-
-Variants of nonlinear feedback control such as feedback linearization or
-sliding mode control are largely ignored in this paper.
-Once the decision to use feedback control
-based on a set of observable quantities and with particular outputs is made,
-one can try out the various linear and nonlinear feedback control paradigms
-to see what works well.
-
-Variants of function approximation methods such as lookup tables, fuzzy logic,
-sigmoidal neural networks, radial basis functions, and locally weighted regression
-are also largely ignored. Once the decision to use a function approximator
-and what the inputs and outputs are has been made, one can try out the various
-approaches to see what works well.
-
-Same for optimization methods.
-
-Same for constraint enforcement (such as avoiding self-collisions)
-in either optimization or feedback control (barrier
-Lyapunov functions (BLF)~\cite{IEEE06911561}).
-
-Stability proofs of any of these methods should be viewed skeptically due
-to the unmodeled operator-exoskeleton and exoskeleton-world contact dynamics, 
-actuator unmodeled dynamics, joint play and exoskeleton
-structural deformation, and controller
-time delays which are typically ignored, especially in proofs involving
-passivity arguments or Lyapunov functions.
-We ignore stability proofs as well.
-
-\section{What are some design philosophy alternatives?}
-
-\subsection{``Invisible'' exoskeleton}
-
-Can we build an exoskeleton allows the operator to behave naturally
-and exerts little or no force on the operator?
-
-This could be achieved through active control using feedback of the
-operator-exoskeleton forces. Additional gravity, friction, actuator, and
-inertial compensation (inverse dynamics) can improve performance.
-
-\subsection{``Natural'' exoskeleton}
-
-Can we build an exoskeleton allows the operator to behave naturally?
-The operator can feel the exoskeleton, but can overpower it when necessary.
-
-Bypass valves and clutches could allow the operator to physically ``take over''
-and push the relatively lightweight leg mounted exoskeleton around.
-
-``Passive'' degrees of freedom can have physical or ``virtual'' springs
-applied to assist in support (all abduct/adduct DOF, rotational DOF).
-These aids can be clutched out or turned off when necessary.
-``Virtual'' springs (springs implemented actively through exoskeleton
-actuators) can be modulated.
-
-\subsection{``Symbiotic'' exoskeleton}
-
-Can we build an exoskeleton that allows the operator/exoskeleton team
-to be effective, but not necessarily behave naturally? 
-The operator needs to learn to ``fly'' the suit,
-just as operators learn to operate parachutes,
-wing suits, diving equipment, high and low temperature protective
-suits, firefighting equipment, and high altitude flight
-and parachute suits.
-
-The exoskeleton can adapt its behavior to the task.
-For example, for jumping, the exoskeleton can operate like a pogo 
-stick:\\
-\url{https://www.youtube.com/watch?v=lbp41vWP4o4},\\ 
-and for running it could act like jumping stilts:\\
-\url{https://www.youtube.com/watch?v=9ZOd7yEyhwI},
-
-\section{Effects of underlying exoskeleton actuator technology}
-
-One basic distinction is whether the actuators are thought of as position
-or force sources. Often this is revealed by the design of the ``low level''
-controller. Note that this is often somewhat confusing. Electric motors
-are torque sources (low impedance), but when a high gear ratio transmission
-is added (such as a harmonic drive) the whole system becomes high impedance
-and thus is best thought of as a position (or velocity) source. Therefore,
-position or velocity 
-control is performed by the low level controller. Hydraulic actuators
-are high impedance, but when a force or load sensor is added (or piston differential
-oil pressure is used as a force sensor), the whole system becomes low impedance
-and is best thought of as a force source. Force (or joint torque) control is
-performed by the low level controller. Force or torque sensing can be added to an
-electromechanical drive train (electric motor plus gears or electric
-motor plus ballscrew)
-to make it lower impedance and more like a force source as well.
-
-\section{Recommended Control Architecture}
-\label{sec:architecture}
-
-We recommend the following hierarchical control architecture~\cite{IEEE06907051}:
-
-{\bf Low Level Control:} Present interface to actuators and
-high bandwidth feedback control, degrees of freedom,
-or synergies of either:
-\begin{itemize}
-\item
-Position control.
-\item
-Pure force or torque control.
-\item
-Impedance control: specify desired position (for spring), and generate
-stiffness, damping, and potentially a modified inertia or moment of inertia.
-Modifying inertia is more dangerous than providing active stiffness
-or damping.
-Or (almost equivalently),
-specify a desired force and change the force according to position, velocity,
-and acceleration.
-\item
-Online optimization of accelerations, actuator forces and torques, and contact forces
-using quadratic programming.
-\end{itemize} 
-One important synergy is coordinated control of the center of mass location and
-velocity for balance.
-PID or fancier feedback control laws are implemented at this level.
-Gravity, friction, actuator dynamics, and other feedforward disturbance
-compensation are implemented at this level, as well as any inverse dynamics
-or other types of feedforward reference tracking control.
-Disturbance observers for model errors and contact/wind forces
-are implemented at this level.
-Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
-or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
-Often, this controller is provided by the manufacturer of the robot
-or exoskeleton and runs on a separate computer system from the
-``user's'' or task/application controller~\cite{Sarcos,Atlas}.
-
-{\bf ``Phoneme'' Control:} Mid-level controllers
-for relatively constant phases of behavior.
-Often involves position or velocity targets (goto or
-stay-at X), trajectory references, ...
-Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
-Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
-Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
-Finer resolution may be needed. For example a walking stance period may
-be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
-The first and second halves of stance are divided by when the center of mass
-passes over the ankle.
-Local inverse kinematics (finding small joint motions for small target changes)
-is performed at this level.
-This level often provides the behavioral primitives or ``verbs'' for robot programming.
-
-{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
-units.
-Often involves finite state machines, timers, if-then condition testing, gain
-scheduling, ...
-Examples: walk-to-location(x=7.8m, y=2.3m), run-to-location(speed=3.7m/s),
-horizontal-jump(height= 2m, length= 5m).
-Global inverse kinematics (selection of solution branch) is performed at this level.
-This level often determines the control flow of a behavior.
-
-{\bf Highest Level Control} aka {\bf Behavior Selection:} 
-Answer the question: ``What do I do next?''
-Select behaviors and behavioral parameters (targets,
-speeds, durations).
-Often involves finite state machines. timers, if-then condition testing, ...
-This level often involves a great deal of human operator online interaction,
-or is completely done by the operator.
-
-{\bf Error Handling:} Errors must be detected and handled by parameter (target)
-adjustment and/or behavior switching
-at all of these levels.
-
-{\bf State Estimation:}
-In this architecture, state estimation to deal with sensor noise and incomplete or
-redundant sensing is a separate process and is as independent as possible
-from the control method~\cite{certainty-eq,separation-prin}.
-This makes evaluating different control methods much easier.
-
-The recommended control architecture is intuitive, abstracts the hardware 
-and lower level control in a straightforward way, is easy to program, and is
-commonly used. Variants of this architecture were used by almost all teams
-in the DARPA Robotics Challenge~\cite{}.
-
-Closely related control architectures combine Phoneme and Word control,
-or combine Phoneme, Word, and Behavior Selection.
-This is not a major design change.
-
-An alternative behavioral architecture might not have explicit levels but instead
-has a ``soup'' of behaviors that compete or cooperate for the ability to
-control the robot or exoskeleton. Subsumption architecture and other behavior-based
-architectures are closer to the ``soup'' approach and to some extent
-use inhibition and excitation
-among behaviors to select active behaviors.
-
-Another alternative behavioral architecture does not have explicit separate behaviors,
-but one or a small number of functions or policies. 
-We are seeing more of these architectures
-due to the success and popularity of function approximation such as
-deep learning.
-
-An alternative approach to state estimation is to design state estimation
-and control laws simultaneously~\cite{LQG-LTR}. Another alternative is
-to have a controller with internal state, and design or learn a policy
-that works well, but does not have an explicit state estimation 
-part~\cite{lead-lag}.
-
-\section{Exoskeleton dynamics}
-
-Exoskeletons are similar to robots, but there are several important differences.
-There are two sets of contact forces: operator-exoskeleton (which we denote
-with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
-with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
-rigidly held by other types of physical constraints, we can lump the operator
-body parts with the exoskelton parts which they are attached to in terms
-of link inertias, locations of center of mass, and moments of inertia,
-and just think of the whole system as one robot.
-We can also separate out the operator dynamics if this is more convenient
-(we separate them out and ignore them in what follows).
-If the operator is loosely connected to the exoskeleton, this introduces a
-huge modeling and control problem, and would also be very dangerous for the
-operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
-
-The dynamics of a revolute exoskeleton are:
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
-\label{eq:dynamics}
-\end{equation}
-where $\vq$ are the joint angles and position of the ``root'' of
-the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
-velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
-centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
-friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
-contact forces)
-are the exoskeleton-world
-contact forces 
-expressed as exoskeleton joint torques, 
-$\vtau$ are the exoskeleton
-joint torques,
-and $\mJ_o^{\tr}(\vq) \vf_o$
-are the operator-exoskelton forces expressed
-as exoskeleton joint torques.
-$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
-that correspond to root velocities, since those dimensions are not actuated.
-
-To understand these equations better, it is useful to have all Jacobians equal
-to the identity matrix. This means the operator is directly applying torques
-at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
-and so is the world ($\vtau_w$) (a very unrealistic assumption). In this case:
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \vtau_w = \vtau + \vtau_o
-\end{equation}
-
-Note that we are ignoring the operator dynamics and operator muscle and tissue
-forces resulting in operator joint torques. We will add these in later (or maybe
-we won't ???).
-
-We are also ignoring actuator dynamics, which we can handle by augmenting 
-the state vector.
-
-[SOMETHING ABOUT MODELING HUMAN + EXOSKELETON + WORLD ]
-
-\section{Test case: Atlas humanoid robot}
-
-Gazebo and SDFast models.
-
-\section{Passive control with no active control}
-
-One control option is to use only passive controls, and no active controls
-or actuators.
-Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
-to reduce forces. 
-An operator would rely on mechanical backdrivability to drive
-exokeleton to do tasks.
-In this case, 
-$\vtau$ is generated by passive devices such as springs or dampers.
-
-\section{Open loop control}
-
-Another control option is open loop control, in which actuator commands are
-constant, or a function of time or phase of a behavior or task
-(at least on the small time scale).
-Commands or behaviors may be selected at a relatively coarse time interval
-by the operator or by the exoskeleton controller.
-This approach typically 
-uses position and velocity control in either joint or
-Cartesian coordinates
-with either pre-generated references, selected targets,
-or online generated trajectories to move to targets or perform tasks.
-The operator relies on mechanical backdrivability 
-to correct exoskeleton behavior.
-In this case, it is common to generate the exoskeleton actuator commands using
-a time or phase index:
-$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
-dependent feedforward torque vector.
-
-Recognizing/estimating phase goes here.
-Phase Triggered References could go here.
-\begin{verbatim}
-H.Kawamoto and Y.Sankai, “Power Assist Method Based on Phase
-Sequence Driven by Interaction between Human and Robot Suit,”
-Proceedings of the IEEE International Workshop on Robot and Human
-Interactive Communication, pp. 491–496, 2004.
-\end{verbatim}
-
-Motion capture, motion libraries goes here.
-
-Often finite state machines whose states correspond to left stance, double support 1,
-right stance, and double support 2 are used to select behaviors. Gait transitions
-can be triggered by foot touchdown and liftoff events, or
-displacement of the Center of Mass (CoM).
-
-Central pattern generator
-
-Iterative learning control to acquire $\vtau_{ff}(t)$ for desired trajectories.
-
-\section{Active gravity compensation}
-
-Active gravity compensation is similar 
-to counterbalancing or physical gravity compensation, except it is done
-with exoskeleton actuators to cancel the exoskeleton weight (but not inertia).
-\begin{equation}
-\vtau = \hat{\mG}(\hat{\vq})
-\end{equation}
-where $\vtau$ are the exoskeleton joint torques, 
-$\hat{\mG}()$ are the gravitational forces calculated using an imperfect model, 
-and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
-The operator relies on mechanical backdrivability to drive
-the exokeleton to do tasks.
-
-Sarcos Primus Humanoid videos of active gravity compensation applied
-to a humanoid robot:\\
-CMU: cancel gravity and friction: \url{https://www.youtube.com/watch?v=Ac5cowTwOPw}\\
-ATR: cancel gravity only: \url{https://www.youtube.com/watch?v=KNxxLm4sPys}
-
-A simple approximation for a lower body exoskeleton with a heavy payload at the
-torso is:
-\begin{equation}
-\vtau \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
-\end{equation}
-where $\hat{\mJ}$ is the estimated Jacobian matrix,
-$\hat{m}$ is the estimated
-weight of the exoskeleton (which could include the operator),
-and $\hat{\vg}$ is the gravity vector (it is estimated due to possible orientation
-errors in the controller's estimate of vertical).
-All weight is assumed to be concentrated at the center of mass of the full system.
-
-\section{Friction compensation}
-
-Friction can also be compensated.
-\begin{equation}
-\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}}) 
-\end{equation}
-are the estimated friction torques at the estimated exoskeleton configuration
-and velocity.
-Note that these may be load dependent, and may require additional state to
-correctly handle stiction and hysteresis effects.
-
-Sarcos Primus Humanoid video of friction compensation applied
-to a humanoid robot:\\
-CMU: cancel gravity and friction: \url{https://www.youtube.com/watch?v=Ac5cowTwOPw}\\
-
-\section{Online inverse dynamics control}
-\label{sec:invdyn}
-
-The goal here is to map from desired motion (in this case desired
-accelerations) to exoskeleton actuator commands (and desired operator
-and world contact forces).
-The exoskeleton actuator torques
-attempt to match the torques caused by inertial, Coriolis, centripetal,
-gravitational, frictional, and exoskeleton-world contact forces.
-\begin{equation}
-\vtau = \vtau_{\invdyn} 
-= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
-+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
-+ \hat{\mG}(\hat{\vq})
-+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
-+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
-\end{equation}
-If there are actuator dynamics or other dynamics, we can include those
-dynamics as well.
-
-These force/torque estimates rely on estimated dynamic models and their parameters.
-If the models and estimated state (position and velocity) are perfect, we get
-perfect cancellation of most of the dynamics:
-\begin{equation}
-\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
-\end{equation}
-If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
-of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
-\begin{equation}
-\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
-\label{eq:cancel}
-\end{equation}
-$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
-the operator can directly modify the trajectory.
-
-Simplifying equation~\ref{eq:cancel} by 
-assuming the operator directly applies torques at the
-exoskeleton joints we get:
-\begin{equation}
-\mM (\vq) \ddot{\vq} = \vtau_o
-\end{equation}
-The operator sees the true inertia of the exoskeleton, but no gravitational
-loads and no other dynamics (Coriolis, centripetal, or frictional).
-
-Note that we cannot change the apparent inertia of the exoskeleton without
-some form of acceleration or force feedback.
-With acceleration feedback we can make the exoskeleton have a different inertia.
-Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
-\begin{equation}
-\vtau = \vtau_{\invdyn} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
-\label{eq:change-inertia1}
-\end{equation}
-sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
-the state estimation produces not only perfect positions and velocities but
-also perfect acceleration estimates. In the case 
-where the operator directly applies torques at the
-exoskeleton joints we get:
-\begin{equation}
-\mM_d (\vq) \ddot{\vq} = \vtau_o
-\end{equation}
-In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
-issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
-to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
-cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
-(if the straps are tight enough).
-
-With force feedback we can also make the exoskeleton have a different inertia.
-Setting $\ddot{\vq}_d$ to zero and adding force feedback:
-\begin{equation}
-\vtau = \vtau_{\invdyn} + \mK \hat{\vf}_o
-\end{equation}
-In the case where the force measurements are perfect and 
-the operator directly applies torques at the
-exoskeleton joints we get:
-\begin{equation}
-(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
-\end{equation}
-By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
-\begin{equation}
-\mM_d (\vq) \ddot{\vq} = \vtau_o
-\end{equation}
-
-Online inverse dynamics control is a generalization of 
-active gravity compensation to include
-inertial, Coriolis, and centripetal forces (forces due to acceleration and
-velocity), and potentially frictional forces.
-Online inverse dynamics can be added to open loop behavior execution
-and many other types of control such as operator force
-feedback control, impedance control, and admittance control.
-Computed torque control and feedback linearization are forms of online
-inverse dynamics.
-
-A key question is where does desired acceleration come from?
-\begin{verbatim}
-  Operator force control.
-  Receding Horizon Control of optimal trajectory of 
-    idealized LIPM+flywheel model (Atkeson)
-  intent prediction (see above)
-  neuromuscular model (Geyer, Wang, Van de Panne)
-  hybrid: Optimal control + neuromuscular model (Atkeson NSF Proposal)
-    - NMM provides bias through changes to optimization criterion
-    - NMM provides hard constraints
-    - NMM replaces components of model-based optimization scheme
-\end{verbatim}
-
-\section{Operator-Exoskeleton Impedance Control}
-
-What is impedance?
-\url{https://en.wikipedia.org/wiki/Mechanical_impedance}
-Stiffness, damping, and mass are components of an impedance, as
-they map kinematic variables (position, velocity, and acceleration)
-to forces and torques.
-
-The goal here is to make the exoskeleton imitate a desired linear dynamic system
-from the point of view of the operator.
-A future white paper will discuss how to make the exoskeleton 
-imitate a desired linear dynamic system 
-from the point of view of external perturbations.
-In the case where there is no force sensing between the operator and the
-exoskeleton, the exoskeleton controller maps from exoskeleton configuration
-and velocity to desired joint/actuator/synergy forces.
-In this case the actuators need to generate forces and torques rather than
-positions or angles, or linear or angular velocities.
-
-Here we make the exoskeleton appear as a locally linear impedance in joint
-coordinates to the operator.
-We can also make the exoskeleton appear as a locally linear impedance to the
-operator in some other coordinate system, such as Cartesian coordinates.
-A different goal is to make the exoskeleton appear as a locally linear impedance to
-the outside world in some coordinate system.
-
-To make the exoskeleton appear as a locally linear impedance in joint
-coordinates to the operator, we add position and velocity feedback to $\vtau$:
-\begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\label{eq:impedance}
-\end{equation}
-so if the dynamic models and state estimation are perfect we get this impedance:
-\begin{equation}
-\mM \ddot{\vq} - \mK_{\vq} ( \vq - \vq_d ) - \mK_{\dot{\vq}} \dot{\vq} = \vtau_o
-\end{equation}
-and the operator sees the desirecd impedance.
-
-Although impedance control is extensively utilized in rehabilitation
-robotics [18], limited number of studies has been focused on
-this method for power augmentation [19, 20].
-\begin{verbatim}
-[18] R. Riener, L. Lunenburger, S. Jezernik, M. Anderschitz, G. Colombo,
-and V. Dietz, "Patient-cooperative strategies for robot-aided treadmill
-training: first experimental results," Neural Systems and Rehabilitation
-Engineering, IEEE Transactions on, vol. 13, pp. 380-394, 2005.
-[19] B.-K. Lee, H.-D. Lee, J.-y. Lee, K. Shin, J.-S. Han, and C.-S. Han,
-"Development of dynamic model-based controller for upper limb
-exoskeleton robot," in Robotics and Automation (ICRA), 2012 IEEE
-International Conference on, 2012, pp. 3173-3178.
-[20] W. Yu, J. Rosen, and X. Li, "PID admittance control for an upper limb
-exoskeleton," in American Control Conference (ACC), 2011, 2011, pp.
-1124-1129.
-\end{verbatim}
-
-\section{Force feedback and get out of the way control (inverse dynamics version)}
-
-The goal here is to use force sensing between the operator and the
-exoskeleton to ultimately generate exoskeleton velocities or angular velocities,
-imitating a desired linear dynamic system.
-This type of control is often referred to as admittance control as the
-exoskeleton maps contact forces into joint motion.
-Inverse dynamics control
-can be used to improve this type of control performance.
-
-What is admittance?
-\url{https://en.wikipedia.org/wiki/Admittance}
-Compliance, inverse damping, inverse mass are components of an admittance,
-as they map force and torques to kinematic variables (position, velocity, and
-acceleration).
-
-We can implement operator force control 
-if measurements of the operator contact forces with
-the exoskeleton are available.
-The exoskeleton moves as to create as little force on the operator as possible 
-assuming the actuators are torque sources:
-\begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_o^{\tr} {\vf_o}_d
-+ \mJ_o^{\tr} \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
-\end{equation}
-$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
-force $\vf_o$
-Note that this does not require inverting the Jacobian matrix.
-We will see that the position control version does invert the Jacobian matrix.
-
-Applying this to force between the operator and the exoskeleton, and assuming
-that the operator directly applies exoskeleton joint torques,
-\begin{equation}
-\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
-\end{equation}
-So if our dynamic models and state estimation are perfect, we get:
-\begin{equation}
-\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
-\end{equation}
-
-One can add force damping terms, which sometimes help:
-\begin{equation}
-\vtau = \vtau_{\invdyn} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
-- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
-\end{equation}
-
-Note that impedance control can also be implemented using force control
-by setting:
-\begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-in addition to directly including the stiffness and damping terms in
-equation~\ref{eq:impedance}.
-
-\section{Mapping External Forces To Operator Forces Using Force Control}
-
-So far we have ignored or canceled the effect of 
-the exoskeleton contact with the outside world.
-A different control objective could be to map the external forces to 
-desired operator forces:
-\begin{equation}
-{\vf_o}_d = F_{ow} ( \vf_w )
-\end{equation}
-and cancel the external forces not transferred in the inverse dynamics.
-Everything in the above equation is estimated, so we are dropping the hats 
-$\hat{\mbox{}}$.
-
-For example, we could want the operator to feel a scaled (scale factor $\alpha$) 
-version of the external
-forces, as if they were contacting the operator directly.
-\begin{equation}
-{\vf_o}_d = \alpha J_o^{-\tr}(\vq) J_w^{\tr}(\vq) \vf_w
-\end{equation}
-and we scale the $\mJ_w^{\tr}(\vq) \vf_w$ term in the inverse dynamics
-by $(1-\alpha)$
-We are assuming we have any necessary force sensing between the exoskeleton and
-the world.
-
-If the operator applies joint torques, this simplifies to
-\begin{equation}
-{\vtau_o}_d = \alpha J_w^{\tr}(\vq) ( \vf_w )
-\end{equation}
-and the inversion of a Jacobian matrix is unnecessary.
-
-This would provide a way of enabling the operator to perceive and control
-external contact forces.
-
-\section{Autonomous exoskeleton control}
-
-\subsection{Autonomous exoskeleton position/velocity control}
-
-A control objective could be to have the exoskeleton to move autonomously,
-which is useful if the operator would like to rest, is attending to some other
-task, or is disabled.
-
-This is usually done by using stored trajectories or trajectories generated
-online that specify desired position $\vq_d$, 
-desired velocity $\dot{\vq}_d$, and
-desired acceleration $\ddot{\vq}_d$.
-The feedforward method is to compute the needed torques (potentially offline)
-using inverse
-dynamics (including the operator link inertias, link centers of mass, and link
-moments of inertia in the dynamic model) based on the desired postion, 
-desired velocity, and desired acceleration. Trajectory tracking errors are compensated
-for by feedback control designed using some other method. In most robotics
-fixed gain independent joint linear control is used.
-This method is safest in the sense that the feedback controller can be tested
-and verified independently of the desired trajectories and dynamic models.
-
-Computed torque approaches use online inverse dynamics computed from the estimated
-position, estimated velocity, and desired acceleration.
-In this case the feedback control to compensate for trajectory tracking errors can
-be independent of the inverse dynamics and its modeling errors, or the desired
-acceleration can be modified with feedback terms.
-\begin{equation}
-\ddot{\vq}_d^* = \ddot{\vq}_{d} + \mK_{\vq}^{\ddot{\vq}} ( \hat{\vq} - \vq_d ) 
-+ \mK_{\dot{\vq}}^{\ddot{\vq}} ( \hat{\dot{\vq}} - \dot{\vq_d} )
-\label{eq:computed-torque-control}
-\end{equation}
-The problem with putting the inverse dynamics in the feedback loop 
-(equation~\ref{eq:computed-torque-control})
-is that the
-effective gains of this control vary widely with the configuration $\vq$.
-Unmodeled dynamics (that also vary with $\vq$) limit the size of the gain matrices
-$\mK_{\vq}^{\ddot{\vq}}$ and $\mK_{\dot{\vq}}^{\ddot{\vq}}$. Because so many quantities
-are varying quite a bit, constant gain matrices may be quite small for a worst case
-design. It is also very difficult to evaluate and 
-verify this type of nonlinear feedback controller.
-For these reasons most current approaches to humanoid
-robots design the feedback controller
-separately (and manually) 
-and use constant gain independent joint linear feedback control.
-
-\subsection{Autonomous exoskeleton-world force control}
-
-A control objective could be to have the exoskeleton to autonomously
-apply a desired force.
-One could add
-\begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
-\end{equation}
-to a gravity compensated control or to control including inverse dynamics.
-
-Active force control could be added to improve the performance:
-\begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
-\end{equation}
-
-\subsection{Autonomous exoskeleton-world impedance control}
-
-Another control objective would be to allow the exoskeleton-world contact to
-be compliant or have an assigned impedance.
-Let's describe a small position offset of the exoskeleton-world
-contact points as $\vq_w$. We know that:
-\begin{equation}
-\dot{\vq}_w = \mJ_w \dot{\vq}
-\end{equation}
-and that
-\begin{equation}
-\vtau = \mJ_w^{\tr} \vf_w 
-\end{equation}
-This allows us to relate stiffness and damping in the external world coordinates
-with the stiffness and damping in exoskeleton joint coordinates.
-Here is the relationship for damping:
-\begin{equation}
-\vtau = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w \dot{\vq}
-\end{equation}
-So the damping in joint coordinates is $\vtau = \mK_{\dot{\vq}} \dot{\vq}$ and
-therefore
-\begin{equation}
-\mK_{\dot{\vq}} = \mJ_w^{\tr} \mK_{\dot{w}} \mJ_w
-\end{equation}
-The mapping is the same for stiffnes:
-\begin{equation}
-\mK_{\vq} = \mJ_w^{\tr} \mK_{w} \mJ_w
-\end{equation}
-So we should assign the exoskeleton joint position gains to be $\mK_{\vq}$ and
-the joint velocity gains to be $\mK_{\dot{\vq}}$ in equation~\ref{eq:impedance}.
-
-Similarly, we can use exoskeleton-world force control to improve performance,
-setting
-\begin{equation}
-{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
-\end{equation}
-and add the force error to equation~\ref{eq:impedance}: 
-\begin{equation}
-\mJ_w^{\tr} \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
-\end{equation}
-
-\section{Using Online Optimization: Quadratic Programming}
-\label{sec:qp}
-
-The dominant paradigm in force controlled (typically
-hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
-such as torque and center of pressure limits using online quadratic programming
-to solve the inverse dynamics equations. This approach can also be used
-to implement and trade off among the many control objectives described so far.
-
-Quadratic programming~\cite{Wikipedia} solves the following problem:
-Minimize
-\begin{equation}
-\vx^{\tr} \mQ \vx + c^{\tr} \vx
-\end{equation}
-with inequality constraints
-\begin{equation}
-\mA_1 \vx \le \vb_1
-\end{equation}
-and equality constraints
-\begin{equation}
-\mA_2 \vx = \vb_2
-\end{equation}
-The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
-torques $\vtau$, and contact forces $\vf$.
-For robots the discrepancy between actual and desired accelerations, joint
-torques, and contact forces are all minimized:
-\begin{equation}
-C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
-+ \vf^{\tr} \mW_{\vf} \vf
-\end{equation}
-$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
-different errors and error directions.
-The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
-Torque limits, friction limits, center of pressure location constraints,
-and unidirectional contact constraints are all
-enforced as inequality constraints.
-Variants of this optimization
-were performed every 1-2ms by various teams to control the
-Atlas robots in the DARPA Robotics Challenge.
-
-This approach can be used to achieve desired joint or synergy (such as center
-of mass) accelerations, center of mass dynamic behavior, center of pressure location,
-weight distribution, maintenance of a desired body part position, orientation,
-velocity, or trajectory. and joint/actuator position and velocity limits (indirectly).
-
-In the case of exoskeletons, 
-part of the optimization criterion could be to
-minimize contact forces on the operator:
-\begin{equation}
-C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
-\end{equation}
-Operator force feedback and impedance control could be implemented by specifying a
-desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
-\begin{equation}
-\vf_o = {\vf_o}_d
-\end{equation}
-or by adding a ``soft'' constraint in the form of a new term in the optimization
-criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
-\end{equation}
-
-To implement operator impedance control we set the desired operator contact force to:
-\begin{equation}
-{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
-\end{equation}
-Assuming that the operator directly applies exoskeleton joint torques, this becomes
-the equality constraint
-\begin{equation}
-\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-or is added to the optimization criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) 
-+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
-\end{equation}
-with
-\begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-
-\subsection{Full Atlas robot implementation.}
-
-\section{Underactuated and overactuated control}
-
-There are two kinds of underactuation we want to consider. The first
-is ``floating body dynamics'', where the robot is not bolted to the floor.
-Humanoid robots have floating body dynamics. A ``root'' location (usually somewhere
-on the pelvis) is chosen, and the position, velocity, orientation, and angular
-velocity of that point and the rigid body it is on
-are included in the state vector $\vq$. The corresponding dimensions of the
-velocity vector $\dot{\vq}$ are not actuated. We have been sweeping this
-issue under the rug so far. It is now useful to be more explicit. A selection matrix
-$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
-\begin{equation}
-\vtau = \mS \vu
-\end{equation}
-The rows in $\mS$ corresponding to the root linear and angular velocities
-are all zero.
-
-The second kind of actuation appears in exoskeletons where we rely on the
-operator to actuate a particular joint. For example, many lower body
-exoskeltons only actuate the joints in the sagittal plane, leaving lateral
-and twist (yaw) joints unactuated. We can actually handle this by using
-the matrix $\mS$ and reducing the size of the $\vu$ vector.
-
-We can also have passive actuation. This we can include by augmenting the
-dynamics equations to include the force equation of the passive element,
-such as a nonlinear spring.
-
-We can have multijoint actuation, which can be handled in $\mS$ as well.
-
-We can have redundant actuation (more than one actuator across any particular
-joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
-
-The modified dynamics equations are
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_o^{\tr}(\vq) \vf_o
-\label{eq:dynamics2}
-\end{equation}
-Note that $\ddot{\vq}$ and the new actuation vector $\vu$
-do not have the same dimensionality. What difference does that make?
-
-Underactuated joints reduce our ability to control behaviors directly
-through the actuators. In some cases we can control unactuated degrees of freedom
-through indirect effects over time, and in some cases those degrees of freedom
-are not ``controllable'' in the technical sense. Remember that we have a human
-operator that fully actuates the exoskeleton, so having some degrees of freedom
-not controllable by the actuators is not fatal to a design.
-
-The {\bf no control} case is unchanged with underactuation.
-
-{\bf Open loop control} continues to work with underactuation
-as long as a drive function can be found
-to generate the desired behavior.
-
-{\bf Gravity compensation} and {\bf friction compensation}
-still work on the actuated degrees
-of freedom. The operator needs to carry the load on the unactuated degrees of freedom.
-
-Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
-no longer works if $\mS$ is not invertable (not square or not full rank).
-In the underactuated case $\mS$ is not square. In this case {\bf online optimization}
-as described in Section~\ref{sec:qp} is used to ``perform inverse dynamics'' by
-finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
-and actuator commands $\vu$.
-
-Another approach to solving this problem is prioritized control, which does
-not include optimization~\cite{}. For reasons we will describe, we believe
-online optimization is much more flexible and can handle programming exoskeleton
-behavior much better, so we will not discuss prioritized control further.
-
-Operator {\bf force feedback} is implemented using online optimization by
-add the following ``hard'' constraint
-to the equality constraints in the quadratic program:
-\begin{equation}
-\vf_o = {\vf_o}_d 
-\end{equation}
-or adding a ``soft'' constraint as a term
-\begin{equation}
-(\vf_o - {\vf_o}_d)^{\tr} W_{\vf_o} (\vf_o - {\vf_o}_d)
-\end{equation}
-in the optimization criterion. $W_{\vf_o}$ is a weight to indicate how much operator
-force is acceptable (if ${\vf_o}_d = 0$) or how much operator force error is
-acceptable.
-Force damping can be implemented by adding a similar ``soft'' constraint:
-\begin{equation}
-\dot{\vf}_o^{\tr} W_{\dot{\vf}_o} \dot{\vf}_o
-\end{equation}
-to the optimization criterion.
-
-{\bf Impedance control} is implemented using online optimization by adding
-the impedance equations as ``hard'' equality constraints (to achieve them exactly)
-\begin{equation}
-\mJ_o^{\tr} {\vf_o} = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-or by adding the ``soft'' constraint ``equation error'' to the optimization criterion
-weighted by a weight factor $W_{imp}$ to indicate
-how large an error is tolerated:
-\begin{equation}
-\ve = \mJ_o^{\tr} {\vf_o} - \mK_{\vq}( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
-\end{equation}
-and
-\begin{equation}
-\ve^{\tr} W_{imp} \ve
-\end{equation}
-
-{\bf Mapping external forces to operator forces} can be done either with the
-``hard'' equality constraint:
-\begin{equation}
-\vf_o = F_{ow} ( \vf_w )
-\end{equation}
-or using the ``soft'' constraint term:
-\begin{equation}
-(\vf_o - F_{ow} ( \vf_w ))^\tr \mW_{fmap}  (\vf_o - F_{ow} ( \vf_w ))
-\end{equation}
-
-Similar techniques are used to handle {\bf autonomous exoskeleton position, velocity,
-force, and impedance control.}
-
-\subsection{How can the operator share control with autonomous control?}
-
-The use of online optimization makes sharing control with an operator easy.
-Dimensions (even powered dimensions) that are under the operator's control have
-low weights in the online optimization. It is also easy to map operator controls
-such as joysticks, sliders, and alternative interfaces to almost any desired
-control outcome by including these measurements in the online optimization.
-The operator and the controller can ``share'' or ``blend''
-control by making their respective
-weight matrices similar.
-
-\section{Operator-Exoskeleton Force Control using position controlled actuation}
-
-We will now consider position controlled actuators.
-At this point our dynamics formulation is no longer appropriate, since we
-are assuming the actuators are now position rather than torque sources.
-
-\subsection{Get out of the way control with position sources}
-
-The exoskeleton moves as to create as little force as possible between the
-operator and the exoskeleton. The actuators in this case may be position sources or
-force sources with high servo gains,
-but ultimately operator-exoskeleton forces are mapped
-to exoskeleton velocities or angular velocities.
-
-\begin{equation}
-\dot{\vq}_d = \mJ^{-1} (\mK_1 \vf)
-\label{eq:gootwcps}
-\end{equation}
-where $\dot{\vq}_d$ are the commanded exoskeleton joint velocities,
-$\mJ$ is an appropriate Jacobian matrix, $\mK_1$ is a gain matrix,
-and $\vf$ is the force vector to be controlled~\cite{IEEE06990981}.
-
-Inverting a Jacobian matrix is problematic when the matrix is
-nearly singular.
-Using a fixed or gain scheduled gain matrix may make more sense.
-\begin{equation}
-\dot{\vq}_d = \mK_3 \vf
-\end{equation}
-Simplifying equation~\ref{eq:gootwcps}
-by assuming the forces to be controlled
-are directly applied to the exoskeleton joints, and thus the Jacobian matrix
-is the identity matrix, gives us a similar equation:
-\begin{equation}
-\dot{\vq}_d = \mK_1 \vtau_o
-\end{equation}
-and we avoid inverting a Jacobian and problems with singularities.
-
-\subsection{Operator force feedback with position sources}
-
-We introduce a desired force to allow for more complex force control.
-\begin{equation}
-\dot{\vq}_d = \mJ^{-1} (\mK_1 (\vf - \vf_d))
-\end{equation}
-
-The exoskeleton could move as to provide a scaled version of exoskeleton-world
-contact forces, or some other (usually simple) mapping. 
-\begin{equation}
-\vf_d = \alpha \mJ_{ow} \vf_w
-\end{equation}
-
-\section{Specific Proposed and Implmented Exoskeleton Control Schemes}
-
-\subsection{Sensitivity Amplification Control}
-
-Sensitivity Amplification Control (SAC) used to control BLEEX
-is a dynamic cancellation technique (a.k.a.
-inverse dynamics, computed torque, or feedback linearization) that also tries
-to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
-using acceleration feedback. It is not clearly stated but it appears that the
-acceleration is the result of double differentiating position, as there do not
-appear to be velocity sensors on BLEEX. Low pass filtering is applied to reduce
-the high frequency noise amplified by the double differentiation process.
-
-\subsection{Integral admittance control}
-
-{\it [Exoskeletons] have the implicit property
-of causing a virtual modification of the dynamic response of
-the human limb. We use this property of the exoskeletons
-action to formulate a unified control design framework called
-Integral Admittance [torque to angle] Shaping, which designs exoskeleton con-
-trollers capable of producing the desired dynamic response
-for the assisted limb. In this framework, a virtual increase
-in the admittance of the limb is produced by coupling it
-to an exoskeleton that exhibits active behavior. Specifically,
-our framework shapes the magnitude profile of the integral
-admittance (i.e. torque-to-angle relationship) of the coupled
-human-exoskeleton system, such that the desired assistance is
-achieved. This framework also ensures that the coupled stability
-and passivity are guaranteed.}~\cite{Nagarajan_etal_2015}
-
-{\it ... the impedance of the coupled human-exoskeleton
-system needs to be reduced below that of the unassisted
-human limb. This implies that the exoskeleton needs to
-cancel its own impedance first and then compensate for
-at least a part of the human limb’s impedance. Therefore,
-the desired exoskeleton behavior must be that of a negative
-impedance ... Consequently,
-the feedback gains will all be
-positive. In other words, the exoskeleton controller uses
-positive feedback, and hence the exoskeleton exhibits active
-behavior, which is capable of performing net positive work
-on the limb. ...
-However, positive feedback naturally raises the question of
-stability, and so we now explain how coupled stability can be
-achieved. Although the exoskeleton exhibits active behavior,
-which can be potentially destabilizing, the controller can be
-designed such that the coupled human-exoskeleton system
-is stable and passive. ... 
-}~\cite{Nagarajan_etal_2015}
-
-{\it
-... Inertia compensation is more complex ...
-It can be shown that using only positive acceleration
-feedback ..., the gain margin of the coupled system
-reduces to the moment of inertia of the exoskeleton, which
-implies that the exoskeleton controller ... can at the most
-compensate for the exoskeleton’s own moment of inertia
-before going unstable. This implies that the moment of
-inertia of the coupled human-exoskeleton system cannot be
-reduced below that of the unassisted human limb, without
-compromising coupled stability. However, using low-pass
-filtered acceleration feedback, it can be shown that inertia 
-reduction can be achieved. ...
-This work uses filtered acceleration
-feedback with a second-order low-pass Butterworth filter. ...
-}~\cite{Nagarajan_etal_2015}
-
-This is a form of frequency domain virtual model control.
-
-\begin{verbatim}
-Kalman filter:
-P. Canet, “Kalman filter estimation of angular velocity and accelera-
-tion: On-line implementation,” McGill University, Montr ́eal, Canada,
-Tech. Rep. TR-CIM-94-15, Nov. 1994.
-\end{verbatim}
-
-\subsection{Dual Control Appproach}
-
-This approach implements a passive swing leg in addition to Sensitivity 
-Amplification Control for the stance leg.
-
-{\it The robot utilized the dual-mode control scheme, which is
-comprised of the active control for the stance phase and the
-passive control (using bypass valves) for the swing phase, to achieve high walking
-speed in the swing phase while supporting heavy loads in
-the stance phase. To reduce the sudden change of the torque
-command at the transition from the swing phase to the stance
-phase, a smoothing method is adopted. We also implemented
-a pre-transition method to take a foot off quickly for fast
-walking by predicting the change from the swing to the stance
-in advance.}~\cite{IEEE07222598}
-
-Bypass valves implement what Sarcos calls ``dangle''.
-
-Stance: virtual joint torque control method Kazerooni et al.(2005)
-When contact location between the wearer and the
-exoskeleton is not fixed and difficult to estimate, this method
-has been shown to be an effective method to generate the
-locomotion for an exoskeleton robot.[3],[4],[21]
-\begin{verbatim}
-H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
-the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
-Conf. Robot. Autom., pp. 4364-4371, 2005.
-Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
-for Human Performance Amplification,” University of California,2003.
-582
-Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
-Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
-Control,” Journal of computers, vol.9, No.1,202-208. 2014.
-\end{verbatim}
-
-To reduce
-sudden changes(command jump) at the phase transitions, a
-smoothing method [5],[19] is introduced and a pre-transition
-method is used to solve the swing delay due to the internal
-pressure(approx 5 bar).
-\begin{verbatim}
-H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
-Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
-Robotics Research, pp. 561-573, 2006.
-oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
-Based Control for a Rotary Series Elastic Actuator Assisting the knee
-\end{verbatim}
-
-Transition Control
-
-1) Smoothing Method: During transitions of the gait
-phase, discontinuity of the control command torque are
-occurred by the different condition for the fixed coordinate
-which is on the backpack at the swing phase or the foot at
-the stance phase. In this paper, to reduce this sudden change
-due to gait phase changes, a smoothing method is proposed
-as shown in Fig. 7. An exponential function is considered
-as the weighting function for smoothing as shown in (6).
-In this case, the weighting is small at the initial stage but
-it exponentially converges to one for supporting the load
-quickly.
-
-2) Pre-transition Method: The pre-transition method is
-that the passive mode is executed in the pre-swing phase
-prior to toe off. This dramatically reduces the moving
-
-\subsection{Ground Reaction Force Control}
-
-{\it
-The ground reaction
-forces (GRF) magnitude and direction is used to command the
-actuators. In some research the GRF sensors are used
-together with other sensors in the control architectures [17],
-while in other exoskeleton the control system is based on the
-GRF merely [RoboKnee, Honda].
-}
-\begin{verbatim}
-17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
-"Intention-based walking support for paraplegia patients with Robot
-Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
-\end{verbatim}
-
-\subsection{Virtual model control}
-
-The goal here is to make the exoskeleton imitate a desired (usually nonlinear)
-dynamic system,
-For running it could be a pogo stick or trampoline, for example.
-
-There are several possible versions of this type of control:
-\begin{enumerate}
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton actuator or internal forces.
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton acceleration.
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton-world contact forces.
-\item
-Map from exoskeleton-world contact forces and system state to exoskeleton actuator or internal forces.
-\item
-Map from exoskeleton-world contact forces and system state to exoskeleton acceleration.
-\item
-Some combination of the above.
-\end{enumerate}
-
-\begin{verbatim}
-virtual model approach: hopper, compass gait (J. Pratt)
-estimate trajectory (jumping) vs. program behavior (trampoline, hopper,
-   compass gait)
-- J' control + behavior
-- Inverse dynamics + behavior
-\end{verbatim}
-
-\subsection{Task Specific Control}
-
-Task specific control can involve switching low level control modes, or abstracting
-the behavior of the exoskeleton with a hierarchy:
-\begin{enumerate}
-\item
-Task specific control: Generate desired motions and contact forces between
-the operator and the exoskeleton, and the exoskeleton and the world.
-\item
-Sometimes there are intermediate levels of control.
-\item
-Exoskeleton control: Control the exoskeleton to generate the desired motions and
-desired contact forces, with the desired impedance or admittance.
-\end{enumerate}
-
-\section{State Estimation}
-
-DISCUSS HOW TO HANDLE SENSOR NOISE
-
-Estimating acceleration using double differentiation with low pass filtering
-currently works well on real exoskeletons.
-However, impacts, shock waves, and the fact that neither
-the operator's body parts or the exoskeletons parts are rigid bodies and the
-joints are not well defined suggest this approach can be improved with actual
-acceleration measurements.
-It is now possible to distribute MEMs IMUs throughout the exoskeleton to
-improve both velocity and 
-acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
-Sensitivity Amplification Control.
-
-Role of phase estimation, behavioral clocks, and phase reset events
-such as impacts.
-
-Disturbance observers~\cite{IEEE06197032}.
-
-\subsection{Full Atlas robot.}
-
-\section{Robustness to parametric modeling error}
-
-\section{Robustness to unmodeled dynamics}
-
-\subsection{Example unmodeled dynamics}
-
-Examples:
-\begin{itemize}
-\item
-A) hydraulic actuator dynamics.
-\item
-B) delay. 
-\item
-C) elastic straps sinusoidal oscillation.
-\item
-D) loose straps (bang coast bang oscillation)
-\end{itemize}
-
-\subsection{1 joint rotary example.}
-
-\subsection{2 joint rotary example: stance.}
-
-\subsection{2 joint rotary example: swing.}
-
-\subsection{4 link body example.}
-
-\subsection{Full Atlas robot.}
-
-\section{Survey Existing Exoskeletons}
 
@@
 
-\input{exos/exo-ul7}
+% \input{exos/exo-ul7}
 
@@
 
-1) We recommend a particular control 
-overall architecture in Section~\ref{sec:architecture}.
-We do not expect this to be controversial.
-
-2) We recommend actuation that can be treated as force or torque sources.
-This is a potentual problem for revolute electric actuation, as it
-is difficult to get necessary performance at low gear ratios (20-50)
-and fit the necessary torque sensing,
-
-3) We recommend the use of dynamic models of the exoskeleton, as has been done
-for the last decade starting with the BLEEX exoskeleton.
-We do not expect this to be controversial.
-
-4) We recommend using online optimization as part of the Lowest Level Controller
-in the form of quadratic programming. Online optimization
-was tested by the top biped robot
-teams in the DARPA Robotics Challenge, and worked well.
-A possible objection to online optimization
-is that it requires more substantial computing
-resources than current exoskeleton control methods. We feel the feasibility
-of this approach has been demonstrated and is easily achievable. Perceptual
-computing costs will dwarf computing costs for control, 
-so the computation cost is a non-issue.
-
-5) We recommend applying as many sensors as possible, and asking the performers
-to make it easy to add more by making the sensor network available in the design.
-This maximizes the probability of success
-with respect to control architectures and algorithms by enabling multiple control
-approaches to be implemented, refined, and support each other.
-In particular, we would like to see force sensing between the operator and the
-exoskeleton, including at the feet, as much force sensing between the exoskeleton 
-and the world as possible, but minimally full six-axis force/torque sensing at
-the exoskeleton feet (where they touch the ground). We would like to see multiple MEMs
-IMUs (measuring linear acceleration and angular velocity) installed across 
-the exoskeleton and at least one high quality fiber optic gyro IMU. 
-We would like to see high quality direct velocity sensing
-such as high count (100,000 counts/revolution) encoders and/or 
-analog rotary or linear tachometers on actuators and joints. 
-We would like to see actuator force or torque sensors
-(load cells or equivalents) with the measurement on the link side (rather than
-the actuator side) of any transmission. Electric current in motors and oil
-pressure in hydraulic pistons can also be used for actuator force estimation,
-but because these measurements are on the actuator side before the transmission
-and in the case of hydraulics before the oil seals on the piston, these measurements
-are greatly contaminated by friction. On the Atlas humanoid we typically saw 10Nm
-joint torque estimation errors for a system that estimated actuator output using
-oil pressure on each side of the piston head.
-A possible objection to this is additional cost. We feel it is a false
-economy to skimp on sensing. Leaving practical sensing out greatly increases
-the risk of poor performance.
-
-6) We recommend a ``symbiotic'' control system design should be used as a backup,
-in case more aggressive design philosophies such as ``invisible'' and ``natural''
-control system designs are not achievable.,
-The ``symbiotic'' control system design
-expects the operator to adapt to the exoskeleton and its control,
-and has the control customized for the operator.
-
-7) We recommend making use of virtual model control specialize
-for the various tasks. This can greatly increase operator/exoskeleton system
-performance over an exoskeleton that just carries a load and is otherwise
-``invisible''.
-
 \bibliographystyle{plain}
@@
 \end{document}
-
-sensitivity to
- modeling error
- sensor error
- unmodeled dynamics
-
-underactuated control problem
- - not fully actuated sagittal only.
- - how necessary is lateral actuation
- - what is the penalty
-  weight
-  drag of actuator
-
-how handle failure and damage after Oct 1
-
-what about a full model of the system
-
-Oct 5-9
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2015-09-24 control.tex
cga updates

@@
 
-\newcommand{\gc}{{\mbox {GravityCompensation}}}
-\newcommand{\invdyn}{{\mbox {\tiny ID}}}
-\newcommand{\invdynw}{{\mbox {\tiny ID+Fw}}}
+\newcommand{\ff}{{\mbox {\tiny ff}}}
+\newcommand{\myo}{{\mbox {\tiny o}}}
+\newcommand{\myx}{{\mbox {\tiny x}}}
+\newcommand{\ox}{{\mbox {\tiny ox}}}
+\newcommand{\ow}{{\mbox {\tiny ow}}}
+\newcommand{\xw}{{\mbox {\tiny xw}}}
+\newcommand{\xinvdyn}{{\mbox {\tiny x-ID}}}
+\newcommand{\xinvdynw}{{\mbox {\tiny x-ID+Fw}}}
+\newcommand{\oxinvdyn}{{\mbox {\tiny ox-ID}}}
+\newcommand{\oxinvdynw}{{\mbox {\tiny ox-ID+Fw}}}
 
@@
 Algorithms\\
-(Draft 2.3)}
+(Draft 3.0)}
 
@@
 \item
+Valve control (for fluidic systems):
+\begin{equation}
+\vu = \mK_{\vq} (\vq - \vq_d) + \mK_{\dot{\vq}} ( \dot{\vq} - \dot{\vq}_d )
++ \mK_{\vtau} ( \vtau - \vtau_d )
++ \mK_{\dot{\vtau}} ( \dot{\vtau} - \dot{\vtau}_d ) + \vu_{\ff}
+\end{equation}
+where $\vu$ is a valve command vector (with a valve command for each actuator),
+$\vq$ is the vector of joint angles, $\dot{\vq}$ are joint velocities,
+$\vtau$ are actutor forces or joint torques, and $\dot{\vtau}$ is the rate of change
+of the forces or torques (the time derivative of $\vtau$). 
+The subscript $d$ indicates a desired value,
+$\mK_{\vq}$ is a joint position gain matrix,
+$\mK_{\dot{\vq}}$ is a joint velocity gain matrix, 
+$\mK_{\vtau}$ is a joint force/torque gain matrix,
+and $\mK_{\dot{\vtau}}$ is a joint force/torque derivative gain matrix.
+$\vu_{\ff}$ are feedforward valve commands.
+This formulation takes into account the most important hydraulic actuator dynamics,
+and allows full state feedback
+(including $\vtau$ as part of the state), pole placement, and linear quadratic
+regulator (LQR) control of
+a third order model of actuator plus rigid body dynamics.
+If electric motor and power amplifier dynamics are taken into account in electric
+motor systems, there would be an analogous equation for the commanded
+motor voltage of
+an analog power amplifier or
+the control input of a PWM amplifier.
+The $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
+by a servo on piston differential pressure in a fluidic system or
+a current servo in an electrical system.
+\item
 Online optimization of accelerations, actuator forces and torques, and contact forces
@@
 
-PID or more complex feedback control laws are implemented by low level control.
+PID or more complex nonlinear feedback control laws such as sliding mode control
+are implemented by low level control. Note that a PID control law can be made
+(smoothly) nonlinear by changing the PID gains on each servo tick.
+Distributed low level control with limited communication typically controls each
+joint or each set of joints (a limb, for example) independently.
+Centralized low level control can support cross coupling of all joints,
+and implement ``modern'' control schemes such as full state feedback,
+pole placement, linear quadratic regulator (LQR) design, linear quadratic
+Gaussian (LQG) design, and linear quadratic Gaussian loop transfer recovery
+(LQG-LTR) design,
 
@@
 
-\section{Exoskeleton dynamics}
+\section{Operator and exoskeleton dynamics}
 
@@
 There are two sets of contact forces: operator-exoskeleton (which we denote
-with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
-with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
+with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
+with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
 rigidly held by other types of physical constraints, we can lump the operator
-body parts with the exoskelton parts which they are attached to in terms
+body parts with the exoskeleton parts which they are attached to in terms
 of link inertias, locations of center of mass, and moments of inertia,
-and just think of the whole system as one robot.
-We can also separate out the operator dynamics if this is more convenient
-(we separate them out and ignore them in what follows).
+and just think of the whole system as one robot (we indicate this with
+the subscript $\ox$ in dynamics equations.
+We can also separate out the operator dynamics (indicated by
+the subscript $\myo$) from the exoskeleton dynamics (indicated by
+the subscript $\myx$) if this is more convenient
+(we separate them out and ignore them in what immediately follows).
 If the operator is loosely connected to the exoskeleton, this introduces a
@@
 
-The dynamics of a revolute exoskeleton standing on the ground are:
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
-\label{eq:dynamics}
-\end{equation}
-where $\vq$ are the joint angles of
-the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
-velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
-centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
-friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
-contact forces)
-are the exoskeleton-world
-contact forces 
+The dynamics of an operator standing on the ground (or on parts of the exoskeleton
+that are on the ground):
+\begin{equation}
+\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
++ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \vtau_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
+\label{eq:odynamics}
+\end{equation}
+where $\vq$ are the operator joint angles,
+$\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
+joint velocities and accelerations,
+$\mM_{\myo}$ is the operator inertia matrix, 
+$\mC_{\myo}()$ are the operator Coriolis and
+centripetal forces, 
+$\mG_{\myo}()$ are the operator gravitational forces,
+$\mJ_{\ow}^{\tr}(\vq) \vf_{\ow}$ (Jacobian matrix multiplied with the
+contact forces) are the operator-world contact forces 
 expressed as exoskeleton joint torques, 
-$\vtau$ are the exoskeleton
-joint torques,
-and $\mJ_o^{\tr}(\vq) \vf_o$
-are the operator-exoskelton forces expressed
-as exoskeleton joint torques.
-
-We do not describe how to derive this equation or identify appropriate parameters
-for an exoskeleton. We can provide this information if requested.
+$\vtau$ are the operator joint torques due to muscle activity and passive tissue
+mechanics,
+and 
+$\mJ_{\ox}^{\tr}(\vq) \vf_{\ox}$
+are the operator-exoskeleton forces expressed
+as operator joint torques.
+We assume the operator joint friction is negligable.
+
+We assume the exoskeleton joints match some or all of the human joints.
+The dynamics of such an exoskeleton standing on the ground or attached to
+the human are:
+\begin{equation}
+\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} = \vtau_{\myx} + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
+\label{eq:xdynamics}
+\end{equation}
+$\mM_{\myx}$ is the exoskeleton inertia matrix,
+$\mC_{\myx}()$ are the exoskeleton Coriolis and centripetal forces,
+$\mG_{\myx}()$ are the exoskeleton gravitational forces,
+$\mF_{\myx}()$ are the exoskeleton friction forces,
+$\mJ_{\xw}^{\tr}(\vq) \vf_{\xw}$ (Jacobian matrix multiplied with the
+contact forces) are the exoskeleton-world
+contact forces expressed as operator joint torques, 
+$\vtau_{\myx}$ are the exoskeleton joint torques,
+and 
+$\mJ_{\ox}^{\tr}(\vq) \vf_{\ow}$
+are the operator-exoskeleton forces expressed
+as operator joint torques.
+
+The dynamics of the combined system when the operator is tightly strapped in
+or otherwise ``rigidly'' attached to the exoskeleton are:
+\begin{equation}
+\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow}
++ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
+= \vtau_{\myo} + \vtau_{\myx}
+\label{eq:oxdynamics}
+\end{equation}
+$\mM_{\ox}$ is the combined operator-exoskeleton inertia matrix,
+$\mC_{\ox}()$ are the combined Coriolis and centripetal forces,
+and $\mG_{\ox}()$ are the combined gravitational forces.
+Note that the forces between the operator and the exoskeleton
+do not appear explicitly in the
+combined dynamics, but they can be computed using either equation~\ref{eq:odynamics}
+or \ref{eq:xdynamics}.
+
+We are ignoring actuator dynamics, which we can handle by augmenting 
+the state vector and adding additional equations.
+
+We do not describe how to derive these equations or identify appropriate 
+model parameters.
+We can provide this information if requested.
 We are assuming that the equations have been derived taking into account constraints
@@
 in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
-We are assuming all degrees of freedom (joints) are actuated until 
-Section~\ref{sec:float}.
+Therefore foot touchdown and liftoff require special handling by controllers
+using these dynamics, which we will not describe in this paper.
+We are assuming all degrees of freedom (joints) are actuated by the operator,
+but not necessarily by the exoskeleton, until we get to 
+``floating body'' dynamics (Section~\ref{sec:float}).
 
@@
 to the identity matrix. This means the operator is directly applying torques
-at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
-and so is the world ($\vtau_w$) (a very unrealistic assumption). In this case:
-\begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \vtau_w = \vtau + \vtau_o
-\end{equation}
-
-Note that we are ignoring the human operator dynamics and operator muscle and tissue
-forces resulting in operator joint torques. Intent prediction will model these
-effects [intent.pdf].
-
-We are also ignoring actuator dynamics, which we can handle by augmenting 
-the state vector and adding additional equations.
+at the exoskeleton joints ($\vtau_{\ox}$) (a realistic assumption), 
+and so is the world ($\vtau_{\ow}$ and $\vtau_{\xw}$)
+(a very unrealistic assumption). In this case,
+the operator dynamics are:
+\begin{equation}
+\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
++ \vtau_{\ow} = \vtau_{\myo} - \vtau_{\ox}
+\end{equation}
+the exoskeleton dynamics are:
+\begin{equation}
+\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \vtau_{\xw} = \vtau_{\myx} + \vtau_{\ox}
+\end{equation}
+and the combined dynamics are:
+\begin{equation}
+\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \vtau_{\ow}
++ \vtau_{\xw} 
+= \vtau_{\myo} + \vtau_{\myx}
+\label{eq:oxdynamics}
+\end{equation}
+
+\section{What do we want from intent prediction?}
+
+For many of the following control systems, no operator intent estimation or
+prediction is needed.
+When we get to inverse dynamics, we need to know a desired acceleration
+$\ddot{\vq}_d$.
+In addition,
+operator force 
+estimates ($\hat{\vf}_{\myo}$ or $\hat{\vtau}_{\myo}$) or
+operator-exoskeleton force 
+estimates ($\hat{\vf}_{\ox}$ or $\hat{\vtau}_{\ox}$) are useful.
+For online optimization, the more we know about the future, the better online
+optimization can perform. We discuss this in Section~\ref{sec:predict}.
 
@@
 
-One control option is to use only passive controls, and no active controls
+One control option is to use only passive controls on a joint or on
+all joints, and no active controls
 or actuators.
@@
 exokeleton to do tasks.
+These passive devices have no controls, and can be included in the dynamics equations.
 In this case, 
-$\vtau$ is generated by passive devices such as springs or dampers.
+$\vtau_{\myx}$ is zero for any passive degree of freedom.
 
@@
 This approach typically 
-uses position and velocity control in either joint or
+uses impedance control in either joint or
 Cartesian coordinates
@@
 a time or phase index:
-$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
+$\vtau_{\myx} = \vtau_{\ff}(t)$ where $\vtau_{\ff}$ is a time or behavioral phase
 dependent feedforward torque vector.
@@
 
-Active gravity compensation is similar 
+Active exoskeleton gravity compensation is similar 
 to counterbalancing or physical gravity compensation, except it is done
-with exoskeleton actuators to cancel the exoskeleton weight (but not inertia).
-\begin{equation}
-\vtau = \hat{\mG}(\hat{\vq})
-\end{equation}
-where $\vtau$ are the exoskeleton joint torques, 
-$\hat{\mG}()$ are the gravitational forces calculated using an imperfect model, 
-and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
+with exoskeleton actuators to cancel the exoskeleton weight (but not inertia):
+\begin{equation}
+\vtau_{\myx} = \hat{\mG}_{\myx}(\hat{\vq})
+\end{equation}
+where $\vtau_{\myx}$ are the exoskeleton joint torques, 
+$\hat{\mG}_{\myx}()$ are the exoskeleton
+gravitational forces calculated using an imperfect model, 
+and $\hat{\vq}$ is the estimated configuration of the operator (and exoskeleton),
 The operator relies on mechanical backdrivability to drive
 the exokeleton to do tasks.
+The weight of the operator can also be supported by the exoskeleton in
+full gravity compensation:
+\begin{equation}
+\vtau_{\myx} = \hat{\mG}_{\ox}(\hat{\vq})
+\end{equation}
+where $\hat{\mG}_{\ox}()$ are the combined operator and exoskeleton
+gravitational forces calculated using an imperfect model,
 
@@
 \begin{equation}
-\vtau \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
-\end{equation}
-where $\hat{\mJ}$ is the estimated Jacobian matrix,
+\vtau_{\myx} \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
+\end{equation}
+where $\hat{\mJ}$ is the estimated Jacobian matrix for the center of mass,
 $\hat{m}$ is the estimated
@@
 \begin{equation}
-\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}}) 
+\vtau_{\myx} = \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}}) 
 \end{equation}
@@
 The exoskeleton actuator torques
-attempt to match the torques caused by inertial, Coriolis, centripetal,
+attempt to match the torques caused by exoskeleton inertial, Coriolis, centripetal,
 gravitational, and frictional forces.
 \begin{equation}
-\vtau = \vtau_{\invdyn} 
-= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
-+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
-+ \hat{\mG}(\hat{\vq})
-+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+\vtau_{\myx} = \vtau_{\xinvdyn} 
+= \hat{\mM}_{\myx}(\hat{\vq}) \ddot{\vq}_d
++ \hat{\mC}_{\myx}(\hat{\vq},\hat{\dot{\vq}})
++ \hat{\mG}_{\myx}(\hat{\vq})
++ \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}})
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} = \vtau_{\invdyn} 
-+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
+\vtau_{\myx} = \vtau_{\xinvdynw} = \vtau_{\xinvdyn} 
++ \hat{\mJ}_{\xw}^{\tr}(\hat{\vq}) \hat{\vf}_{\xw}
 \end{equation}
@@
 \begin{equation}
-\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
+\mM_{\myx}(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
 \end{equation}
 If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
-of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
-\begin{equation}
-\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
+of the exoskeleton with an effective inertia of 
+$(\mM^{-1}_{\myx}(\vq) \mJ_{\ox}^{\tr}(\vq))^{-1}$
+\begin{equation}
+\ddot{\vq} = \mM^{-1}_{\myx}(\vq) \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
 \label{eq:cancel}
@@
 $\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
-the operator can directly modify the trajectory.
+the operator can directly modify the current trajectory.
 
@@
 \begin{equation}
-\mM (\vq) \ddot{\vq} = \vtau_o
+\mM_{\myx} (\vq) \ddot{\vq} = \vtau_{\ox}
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
+\vtau_{\myx} = \vtau_{\xinvdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}_{\myx}(\hat{\vq})) \hat{\ddot{\vq}}
 \label{eq:change-inertia1}
 \end{equation}
-sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
+sets the apparent exoskeleton inertia to $\mM_d$ if the dynamic models are perfect, and
 state estimation produces not only perfect positions and velocities but
@@
 \begin{equation}
-\mM_d (\vq) \ddot{\vq} = \vtau_o
+\mM_d (\vq) \ddot{\vq} = \vtau_{\ox}
 \end{equation}
 In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
-issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
+issue for a matrix) than $\mM_{\myx}$ at any $\vq$. This type of control is vulnerable
 to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
-cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
+cancel the human's dynamics, the human stabilizes a possibly unstable exoskeleton
 (if the straps are tight enough).
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + \mK \hat{\vf}_o
+\vtau_{\myx} = \vtau_{\xinvdynw} + \mK \hat{\vf}_{\ox}
 \end{equation}
@@
 \begin{equation}
-(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
-\end{equation}
-By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
-\begin{equation}
-\mM_d (\vq) \ddot{\vq} = \vtau_o
+(\mK + \mI)^{-1} \mM_{\myx} (\vq) \ddot{\vq} = \vtau_{\ox}
+\end{equation}
+By setting $\mK = \mM^{-1}_d (\vq) \mM_{\myx} (\vq) - \mI$ we can achieve
+\begin{equation}
+\mM_d (\vq) \ddot{\vq} = \vtau_{\ox}
 \end{equation}
@@
 A key question is where does the desired acceleration $\ddot{\vq}_d$ come from?
-This is discussed in the companion paper [intent.pdf].
+This is discussed in a companion paper [intent.pdf].
 
@@
 operator in some other coordinate system, such as Cartesian coordinates.
-A different goal is to make the exoskeleton appear as a locally linear impedance to
+A different goal is to make the exoskeleton or combined operator and
+exoskeleton appear as a locally linear impedance to
 the outside world in some coordinate system.
@@
 To make the exoskeleton appear as a locally linear impedance in joint
-coordinates to the operator, we add position and velocity feedback to $\vtau$:
-\begin{equation}
-\vtau = \vtau_{\invdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+coordinates to the operator, we add position and velocity feedback to $\vtau_{\myx}$:
+\begin{equation}
+\vtau_{\myx} = \vtau_{\xinvdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \label{eq:impedance}
@@
 \begin{equation}
-\mM \ddot{\vq} - \mK_{\dot{\vq}} \dot{\vq} - \mK_{\vq} ( \vq - \vq_d ) = \vtau_o
+\mM_{\myx} \ddot{\vq} + \mK_{\dot{\vq}} \dot{\vq} + \mK_{\vq} ( \vq - \vq_d ) = \vtau_{\ox}
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + \mJ_o^{\tr}(\vq) {\vf_o}_d
-+ \mJ_o^{\tr}(\vq) \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
-\end{equation}
-$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
-force $\vf_o$
+\vtau_{\myx} = \vtau_{\xinvdynw} + \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
++ \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
+\end{equation}
+$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
+force $\vf_{\ox}$
 Note that this does not require inverting the Jacobian matrix.
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
+\vtau_{\myx} = \vtau_{\xinvdynw} + {\vtau_{\ox}}_d - \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d )
 \end{equation}
@@
 \begin{equation}
-\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
+\ddot{\vq} = - \mM^{-1}_{\myx} (\vq) ({\vtau_{\ox}}_d + \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d ))
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
-- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
+\vtau = \vtau_{\xinvdynw} + {\vtau_{\ox}}_d - \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d )
+- \mK_{\dot{\vtau}_{\ox}} \hat{\dot{\vtau}}_{\ox}
 \end{equation}
@@
 \begin{equation}
-{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+{\vtau_{\ox}}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \end{equation}
@@
 \begin{equation}
-{\vf_o}_d = F_{ow} ( \vf_w )
+{\vf_{\ox}}_d = F_{\ox-\xw} ( \vf_{\xw} )
 \end{equation}
@@
 \begin{equation}
-{\vf_o}_d = \alpha J_o^{-\tr}(\vq) J_w^{\tr}(\vq) \vf_w
-\end{equation}
-and we scale the $\mJ_w^{\tr}(\vq) \vf_w$ term in the inverse dynamics
+{\vf_{\ox}}_d = \alpha J_{\ox}^{-\tr}(\vq) J_{\xw}^{\tr}(\vq) \vf_{\xw}
+\end{equation}
+and we scale the $\mJ_{\xw}^{\tr}(\vq) \vf_{\xw}$ term in the inverse dynamics
 by $(1-\alpha)$
@@
 \begin{equation}
-{\vtau_o}_d = \alpha J_w^{\tr}(\vq) ( \vf_w )
+{\vtau_{\ox}}_d = \alpha J_{\xw}^{\tr}(\vq) ( \vf_{\xw} )
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
+\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
+\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\invdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\vtau = \vtau_{\xinvdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
 + \mJ_w^{\tr}(\vq) \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
@@
 operator to actuate a particular joint. For example, many lower body
-exoskeltons only actuate the joints in the sagittal plane, leaving lateral
+exoskeletons only actuate the joints in the sagittal plane, leaving lateral
 and twist (yaw) joints unactuated or passively actuated.
@@
 \mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_o^{\tr}(\vq) \vf_o
++ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
 \label{eq:dynamics2}
@@
 \begin{equation}
-C_e(\vx) = C_r(\vx) + \vf^{\tr}_o \mW_{\vf_o} \vf_o
+C_e(\vx) = C_r(\vx) + \vf^{\tr}_{\ox} \mW_{\vf_{\ox}} \vf_{\ox}
 \end{equation}
 Operator force feedback and impedance control could be implemented by specifying a
-desired operator force ${\vf_o}_d$ and adding a ``hard'' equality constraint:
-\begin{equation}
-\vf_o = {\vf_o}_d
+desired operator force ${\vf_{\ox}}_d$ and adding a ``hard'' equality constraint:
+\begin{equation}
+\vf_{\ox} = {\vf_{\ox}}_d
 \end{equation}
@@
 \begin{equation}
-C_e(\vx) = C_r(\vx) + (\vf_o - {\vf_o}_d)^{\tr} \mW_{\vf_o} (\vf_o - {\vf_o}_d)
+C_e(\vx) = C_r(\vx) + (\vf_{\ox} - {\vf_{\ox}}_d)^{\tr} \mW_{\vf_{\ox}} (\vf_{\ox} - {\vf_{\ox}}_d)
 \end{equation}
@@
 \begin{equation}
-\dot{\vf}_o^{\tr} W_{\dot{\vf}_o} \dot{\vf}_o
+\dot{\vf}_{\ox}^{\tr} W_{\dot{\vf}_{\ox}} \dot{\vf}_{\ox}
 \end{equation}
@@
 \begin{equation}
-\mJ_o^{\tr} {\vf_o} = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\mJ_{\ox}^{\tr} {\vf_{\ox}} = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \end{equation}
@@
 \begin{equation}
-\ve = \mJ_o^{\tr} {\vf_o} - \mK_{\vq}( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\ve = \mJ_{\ox}^{\tr} {\vf_{\ox}} - \mK_{\vq}( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \end{equation}
@@
 \begin{equation}
-{\vf_o}_d = \mJ_o^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
+{\vf_{\ox}}_d = \mJ_{\ox}^{-1} ( \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}} )
 \end{equation}
@@
 \begin{equation}
-\vtau_o = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
+\vtau_{\ox} = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
 \end{equation}
@@
 C_e(\vx) = C_r(\vx) 
-+ (\vtau_o - {\vtau_o}_d)^{\tr} \mW_{\vtau_o} (\vtau_o - {\vtau_o}_d)
++ (\vtau_{\ox} - {\vtau_{\ox}}_d)^{\tr} \mW_{\vtau_{\ox}} (\vtau_{\ox} - {\vtau_{\ox}}_d)
 \end{equation}
@@
 \begin{equation}
-\vf_o = F_{ow} ( \vf_w )
+\vf_{\ox} = F_{ow} ( \vf_w )
 \end{equation}
@@
 \begin{equation}
-(\vf_o - F_{ow} ( \vf_w ))^\tr \mW_{fmap}  (\vf_o - F_{ow} ( \vf_w ))
+(\vf_{\ox} - F_{ow} ( \vf_w ))^\tr \mW_{fmap}  (\vf_{\ox} - F_{ow} ( \vf_w ))
 \end{equation}
@@
 \begin{equation}
-\dot{\vq}_d = \mK_1 \vtau_o
+\dot{\vq}_d = \mK_1 \vtau_{\ox}
 \end{equation}
@@
 
-\subsection{Dual Control Appproach}
+\subsection{Dual Control Approach}
 
@@
 Note that adaptive or learning
-control does not promise to improve the kind of robustness exoskeltons need. 
+control does not promise to improve the kind of robustness exoskeletons need. 
 These approaches promise

### 2015-09-24 control.tex
more updates

@@
 and just think of the whole system as one robot (we indicate this with
-the subscript $\ox$ in dynamics equations.
+the subscript $\ox$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
+terms in the dynamics equations).
 We can also separate out the operator dynamics (indicated by
-the subscript $\myo$) from the exoskeleton dynamics (indicated by
-the subscript $\myx$) if this is more convenient
-(we separate them out and ignore them in what immediately follows).
+the subscript $\myo$ on ICCG terms) from the exoskeleton dynamics (indicated by
+the subscript $\myx$ on ICCG terms) if this is more convenient
+(we separate them out and ignore operator dynamics in what immediately follows).
 If the operator is loosely connected to the exoskeleton, this introduces a
@@
 \begin{equation}
-\vtau_{\myx} = \vtau_{\xinvdynw} + \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
-+ \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
+\vtau_{\myx} = \vtau_{\xinvdynw} - \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
+- \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
+\mJ_{\xw}^{\tr}(\vq) {\vf_{\xw}}_d
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d + \mJ_w^{\tr}(\vq) \mK_{\vf_w} (\vf_w - {\vf_w}_d)
+\mJ_{\xw}^{\tr}(\vq) \mK_{\vf_{\xw}} (\vf_{\xw} - {\vf_{\xw}}_d)
 \end{equation}
@@
 \begin{equation}
-\dot{\vq}_w = \mJ_w(\vq) \dot{\vq}
+\dot{\vq}_w = \mJ_{\xw}(\vq) \dot{\vq}
 \end{equation}
@@
 \begin{equation}
-\vtau = \mJ_w^{\tr}(\vq) \vf_w 
+\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
 \end{equation}
@@
 \begin{equation}
-\vtau = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq) \dot{\vq}
-\end{equation}
-So the damping in joint coordinates is $\vtau = \mK_{\dot{\vq}} \dot{\vq}$ and
+\vtau_{\myx} = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq) \dot{\vq}
+\end{equation}
+So the damping in joint coordinates is $\vtau_{\myx} = \mK_{\dot{\vq}} \dot{\vq}$ and
 therefore
 \begin{equation}
-\mK_{\dot{\vq}}(\vq) = \mJ_w^{\tr}(\vq) \mK_{\dot{w}} \mJ_w(\vq)
+\mK_{\dot{\vq}}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{\dot{w}} \mJ_{\xw}(\vq)
 \end{equation}
@@
 \begin{equation}
-\mK_{\vq}(\vq) = \mJ_w^{\tr}(\vq) \mK_{w} \mJ_w(\vq)
+\mK_{\vq}(\vq) = \mJ_{\xw}^{\tr}(\vq) \mK_{w} \mJ_{\xw}(\vq)
 \end{equation}
@@
 \begin{equation}
-{\vf_w}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
+{\vf_{\xw}}_d = - \mK_w ( \hat{\vq}_w - {\vq_w}_d ) - \mK_{\dot{w}} \hat{\dot{\vq}}_w
 \end{equation}
@@
 \begin{equation}
-\vtau = \vtau_{\xinvdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
-+ \mJ_w^{\tr}(\vq) \mK_{\vf_w} ( \hat{\vf}_w - {\vf_w}_d )
+\vtau_{\myx} = \vtau_{\xinvdyn} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
++ \mJ_{\xw}^{\tr}(\vq) \mK_{\vf_{\xw}} ( \hat{\vf}_{\xw} - {\vf_{\xw}}_d )
 \end{equation}
@@
 is ``floating body dynamics'', where the robot is not bolted to the floor.
+Contacts are made and broken during walking, running, and many other behaviors,
+and contacts slip.
 Humanoid robots (and humans) have floating body dynamics.
@@
 are included in the state vector $\vq$. The corresponding dimensions of the
-velocity vector $\dot{\vq}$ are not actuated. A selection matrix
-$\mS$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
-\begin{equation}
-\vtau = \mS \vu
+velocity vector $\dot{\vq}$ are not actuated. A Jacobian matrix
+$\mJ_{\myx}$ is used to map from actuation variables $\vu$ to joint torques $\vtau$:
+\begin{equation}
+\vtau_{\myx} = \mJ_{\myx} \vu_{\myx}
 \end{equation}
@@
 are all zero.
+To maintain compatability in our joint representation, we will similarly
+map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
+\begin{equation}
+\vtau_{\myo} = \mJ_{\myo} \vu_{\myo}
+\end{equation}
 
@@
 and twist (yaw) joints unactuated or passively actuated. 
+Many ``exoskeletons'' or orthoses only actuate one joint on each limb (typically
+the ankle or knee). In these situations the question arises as to what is
+the scope of control: should the device only consider
+just the actuated joints,
+or should it consider a wider range of operator joints? Many of these devices
+do not have enough sensing (such as an IMU) to even estimate the state of other
+operator joints. We believe that for high performance the entire state of the
+operator must be considered in control, and we describe a way to do it in
+Section~\ref{sec:qp}.
+However, we are not aware of any system that goes this far in terms of control.
 
@@
 
-The underactuation can be handled by reducing the size of the $\vu$ vector
-and changing the matrix $\mS$ to reflect the missing actuation.
+``Missing joint''
+underactuation can be handled by reducing the size of the $\vu_{\myx}$ vector
+and changing the matrix $\mJ_{\myx}$ to reflect the missing actuation.
 We can have redundant actuation (more than one actuator across any particular
-joint). Here we augment $\vu$ and $\mS$ to accomodate extra actuators.
-We can have multijoint actuation, which can be handled in $\vu$ and $\mS$ as well.
+joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accomodate extra actuators.
+We can have multijoint actuation, which can be handled in $\vu_{\myx}$ 
+and $\mJ_{\myx}$ as well.
 
@@
 \begin{equation}
-\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
-+ \mJ_w^{\tr}(\vq) \vf_w = \mS \vu + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
-\label{eq:dynamics2}
-\end{equation}
-Note that $\ddot{\vq}$ and the new actuation vector $\vu$
+\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
++ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \mJ_{\myo} \vu_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
+\label{eq:odynamics2}
+\end{equation}
+\begin{equation}
+\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} = \mJ_{\myx} \vu_{\myx} + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
+\label{eq:xdynamics2}
+\end{equation}
+\begin{equation}
+\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
++ \mF_{\myx}(\vq,\dot{\vq})
++ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow}
++ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
+= \mJ_{\myo} \vu_{\myo} + \mJ_{\myx} \vu_{\myx}
+\label{eq:oxdynamics2}
+\end{equation}
+Note that $\ddot{\vq}$ and the actuation vectors $\vu_{\myo}$ and $\vu_{\myx}$
 do not have the same dimensionality. What difference does that make?
@@
 
-Straightforward {\bf online inverse dynanamics} as described in Section~\ref{sec:invdyn}
-no longer works if $\mS$ is not invertable (not square or not full rank).
-In the underactuated and overactuated cases $\mS$ is not square. 
-In these cases {\bf online optimization}
-as described in the next Section is used to ``perform inverse dynamics'' by
+Full (operator) body
+{\bf online inverse dynamics} as described in Section~\ref{sec:invdyn}
+no longer works if $\mJ_{\myx}$ is not invertable (not square or not full rank).
+In the underactuated and overactuated cases $\mJ_{\myx}$ is not square. 
+For a partial (operator) body
+exoskeleton we can rewrite the dynamics in terms of only those
+joints the exoskeleton actuates, and invert those dynamics to cancel exoskeleton
+effects on those joint. This is what has been done with the BLEEX exoskeleton,
+for example.~\cite{}.
+
+However, {\bf online optimization}
+as described in the next Section can be used to perform full (operator)
+body inverse dynamics by
 finding desirable combinations of accelerations $\ddot{\vq}$, contact forces $\vf$,
-and actuator commands $\vu$.
+and actuator commands $\vu_{\myx}$.
 
@@
 \end{equation}
+
+Note that the same approach could be used to minimize or control internal forces
+such as skeletal forces (forces between bones) at the operator joints to protect
+the ankle, hip, and knee joints, for example.
+MORE GOES HERE.
 
@@
  unmodeled dynamics
+* future (after Oct 1) simulation studies
 
 underactuated control problem
- - not fully actuated sagittal only.
  - how necessary is lateral actuation
- - what is the penalty
+   * we don't know
+ - * what is the penalty
   weight
   drag of actuator
-
-how handle failure and damage after Oct 1
-
-what about a full model of the system
-
-Oct 5-9
+* future (after Oct 1) simulation studies
+
+how handle failure and damage 
+* after Oct 1

### 2015-09-24 exos/exosText_MIT.tex
added Matts stuff but still need to get his bib files

% % Created 2015-09-15 Tue 11:46
% \documentclass[11pt]{article}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{fixltx2e}
% \usepackage{graphicx}
% \usepackage{longtable}
% \usepackage{float}
% \usepackage{wrapfig}
% \usepackage{rotating}
% \usepackage[normalem]{ulem}
% \usepackage{amsmath}
% \usepackage{textcomp}
% \usepackage{marvosym}
% \usepackage{wasysym}
% \usepackage{amssymb}
% \usepackage{hyperref}
% \usepackage{color}
% \usepackage{soul}
% \tolerance=1000
% \usepackage[margin=1in]{geometry}

% \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

% %\author{Alex Ansari}
% %\date{}
% %\title{TALOS}
% %\hypersetup{
% %  pdfkeywords={},
% %  pdfsubject={},
% %  pdfcreator={Emacs 24.3.1 (Org mode 8.2.10)}}

% \begin{document}
\section{MIT Exoskeleton}

Design is based  on fact that evidence from biology and passive walking devices suggest that legged locomotion can be very efficient.  The main physical concept behind this efficiency being that there is a gait cycle for a pair of legs that naturally exchanges energy between gravity and inertia.  The main mechanical design principles centered around using a combination of passive and active elements.  In particular, the MIT design uses {\bf springs, variable impedance joints, and powered actuators.}  The suit generally  functions as a lightweight, underactuated robot which runs in parallel to an operator and supports the weight of a payload.

Additionally, the leg structure of the suit allowed weight from the suit as well as payload to be transferred directly to the ground.

Additionally, there are several other design principles that the MIT team discusses which are especially relevant to this report.  The first is that joint {\bf powers scale linearly with mass}  The second is that altering the gait pattern of the operator has been shown to increase the physiological energy expended during locomotion.  Thus, the specifications for actuation and control for their system are extracted from the angle, torque , and power data of human walking joint patterns.

\subsection{Actuator specifications}

The MIT exoskeleton uses a single actuator located at the hip.  The actuator is designed around a total system weight of 165 kg.  The maximum scaled hip torque for a system with this mass is approximately negative 130 Nm during the stance phase and approximately 100 Nm during the swing phase of the gait.  The hip is chosen as the actuated joint because proximal mass is metabolically less expensive in walking than distal mass.

The specific choice of actuation for the MIT exoskeleton hip joint was a {\bf linear series elastic actuator}.  This choice was made because it provided a lightweight and relatively inexpensive means of implementing force control with a bandwidth similar to that of natural muscle.  A model of the actuator is shown in Figure \ref{fig:mitSEA}.
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/MIT/mitSEA}
  \caption{}
  \vspace{-0.2in}
 \label{fig:mitSEA}   
 \end{figure} 
Like all series elastic designs the actuator shown in Figure \ref{fig:mitSEA} includes a spring in series with the output of the motor drive system of the actuator.  The motor-drive mechanism consists of a {\bf brushed DC motor driving a 3mm ball screw via a 2:1 reduction belt drive}.  The nut of the ball screw is coupled to the actuator output shaft via four compression die springs.  The compression of the die springs are measured using a linear potentiometer, which is how forces at the output are measured.

The motor chosen for the {\bf actuator was a 150W Maxon RE40 Brushed DC motor}, which was chosen based on the bio-mechanical data and for its power and weight ratio.

The force bandwidth for the actuator was characterized during both stance and swing phases of the walking gait cycle.  {\bf During stance, the bandwidth of the actuator was 35 Hz, while the bandwidth during swing was found to be 40Hz.}

 \subsection{Exoskeleton Specifications}

 The MIT exoskeleton design has a total of fourteen degrees of freedom, three at each hip, one at each knee, two at each ankle, and one in each foot.

 A cam mechanism was implemented at the hip joint to enable hip ab/adduction.  This system was necessary because, during abduction, there is was a length difference between the operator and exoskeleton leg which resulted from having non-collocated centers of rotation.  The cam mechanism automatically adjusts the exoskeleton leg length such that the center of rotation of the exoskeleton hip is projected onto the biological hip center of rotation.  A picture of the MIT exoskeleton is shown in Figure \ref{fig:MITsuit}.
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/MIT/MITsuit}
  \caption{}
%  \vspace{-0.2in}
 \label{fig:MITsuit}   
 \end{figure}

 The interface between the suit and operator consisted of shoulder straps, a compliant waist belt, thigh cuffs, and a shoe connection.

 The {\bf variable damper knee mechanism} in the flexion/extension degree of freedom was implemented using a magnetorheological damper.  The damper could exert a {\bf maximum braking torque of 60 Nm and consumed on average 1W of electrical power}.

 A linear spring located at the ankle was designed to capture negative energy during dorsilflexion.  The energy stored was released during plantar flexion to add energy to the system as the foot comes off the ground. A urethane spring was used as a liner spring in a lever compression assembly.  The spring constant of the urethane was 247 kN/m, the lever arm of the assembly was 0.0381 m, which gave the rotary spring a stiffness of 356 Nm/rad.

 A unidirectional spring assembly was also used in the hip ab/adduction degree of freedom.  The spring allowed the hip to freely abduct.  It was calculated that 8 Nm were required during adduction.  To this end, the effective rotary stiffness of the spring assembly was 96 Nm/rad.

The suit was powered using a 48V battery pack.

The suit was instrumented with rotary potentiometers at the hip and knee.  Additionally, strain gauges were placed on the structure of the exoskeleton shank, which measured sagittal bending moment and compression force

 \subsection{Control Specifications}

 The control of the MIT exoskeleton functions around the fact that desired actuation and active damping at the knee are functions of gait cycle, at that the values of these desired functions could be determined from human walking data.  Fundamentally, the control strategy works as a state machine, which uses joint angle and measured forces to implement state transitions.

\begin{figure}[thpb]
\centering
\includegraphics[width=4.in]{exos/figs/MIT/walkingCycle}
  \caption{}
  \vspace{-0.2in}
 \label{fig:walkingCycle} 
 \end{figure} 
The walking cycle around which the MIT control scheme was designed is shown in Figure \ref{fig:walkingCycle}.  The individual control strategies in the various phases of the walking cycle are highlighted in Figure \ref{fig:walkingCycle}.  For the hip there are four distinct strategies.  During the thrust phase, the hip exerts a torque which helps to raise the center of mass of the system.  During the extension spring phase, a virtual spring stiffness is programmed which allows energy to be virtually stored while the center of mass moves forward.  During the swing assist phase the virtual energy is released, resulting in a torque being applied which assists in swinging the leg forward.  Finally, in leg retraction a torque is applied to help with foot placement and weight acceptance.  A diagram which illustrates how the sensors in the shank and hip joint are used to switch between these various phases is show in Figure \ref{fig:hipControl} 
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/MIT/hipControl}
  \caption{}
  \vspace{-0.2in}
 \label{fig:hipControl} 
 \end{figure}

 The knee controller in the MIT exoskeleton effectively functions independently from he hip controller.  During knee strike, the variable damper in the knee exerts a torque proportional to the rotational velocity of the knee joint.  A residual magnetic field remained in the knee joint after it was turned off, creating a resistive torque.  The knee needed to be actively demagnetized at full extension during the last phase of stance to allow the knee to swing freely during the swing phase.  The state machine diagram that illustrates how the various phase transitions for the knee joint were implemented using the exoskeleton's onboard sensors is shown in Figure \ref{fig:kneeControl} 
  \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/MIT/kneeControl}
  \caption{}
  \vspace{-0.2in}
 \label{fig:kneeControl} 
 \end{figure}

%  \begin{figure}[thpb]
%\centering
%\includegraphics[width=3.in]{figs/torqueCon}
%  \caption{}
%  \vspace{-0.2in}
% \label{fig:IHMCTOR}   
% \end{figure}
%

 \subsection{Opinion}

 The MIT hardware and control system present a very well designed and thought out system which focus primarily on efficiency in design.  The series elastic linear ball screw hardware design seems to be a very smart way to implement efficient drives as well as to perform high fidelity force control.  Using a strain gauge in the shank was also a novel means of of measuring ground reaction forces.  This design may offer benefits in terms of repeatability over different operators as foot fall patterns for different users may include significant variations.

 The details on the controllers were somewhat sparse in the sense that, at least in \cite{}, the exact means through which the added torques were derived was not provided.  Overall though, the controllers were intelligently designed in the sense that they explicitly took the biomechanics of walking into account in their design and implementation.  A similar approach for a variety of different behaviors may be very beneficial to the TALOS project.

% \end{document}

% Conor James Walsh and Daniel Paluska and Kenneth Pasch and William Grand and Andrew Valiente and Hugh Herr, "Development of a lightweight, underactuated exoskeleton for load-carrying augmentation," Proceedings of the 2006 IEEE International Conference on Robotics and Automation, 2006, pp. 3485 - 3491.

% Conor James Walsh and Kenneth Pasch and Hugh Herr, "An autonomous, underactuated exoskeleton for load- carrying augmentation, " International Conference on Intelligent Robots and Systems, 2006, pp. 1410-1415.

### 2015-09-24 exos/exosText_bodyExt.tex
added Matts stuff but still need to get his bib files

% % Created 2015-09-15 Tue 11:46
% \documentclass[11pt]{article}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{fixltx2e}
% \usepackage{graphicx}
% \usepackage{longtable}
% \usepackage{float}
% \usepackage{wrapfig}
% \usepackage{rotating}
% \usepackage[normalem]{ulem}
% \usepackage{amsmath}
% \usepackage{textcomp}
% \usepackage{marvosym}
% \usepackage{wasysym}
% \usepackage{amssymb}
% \usepackage{hyperref}
% \usepackage{color}
% \usepackage{soul}
% \tolerance=1000
% \usepackage[margin=1in]{geometry}

% \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

%\begin{document}
\section{Body Extender}

The body extender is one of the few existing full-body exoskeletons.  The Body Extender's main objective is to increase the forceful interaction capabilities, specifically handling heavy loads, of an operator in difficult and unknown environments.  The overall suit is composed of four robotic limbs with kinematics designed to be anthropomorphically similar.  The suit contains twenty-two independently actuated degrees of freedom.  Note that the Body Extender is the only fully actuated suit included in this report.  The fully actuated design of the suit was chosen based on the fact that handling, or rather manipulating heavy objects requires forces and torques that far exceed human capabilities.  Note also that having many actuated degrees of freedom means that fully actuated suits will be very heavy, and thus will require specialized controllers.

The main design objectives for the Body Extender included  requiring that operators be able to use the hardware with a minimal amount of training.  Thus, the system should not require any substantial modification the a human's motor habits.  This objective includes the fact that the mass distribution of the suit, of special consideration in these heavy systems, be similar to that of an unloaded operator.  Additionally, the resistive forces between the operator and suit should always be minimized, even when the suit is not loaded externally.

A picture of the Body Extender exoskeleton and its five main components, four limbs and a torso, are shown in Figure \ref{fig:bodyExt}.
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/bodyExt/bodyExt}
  \caption{}
  \vspace{-0.2in}
 \label{fig:bodyExt}   
 \end{figure}

\subsection{Actuator Specifications}

 Each of the actuators in the Body Extender suit, execpt the forearm pronosupination are dirven by a linear actuator which is composed of an electrical motor, incremental encoder, ball screw, and angular contact ball bearings.  The motors are frameless DC torque controlled motors.  The {\bf peak torque of the motors are 7 Nm, with a maximum continuous torque of 6 Nm}.  Each motor weighs 1.4 kg.  The {\bf linear actuator assembly is able to supply 8000 N with a total weight of 2.4 kg}.

 For joints that have a small range of motion and correspondingly low torque requirements, the linear actuators are used in straightforward lever mechanism configurations.  For degrees of free which require relatively large ranges of motion as well as torque, a motion conversion system is used to transform the linear actuator motion into rotational motion.  The motion conversion unit consisted of a pantograph, two metallic tendons, and an output pulley connected to the output link.  A CAD drawing of the motion conversion unit is shown in Figure \ref{fig:motionConv}. 
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/bodyExt/motionConv}
  \caption{}
%  \vspace{-0.2in}
 \label{fig:motionConv}   
 \end{figure}
 There are two configurations of the motion conversion unit.  The first has a {\bf maximum continuous output of 500 Nm and the second a maximum continuous output of 270 Nm}.  The two units weigh 6 kg and 5 kg respectively, and have {\bf mechanical efficiency of about 85\%}.  {\bf The speed of the actuators are about 60 deg/s}.

 \subsection{Exoskeleton Specifications}

 The Body Extender exoskeleton consists of twenty-two independently actuated degrees of freedom.  The suit has two arms, two legs,and a torso, or central backpack unit.

 Each of the two arms on the system has a servo-amplified degree of freedom for grasping heavy objects.  The grasping force of various objects is achieved using a force sensing trigger mechanism; the grasping force is proportional to the force applied to the trigger.

 The suit contains twenty-two incremental encoders which measure angular displacement of each of the electric motors.  There are also {\bf five six-axis force/torque sensors mounted at the five connecting points between the user and the device; two at the hands, two at the feet, and one on the trunk.}  These sensors directly measure the forces and torques between the user and the suit.  There are additionally two single axis force sensors integrated into the grippers that measure the force applied by the operator on the ``gripper triggers," as well as {\bf a three axis accelerometer integrated into the backpack unit which estimates the trunk orientation}.  Lastly, accelerometers are distributed throughout the system to monitor link dynamics, serving the roll of supporting overall equilibrium and internal stability during manipulation tasks.

{\bf The control software is distributed on one central processing unit as well as sixteen local units located throughout the system}.  The communication between processing units is handled using a high-speed field bus.

{\bf The system is powered using thirteen batteries in series, which provide a nominal voltage of 78 V}.  When the system is run in stand alone mode, the batteries can provide eight hours of continuous work.

 \subsection{Control Specifications}

 The body extender control algorithm centers around a {\bf feedback linearization policy} that depends on measuring the force between the operator and suit, knowing the contact state of the suit and environment, decoupling the control of each of the limbs for each other, and having each limb be fully actuated.  The dynamics for each of the Body Extender limbs are written in standard form as  
\begin{equation}
B_i(q) \ddot{q} + C_i(q,\dot{q})\dot{q} + G_i(q) = \tau + J^T_{iu}F_s + J^T_{ia}F_a,
\end{equation}
where the $i = \{ i,2,3,4\}$ where each integer corresponds to a different limb, $\tau$ represents the actuator torques, $F_s$ the measured forces between the operator and suit, and $F_a$ the unmeasured contact forces on the suit.  The $B_i$, $C_i$, and $G_i$ terms represent the limb inertia, Coriolis and centrifugal, and gravitational torques.

The authors in \cite{} use a decoupling strategy in their controllers which enables the individual limbs in the Body Extender to be somewhat decoupled from each other.  They claim that they accomplish this using the following model:
\begin{equation}
B_i(q) \ddot{q} + C_i(q,\dot{q})\dot{q} + G_{iT}(q) = \tau + J^T_{iu}F_s + J^T_{ia}F_a - B_{iT}{q}\dot{V}_T - C_{iT}(q,\dot{q})V_T,
\end{equation}
where, $V_T$ is the trunk velocity, $G_{iT}$ the gravitational effect of the trunk on the limb, $B_{iT}$ the inertia of the trunk acting on the limb, and $C_{iT}$ the trunk Coriolis and centrifugal term.  Note that $V_T$ and $\dot{V}_T$ are directly measured through the accelerometer and gyroscope in the trunk.

The basis of the feedback linearizing controller in \cite{} operates under two fundamental assumptions, and only appears to relate directly to the case of a single one of the manipulator arms.  The first assumption of the controller is that the mass and inertial properties of an object being manipulated are estimated online.  The second is that the forces measured between the operator and suit can be modeled as
\begin{equation}
F_s = M_p \dot{V}_e + G_p,
\label{eq:forceEq}
\end{equation}
where $M_p$ is an effective mass of the end effector, $\dot{V}_e$ is the acceleration of the end effector, and $G_p$ is an effective gravitational term.  Note that the gravitational term is zero if the arm is not holding an external load.  The importance of these two assumptions will be more clear after the analytical forms of the feedback linearizing controllers are introduced.

The dynamics of the arm when not manipulating a load are
\begin{equation}
B(q) \ddot{q} + C(q,\dot{q})\dot{q} + G(q) = \tau. \notag
\end{equation} 
The feedback linearizing controller for this case has the form
\begin{equation}
\tau = B(q)(\ddot{q}_d + K_v \dot{\tilde{q}} + K_p \tilde{q}) + C(q,\dot{q})\dot{q} + G(q), \notag
\end{equation}
where $\tilde{q} = q_d - q$.  The details are unclear in \cite{}, but it appears that the authors either propose using predefined trajectories to define $\ddot{q}_d$, $\dot{q}_d$, and $q_d$, or use the assumption in \eqref{eq:forceEq} to determine these values.  
%\begin{equation}
%F_s = M_p \dot{V}_e
%\label{eq:assump1}
%\end{equation}
Equation \ref{eq:forceEq} relates the measured force to the derivative of the end effector velocity $\dot{V}_e$.  This end effector velocity can then be related to the desired joint velocities using the analytical jacobian transformation \[ \ddot{q}_d = J(q)^{-1}(\dot{V}_e - \dot{J}(q,\dot{q})\dot{q}).\]  The values for $\dot{q}_d$ and $q_d$ can then be found through integration.  It is not clear the how authors explicitly decouple the human dynamics from the robot in this case.

In the case where the exoskeleton arm is manipulating a load, the feedback linearizing controller has the augmented form
\begin{equation}
\tau = B(q)(\ddot{q}_d + K_v \dot{\tilde{q}} + K_p \tilde{q}) + C(q,\dot{q})\dot{q} + G(q) -J^T(q) F_s -J^T(q)F_a. \notag
\end{equation}
The main difference in this case is that the controller here explicitly compensates for the measured human-robot interaction force $F_s$, and the controller includes the un-sensed forces $F_a$ which arise from the object being manipulated.  The object interaction forces thus need to be estimated, which was the first fundamental assumption on which the control scheme for the Body Extender relied.  Because the point at which the manipulator arm and object are connected is assumed to be well known, and the arm is instrumented with inertial sensors, estimating the interaction forces of the manipulated object is be assumed to be equivalent to estimating its inertial properties.  As outlined in \cite{}, this is accomplished using a standard regressor which uses the input and output of the feedback linearizing controller as the basis for the regression.  The details of this process are left to \cite{}.

\subsection{Opinion}

The Body Extender suit is one of the only full-body exoskeletons available today, and as such is a somewhat remarkable system. In terms of how the Body Extender system is designed and how it is controlled, the suit may be limited. Clearly, for the task of handling heavy objects a fully actuated suit may be the best option, but generally speaking this seems almost like an over engineered solution.  The Body Extender is a heavy, overly clunky suit.  This suit may be alright for quasi-static operations, but it is unlikely that it will be useful for more than pick and place type operations.  Again, seeing as this is basically what it was designed for, that seems alright.

In terms of the controllers highlighted in \cite{}, there are some good ideas.  For example, using a regressor to estimate unknown external forces may be an idea which is directly applicable to the TALOS project.  In the case of the Body Extender this regression is dependent on knowing at the location of the external force  relative to the suit, but the idea in general is interesting.  The idea of using the measured forces to generate desired trajectories around which a stabilizing controller is designed also appears to be a relatively good idea.  Note that, like any of the model-based control approaches outlined in the exoskeleton literature, or robotics literature for that matter, the feedback linearizing controller in this work is dependent on knowing the model of the robot hardware.

In terms of the control of the overall suit, it was difficult to find any control papers about the Body Extender that covered anything other than upper-body single-arm control of a manipulated object.  The decoupling strategy discussed above, decoupling the individual limbs from each other, seems somewhat feasible, but is not the best way to approach coordinated limb control.  It was difficult to find literature documenting the specifics of lower-limb control for the Body Extender.  Presumably a similar strategy to that discussed here, with the addition of gravity compensation for the suit, is used.

% \end{document}

% S. Marcheschi, F. Salsedo, M. Fontana, and M. Bergamasco, ?Body ex- tender: Whole body exoskeleton for human power augmentation,? in Proc. IEEE Int. Conf. Robotics Automation, 2011, pp. 611?616.

% G.P. Rosati Papini, and C.A. Avizzano, "Transparent Force Control for Body Extender," 2012 IEEE RO-MAN: The 21st IEEE International Symposium on Robot and Human Interactive Communication, 2012, pp. 138-143.

### 2015-09-24 exos/exosText_dualMode.tex
added Matts stuff but still need to get his bib files

% % Created 2015-09-15 Tue 11:46
% \documentclass[11pt]{article}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{fixltx2e}
% \usepackage{graphicx}
% \usepackage{longtable}
% \usepackage{float}
% \usepackage{wrapfig}
% \usepackage{rotating}
% \usepackage[normalem]{ulem}
% \usepackage{amsmath}
% \usepackage{textcomp}
% \usepackage{marvosym}
% \usepackage{wasysym}
% \usepackage{amssymb}
% \usepackage{hyperref}
% \usepackage{color}
% \usepackage{soul}
% \tolerance=1000
% \usepackage[margin=1in]{geometry}

% \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

% %\author{Alex Ansari}
% %\date{}
% %\title{TALOS}
% %\hypersetup{
% %  pdfkeywords={},
% %  pdfsubject={},
% %  pdfcreator={Emacs 24.3.1 (Org mode 8.2.10)}}

% \begin{document}
\section{Hydraulic Lower Extremity Exoskeleton}

Exoskeleton suit was originally developed for military uses, specifically to allow operators to perform motions without incurring muscle fatigue.  Their main contributions in the literature re based on a {\bf virtual joint torque control}, which has been shown to be an effective human robot interaction control strategy in situations where the connections between the uses and hardware system are relatively unknown or can dramatically change as a function of time.

Additionally, the hardware system focuses on actuator selection.  The main hardware design principle is focused on developing a suit that can augment a users power and move fast when necessary.  They selected hydraulic actuation because the addition of electric actuator weight on the system leg joints makes the system difficult to move quickly.  On the other hand, hydraulic actuators have high power density relative to weight and size, and generally provide very high force outputs and low impedance.  The tradeoff with hydraulics come in the way of having low accuracy in terms of force control as well as hard to model nonlinear characteristics.

A picture of the Hydraulic Lower Extremity Exoskeleton is presented in Figure \ref{fig:exoSuit}.
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/exoSuit}
  \caption{}
  \vspace{-0.2in}
 \label{fig:exoSuit}   
 \end{figure}

\subsection{Actuator specifications}

The actuator specifications for the Hydraulic Lower Extremity Exoskeleton were designed around human walking data at 4km/h with a 45 kg load.  The resulting calculated hydraulic capacity for the while system was 2050psi and 8lpm.  The individual actuators were designed to be double acting at the hip and knee flexion/extension joints. The {\bf hip maximum thrust was 4 kN} and {\bf knee maximum thrust was 4 kN}.  Each joint joint included three-way servo valves (M 200, Star-Hydraulic) to control rate and direction of bi-directionality and bypass valves to connect the pump path to tank path for fast, energy efficient motion during swing phases.

The hydraulic system for each joint is shown in Figure \ref{fig:hydraulicSys}.
A picture of the Hydraulic Lower Extremity Exoskeleton is presented in Figure \ref{fig:exoSuit}.
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/hydraulicSys}
  \caption{}
  \vspace{-0.2in}
 \label{fig:hydraulicSys}   
 \end{figure}

%\begin{figure}[thpb]
%\centering
%\includegraphics[width=3.in]{figs/seaAssm}
%  \caption{}
%  \vspace{-0.2in}
% \label{fig:IHMCSEA}   
% \end{figure}

 \subsection{Exoskeleton Specifications}

 The Hydraulic Lower Extremity Exoskeleton includes four actively controlled joints in the sagittal plane; the knee and hip flexion/extension degrees of freedom are powered degrees of freedom for each leg.  The hardware also includes eight passive joints: one ab/adduction DOF for each hip and three degrees of freedom for each ankle.

 The hydraulic system of the suit is composed of the four actuators and a hydraulic power unit, consisting of a pump, motor, temperature sensor, and various valves.

 Additionally, various {\bf sensors} are incorporated to measure joint angles, actuator forces , and ground reaction forces.  Three absolute angular encodes are incorporated at the hip, knee, and ankle pitch joints.  In-line load cells are incorporated in each cylinder tube to measure actuator forces, and four single axis load cells are included in foot sensors to be used for gait phase detection.

 \subsection{Control Specifications}

 \begin{figure}[thpb]
\centering
\includegraphics[width=5.in]{exos/figs/hydLowerExrem/dualModeDia}
  \caption{}
  \vspace{-0.2in}
 \label{fig:dualModeDia}   
 \end{figure}
The control design principle for the Hydraulic Lower Extremity Exoskeleton is base around the concept that, for walking gaits, the {\bf stance and swing controllers should be considered separately}.  To this end, the literature for this hardware system introduces an active-passive control method called {\bf Dual-Mode Control}.  Active control is used during stance phases whereas the passive control is used to quickly and freely control the system's legs during swing phases.  Additionally, to handle the inherently discontinuous jumps in the command signals that occur as a function of this dual mode abstraction, a smoothing method is introduced that is used in conjunction with a pre-transition method that solves the swing delay due to internal cylinder pressure.  A diagram which highlights the different modes in this control strategy is shown in Figure \ref{fig:dualModeDia}.

Various components of the overall dual-mode control strategy, at the hardware level, are presented in Figure \ref{fig:suitDia}.  
  \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/suitDia}
  \caption{}
  \vspace{-0.2in}
 \label{fig:suitDia}   
 \end{figure}
This figure contains a number of components, specifically a {\bf gait phase estimation block} which uses the foot sensors to estimate the stance mode of the system, a {\bf human intention block}, which uses angular encoders and joint force sensors to estimate the motion of the operator, a {\bf torque controller}, and a {\bf gravity compensation} block.  Note that these various components of the control system are used to support the general assumption of the dual mode controller, i.e., 
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/torAssist}
  \caption{}
  \vspace{-0.2in}
% \label{fig:suitDia}   
 \end{figure}

 \noindent
where, $GP$ is the gait phase, $GRF$ is the ground reaction force, and $\delta$ is a threshold value.  This equation explicitly states that an active torque will be applied for each leg during stance and that the legs are passive in their respective swing phases.

Like other control methods presented in this report, the dual mode control method in this section uses the fundamental assumption that it is difficult to measure the interaction forces between the user and hardware system directly as a primary design principle.  To this end, the human-robot interaction force is assumed to be completely represented in terms of the joint torque applied by the operator in the hardware.

To calculate the active torque applied to the joints during the active mode control, the robot dynamics are considered independently from the operator, namely,
\begin{equation}
T_a = M(q)\ddot{q} + C(q,\dot{q})\dot{q}+G(g),
\label{dynamics}
\end{equation}   
where $T_a$ is the joint torque applied by the actuators, $M(q)$ is the inertia matrix, $C(q,\dot{q})$ is the centripetal and Coriolis matrix, and $G(q)$ is the gravitational torques. note that this equation neglects friction as well as other actuation dynamics.

Using the force sensors on the hydraulic actuators, the {\bf human-robot interaction torques} can be approximated as
\begin{equation}
\hat{T}_{hm} = M_n(q)\ddot{q} + C_n(q,\dot{q})\dot{q} + G_n(q) - \hat{T}_a, \notag
\end{equation}
where $M_n$, $C_n$, and $G_n$ are the nominal models of the true inertia, centripetal, and gravitational parameters and $\hat{T}_a$ represents the measured actuator torques.  The $M_n(q)\ddot{q} + C_n(q,\dot{q})\dot{q} + G_n(q) $ portion of $\hat{T}_{hr}$ represents the inverse dynamics component of the closed-loop control system.  The overall joint torque controller, which combines compensation to for operator intent as well as balancing the suit under a gravitational load is then written \[ \tau_\text{active} = K(s) \hat{T}_{hm} + G_n(q),\] where $K(s) = K_p +s K_d,$ i.e., there is closed-loop PD control law operating on the error between the predicted robot dynamics and the measured joint torques.  The block diagram for the active mode controller is shown in Figure \ref{fig:blockDia}. 
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/blockDia}
  \caption{}
  \vspace{-0.2in}
 \label{fig:blockDia}   
 \end{figure}
 Note that $G_a$ represents the dynamics of the the system and $G_a'$ the inverse dynamics model in Figure \ref{fig:blockDia}.

In the passive mode, which occurs during the swing phase, the leg of the human-robot system moves quickly and freely under using the passive inertial forces of the overall system.  This is accomplished using bypass valves that connect the pump path to the tank path to achieve mechanical back-drivability.  The foot sensors on each foot are used to trigger the bypass valves and to initiate the pass control mode.

A novel component in the literature of the Hydraulic Lower Extremity Exoskeleton controller is how the transitions between the active and passive control modes are handled.  In particular, an explicit smoothing method is used to address the discontinuity in the command torque at phase transitions.  One solution which is proposed is to use an exponential weighting function of the form
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/hydLowerExrem/weightingForm}
  \caption{}
%  \vspace{-0.2in}
% \label{fig:suitDia}   
 \end{figure}

 \noindent
 where $\tau_\text{inv}$ is the inverse dynamics torque associated with the robot model, $f_\text{weighting}$ is the weighting function, $a$ is a sensitivity factor, and $N$ describes the duration of the transition period.  The transition duration period as well as sensitivity factor appear to be hand tuned parameters.

In addition to phase transition smoothing a pre-transition control method which happens at the end of the stance phase prior to the passive swing phase is also used to effect smooth phase transitioning.  The pre-transition control works by measuring the ground reaction force and setting a minimum threshold value.  When the GRF drops below the threshold, the passive swing control mode is entered, i.e., the bypass valve at the actuators are opened.  The objective of this method is to reduce the phase transition delay due to residual pressure in the foot sensors.

 \subsection{Opinion}

The design of the Hydraulic Lower Extremity Exoskeleton has a lot going for it.  The best part of this hardware as well as control design is the fact that the creators of it are doing something relatively straightforward.  The actuators are composed of what appears to be off-the-shelf components, and the addition of bypass valves to take advantage of passive dynamics during walking to produce more energy conscious behaviors is a good idea.

One potential drawback of this hardware is the power system.  The documentation in \cite{} was sparse with respect to the exact specifications of the motor and energy storage device which powers the hydraulic system.

The control system for this hardware followed a relatively straight-forward design, effectively creating a ``get out of the way design" using the torque sensed at the joints only.  In the short-term, this type of policy seems to be a good strategy.  The other highlight of the outlined control policy was the smoothing function which was used during phase transitions in the walking gait cycle.  A similar type of smoothing will most likely be beneficial in terms of operator experience for a varity of different operational modes in which different phase transitions occur, e.g., running, jumping, etc.

 % \end{document}

 % Hongchul Kim and Changhoon Seo and Young June Shin and Jung Kim and Youn Sik Kang, "Locomotion Control Strategy of Hydraulic Lower Extremity Exoskeleton Robot," IEEE International Conference on Advanced Intelligent Mechatronics, 2015, pp. 577-582.

### 2015-09-24 exos/exosText_ihmc.tex
added Matts stuff but still need to get his bib files

% % Created 2015-09-15 Tue 11:46
% \documentclass[11pt]{article}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{fixltx2e}
% \usepackage{graphicx}
% \usepackage{longtable}
% \usepackage{float}
% \usepackage{wrapfig}
% \usepackage{rotating}
% \usepackage[normalem]{ulem}
% \usepackage{amsmath}
% \usepackage{textcomp}
% \usepackage{marvosym}
% \usepackage{wasysym}
% \usepackage{amssymb}
% \usepackage{hyperref}
% \usepackage{color}
% \usepackage{soul}
% \tolerance=1000
% \usepackage[margin=1in]{geometry}

% \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

% %\author{Alex Ansari}
% %\date{}
% %\title{TALOS}
% %\hypersetup{
% %  pdfkeywords={},
% %  pdfsubject={},
% %  pdfcreator={Emacs 24.3.1 (Org mode 8.2.10)}}

% \begin{document}
\section{IHMC Mobility Assist Exoskeleton (MAE)}
Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the literature for this device does include a ``performance enhancement" control mode and as such is within the scope of this document.  The main design principle highlighted in the IHMC MAE hardware is a recurring theme throughout the exoskeleton literature, which is that directly measuring user intent can be difficult for a number of reasons.  This issue is resolved by using a novel control and feedback system, which in the IHMC MAE is primarily based on the use of a novel actuator design.

The IHMC MAE uses novel series elastic actuators  (SEAs) in their hardware design mostly because they offer the ability to perform high-fidelity impedance control.  In series elastic actuators, a compliant element is placed in series after the output a motor's drive mechanism. The measured compression of the compliant element is used to calculate the force, or torque, acting on the output of the actuator.  { \bf Series elastic actuators} are generally viewed as a good way to provide {\bf accurate force feedback} information and to provide a {\bf low mechanical impedance}.  The major disadvantage of SEAs is that they have a relatively low bandwidth at high forces, as the compliant element needs time to compress before a force can be sensed and thus inherently include a delay between the time a force is applied and when the control system can react.

\subsection{Actuator specifications}

 Actuator designed around data collected from clinical gait analysis.  Specifically, the actuators at the hip and knee were designed to be able to provide 40Nm of peak torque during the stance phase of the walking gat cycle.  The rotary SEA was composed of a Moog BN34-25EU-02 brushless DC motored with the 1:100 harmonic drive, an SHD-20.

The spring mechanism which sits between the gearbox and joint output in the IHMC MAE design is shown in Figure \ref{fig:IHMCSEA}.  The mechanism shown uses linear die springs and an angular encoder to measure the torque in the out of the joint.  The choice of linear die spring was based on the spring's predictable force to displacement function as well as favorable energy storage to weight characteristics.  
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/ihmc/seaAssm}
  \caption{}
  \vspace{-0.2in}
 \label{fig:IHMCSEA}   
 \end{figure}

Once assembled, the overall actuator had a  {\bf torque limit of 80 Nm}, with a {\bf velocity limit of 6.8 rad/s}.  As noted, the {\bf bandwidth of the actuator} is its main drawback, which limits the torque control to approximately {\bf 10 Hz at amplitudes greater than 15 Nm} and can track up to {\bf 30Hz for lower magnitude forces}.

 \subsection{Exoskeleton Specifications}

 The IHMC MAE is a lower extremity exoskeleton with a total of ten degrees of freedom (DOF), with five each corresponding to the two legs of the system.  Of the ten DOFs only six total are actively actuated, the other four are passive.  Each of the powered joints is powered using the SEA joint discussed in the previous subsection.  A labeled diagram of the IMHC system is shown in Figure \ref{fig:IHMCSYS}.

 The hip ad/abduction as well as flexion/extension DOFs are actively actuated.  The hip yaw DOF consists of a curved spring-loaded roller bearing whose center of rotation is approximately at the user's hip joint.

 The knee flexion/extension joint is connected to the hip flexion/extension joint using telescoping tubes and is also actuated.

 The final DOF in the IHMC design is the dorsal flexion ankle joint.  This joint is spring loaded such that it provides unidirectional torque to provide toe clearance during walking.

 In its current configuration, the IHMC exoskeleton in controlled by an off-board computer and is powered using a tether.

 The suit {\bf includes position and force sensors in the actuators in addition to two foot switches per foot}.  The foot switches are used to detect whether the system in in single or double support phase, as well as providing information on how the overall system load is approximately distributed on the stance legs.

 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/ihmc/ihmcSys}
  \caption{}
  \vspace{-0.2in}
 \label{fig:IHMCSYS}   
 \end{figure}

 \subsection{Control Specifications}

 The noted control approaches of the IHMC MAE are both relatively straight forward implementations of low-level torque based control strategies.  The first control scheme takes full advantage of the SEA actuators ability to directly measure output torque and subsequently to perform closed-loop torque control, which is pictured in \ref{fig:IHMCTOR}.     
  \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/ihmc/torqueCon}
  \caption{}
  \vspace{-0.2in}
 \label{fig:IHMCTOR}   
 \end{figure}
 The pictured control scheme takes in a desired torque, compares the desired value to the output torque measured at each individual actuator, then uses this value as well as its derivative to compute an actuator command signal, which in this case is a current.  The largest question in this controller, as presented in the literature, is how to define the desired torque term $\tau_\text{des}$.

 The second low-level control approach presented in the IHMC MAE literature is a combination of a position-based controller with an inner-loop torque controller.  This control scheme is presented in Figure \ref{fig:IHMCPOS}.
  \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/ihmc/positionCon}
  \caption{}
  \vspace{-0.2in}
 \label{fig:IHMCPOS}   
 \end{figure}
 The controller in Figure \ref{fig:IHMCPOS} shows an outer-loop position-based PD controller which takes in desired joint trajectories and compares them to measured joint positions.  The values are used to generate desired torque values which are fed into an inner closed-loop torque control loop.  This control scheme is a good low-level choice if high-impedance very accurate motions of the suit are desired.  Similarly to the torque-only control loop above, the main question in this control scheme is how the desired joint trajectories are defined.

 One possibility for defining either the desired torques in the first control scheme or the desired joint trajectories in the second is to use clinical gait analysis data.  In this approach the controllers highlighted in this subsection would act to play back nominal joint trajectory data.  The foot sensors in the system would be used to detect phase transitions and to dynamically adjust the prerecorded data to the user in realtime.

 \subsection{Opinion}

 The actuator design and use of series-elastic actuation is in general a good idea for many applications, but may be a problem for the highly dynamic maneuvers required by operators in the TALOS project.  In a large sense, bandwidth is everything in these suits due to the fact that behavior prediction is very complicated.  Simply put, 10Hz control bandwidth at the actuator may be too slow to safely react in dynamically changing environments.

 The other issue with the IHMC work is that their control approach only addresses the lowest-level of control, and does not in general specify exactly how to generalize the process of of generating reference trajectories, whether they be torque or joint trajectories.  That said, other researchers do explicitly provided general methods for generating desired trajectories in what amounts to a mid-level of a hierarchical control scheme.  Thus combining other approaches with the low-level highly accurate torque control provided by the novel SEAs presented in this section could offer a potentially good solution at the actuator level.

% \end{document}

% Hian Kai Kwa and Jerryll H. Noorden and Mathew Missel and Travis Craig and Jerry E. Pratt and Peter D. Neuhaus, "Development of the IHMC Mobility Assist Exoskeleton,"  International Conference on Robotics and Automation, 2009.

### 2015-09-24 exos/exosText_roboKnee.tex
added Matts stuff but still need to get his bib files

% % Created 2015-09-15 Tue 11:46
% \documentclass[11pt]{article}
% \usepackage[utf8]{inputenc}
% \usepackage[T1]{fontenc}
% \usepackage{fixltx2e}
% \usepackage{graphicx}
% \usepackage{longtable}
% \usepackage{float}
% \usepackage{wrapfig}
% \usepackage{rotating}
% \usepackage[normalem]{ulem}
% \usepackage{amsmath}
% \usepackage{textcomp}
% \usepackage{marvosym}
% \usepackage{wasysym}
% \usepackage{amssymb}
% \usepackage{hyperref}
% \usepackage{color}
% \usepackage{soul}
% \tolerance=1000
% \usepackage[margin=1in]{geometry}

% \newcommand{\hilight}[1]{\colorbox{yellow}{#1}}

% %\author{Alex Ansari}
% %\date{}
% %\title{TALOS}
% %\hypersetup{
% %  pdfkeywords={},
% %  pdfsubject={},
% %  pdfcreator={Emacs 24.3.1 (Org mode 8.2.10)}}

% \begin{document}
\section{RoboKnee}

The RoboKnee device was built as a prototype system to enhance human strength, endurance and speed.  The main design objectoves in the RoboKnee system were to enhance human performance using a low-impedance mechanical system that incorporates a natural user interface, long life, and is comfortable to wear.  The main control objective of the suit is to provide a ``get out of the way" control scheme, wear the system inherently interacts with the user through a low-impedance interface.

The design of a linear series elastic actuator and a novel control method for controlling a single powered joint at the knee of the operator are presented in this section.

\subsection{Actuator specifications}

The RoboKnee actuator was designed to specifically provide low impedance and high force feedback fidelity.  Series elastic actuators are a more robust option for measuring output forces and torques than using notoriously delicate and expensive load cells.

The actuator used in the RoboKnee device is shown in Figure \ref{fig:roboAct}.  The actuator consists of a drive train subassembly and an output carriage subassembly. The two subassemblies are coupled through die compression springs.  A servo motor directly drives a ball screw assembly to drive a ball nut flange which pushes on a spring retaining plate.  Secondary spring retaining plate is coupled to an output assembly.  Measuring the deflection in the die springs makes it possible to measure the force applied to the output of the linear SEA.  
\begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/roboKnee/roboSEA}
  \caption{}
  \vspace{-0.2in}
 \label{fig:roboAct}   
 \end{figure}

The linear actuator pictured in Figure \ref{fig:roboAct} { \bf weighs 1.13 kg, has a stroke length of 30.5 cm and diameter of 5.8 cm}.  The {\bf maximum speed of the actuator is 28 cm/s}.  The actuators {\bf maximum continuous force is 565N, with a peak maximum force of 1330 N}, with {\bf maximum continuous power of 164 W, and maximum peak power of 634 W}.

The force control bandwidth of the RoboKnee actuator is dependent on the magnitude of the force, as significant time delays can occur for larger forces due to the fact that there is an inherent time delay due to the time it takes for the spring to compress.  Thus, the {\bf small force control bandwidth of the system is 35 Hz} and the {\bf high force bandwidth is around 7.5 Hz}.

 \subsection{Exoskeleton Specifications}

 The RoboKnee mechanism consists of a single actuated degree of freedom at the flexion/extension joint of the operator.  Note that the mechanism itslef does not provide a pathway for transferring weight from a load directly to the ground.  The mechanism transfers loads directly to the users musculoskeletal structure, and focuses primarily on augmenting the torque supplied to the users knee joint.

 The RoboKnee mechanism itself is composed of an off-the-shelf knee brace modified with structural pieces to extend the brace and provide actuator attachment points.  The RoboKnee mechanism is shown in \ref{fig:roboKnee}.  
 \begin{figure}[thpb]
\centering
\includegraphics[width=3.in]{exos/figs/roboKnee/roboBrace}
  \caption{}
  \vspace{-0.2in}
 \label{fig:roboKnee}   
 \end{figure}
 There exact length between attachment points of the actuator is not listed in the literature, but it appears that the attachment points are at minimum the stroke displacement of the actuator  away from each other, so a {\bf minimum moment arm of 30.5 cm} is a reasonable approximation.

 The RoboKnee mechanism uses a linear potentiometer spanning the spring retaining plates to measure displacement in the springs and subsequently to measure force at the actuator output.  The forces between the users foot and the ground are also measured using single axis load cells.  Potentiometers which measure the knee joint angle are also employed in this design.

 The single actuator as well as control and sensor electronics are run using {\bf 4 kg of nickel-metal-hydride batteries, which give the system only about 30-60 minutes of heavy use}.

 \subsection{Control Specifications}

 The control scheme for the RoboKnee device is a hierarchical approach, which uses a straightforward mid-level force generation scheme coupled with a low-level closed-loop force-based control loop.  The low-level, or rather joint level control is a PD control loop.

 The mid-level controller of the RoboKnee uses several simplifying assumptions to perform ``force amplification" to offload a percentage of the torque required by the user to actuate the system's knee joint.  This is achieved through a positive force feedback amplification which is based on measuring the ground reaction force and using this value to calculate the torque that would be required to produce this force in a static situation, namely, \[ \tau = R \times F,\] where $R$ is the vector from the ground reaction force to the users knee and $F$ is the ground reaction force.  Note that in the RoboKnee implementation this value can only be estimated due to the fact that the ankle joint angle and the ground reaction force is assumed to be purely vertical (which is a result of the single axis force sensors on the users feet).

 Once the estimated joint torque $\tau$ is calculated, an amplification factor is used in the RoboKnee control scheme to decide how much of the required knee torque will be provided by the control system.  For example, if the amplification factor were chosen to be one the actuator would completely compensate for the sensed ground reaction forces, and if it were set to zero to exoskeleton would provide no force.

 \subsection{Opinion}

 The RoboKnee is a relatively simple system that highlights several intelligent hardware and control design concepts.  The use of a linear ball screw with a series elastic element that enables high-fidelity force control is a very good idea. The idea of using the measured ground reaction forces to produce the desired knee joint torque is also interesting in the sense that it is very simple and relies on a minimal amount of information.

 The overall design of the RoboKnee mechanism is most likely not the greatest, as it significantly increases the effective volume of the operator-system leg, but as a prototype this concept offers interesting benefits.

 % \end{document}

 % Jerry E. Pratt and Benjamin T. Krupp and Christopher J. Morse and Steven Collins, "The RoboKnee: An Exosskeleton for Enhancing Strength and Endurance During Walking," International Conference on Robotics and Automation, 2004, pp. 2430-2435.

### 2015-09-25 control.tex
cga updates

@@
 %\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
-A later white paper will discuss
+A later white paper could discuss
 possible combinations of response to user and exoskeleton
@@
 
-A later white paper will discuss how to handle malfunctions and damage to the
-system. Basically this boils down to fault detection, and switchover to
+A later white paper could discuss how to handle exoskeleton
+malfunctions and damage to the
+system. Basically this boils down to fault detection, and switch over to
 an appropriate safe mode of operation.
 
-A later white paper will evaluate using simulation
+A later white paper could evaluate using simulation
 the various control approaches' robustness to modeling error and unmodeled
@@
 This level provides an idealization of the exoskeleton in two ways: 
-What degrees of freedom (DOF) are controlled, and how is a degree of
-freedom controlled?
+{\it What degrees of freedom (DOF) are controlled?,} and 
+{\it How is a degree of freedom controlled?}
 
@@
 $\vq$ is the vector of joint angles, $\dot{\vq}$ are joint velocities,
-$\vtau$ are actutor forces or joint torques, and $\dot{\vtau}$ is the rate of change
+$\vtau$ are actuator forces or joint torques, and $\dot{\vtau}$ is the rate of change
 of the forces or torques (the time derivative of $\vtau$). 
@@
 the control input of a PWM amplifier.
-The $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
+The torque error term $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
 by a servo on piston differential pressure in a fluidic system or
-a current servo in an electrical system.
+an electrical current servo in an electric system.
 \item
@@
 There are two sets of contact forces: operator-exoskeleton (which we denote
-with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
-with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
+with a subscript ${\mbox {ox}}$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
+with a subscript ${\mbox {xw}}$ as in $\vf_{\xw}$). If the operator is tightly strapped or
 rigidly held by other types of physical constraints, we can lump the operator
@@
 and just think of the whole system as one robot (we indicate this with
-the subscript $\ox$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
-terms in the dynamics equations).
+the subscript ${\mbox {ox}}$ on inertial, Coriolis and centripetal, and gravitational (ICCG)
+terms in the dynamics equations as in $\mM_{\ox}$).
 We can also separate out the operator dynamics (indicated by
-the subscript $\myo$ on ICCG terms) from the exoskeleton dynamics (indicated by
-the subscript $\myx$ on ICCG terms) if this is more convenient
+the subscript ${\mbox {o}}$ on ICCG terms as in $\mM_{\myo}$) from the exoskeleton dynamics (indicated by
+the subscript ${\mbox {x}}$ on ICCG terms as in $\mM_{\myx}$) if this is more convenient
 (we separate them out and ignore operator dynamics in what immediately follows).
@@
 as operator joint torques.
-We assume the operator joint friction is negligable.
+We assume the operator joint friction is negligible.
 
@@
 = \vtau_{\myo} + \vtau_{\myx}
-\label{eq:oxdynamics}
 \end{equation}
@@
 Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
-to reduce forces. 
+to reduce operator forces. 
 An operator would rely on mechanical backdrivability to drive
-exokeleton to do tasks.
+exoskeleton to do tasks.
 These passive devices have no controls, and can be included in the dynamics equations.
@@
 
+\section{Fixed feedforward control with feedback control}
+
+We can add our favorite feedback control scheme to open loop control.
+Now the control is closed loop, but the feedforward control to achieve
+a trajectory is designed and executed independently of any feedback control.
+In most practical robotics manually designed
+fixed gain independent joint linear (PID) feedback control is used.
+As previously mentioned, more complex feedback control approaches such as
+approaches that coordinate across joints
+(full state feedback,
+pole placement, linear quadratic regulator (LQR) design, linear quadratic
+Gaussian (LQG) design, and linear quadratic Gaussian loop transfer recovery
+(LQG-LTR) design)
+or nonlinear alternatives that handle discontinuous control problems such as
+sliding mode control can also be used.
+Feedback gains can be ``gain scheduled'', which means that feedback gains are
+pre-tabulated and looked up and deployed
+based on the current position, velocity, or some other
+quantity (temperature, for example).
+
 \section{Active gravity compensation}
 
+Now we will examine feedforward control that varies and is computed based on
+the current state of the exoskeleton.
 Active exoskeleton gravity compensation is similar 
@@
 The operator relies on mechanical backdrivability to drive
-the exokeleton to do tasks.
+the exoskeleton to do tasks.
 The weight of the operator can also be supported by the exoskeleton in
@@
 
-Friction can also be compensated.
+Friction can also be compensated using feedforward control.
 \begin{equation}
@@
 and velocity.
-Note that these may be load dependent, and may require additional state to
+Note that friction may be load dependent, and may require additional state to
 correctly handle stiction and hysteresis effects.
@@
 \url{https://en.wikipedia.org/wiki/Admittance}
-Compliance, inverse damping, inverse mass are components of an admittance,
+Compliance, inverse damping, and inverse mass are components of an admittance,
 as they map force and torques to kinematic variables (position, velocity, and
@@
 \end{equation}
-$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
-force $\vf_{\ox}$
+$\mK_{\vf_{\ox}}$ is a gain matrix. $\mJ_{\ox}$ is the Jacobian matrix for the operator
+force $\vf_{\ox}$.
 Note that this does not require inverting the Jacobian matrix.
@@
 \begin{equation}
-{\vf_{\ox}}_d = F_{\ox-\xw} ( \vf_{\xw} )
+{\vf_{\ox}}_d = \mF_{\ox-\xw} ( \vf_{\xw} )
 \end{equation}
@@
 dynamics (including the operator link inertias, link centers of mass, and link
-moments of inertia in the dynamic model) based on the desired postion, 
+moments of inertia in the dynamic model) based on the desired position, 
 desired velocity, and desired acceleration. Trajectory tracking errors are compensated
@@
 \end{equation}
-The mapping is the same for stiffnes:
+The mapping is the same for stiffness:
 \begin{equation}
@@
 \end{equation}
-The rows in $\mS$ corresponding to the root linear and angular velocities
+The rows in $\mJ_{\myx}$ corresponding to the root linear and angular velocities
 are all zero.
-To maintain compatability in our joint representation, we will similarly
+To maintain compatibility in our joint representation, we will similarly
 map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
@@
 
-The second kind of underactuation appears in exoskeletons where we rely on the
+The second kind of underactuation appears in exoskeletons when we rely on the
 operator to actuate a particular joint. For example, many lower body
@@
 We can have redundant actuation (more than one actuator across any particular
-joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accomodate extra actuators.
-We can have multijoint actuation, which can be handled in $\vu_{\myx}$ 
+joint). Here we augment $\vu_{\myx}$ and $\mJ_{\myx}$ to accommodate extra actuators.
+We can have multi-joint actuation, which can be handled in $\vu_{\myx}$ 
 and $\mJ_{\myx}$ as well.
@@
 
-{\bf Open loop control} continues to work with underactuation
+{\bf Open loop control} 
+and
+{\bf fixed feedforward control with feedback control}
+continue to work with underactuation
 as long as a drive function can be found
@@
 {\bf online inverse dynamics} as described in Section~\ref{sec:invdyn}
-no longer works if $\mJ_{\myx}$ is not invertable (not square or not full rank).
+no longer works if $\mJ_{\myx}$ is not invertible (not square or not full rank).
 In the underactuated and overactuated cases $\mJ_{\myx}$ is not square. 
@@
 joints the exoskeleton actuates, and invert those dynamics to cancel exoskeleton
-effects on those joint. This is what has been done with the BLEEX exoskeleton,
+effects on those joints. This is what has been done with the BLEEX exoskeleton,
 for example.~\cite{}.
@@
 commands $\vu$, and contact forces $\vf$.
-For robots the discrepancy between actual and desired accelerations, actuator command
-magnitude, and contact force magnitude are all minimized:
-\begin{equation}
-C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
-+ \vf^{\tr} \mW_{\vf} \vf
-\end{equation}
-$\mW_{\vq}$, $\mW_{\vtau}$, and $\mW_{\vf}$ are weight matrices to trade off the
+We could optimize only taking into account the exoskeleton and its dynamics,
+minimizing the discrepancy between actual and desired accelerations, actuator command
+magnitude, and contact force magnitude:
+\begin{equation}
+C_{\myx}(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d)
++ \vu_{\myx}^{\tr} \mW_{\vu_{\myx}} \vu_{\myx}
++ \vf_{\ox}^{\tr} \mW_{\vf_{\ox}} \vf_{\ox}
++ \vf_{\xw}^{\tr} \mW_{\vf_{\xw}} \vf_{\xw}
+\end{equation}
+$\mW_{\vq}$, $\mW_{\vu_{\myx}}$, $\mW_{\vf_{\ox}}$,
+and $\mW_{\vf_{\xw}}$ are weight matrices to trade off the
 different errors/magnitudes and error/magnitude directions.
-The dynamics (equation~\ref{eq:dynamics2}) are enforced as an equality constraint.
+The dynamics (equation~\ref{eq:xdynamics2}) are enforced as an equality constraint.
 Actuator limits, friction limits, center of pressure location constraints,
@@
 
-This approach can be used to achieve desired joint or synergy (such as center
+Alternatively, we could optimize the operator-exoskeleton combination,
+minimizing:
+\begin{equation}
+C_{\ox}(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) 
++ \vu^{\tr}_{\myx} \mW_{\vu_{\myx}} \vu_{\myx}
++ \vf_{\ow}^{\tr} \mW_{\vf_{\ow}} \vf_{\ow}
++ \vf_{\ox}^{\tr} \mW_{\vf_{\ox}} \vf_{\ox}
++ \vf_{\xw}^{\tr} \mW_{\vf_{\xw}} \vf_{\xw}
+\end{equation}
+In this case we would enforce the combined 
+dynamics (equation~\ref{eq:oxdynamics2}) as an equality constraint.
+Note that we assume the operator muscle command
+$\vu_{\myo}$ is fixed.
+We include the operator-exoskeleton force $\vf_{\ox}$ so it can be minimized or set to
+a desired value.
+
+These approaches can be used to achieve desired joint or synergy (such as center
 of mass) accelerations, center of mass dynamic behavior, center of pressure location,
@@
 
-\subsection{Controlling operator forces and perceived exoskeleton impedance}
+\subsection{What does online optimization want from operator intent prediction?}
+\label{sec:predict}
+
+Online optimization for the current instant in time tries to match
+actual accelerations to desired accelerations, so some estimate of the desired
+acceleration $\ddot{\vq}_d$ is useful. In addition, to accurately estimate the dynamics,
+we need an estimate of the operator-exoskeleton force $\vf_{\ox}$, and the 
+operator-world force $\vf_{\ow}$. These estimates could be deterministic values
+or probability distributions. Probability distributions are more expensive to
+optimize.
+
+There are more complex versions of online optimization that consider a future
+time horizon, and use some form of receding horizon control (RHC or MPC) to
+optimize a time window on each servo tick. In this case knowledge of
+future desired accelerations is useful, as well as future operator forces.
+This knowledge could be in the form of deterministic trajectories, 
+probability distributions of desired accelerations and operator 
+forces at each future time,
+or probability distributions for trajectories of desired accelerations and
+operator forces,
+which could take into account the dependence of future desired accelerations and 
+operator forces
+on previous accelerations and forces.
+
+\subsection{Controlling operator forces}
 
@@
 part of the optimization criterion could be to
-minimize contact forces on the operator:
-\begin{equation}
-C_e(\vx) = C_r(\vx) + \vf^{\tr}_{\ox} \mW_{\vf_{\ox}} \vf_{\ox}
+minimize contact forces on the operator by including a term penalizing those forces:
+\begin{equation}
+\vf^{\tr}_{\ox} \mW_{\vf_{\ox}} \vf_{\ox}
 \end{equation}
@@
 \begin{equation}
-C_e(\vx) = C_r(\vx) + (\vf_{\ox} - {\vf_{\ox}}_d)^{\tr} \mW_{\vf_{\ox}} (\vf_{\ox} - {\vf_{\ox}}_d)
+(\vf_{\ox} - {\vf_{\ox}}_d)^{\tr} \mW_{\vf_{\ox}} (\vf_{\ox} - {\vf_{\ox}}_d)
 \end{equation}
@@
 
-Note that the same approach could be used to minimize or control internal forces
-such as skeletal forces (forces between bones) at the operator joints to protect
-the ankle, hip, and knee joints, for example.
-MORE GOES HERE.
-
-We can implement operator {\bf impedance control} in several ways:
+\subsection{Reducing operator internal forces}
+
+Underactuated exoskeletons may rely on operator bones and tissue to transmit forces.
+In this case, there is a risk of
+tearing tendons, ligaments, and muscle, bruising soft tissue, damaging joints, 
+and fracturing or breaking bones. Online optimization can help minimize forces
+within the operator's body. Note that it is not net joint forces (the minimal
+joint forces that are computed by inverse dynamics) that matter.
+It is local internal forces, such as forces within a bone or soft tissue.
+Since internal forces can cancel each other,
+local internal forces are usually larger than internal forces computed
+from net joint torques.
+We also note that humans often coactivate their muscles to increase joint stiffness,
+and to maintain structural integrity of their bodies under load. A prediction
+of minimal muscle activation to achieve net joint torques will often underestimate
+the amount of muscle activation, muscle forces, and the resulting forces in and on 
+bones and soft tissue.
+
+To minimize internal forces, we will create a set of internal ``virtual'' sensors
+that ``measure'' internal forces at key points, resulting in vector of internal
+forces $\vb$. In reality, we will use a neuromuscular model of the individual
+operator, their previous muscle activations, and predictions of their future
+muscle activations to calculate $\vb$.
+We can include a penalty on the deviation of $\vb$ from a desired value $\vb_d$
+in online optimization:
+\begin{equation}
+(\vb - \vb_d)^{\tr} \mW_{\vb} (\vb - \vb_d)
+\end{equation}
+
+\subsection{Controlling perceived exoskeleton impedance}
+
+We can control the exoskeleton impedance the operator feels in several ways:
 We can add 
@@
 \end{equation}
-or is added to the optimization criterion:
-\begin{equation}
-C_e(\vx) = C_r(\vx) 
-+ (\vtau_{\ox} - {\vtau_{\ox}}_d)^{\tr} \mW_{\vtau_{\ox}} (\vtau_{\ox} - {\vtau_{\ox}}_d)
+or a penalty term is added to the optimization criterion:
+\begin{equation}
+(\vtau_{\ox} - {\vtau_{\ox}}_d)^{\tr} \mW_{\vtau_{\ox}} (\vtau_{\ox} - {\vtau_{\ox}}_d)
 \end{equation}
@@
 \begin{equation}
-\vf_{\ox} = F_{ow} ( \vf_w )
+\vf_{\ox} = \mF_{\ox-\xw} ( \vf_{\xw} )
 \end{equation}
@@
 \begin{equation}
-(\vf_{\ox} - F_{ow} ( \vf_w ))^\tr \mW_{fmap}  (\vf_{\ox} - F_{ow} ( \vf_w ))
+(\vf_{\ox} - \mF_{\ox-\xw} ( \vf_{\xw} ))^\tr \mW_{fmap} (\vf_{\ox} - \mF_{\ox-\xw} ( \vf_{\xw} ))
 \end{equation}
@@
 \end{equation}
-
-\section{Specific Proposed and Implmented Exoskeleton Control Schemes}
-
-\subsection{Sensitivity Amplification Control}
-
-Sensitivity Amplification Control (SAC) used to control BLEEX
-is a dynamic cancellation technique (a.k.a.
-inverse dynamics, computed torque, or feedback linearization) that also tries
-to modify the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
-using acceleration feedback. It is not clearly stated but it appears that the
-acceleration is the result of double differentiating position, as there do not
-appear to be velocity sensors on BLEEX. Low pass filtering is applied to reduce
-the high frequency noise amplified by the double differentiation process.
-
-\subsection{Integral admittance control}
-
-{\it [Exoskeletons] have the implicit property
-of causing a virtual modification of the dynamic response of
-the human limb. We use this property of the exoskeletons
-action to formulate a unified control design framework called
-Integral Admittance [torque to angle] Shaping, which designs exoskeleton con-
-trollers capable of producing the desired dynamic response
-for the assisted limb. In this framework, a virtual increase
-in the admittance of the limb is produced by coupling it
-to an exoskeleton that exhibits active behavior. Specifically,
-our framework shapes the magnitude profile of the integral
-admittance (i.e. torque-to-angle relationship) of the coupled
-human-exoskeleton system, such that the desired assistance is
-achieved. This framework also ensures that the coupled stability
-and passivity are guaranteed.}~\cite{Nagarajan_etal_2015}
-
-{\it ... the impedance of the coupled human-exoskeleton
-system needs to be reduced below that of the unassisted
-human limb. This implies that the exoskeleton needs to
-cancel its own impedance first and then compensate for
-at least a part of the human limb’s impedance. Therefore,
-the desired exoskeleton behavior must be that of a negative
-impedance ... Consequently,
-the feedback gains will all be
-positive. In other words, the exoskeleton controller uses
-positive feedback, and hence the exoskeleton exhibits active
-behavior, which is capable of performing net positive work
-on the limb. ...
-However, positive feedback naturally raises the question of
-stability, and so we now explain how coupled stability can be
-achieved. Although the exoskeleton exhibits active behavior,
-which can be potentially destabilizing, the controller can be
-designed such that the coupled human-exoskeleton system
-is stable and passive. ... 
-}~\cite{Nagarajan_etal_2015}
-
-{\it
-... Inertia compensation is more complex ...
-It can be shown that using only positive acceleration
-feedback ..., the gain margin of the coupled system
-reduces to the moment of inertia of the exoskeleton, which
-implies that the exoskeleton controller ... can at the most
-compensate for the exoskeleton’s own moment of inertia
-before going unstable. This implies that the moment of
-inertia of the coupled human-exoskeleton system cannot be
-reduced below that of the unassisted human limb, without
-compromising coupled stability. However, using low-pass
-filtered acceleration feedback, it can be shown that inertia 
-reduction can be achieved. ...
-This work uses filtered acceleration
-feedback with a second-order low-pass Butterworth filter. ...
-}~\cite{Nagarajan_etal_2015}
-
-This is a form of frequency domain virtual model control.
-
-\begin{verbatim}
-Kalman filter:
-P. Canet, “Kalman filter estimation of angular velocity and accelera-
-tion: On-line implementation,” McGill University, Montr ́eal, Canada,
-Tech. Rep. TR-CIM-94-15, Nov. 1994.
-\end{verbatim}
-
-\subsection{Dual Control Approach}
-
-This approach implements a passive swing leg in addition to Sensitivity 
-Amplification Control for the stance leg.
-
-{\it The robot utilized the dual-mode control scheme, which is
-comprised of the active control for the stance phase and the
-passive control (using bypass valves) for the swing phase, to achieve high walking
-speed in the swing phase while supporting heavy loads in
-the stance phase. To reduce the sudden change of the torque
-command at the transition from the swing phase to the stance
-phase, a smoothing method is adopted. We also implemented
-a pre-transition method to take a foot off quickly for fast
-walking by predicting the change from the swing to the stance
-in advance.}~\cite{IEEE07222598}
-
-Bypass valves implement what Sarcos calls ``dangle''.
-
-Stance: virtual joint torque control method Kazerooni et al.(2005)
-When contact location between the wearer and the
-exoskeleton is not fixed and difficult to estimate, this method
-has been shown to be an effective method to generate the
-locomotion for an exoskeleton robot.[3],[4],[21]
-\begin{verbatim}
-H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
-the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
-Conf. Robot. Autom., pp. 4364-4371, 2005.
-Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
-for Human Performance Amplification,” University of California,2003.
-582
-Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
-Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
-Control,” Journal of computers, vol.9, No.1,202-208. 2014.
-\end{verbatim}
-
-To reduce
-sudden changes(command jump) at the phase transitions, a
-smoothing method [5],[19] is introduced and a pre-transition
-method is used to solve the swing delay due to the internal
-pressure(approx 5 bar).
-\begin{verbatim}
-H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
-Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
-Robotics Research, pp. 561-573, 2006.
-oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
-Based Control for a Rotary Series Elastic Actuator Assisting the knee
-\end{verbatim}
-
-Transition Control
-
-1) Smoothing Method: During transitions of the gait
-phase, discontinuity of the control command torque are
-occurred by the different condition for the fixed coordinate
-which is on the backpack at the swing phase or the foot at
-the stance phase. In this paper, to reduce this sudden change
-due to gait phase changes, a smoothing method is proposed
-as shown in Fig. 7. An exponential function is considered
-as the weighting function for smoothing as shown in (6).
-In this case, the weighting is small at the initial stage but
-it exponentially converges to one for supporting the load
-quickly.
-
-2) Pre-transition Method: The pre-transition method is
-that the passive mode is executed in the pre-swing phase
-prior to toe off. This dramatically reduces the moving
-
-\subsection{Ground Reaction Force Control}
-
-{\it
-The ground reaction
-forces (GRF) magnitude and direction is used to command the
-actuators. In some research the GRF sensors are used
-together with other sensors in the control architectures [17],
-while in other exoskeleton the control system is based on the
-GRF merely [RoboKnee, Honda].
-}
-\begin{verbatim}
-17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
-"Intention-based walking support for paraplegia patients with Robot
-Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
-\end{verbatim}
-
-\subsection{Virtual model control}
-
-The goal here is to make the exoskeleton imitate a desired (usually nonlinear)
-dynamic system,
-For running it could be a pogo stick or trampoline, for example.
-
-There are several possible versions of this type of control:
-\begin{enumerate}
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton actuator or internal forces.
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton acceleration.
-\item
-Map from operator-exoskeleton contact forces and system state
-to exoskeleton-world contact forces.
-\item
-Map from exoskeleton-world contact forces and system state to exoskeleton actuator or internal forces.
-\item
-Map from exoskeleton-world contact forces and system state to exoskeleton acceleration.
-\item
-Some combination of the above.
-\end{enumerate}
-
-\begin{verbatim}
-virtual model approach: hopper, compass gait (J. Pratt)
-estimate trajectory (jumping) vs. program behavior (trampoline, hopper,
-   compass gait)
-- J' control + behavior
-- Inverse dynamics + behavior
-\end{verbatim}
-
-\subsection{Task Specific Control}
-
-Task specific control can involve switching low level control modes, or abstracting
-the behavior of the exoskeleton with a hierarchy:
-\begin{enumerate}
-\item
-Task specific control: Generate desired motions and contact forces between
-the operator and the exoskeleton, and the exoskeleton and the world.
-\item
-Sometimes there are intermediate levels of control.
-\item
-Exoskeleton control: Control the exoskeleton to generate the desired motions and
-desired contact forces, with the desired impedance or admittance.
-\end{enumerate}
 
@@
 we use extended Kalman Filtering (EKF) to perform state estimation.
-Currently, using the full body dynamics (equation~\ref{eq:dynamics2})
+Currently, using the full body dynamics 
+(equations~\ref{eq:xdynamics2} or \ref{eq:oxdynamics2})
 and any additional constraint equations for ground contacts (feet on the ground):
@@
 For higher level (overall) state estimation,
-we developed a center of mass location and velocity estimator, and its
-application in full body inverse dynamics control, fall detection
-and fall prevention for humanoid robots and exoskeletons~\cite{}.
+we developed a center of mass location and velocity estimator~\cite{}.
 The center of mass filter is modeled as a multiple model EKF with contact switching
@@
 where no safety belay was
-allowed, our fall early warning system
+allowed, our fall early warning system based on this estimator
 successfully saved our robot from
@@
 
-\section{Robustness to parametric modeling error and unmodelled dynamics}
+\section{Robustness to parametric modeling error and unmodeled dynamics}
 
@@
 
-This question needs to be addressed using simulation, and is deferred to a future
-white paper [robust.pdf].
+This question needs to be addressed using simulation, and is deferred to a 
+possible future white paper [robust.pdf].
 
@@
 (except in some contrived examples) and errors do not grow exponentially in
-response to feedforward control errors.
+response to feedforward control errors (except for unstable and chaotic systems).
 So feedforward control is technically robust,  
-and we don't have to worry about it blowing up (except for chaotic systems).
+and we don't have to worry about it going wildly out of control.
 We do have to worry about performance.
@@
 We can ``verify'' designs using simulation, but making accurate simulations of
-the effects of contact and unmodelled dynamics is also a current research issue.
+the effects of contact and unmodeled dynamics is also a current research issue.
 There is no getting around the fact that our operators will be like the test
@@
 These approaches promise
-to improve performance, and try to track (slowly) changes in the operator, exoskeleton,
+to improve performance, and try to track (slow) changes in the operator, exoskeleton,
 or contacts to maintain performance. If something changes quickly, we need robust
 control design.
+
+\section{Specific Proposed and Implemented Exoskeleton Control Schemes}
+
+\subsection{Sensitivity Amplification Control}
+
+Sensitivity Amplification Control (SAC) used to control BLEEX
+is a dynamic cancellation technique (a.k.a.
+inverse dynamics, computed torque, or feedback linearization) that also tries
+to modify the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
+using acceleration feedback. It is not clearly stated but it appears that the
+acceleration is the result of double differentiating position, as there do not
+appear to be velocity sensors on BLEEX. Low pass filtering is applied to reduce
+the high frequency noise amplified by the double differentiation process.
+
+\subsection{Integral admittance control}
+
+{\it [Exoskeletons] have the implicit property
+of causing a virtual modification of the dynamic response of
+the human limb. We use this property of the exoskeletons
+action to formulate a unified control design framework called
+Integral Admittance [torque to angle] Shaping, which designs exoskeleton con-
+trollers capable of producing the desired dynamic response
+for the assisted limb. In this framework, a virtual increase
+in the admittance of the limb is produced by coupling it
+to an exoskeleton that exhibits active behavior. Specifically,
+our framework shapes the magnitude profile of the integral
+admittance (i.e. torque-to-angle relationship) of the coupled
+human-exoskeleton system, such that the desired assistance is
+achieved. This framework also ensures that the coupled stability
+and passivity are guaranteed.}~\cite{Nagarajan_etal_2015}
+
+{\it ... the impedance of the coupled human-exoskeleton
+system needs to be reduced below that of the unassisted
+human limb. This implies that the exoskeleton needs to
+cancel its own impedance first and then compensate for
+at least a part of the human limb’s impedance. Therefore,
+the desired exoskeleton behavior must be that of a negative
+impedance ... Consequently,
+the feedback gains will all be
+positive. In other words, the exoskeleton controller uses
+positive feedback, and hence the exoskeleton exhibits active
+behavior, which is capable of performing net positive work
+on the limb. ...
+However, positive feedback naturally raises the question of
+stability, and so we now explain how coupled stability can be
+achieved. Although the exoskeleton exhibits active behavior,
+which can be potentially destabilizing, the controller can be
+designed such that the coupled human-exoskeleton system
+is stable and passive. ... 
+}~\cite{Nagarajan_etal_2015}
+
+{\it
+... Inertia compensation is more complex ...
+It can be shown that using only positive acceleration
+feedback ..., the gain margin of the coupled system
+reduces to the moment of inertia of the exoskeleton, which
+implies that the exoskeleton controller ... can at the most
+compensate for the exoskeleton’s own moment of inertia
+before going unstable. This implies that the moment of
+inertia of the coupled human-exoskeleton system cannot be
+reduced below that of the unassisted human limb, without
+compromising coupled stability. However, using low-pass
+filtered acceleration feedback, it can be shown that inertia 
+reduction can be achieved. ...
+This work uses filtered acceleration
+feedback with a second-order low-pass Butterworth filter. ...
+}~\cite{Nagarajan_etal_2015}
+
+This is a form of frequency domain virtual model control.
+
+\begin{verbatim}
+Kalman filter:
+P. Canet, “Kalman filter estimation of angular velocity and accelera-
+tion: On-line implementation,” McGill University, Montr ́eal, Canada,
+Tech. Rep. TR-CIM-94-15, Nov. 1994.
+\end{verbatim}
+
+\subsection{Dual Control Approach}
+
+This approach implements a passive leg swing in addition to Sensitivity 
+Amplification Control for the stance leg.
+
+{\it The robot utilized the dual-mode control scheme, which is
+comprised of the active control for the stance phase and the
+passive control (using bypass valves) for the swing phase, to achieve high walking
+speed in the swing phase while supporting heavy loads in
+the stance phase. To reduce the sudden change of the torque
+command at the transition from the swing phase to the stance
+phase, a smoothing method is adopted. We also implemented
+a pre-transition method to take a foot off quickly for fast
+walking by predicting the change from the swing to the stance
+in advance.}~\cite{IEEE07222598}
+
+%Bypass valves implement what Sarcos calls ``dangle''.
+
+Stance: virtual joint torque control method Kazerooni et al.(2005)
+When contact location between the wearer and the
+exoskeleton is not fixed and difficult to estimate, this method
+has been shown to be an effective method to generate the
+locomotion for an exoskeleton robot.[3],[4],[21]
+\begin{verbatim}
+H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
+the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
+Conf. Robot. Autom., pp. 4364-4371, 2005.
+Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
+for Human Performance Amplification,” University of California,2003.
+582
+Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
+Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
+Control,” Journal of computers, vol.9, No.1,202-208. 2014.
+\end{verbatim}
+
+To reduce
+sudden changes(command jump) at the phase transitions, a
+smoothing method [5],[19] is introduced and a pre-transition
+method is used to solve the swing delay due to the internal
+pressure(approx 5 bar).
+\begin{verbatim}
+H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
+Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
+Robotics Research, pp. 561-573, 2006.
+oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
+Based Control for a Rotary Series Elastic Actuator Assisting the knee
+\end{verbatim}
+
+Transition Control
+
+1) Smoothing Method: During transitions of the gait
+phase, discontinuity of the control command torque are
+occurred by the different condition for the fixed coordinate
+which is on the backpack at the swing phase or the foot at
+the stance phase. In this paper, to reduce this sudden change
+due to gait phase changes, a smoothing method is proposed
+as shown in Fig. 7. An exponential function is considered
+as the weighting function for smoothing as shown in (6).
+In this case, the weighting is small at the initial stage but
+it exponentially converges to one for supporting the load
+quickly.
+
+2) Pre-transition Method: The pre-transition method is
+that the passive mode is executed in the pre-swing phase
+prior to toe off. This dramatically reduces the moving
+
+\subsection{Ground Reaction Force Control}
+
+{\it
+The ground reaction
+forces (GRF) magnitude and direction is used to command the
+actuators. In some research the GRF sensors are used
+together with other sensors in the control architectures [17],
+while in other exoskeleton the control system is based on the
+GRF merely [RoboKnee, Honda].
+}
+\begin{verbatim}
+17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
+"Intention-based walking support for paraplegia patients with Robot
+Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
+\end{verbatim}
+
+\subsection{Virtual model control}
+
+The goal here is to make the exoskeleton imitate a desired (usually nonlinear)
+dynamic system,
+For running it could be a pogo stick or trampoline, for example.
+
+There are several possible versions of this type of control:
+\begin{enumerate}
+\item
+Map from operator-exoskeleton contact forces and system state
+to exoskeleton actuator or internal forces.
+\item
+Map from operator-exoskeleton contact forces and system state
+to exoskeleton acceleration.
+\item
+Map from operator-exoskeleton contact forces and system state
+to exoskeleton-world contact forces.
+\item
+Map from exoskeleton-world contact forces and system state to exoskeleton actuator or internal forces.
+\item
+Map from exoskeleton-world contact forces and system state to exoskeleton acceleration.
+\item
+Some combination of the above.
+\end{enumerate}
+
+\begin{verbatim}
+virtual model approach: hopper, compass gait (J. Pratt)
+estimate trajectory (jumping) vs. program behavior (trampoline, hopper,
+   compass gait)
+- J' control + behavior
+- Inverse dynamics + behavior
+\end{verbatim}
+
+\subsection{Task Specific Control}
+
+Task specific control can involve switching low level control modes, or abstracting
+the behavior of the exoskeleton with a hierarchy:
+\begin{enumerate}
+\item
+Task specific control: Generate desired motions and contact forces between
+the operator and the exoskeleton, and the exoskeleton and the world.
+\item
+Sometimes there are intermediate levels of control.
+\item
+Exoskeleton control: Control the exoskeleton to generate the desired motions and
+desired contact forces, with the desired impedance or admittance.
+\end{enumerate}
 
@@
 
-sensitivity to
- modeling error
- sensor error
- unmodeled dynamics
-* future (after Oct 1) simulation studies
-
-underactuated control problem
- - how necessary is lateral actuation
-   * we don't know
- - * what is the penalty
-  weight
-  drag of actuator
-* future (after Oct 1) simulation studies
-
-how handle failure and damage 
-* after Oct 1
+I have incorporated your suggestions about robustness and underactuation
+in the white paper
+www.cs.cmu.edu/~cga/exo/control.pdf
+
+State estimation (to deal with sensor noise and to fuse disparate sensors)
+is discussed in section 21 and mentioned in 5.5, 5.8,
+
+Robustness (sensitivity to modeling error and unmodeled dynamics) is
+discussed in section 22
+
+Underactuation is discussed in section 18 and the following sections.
+
+Shared control is discussed in section 19.5
+
+********************************************
+
+A future white paper could explore using simulation studies to experimentally
+assess robustness of various control approaches.
+
+A future white paper could explore using simulation studies to numerically
+estimate:
+1) how necessary is lateral actuation
+2) what is the penalty (weight, drag of actuator, complexity, maintenance)
+
+A future white paper could explore how to handle failure and damage, again using
+simulation.

### 2015-09-25 design-goals.tex
cga updates

@@
 
-Can we build an exoskeleton allows the operator to behave naturally
+Can we build an exoskeleton that allows the operator to behave naturally
 and exerts little or no force on the operator?
@@
 
-Can we build an exoskeleton allows the operator to behave naturally?
+Can we build an exoskeleton that allows the operator to behave naturally?
 The operator can feel the exoskeleton, but can overpower it when necessary.

### 2015-09-25 control.tex
cga updates

@@
 
-\subsection{Full Atlas robot implementation.}
-
-Discuss or point to DRC papers by multiple groups.
+\subsection{Full robot implementations.}
+
+Point to DRC papers by multiple groups.
+
+Figure~\ref{f:cga1} shows the architecture of our robot control based on hierarchical optimization. 
+It was successfully tested on our SARCOS Primus Humanoid
+and ATLAS (Boston Dynamics) platforms.
+Our ATLAS robot demonstrated rough terrain walking, ladder climbing, and full body manipulation in the DARPA Robotics Challenge (DRC) Trials of 2013.
+This work also received
+a ``Best Oral Paper Award'' at Humanoids 2013~\cite{Feng/etal13}. Figure~\ref{fig:terrain_pic} shows a sequence of snapshots of ATLAS 
+walking on rough terrain and plots of measured trajectories of the feet and the center of mass (CoM) for the DRC Trials.
+Figure~\ref{fig:ladder_pic} shows similar data for ladder climbing.
+In the DARPA Robotics Challenge Finals in 2015 our robot was the only
+biped that did not fall down or need to be rescued by 
+humans during the two days of the evaluations~\cite{cga-hum15}.
+Our robot performance was also the most consistent across the two days.
+Because of the way the evaluations were run, we have only limited data
+from the actual evaluations. Videos of our robot's performance during the
+DRC Finals are available~\cite{cga-web-page}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.75\textwidth]{Figures/cga1_modified}
+\caption{Controller architecture based on hierarchical optimization.}
+\label{f:cga1}
+\end{figure}
+
+\begin{figure}[t]
+  \begin{center}
+    {\includegraphics[height=3.8cm]{Figures/atlas_terrain}}
+    \caption{ATLAS robot practicing for the terrain task of the DRC.
+Photo snapshots taken every 5 seconds.
+Measured trajectories in the transverse (a) and sagittal planes (b) of the left (red) and right foot (green) and the center of mass (CoM, blue) are shown.
+The robot walks in a straight line along $y=0$ in reality.
+Our state estimator at that time drifted significantly.
+This has been fixed by adding vision-based SLAM.
+      }\label{fig:terrain_data} 
+\label{fig:terrain_pic} 
+  \end{center}
+\end{figure}
+
+\begin{figure}[t]
+  \begin{center}
+    {\includegraphics[height=4cm]{Figures/atlas_ladder}}
+    \caption{
+      ATLAS climbing the top half of the same 
+      ladder as used in the DRC. Photo snapshots taken every 13 seconds. The
+      top row shows repositioning of the hook hands, and the bottom row shows
+      stepping up one tread.
+      The two plots show ATLAS climbing the first five treads during
+      the actual run at the DRC Trials.
+Axis definitions and color codes for foot and CoM trajectories are the same as in Fig.~\ref{fig:terrain_data}.
+Left and right hand positions are plotted with cyan and magenta dashed lines.
+      }\label{fig:ladder_data} 
+\label{fig:ladder_pic} 
+  \end{center}
+\end{figure}
+
+In the controller architecture, the {\bf Behavior Library} (Fig.~\ref{f:cga1}) contains
+learned behaviors and behaviors created from demonstration or
+motion capture of humans or other robots, and also acts as a cache
+for planning. The output of the library is used to prime, speed up,
+and guide the real time optimization process, and can be motion
+or force trajectories in joint or task coordinates ($\vx_d(t)$), as well as
+possibly including value, heuristic cost, or utility functions ($V_x(t)$)
+that can be used
+to guide search and optimization. 
+Libraries can be built in advance, or when task specification are known.
+The rest of the controller can
+also operate without information from the library if suitable information
+is not available.
+
+The {\bf Multistep Low-Order Dynamics Planner} uses
+receding horizon control to continually optimize trajectories of
+motion and force of a simplified model of the current task.
+In walking, this is usually center of mass ($COM_d(t)$),
+angular momentum ($\mL_d(t)$),
+contact force ($\mbox{contacts}_d(t)$) trajectories,
+and value, heuristic cost, or utility functions ($V_c(t)$)
+specifying tradeoffs between these quantities.
+We optimize with a time horizon of several seconds
+using Differential Dynamic Programming~\cite{ddp}.
+
+For walking, our planner is similar in spirit to Kajita's Preview Control for a Linear Inverted
+Pendulum Model (LIPM)~\cite{kajita03}. We are also using a CoM
+model, reason about Zero Moment Point (ZMP), and use future information 
+to guide the current trajectory. However, our planner can be generalized to 
+nonlinear models and adjust foot steps while optimizing the CoM 
+trajectory. 
+%We explicitly add the vertical height in our  CoM model to handle height variations on rough terrain. 
+Like capture point methods~\cite{capture_pt}, we take the 
+next few steps into consideration but
+do not plan to come to rest at the end. Ogura et al.~\cite{wabian2} 
+investigated generating human-like walking with heel-strike and toe-off by 
+parameterizing the swing foot trajectory. In contrast, we use very simple rules to guide the low level controller to achieve the same behaviors.
+
+The {\bf Full Body QP Optimization} uses quadratic programming (QP) with a
+very short time horizon (1ms on SARCOS and 2ms on ATLAS) to optimize the 
+desired full state of the robot ($\vtheta_d$, $\dot{\vtheta}_d$),
+joint torques ($\vtau_d$), and contact forces ($\vf_d$).
+Currently, we pre-specify gain matrices 
+($\mK_{\theta}$, $\mK_{\dot{\theta}}$, $\mK_{\tau}$,
+$\mK_{\mbox{f}}$)
+for the robot controller, but plan to explore how to optimize gains
+as well as trajectories.
+This module enforces constraints
+such as robot kinematics and dynamics 
+and joint, velocity, torque, actuator, friction cone,
+and center of pressure limits, and resolves redundancies.
+The module performs inverse kinematics and dynamics to provide us with
+compliant motions and dynamic behaviors, while compensating for the
+effects of modeling errors.
+
+For the QP, we use a formulation developed in our group 
+\cite{Stephens-thesis,Whitman10,Whitman-thesis}. Unlike~\cite{mistry10,usc_id1} who 
+use orthogonal decomposition to project the allowable motions into the null 
+space of the constraint Jacobian, and minimize costs in the contact constraints and the commands, we directly 
+optimize a quadratic cost in terms of state accelerations, torques and contact 
+forces on the full robot model. This design choice allows us to trade off physical quantities of 
+interest. We are also able to directly reason about inequality constraints 
+such as center of pressure within the support polygon, friction, and torque 
+limits. Although it becomes a bigger QP problem, we are still able to solve it 
+in real time. Hutter el al.~\cite{eth_id} resolved redundancy in inverse 
+dynamics using a kinematic task prioritization approach that ensures lower 
+priority tasks always exist in the null space of higher priority ones. In 
+contrast to their strictly hierarchical approach, we minimize a sum of 
+weighted terms. 
+We can directly specify the relative importance of the terms by adjusting 
+the weights.
+
+%One popular approach to controlling humanoid robots is through inverse 
+%kinematics with stiff joint position tracking. On the other hand, 
+%inverse dynamics
+%based approaches have gained increasing acceptance by providing compliant 
+%motions and robustness to external perturbations. 
+%However, the performance of inverse dynamics methods 
+%is heavily dependent on high quality dynamic models, which are often very 
+%difficult to produce for a physical robot. In contrast, inverse
+%kinematics based approaches 
+%only require kinematic models, which are much easier to generate in practice. 
+%This was one reason we supplemented our original two part implementation with 
+%an inverse kinematics controller for the physical robot.
+
+The {\bf Low-Level Robot Controller} is a blend of our work and controllers
+provided by the robot manufacturers. For SARCOS
+and ATLAS, hydraulic valve commands for each
+joint $i$ are generated by high rate servo controllers provided
+by the manufacturers,
+\begin{equation}
+v_i = k_{\theta,i} ( \theta_i - \theta_{d,i} )
++ k_{\dot{\theta},i} ( \dot{\theta}_i - \dot{\theta}_{d,i} )
++ k_{\tau,i} ( \tau_i - \tau_{d,i} )
++ v_{\ff,i}
+\end{equation}
+where $v_i$ is the valve command,
+$\theta_i$ is the joint angle,
+$\dot{\theta}_i$ is the joint angular velocity,
+$\tau_i$ is the joint torque,
+and $v_{\ff,i}$ is a feedforward valve command.
+Subscripts $d$ indicate desired values, and
+quantities denoted with $k$ are scalar gains.
+This joint level servo runs at 5kHz on SARCOS and at
+1kHz on ATLAS.
+
+We augment this independent joint control by coupling joints
+and providing force control based on force/torque sensors in the wrists,
+ankles, and skin of the robot,
+\begin{equation}
+\vv = \mK_{\theta} ( \vtheta - \vtheta_{d} )
++ \mK_{\dot{\theta}} ( \dot{\vtheta} - \dot{\vtheta}_{d} )
++ \mK_{\tau} ( \vtau - \vtau_{d} )
++ \mK_{\mbox{f}} ( \vf - \vf_{d} ) 
++ \vv_{\ff}
+\end{equation}
+where $\vv$, $\vtheta$, $\dot{\vtheta}$, $\vtau$, and $\vv_{\ff}$
+are the corresponding vector quantities, and the $\mK$ are matrix gains
+coupling all joints. $\vf$ are sensed contact forces.
+The coupled controller runs at 1kHz on SARCOS and 500Hz on ATLAS.
+
+Figure~\ref{f:cga1} does not include our state estimator, which provides
+appropriate state information to all other modules. The details of this work on state estimation are
+described in~\cite{xinji_com_est,Xinjilefu-thesis}.
 

### 2015-09-25 control.tex
cga updates

@@
 Algorithms\\
-(Draft 3.0)}
+(Draft 3.1)}
 
@@
 I have incorporated your suggestions about robustness and underactuation
-in the white paper
+in the white paper draft 3.1 now at:
 www.cs.cmu.edu/~cga/exo/control.pdf

### 2015-09-28 control.tex
more updates

@@
 
+This recent paper discusses shared control \url{http://www.ri.cmu.edu/pub_files/2015/7/Javdani15Hindsight.pdf}
+
 \subsection{Full robot implementations.}
@@
 
+We use the operator state (the set of all joint angles and velocities
+as well as a root position and velocity to represent operator location
+and orientation in the world) as the basis of inverse dynamics and
+online optimization-based control.
+In a full-body exoskeleton
+(not necessarily fully actuated), the
+exoskeleton can be used to measure operator positions, and velocities.
+For partial exoskeletons,
+it may be the case that additional sensors such as distributed 
+angle sensors (goniometers), tachometers, or inertial measurement units (IMUs)
+are attached to the operator or part of the operator's clothing.
+A major goal of state estimation is to process these measurements to
+produce a useful estimate of operator state.
+
 We have not seen a lot of discussion of state estimation of exoskeletons in
 the literature.
-We will discuss our approach to state estimation on humanoid robots here.
+In lieu of that,
+we will discuss our approach to state estimation on humanoid robots here.
 State estimation to deal with sensor noise,

### 2015-09-28 exos/bleex.tex
cga-fixes
Added bibliographies by section to survey.tex. Use biber to build survey.tex
edits to survey document

@@
-\subsection{BLEEX}
+\section{BLEEX}
 \label{exo:bleex}
+\begin{refsection}[exos/bleex.bib]
 
@@
 
-Assuming no outside disturbances, the BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as $d$ (assuming no outside disturbances).
+The BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as $d$ (assuming no outside disturbances).
 Neglecting gravity, the exokeleton angular velocity is modeled as
 \[v = G r + S d ,\] 
-where $v$ represents the angular velocity of the exokeleton, $G$ is the transfer function from actuator inputs ($G$ is the exoskeleton's dynamics), $r$ is the actuator input, and $S$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
+where $G$ is the transfer function from actuator inputs ($G$ is the exoskeleton's dynamics), $r$ is the actuator input, and $S$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
 
@@
 To achieve a large sensitivity, BLEEX uses $C = (1-\alpha^{-1})G^{-1}$ so that $\alpha$ provides a direct (scalar) amplification factor.
-A low pass filter is often added to the $C$ term in order to damp out high frequency dynamics of the exoskeleton, which are not captured in these models.
+A low-pass filter is often added to the $C$ term in order to damp out high frequency dynamics of the exoskeleton, which are not captured in these models.
 Note that the controller depends on an inverse dynamic model of the exoskeleton, $G^{-1}$.
@@
 
-\bibliographystyle{plain}
-\bibliography{exos/bleex}
+\printbibliography[heading=subbibliography]
 
 The figures in this section were obtained from \cite{sesitivityAmpPaper2005}. Materials presented are based on the references above.
+
+\end{refsection}

### 2015-09-28 exos/exosText_MIT.tex
edits to survey document

@@
 
-Design is based  on fact that evidence from biology and passive walking devices suggest that legged locomotion can be very efficient.
+The MIT Exoskeleton is based on evidence from biology and passive walking devices that suggest that legged locomotion can be very efficient.
 The main physical concept behind this efficiency being that there is a gait cycle for a pair of legs that naturally exchanges energy between gravity and inertia.
-The main mechanical design principles centered around using a combination of passive and active elements.
-In particular, the MIT design uses {\bf springs, variable impedance joints, and powered actuators.}  The suit generally  functions as a lightweight, underactuated robot which runs in parallel to an operator and supports the weight of a payload.
+The main mechanical design principles center around the use of a combination of passive and active elements.
+In particular, the MIT design uses {\bf springs, variable impedance joints, and powered actuators.}  The suit generally  functions as a lightweight, underactuated robot that runs in parallel to an operator and supports the weight of a payload.
+Additionally, the leg structure of the suit allows weight from the suit and payload to be transferred directly to the ground.
 
-Additionally, the leg structure of the suit allowed weight from the suit as well as payload to be transferred directly to the ground.
-
-Additionally, there are several other design principles that the MIT team discusses which are especially relevant to this report.
-The first is that joint {\bf powers scale linearly with mass}  The second is that altering the gait pattern of the operator has been shown to increase the physiological energy expended during locomotion.
+Of the several other design principles discussed, the MIT team emphasizes that joint {\bf powers scale linearly with mass}.
+Also, the team points out that alterations in the operator's gait pattern have been shown to increase the physiological energy expended during locomotion.
 Thus, the specifications for actuation and control for their system are extracted from the angle, torque , and power data of human walking joint patterns.
@@
 The actuator is designed around a total system weight of 165 kg.
-The maximum scaled hip torque for a system with this mass is approximately negative 130 Nm during the stance phase and approximately 100 Nm during the swing phase of the gait.
+The maximum scaled hip torque for a system with this mass is approximately $-130$ Nm during the stance phase and approximately 100 Nm during the swing phase of the gait.
 The hip is chosen as the actuated joint because proximal mass is metabolically less expensive in walking than distal mass.
@@
 The specific choice of actuation for the MIT exoskeleton hip joint was a {\bf linear series elastic actuator}.
-This choice was made because it provided a lightweight and relatively inexpensive means of implementing force control with a bandwidth similar to that of natural muscle.
+This choice provided a lightweight and relatively inexpensive means of implementing force control with a bandwidth similar to that of natural muscle.
 A model of the actuator is shown in Figure \ref{fig:mitSEA}.
@@
  \end{figure} 
-Like all series elastic designs the actuator shown in Figure \ref{fig:mitSEA} includes a spring in series with the output of the motor drive system of the actuator.
+Like all series elastic designs, the actuator shown in Figure \ref{fig:mitSEA} includes a spring in series with the output of the motor drive system of the actuator.
 The motor-drive mechanism consists of a {\bf brushed DC motor driving a 3mm ball screw via a 2:1 reduction belt drive}.
@@
 The compression of the die springs are measured using a linear potentiometer, which is how forces at the output are measured.
-
-The motor chosen for the {\bf actuator was a 150W Maxon RE40 Brushed DC motor}, which was chosen based on the bio-mechanical data and for its power and weight ratio.
-
-The force bandwidth for the actuator was characterized during both stance and swing phases of the walking gait cycle.
+The team selected a {\bf 150W Maxon RE40 Brushed DC motor} for the actuator  based on the bio-mechanical data and its power and weight ratio.
+The force bandwidth for this actuator was characterized during both stance and swing phases of the walking gait cycle.
 {\bf During stance, the bandwidth of the actuator was 35 Hz, while the bandwidth during swing was found to be 40Hz.}
@@
  The MIT exoskeleton design has a total of fourteen degrees of freedom, three at each hip, one at each knee, two at each ankle, and one in each foot.
-
- A cam mechanism was implemented at the hip joint to enable hip ab/adduction.
-This system was necessary because, during abduction, there is was a length difference between the operator and exoskeleton leg which resulted from having non-collocated centers of rotation.
+A cam mechanism implemented at the hip joint enables hip ab/adduction.
+This system was necessary because, during abduction, there is a length difference between the operator and exoskeleton leg which results from having non-collocated centers of rotation.
 The cam mechanism automatically adjusts the exoskeleton leg length such that the center of rotation of the exoskeleton hip is projected onto the biological hip center of rotation.
@@
  \end{figure} 
+% 
+ The interface between the suit and operator consists of shoulder straps, a compliant waist belt, thigh cuffs, and a shoe connection.
 
- The interface between the suit and operator consisted of shoulder straps, a compliant waist belt, thigh cuffs, and a shoe connection.
+ The exoskeleton implements a {\bf variable damper knee mechanism} in the flexion/extension degree of freedom using a magnetorheological damper.
+The damper can exert a {\bf maximum braking torque of 60 Nm and consums on average 1W of electrical power}.
+Additionally, a linear spring at the ankles captures negative energy during dorsilflexion.
+The energy stored is released during plantar flexion to add energy to the system as the foot comes off the ground.
+A urethane spring (247 kN/m) is used as a liner spring in a lever compression assembly.
+With a lever arm of 0.0381 m, the assembly yields a spring rotary stiffness of 356 Nm/rad.
 
- The {\bf variable damper knee mechanism} in the flexion/extension degree of freedom was implemented using a magnetorheological damper.
-The damper could exert a {\bf maximum braking torque of 60 Nm and consumed on average 1W of electrical power}.
+ A unidirectional spring assembly is also used in the hip ab/adduction degree of freedom.
+The spring allows the hip to freely abduct.
+The effective rotary stiffness of the spring assembly is 96 Nm/rad in order to provide the 8 Nm required during adduction.
 
- A linear spring located at the ankle was designed to capture negative energy during dorsilflexion.
-The energy stored was released during plantar flexion to add energy to the system as the foot comes off the ground.
-A urethane spring was used as a liner spring in a lever compression assembly.
-The spring constant of the urethane was 247 kN/m, the lever arm of the assembly was 0.0381 m, which gave the rotary spring a stiffness of 356 Nm/rad.
-
- A unidirectional spring assembly was also used in the hip ab/adduction degree of freedom.
-The spring allowed the hip to freely abduct.
-It was calculated that 8 Nm were required during adduction.
-To this end, the effective rotary stiffness of the spring assembly was 96 Nm/rad.
-
-The suit was powered using a 48V battery pack.
-
-The suit was instrumented with rotary potentiometers at the hip and knee.
-Additionally, strain gauges were placed on the structure of the exoskeleton shank, which measured sagittal bending moment and compression force
+The suit uses a 48V battery pack and is instrumented with \textbf{rotary potentiometers} at the hip and knee.
+Additionally, \textbf{strain gauges} were placed on the structure of the exoskeleton shank to measured sagittal bending moment and compression force.
 
@@
 
- The control of the MIT exoskeleton functions around the fact that desired actuation and active damping at the knee are functions of gait cycle, at that the values of these desired functions could be determined from human walking data.
-Fundamentally, the control strategy works as a state machine, which uses joint angle and measured forces to implement state transitions.
+The MIT exoskeleton controllers use the fact that desired actuation and active damping at the knee are functions of gait cycle.
+Hence, the values of these desired functions can be determined from human walking data.
+Fundamentally, the control strategy works as a state machine that uses joint angle and measured forces to implement state transitions.
 
@@
 For the hip there are four distinct strategies.
-During the thrust phase, the hip exerts a torque which helps to raise the center of mass of the system.
-During the extension spring phase, a virtual spring stiffness is programmed which allows energy to be virtually stored while the center of mass moves forward.
+During the thrust phase, the hip exerts a torque that helps to raise the center of mass of the system.
+During the extension spring phase, a virtual spring stiffness allows energy to be virtually stored while the center of mass moves forward.
 During the swing assist phase the virtual energy is released, resulting in a torque being applied which assists in swinging the leg forward.
 Finally, in leg retraction a torque is applied to help with foot placement and weight acceptance.
-A diagram which illustrates how the sensors in the shank and hip joint are used to switch between these various phases is show in Figure \ref{fig:hipControl}
+A diagram that illustrates how the sensors in the shank and hip joint are used to switch between these various phases is show in Figure \ref{fig:hipControl}
  \begin{figure}[thpb]
@@
 During knee strike, the variable damper in the knee exerts a torque proportional to the rotational velocity of the knee joint.
-A residual magnetic field remained in the knee joint after it was turned off, creating a resistive torque.
-The knee needed to be actively demagnetized at full extension during the last phase of stance to allow the knee to swing freely during the swing phase.
-The state machine diagram that illustrates how the various phase transitions for the knee joint were implemented using the exoskeleton's onboard sensors is shown in Figure \ref{fig:kneeControl}
+A residual magnetic field remains in the knee joint after it is turned off, creating a resistive torque.
+The knee therefore needs to be actively demagnetized at full extension during the last phase of stance to allow it to swing freely during the subsequent swing phase.
+The state machine diagram that illustrates how the various phase transitions for the knee joint were implemented using the exoskeleton's on-board sensors is in Figure \ref{fig:kneeControl}
   \begin{figure}[thpb]
@@
 
- The MIT hardware and control system present a very well designed and thought out system which focus primarily on efficiency in design.
+ The MIT hardware and control system present a very well designed and thought-out system that focuses primarily on efficiency in design.
 The series elastic linear ball screw hardware design seems to be a very smart way to implement efficient drives as well as to perform high fidelity force control.
-Using a strain gauge in the shank was also a novel means of of measuring ground reaction forces.
+The use of a strain gauge in the shank was also a novel means of measuring ground reaction forces.
 This design may offer benefits in terms of repeatability over different operators as foot fall patterns for different users may include significant variations.
@@
 1410-1415.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-28 exos/exosText_dualMode.tex
edits to survey document

@@
 
-Exoskeleton suit was originally developed for military uses, specifically to allow operators to perform motions without incurring muscle fatigue.
-Their main contributions in the literature re based on a {\bf virtual joint torque control}, which has been shown to be an effective human robot interaction control strategy in situations where the connections between the uses and hardware system are relatively unknown or can dramatically change as a function of time.
+The Hydraulic Lower Extremity Exoskeleton suit was originally developed for military uses, specifically to allow operators to perform motions without incurring muscle fatigue.
+The exoskeleton uses a new {\bf virtual joint torque control} approach, which has been shown to be effective for controlling human robot interaction in situations where the connections between the users and hardware system are relatively unknown or can dramatically change as a function of time.
 
-Additionally, the hardware system focuses on actuator selection.
-The main hardware design principle is focused on developing a suit that can augment a users power and move fast when necessary.
-They selected hydraulic actuation because the addition of electric actuator weight on the system leg joints makes the system difficult to move quickly.
-On the other hand, hydraulic actuators have high power density relative to weight and size, and generally provide very high force outputs and low impedance.
-The tradeoff with hydraulics come in the way of having low accuracy in terms of force control as well as hard to model nonlinear characteristics.
+Additionally, the main hardware design principle is focused on developing a suit that can augment user power and yet move fast when necessary.
+The design team selected hydraulic actuators because the weight of electric actuators on leg joints would make the system difficult to move quickly.
+Hydraulic actuators have high power density relative to weight and size and generally provide very high force outputs and low impedance.
+As a trade-off, hydraulics have reduced accuracy in terms of force control and it is difficult to model their nonlinear characteristics.
 
@@
 
-The actuator specifications for the Hydraulic Lower Extremity Exoskeleton were designed around human walking data at 4km/h with a 45 kg load.
-The resulting calculated hydraulic capacity for the while system was 2050psi and 8lpm.
+The Hydraulic Lower Extremity Exoskeleton's actuators were designed based on human walking data obtained at 4km/h with a 45 kg load.
+Based on this data, the hydraulic capacity system was designed to provide 2050psi and 8lpm.
 The individual actuators were designed to be double acting at the hip and knee flexion/extension joints.
-The {\bf hip maximum thrust was 4 kN} and {\bf knee maximum thrust was 4 kN}.
-Each joint joint included three-way servo valves (M 200, Star-Hydraulic) to control rate and direction of bi-directionality and bypass valves to connect the pump path to tank path for fast, energy efficient motion during swing phases.
+They provide a {\bf hip maximum thrust of 4 kN} and {\bf knee maximum thrust of 4 kN}.
+Each joint joint includes three-way servo valves (M 200, Star-Hydraulic) to control rate and direction of bi-directionality, and bypass valves to connect the pump path to tank path for fast, energy-efficient motion during swing phases.
 
@@
 
- The Hydraulic Lower Extremity Exoskeleton includes four actively controlled joints in the sagittal plane; the knee and hip flexion/extension degrees of freedom are powered degrees of freedom for each leg.
-The hardware also includes eight passive joints: one ab/adduction DOF for each hip and three degrees of freedom for each ankle.
+ The Hydraulic Lower Extremity Exoskeleton includes four actively controlled joints in the sagittal plane, with powered knee and hip flexion/extension.
+The eight passive joints include one ab/adduction DOF for each hip and 3 DOF for each ankle.
 
- The hydraulic system of the suit is composed of the four actuators and a hydraulic power unit, consisting of a pump, motor, temperature sensor, and various valves.
+ The hydraulic system is composed of the four actuators and a hydraulic power unit, consisting of a pump, motor, temperature sensor, and various valves.
 
- Additionally, various {\bf sensors} are incorporated to measure joint angles, actuator forces , and ground reaction forces.
-Three absolute angular encodes are incorporated at the hip, knee, and ankle pitch joints.
-In-line load cells are incorporated in each cylinder tube to measure actuator forces, and four single axis load cells are included in foot sensors to be used for gait phase detection.
+ On-board sensors measure \textbf{joint angles}, \textbf{actuator forces}, and \textbf{ground reaction forces}.
+Three \textbf{absolute angular encoders} are incorporated at the hip, knee, and ankle pitch joints.
+In-line \textbf{load cells} are incorporated in each cylinder tube to measure actuator forces, and four single axis load cells are included in foot sensors to be used for gait phase detection.
 
@@
  \end{figure}
-The control design principle for the Hydraulic Lower Extremity Exoskeleton is base around the concept that, for walking gaits, the {\bf stance and swing controllers should be considered separately}.
-To this end, the literature for this hardware system introduces an active-passive control method called {\bf Dual-Mode Control}.
+The control design principle for the Hydraulic Lower Extremity Exoskeleton is based on the concept that, for walking gaits, the {\bf stance and swing controllers should be considered separately}.
+To this end, the control designers developed an active-passive control method called {\bf Dual-Mode Control}.
 Active control is used during stance phases whereas the passive control is used to quickly and freely control the system's legs during swing phases.
-Additionally, to handle the inherently discontinuous jumps in the command signals that occur as a function of this dual mode abstraction, a smoothing method is introduced that is used in conjunction with a pre-transition method that solves the swing delay due to internal cylinder pressure.
-A diagram which highlights the different modes in this control strategy is shown in Figure \ref{fig:dualModeDia}.
+Additionally, to handle the inherently discontinuous jumps in the command signals that occur in this dual mode abstraction, a smoothing method is used in conjunction with a pre-transition method that solves for the swing delay due to internal cylinder pressure.
+A diagram which highlights the different modes in this control strategy is in Figure \ref{fig:dualModeDia}.
 
-Various components of the overall dual-mode control strategy, at the hardware level, are presented in Figure \ref{fig:suitDia}.  
+Various components of the dual-mode control strategy, at the hardware level, are presented in Figure \ref{fig:suitDia}.  
   \begin{figure}[thpb]
@@
  \end{figure}
-This figure contains a number of components, specifically a {\bf gait phase estimation block} which uses the foot sensors to estimate the stance mode of the system, a {\bf human intention block}, which uses angular encoders and joint force sensors to estimate the motion of the operator, a {\bf torque controller}, and a {\bf gravity compensation} block.
+This figure contains a {\bf gait phase estimation block}, which uses the foot sensors to estimate the stance mode of the system, a {\bf human intention block}, which uses angular encoders and joint force sensors to estimate the motion of the operator, a {\bf torque controller}, and a {\bf gravity compensation} block.
 Note that these various components of the control system are used to support the general assumption of the dual mode controller, i.e.,
@@
  \end{figure}
-
+ %
  \noindent
@@
 
-Like other control methods presented in this report, the dual mode control method in this section uses the fundamental assumption that it is difficult to measure the interaction forces between the user and hardware system directly as a primary design principle.
+Like other controllers presented in this report, the dual mode control method in this section uses the fundamental assumption that it is difficult to measure the interaction forces between the user and hardware system directly as a primary design principle.
 To this end, the human-robot interaction force is assumed to be completely represented in terms of the joint torque applied by the operator in the hardware.
@@
 \begin{equation}
-T_a = M(q)\ddot{q} + C(q,\dot{q})\dot{q}+G(g),
+\vT_a = \mM(\vq)\ddot{\vq} + \mC(\vq,\dot{\vq})\dot{\vq}+\mG(\vq),
 \label{dynamics}
 \end{equation}   
-where $T_a$ is the joint torque applied by the actuators, $M(q)$ is the inertia matrix, $C(q,\dot{q})$ is the centripetal and Coriolis matrix, and $G(q)$ is the gravitational torques.
+where $\vT_a$ is the joint torque applied by the actuators, $\mM(\vq)$ is the inertia matrix, $\mC(\vq,\dot{\vq})$ is the centripetal and Coriolis matrix, and $\mG(\vq)$ is the gravitational torques.
 note that this equation neglects friction as well as other actuation dynamics.
@@
 \begin{equation}
-\hat{T}_{hm} = M_n(q)\ddot{q} + C_n(q,\dot{q})\dot{q} + G_n(q) - \hat{T}_a, \notag
+\hat{\vT}_{hm} = \mM_n(\vq)\ddot{\vq} + \mC_n(\vq,\dot{\vq})\dot{\vq} + \mG_n(\vq) - \hat{\vT}_a, \notag
 \end{equation}
-where $M_n$, $C_n$, and $G_n$ are the nominal models of the true inertia, centripetal, and gravitational parameters and $\hat{T}_a$ represents the measured actuator torques.
-The $M_n(q)\ddot{q} + C_n(q,\dot{q})\dot{q} + G_n(q) $ portion of $\hat{T}_{hr}$ represents the inverse dynamics component of the closed-loop control system.
-The overall joint torque controller, which combines compensation to for operator intent as well as balancing the suit under a gravitational load is then written \[ \tau_\text{active} = K(s) \hat{T}_{hm} + G_n(q),\] where $K(s) = K_p +s K_d,$ i.e., there is closed-loop PD control law operating on the error between the predicted robot dynamics and the measured joint torques.
+where $\mM_n$, $\mC_n$, and $\mG_n$ are the nominal models of the true inertia, centripetal, and gravitational parameters and $\hat{\vT}_a$ represents the measured actuator torques.
+The $\mM_n(\vq)\ddot{\vq} + \mC_n(\vq,\dot{\vq})\dot{\vq} + \mG_n(\vq) $ portion of $\hat{\vT}_{hr}$ represents the inverse dynamics component of the closed-loop control system.
+The overall joint torque controller, which combines compensation for operator intent as well as balancing of the suit under a gravitational load, is then written \[ {\bf\tau}_\text{active} = K(s) \hat{\vT}_{hm} + \mG_n(\vq),\] where $K(s) = K_p +s K_d,$ i.e., there is closed-loop PD control law operating on the error between the predicted robot dynamics and the measured joint torques.
 The block diagram for the active mode controller is shown in Figure \ref{fig:blockDia}.
@@
  \end{figure}
- Note that $G_a$ represents the dynamics of the the system and $G_a'$ the inverse dynamics model in Figure \ref{fig:blockDia}.
+ Note that $\mG_a$ represents the dynamics of the the system and $\mG_a'$ the inverse dynamics model in Figure \ref{fig:blockDia}.
 
  In the passive mode, which occurs during the swing phase, the leg of the human-robot system moves quickly and freely under using the passive inertial forces of the overall system.
-This is accomplished using bypass valves that connect the pump path to the tank path to achieve mechanical back-drivability.
-The foot sensors on each foot are used to trigger the bypass valves and to initiate the pass control mode.
+This is accomplished using bypass valves that connect the pump and tank paths to achieve mechanical back-drivability.
+The foot sensors trigger the bypass valves and to initiate the pass control mode.
 
-A novel component in the literature of the Hydraulic Lower Extremity Exoskeleton controller is how the transitions between the active and passive control modes are handled.
-In particular, an explicit smoothing method is used to address the discontinuity in the command torque at phase transitions.
-One solution which is proposed is to use an exponential weighting function of the form
+The Hydraulic Lower Extremity Exoskeleton uses a novel scheme to manage transitions between the active and passive control modes.
+In particular, the controller applies an explicit smoothing method to address discontinuities in the command torque at phase transitions.
+The implementation proposed uses an exponential weighting function of the form
  \begin{figure}[thpb]
@@
  \noindent
- where $\tau_\text{inv}$ is the inverse dynamics torque associated with the robot model, $f_\text{weighting}$ is the weighting function, $a$ is a sensitivity factor, and $N$ describes the duration of the transition period.
+ where ${\bf\tau}_\text{inv}$ is the inverse dynamics torque associated with the robot model, $f_\text{weighting}$ is the weighting function, $a$ is a sensitivity factor, and $N$ describes the duration of the transition period.
 The transition duration period as well as sensitivity factor appear to be hand tuned parameters.
 
-In addition to phase transition smoothing a pre-transition control method which happens at the end of the stance phase prior to the passive swing phase is also used to effect smooth phase transitioning.
+In addition to phase transition smoothing, a pre-transition control method is applied at the end of the stance phase prior to the passive swing phase in order to affect smooth phase transitioning.
 The pre-transition control works by measuring the ground reaction force and setting a minimum threshold value.
-When the GRF drops below the threshold, the passive swing control mode is entered, i.e., the bypass valve at the actuators are opened.
+When the GRF drops below the threshold the system enters a passive swing control mode, i.e., the bypass valve at the actuators are opened.
 The objective of this method is to reduce the phase transition delay due to residual pressure in the foot sensors.
@@
 577-582.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-28 survey.tex
Added bibliographies by section to survey.tex. Use biber to build survey.tex
added Matts stuff but still need to get his bib files
edits to survey document

@@
 \usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
-\usepackage{cite}
+% \usepackage{cite}
 \usepackage{graphicx}
@@
 \usepackage{hyperref}
+\usepackage[backend=biber]{biblatex}
+\addbibresource{exos/bleex.bib}
+\addbibresource{exos/hal.bib}
+\addbibresource{exos/xor.bib}
 
@@
 \input{latex-commands}
+
+\newcommand{\vF}{{\bf F}}
+\newcommand{\vV}{{\bf V}}
+\newcommand{\vG}{{\bf G}}
+\newcommand{\vT}{{\bf T}}
 
@@
 
+\input{exos/exosText_bodyExt}
+
+\input{exos/exosText_dualMode}
+
+\input{exos/exosText_ihmc}
+
+\input{exos/exosText_MIT}
+
+\input{exos/exosText_roboKnee}
+
 % \input{exos/exo-ul7}
@@
 
-\bibliographystyle{plain}
-\bibliography{exo}
+% \bibliographystyle{plain}
+% \bibliography{exo}
 
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:
+
+Notes:
+
+I copied figure captions and numbers. Matt did not.  Coordinate?
+
+Coordinate notes about sEMG and user intention -- BLEEX;
+
+I don't point out what the terms are in the dynamics equations.
+
+See if the Hydraulic Lower Extremity Exo is sensitivity amp and if so change opinion so we don't favor the approach.
+
+Discuss IHMC -- issues with high impedance torque control in practice.
+
+Talk a lot about drawbacks of SEAs in IHMC.  Are these included in MIT?

### 2015-09-28 exos/exosText_bodyExt.tex

@@
 
-The body extender is one of the few existing full-body exoskeletons.
-The Body Extender's main objective is to increase the forceful interaction capabilities, specifically handling heavy loads, of an operator in difficult and unknown environments.
+The Body Extender is one of the few existing full-body exoskeletons.
+The Body Extender's main objective is to increase the forceful interaction capabilities, specifically the heavy load handling capabilities, of an operator in difficult and unknown environments.
 The overall suit is composed of four robotic limbs with kinematics designed to be anthropomorphically similar.
@@
 
-The main design objectives for the Body Extender included  requiring that operators be able to use the hardware with a minimal amount of training.
-Thus, the system should not require any substantial modification the a human's motor habits.
-This objective includes the fact that the mass distribution of the suit, of special consideration in these heavy systems, be similar to that of an unloaded operator.
-Additionally, the resistive forces between the operator and suit should always be minimized, even when the suit is not loaded externally.
+As a primary objective, the Body Extender aims to allow operators to use the hardware with minimal training.
+Thus, the system is designed so as not to require substantial modifications to human motor habits.
+The effective mass distribution of the suit, of special consideration in these heavy systems, is designed to be similar to that of an unloaded operator.
+Like most other strength augmentation exoskeletons, the Body Extender is intended to minimize resistive forces between the operator and suit, even when the suit is not loaded externally.
 
-A picture of the Body Extender exoskeleton and its five main components, four limbs and a torso, are shown in Figure \ref{fig:bodyExt}.
+A picture of the Body Extender exoskeleton and its five main components, four limbs and a torso, is shown in Figure \ref{fig:bodyExt}.
 \begin{figure}[thpb]
@@
 
- Each of the actuators in the Body Extender suit, execpt the forearm pronosupination are dirven by a linear actuator which is composed of an electrical motor, incremental encoder, ball screw, and angular contact ball bearings.
+ Except for forearm pronosupination, each of the actuators in the Body Extender suit are driven by a linear actuator that is composed of an electrical motor, incremental encoder, ball screw, and angular contact ball bearings.
 The motors are frameless DC torque controlled motors.
-The {\bf peak torque of the motors are 7 Nm, with a maximum continuous torque of 6 Nm}.
+The {\bf peak torque of the motors is 7 Nm, with a maximum continuous torque of 6 Nm}.
 Each motor weighs 1.4 kg.
@@
 
- For joints that have a small range of motion and correspondingly low torque requirements, the linear actuators are used in straightforward lever mechanism configurations.
-For degrees of free which require relatively large ranges of motion as well as torque, a motion conversion system is used to transform the linear actuator motion into rotational motion.
-The motion conversion unit consisted of a pantograph, two metallic tendons, and an output pulley connected to the output link.
+ Joints that have a small range of motion and low torque requirements use linear actuators in straightforward lever mechanism configurations.
+For degrees of freedom which require relatively large ranges of motion as well as torque, a motion conversion system transforms the linear actuator motion into rotational motion.
+The motion conversion unit consists of a pantograph, two metallic tendons, and an output pulley connected to the output link.
 A CAD drawing of the motion conversion unit is shown in Figure \ref{fig:motionConv}.
@@
 The two units weigh 6 kg and 5 kg respectively, and have {\bf mechanical efficiency of about 85\%}.
-{\bf The speed of the actuators are about 60 deg/s}.
+{\bf The speed of the actuators is approximately 60 deg/s}.
 
@@
 
- The Body Extender exoskeleton consists of twenty-two independently actuated degrees of freedom.  The suit has two arms, two legs,and a torso, or central backpack unit.
+ The Body Extender's twenty-two actuated degrees of freedom include two arms, two legs, and a torso or central backpack unit.
+The arms have a servo-amplified degree of freedom for grasping heavy objects.
+Grasping forces are achieved using a force sensing trigger mechanism; the force is proportional to the force applied to the trigger.
 
- Each of the two arms on the system has a servo-amplified degree of freedom for grasping heavy objects.
-The grasping force of various objects is achieved using a force sensing trigger mechanism; the grasping force is proportional to the force applied to the trigger.
-
- The suit contains twenty-two incremental encoders which measure angular displacement of each of the electric motors.
+ The suit contains twenty-two incremental encoders that measure angular displacement of each of the electric motors.
 There are also {\bf five six-axis force/torque sensors mounted at the five connecting points between the user and the device; two at the hands, two at the feet, and one on the trunk.}  These sensors directly measure the forces and torques between the user and the suit.
 There are additionally two single axis force sensors integrated into the grippers that measure the force applied by the operator on the ``gripper triggers," as well as {\bf a three axis accelerometer integrated into the backpack unit which estimates the trunk orientation}.
-Lastly, accelerometers are distributed throughout the system to monitor link dynamics, serving the roll of supporting overall equilibrium and internal stability during manipulation tasks.
+Lastly, \textbf{accelerometers are distributed throughout the system} to monitor link dynamics, supporting overall equilibrium and internal stability during manipulation tasks.
 
-{\bf The control software is distributed on one central processing unit as well as sixteen local units located throughout the system}.
+{\bf The control software is distributed between a central processing unit and sixteen local units located throughout the system}.
 The communication between processing units is handled using a high-speed field bus.
@@
 {\bf The system is powered using thirteen batteries in series, which provide a nominal voltage of 78 V}.
-When the system is run in stand alone mode, the batteries can provide eight hours of continuous work.
+When the system is run in stand-alone mode, the batteries can provide eight hours of continuous work.
 
@@
 \begin{equation}
-B_i(q) \ddot{q} + C_i(q,\dot{q})\dot{q} + G_i(q) = \tau + J^T_{iu}F_s + J^T_{ia}F_a,
+\mB_i(q) \ddot{q} + \mC_i(q,\dot{q})\dot{q} + \mG_i(q) = {\bf \tau} + \mJ^T_{iu}{\bf F_s} + \mJ^T_{ia}{\bf F_a},
 \end{equation}
-where the $i = \{ i,2,3,4\}$ where each integer corresponds to a different limb, $\tau$ represents the actuator torques, $F_s$ the measured forces between the operator and suit, and $F_a$ the unmeasured contact forces on the suit.
-The $B_i$, $C_i$, and $G_i$ terms represent the limb inertia, Coriolis and centrifugal, and gravitational torques.
+where $i$ is an integer corresponding to a different limb, ${\bf \tau}$ represents the actuator torques, ${\bf F_s}$ the measured forces between the operator and suit, and ${\bf F_a}$ the unmeasured contact forces on the suit.
+The $\mB_i$, $\mC_i$, and $\mG_i$ terms represent the limb inertia, Coriolis and centrifugal, and gravitational torques.
 
@@
 \begin{equation}
-B_i(q) \ddot{q} + C_i(q,\dot{q})\dot{q} + G_{iT}(q) = \tau + J^T_{iu}F_s + J^T_{ia}F_a - B_{iT}{q}\dot{V}_T - C_{iT}(q,\dot{q})V_T,
+\mB_i(q) \ddot{q} + \mC_i(q,\dot{q})\dot{q} + \mG_{iT}(q) = {\bf \tau} + \mJ^T_{iu}F_s + \mJ^T_{ia}F_a - \mB_{iT}{q}\dot{V}_T - \mC_{iT}(q,\dot{q}){\bf V_T},
 \end{equation}
-where, $V_T$ is the trunk velocity, $G_{iT}$ the gravitational effect of the trunk on the limb, $B_{iT}$ the inertia of the trunk acting on the limb, and $C_{iT}$ the trunk Coriolis and centrifugal term.
-Note that $V_T$ and $\dot{V}_T$ are directly measured through the accelerometer and gyroscope in the trunk.
+where, ${\bf V_T}$ is the trunk velocity, $\mG_{iT}$ the gravitational effect of the trunk on the limb, $\mB_{iT}$ the inertia of the trunk acting on the limb, and $\mC_{iT}$ the trunk Coriolis and centrifugal term.
+Note that ${\bf V_T}$ and $\dot{{\bf V}}_{\bf T}$ are directly measured through the accelerometer and gyroscope in the trunk.
 
-The basis of the feedback linearizing controller in \cite{} operates under two fundamental assumptions, and only appears to relate directly to the case of a single one of the manipulator arms.
+The feedback linearizing controller in \cite{} operates under two fundamental assumptions and only appears to relate directly to the case of a single manipulator arm.
 The first assumption of the controller is that the mass and inertial properties of an object being manipulated are estimated online.
@@
 \begin{equation}
-F_s = M_p \dot{V}_e + G_p,
+\vF_s = \mM_p \dot{\vV}_e + \vG_p,
 \label{eq:forceEq}
 \end{equation}
-where $M_p$ is an effective mass of the end effector, $\dot{V}_e$ is the acceleration of the end effector, and $G_p$ is an effective gravitational term.
+where $\mM_p$ is an effective mass of the end effector, $\dot{\vV}_e$ is the acceleration of the end effector, and $\mG_p$ is an effective gravitational term.
 Note that the gravitational term is zero if the arm is not holding an external load.
@@
 \begin{equation}
-B(q) \ddot{q} + C(q,\dot{q})\dot{q} + G(q) = \tau. \notag
+\mB(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq})\dot{\vq} + \mG(\vq) = {\bf \tau}. \notag
 \end{equation} 
@@
 \begin{equation}
-\tau = B(q)(\ddot{q}_d + K_v \dot{\tilde{q}} + K_p \tilde{q}) + C(q,\dot{q})\dot{q} + G(q), \notag
+{\bf \tau} = \mB(\vq)(\ddot{\vq}_d + K_v \dot{\tilde{\vq}} + K_p \tilde{\vq}) + \mC(\vq,\dot{\vq})\dot{\vq} + \mG(\vq), \notag
 \end{equation}
-where $\tilde{q} = q_d - q$.
-The details are unclear in \cite{}, but it appears that the authors either propose using predefined trajectories to define $\ddot{q}_d$, $\dot{q}_d$, and $q_d$, or use the assumption in \eqref{eq:forceEq} to determine these values.
+where $\tilde{\vq} = \vq_d - \vq$.
+The details are unclear in \cite{}, but it appears that the authors either propose using predefined trajectories to define $\ddot{\vq}_d$, $\dot{\vq}_d$, and $\vq_d$, or use the assumption in \eqref{eq:forceEq} to determine these values.
 %\begin{equation}
-%F_s = M_p \dot{V}_e
+%\vF_s = \mM_p \dot{V}_e
 %\label{eq:assump1}
 %\end{equation}
-Equation \ref{eq:forceEq} relates the measured force to the derivative of the end effector velocity $\dot{V}_e$.
-This end effector velocity can then be related to the desired joint velocities using the analytical jacobian transformation \[ \ddot{q}_d = J(q)^{-1}(\dot{V}_e - \dot{J}(q,\dot{q})\dot{q}).\]  The values for $\dot{q}_d$ and $q_d$ can then be found through integration.
+Equation \ref{eq:forceEq} relates the measured force to the derivative of the end effector velocity $\dot{\vV}_e$.
+This end effector velocity can then be related to the desired joint velocities using the analytical Jacobian transformation \[ \ddot{\vq}_d = \mJ(\vq)^{-1}(\dot{V}_e - \dot{J}(\vq,\dot{\vq})\dot{\vq}).\]  The values for $\dot{\vq}_d$ and $\vq_d$ can then be found through integration.
 It is not clear the how authors explicitly decouple the human dynamics from the robot in this case.
@@
 \begin{equation}
-\tau = B(q)(\ddot{q}_d + K_v \dot{\tilde{q}} + K_p \tilde{q}) + C(q,\dot{q})\dot{q} + G(q) -J^T(q) F_s -J^T(q)F_a. \notag
+{\bf \tau} = \mB(\vq)(\ddot{\vq}_d + K_v \dot{\tilde{\vq}} + K_p \tilde{\vq}) + \mC(\vq,\dot{\vq})\dot{\vq} + \mG(\vq) -\mJ^T(\vq) \vF_s -\mJ^T(\vq)\vF_a. \notag
 \end{equation}
-The main difference in this case is that the controller here explicitly compensates for the measured human-robot interaction force $F_s$, and the controller includes the un-sensed forces $F_a$ which arise from the object being manipulated.
-The object interaction forces thus need to be estimated, which was the first fundamental assumption on which the control scheme for the Body Extender relied.
+The main difference in this case is that the controller here explicitly compensates for the measured human-robot interaction force $\vF_s$, and the controller includes the un-sensed forces $\vF_a$ which arise from the object being manipulated.
+The object interaction forces thus need to be estimated.
+The ability to estimate these forces is the first fundamental assumption on which the Body Extender's control scheme relies.
 Because the point at which the manipulator arm and object are connected is assumed to be well known, and the arm is instrumented with inertial sensors, estimating the interaction forces of the manipulated object is be assumed to be equivalent to estimating its inertial properties.
@@
 In terms of how the Body Extender system is designed and how it is controlled, the suit may be limited.
-Clearly, for the task of handling heavy objects a fully actuated suit may be the best option, but generally speaking this seems almost like an over engineered solution.
+Clearly, for the task of handling heavy objects a fully actuated suit may be the best option, but, generally speaking, this seems almost like an over engineered solution.
 The Body Extender is a heavy, overly clunky suit.
@@
 In terms of the control of the overall suit, it was difficult to find any control papers about the Body Extender that covered anything other than upper-body single-arm control of a manipulated object.
-The decoupling strategy discussed above, decoupling the individual limbs from each other, seems somewhat feasible, but is not the best way to approach coordinated limb control.
+The strategy of decoupling the individual limbs from each other, seems somewhat feasible, but is not the best way to approach coordinated limb control.
 It was difficult to find literature documenting the specifics of lower-limb control for the Body Extender.
@@
 138-143.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-28 exos/exosText_ihmc.tex

@@
 \section{IHMC Mobility Assist Exoskeleton (MAE)}
-Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the literature for this device does include a ``performance enhancement" control mode and as such is within the scope of this document.
-The main design principle highlighted in the IHMC MAE hardware is a recurring theme throughout the exoskeleton literature, which is that directly measuring user intent can be difficult for a number of reasons.
-This issue is resolved by using a novel control and feedback system, which in the IHMC MAE is primarily based on the use of a novel actuator design.
+Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the device is also provides a ``performance enhancement" control mode, which is within the scope of this document.
+The main design principle highlighted in the IHMC MAE hardware (a recurring theme throughout exoskeleton literature) is that directly measuring user intent can be difficult for a number of reasons.
+The IHMC designers resolve the issue using a new control and feedback system, which for the IHMC MAE, is primarily based on a unique actuation scheme.
 
@@
 
- Actuator designed around data collected from clinical gait analysis.
-Specifically, the actuators at the hip and knee were designed to be able to provide 40Nm of peak torque during the stance phase of the walking gat cycle.
-The rotary SEA was composed of a Moog BN34-25EU-02 brushless DC motored with the 1:100 harmonic drive, an SHD-20.
+The IHMC actuators are designed based on data collected from clinical gait analysis.
+Specifically, the actuators at the hip and knee were designed to be able to provide 40Nm of peak torque during the stance phase of the walking gait cycle.
+The rotary SEA is composed of a Moog BN34-25EU-02 brushless DC motor with a 1:100 harmonic drive (SHD-20).
 
-The spring mechanism which sits between the gearbox and joint output in the IHMC MAE design is shown in Figure \ref{fig:IHMCSEA}.
-The mechanism shown uses linear die springs and an angular encoder to measure the torque in the out of the joint.
+The spring mechanism, which sits between the gearbox and joint output, is shown in Figure \ref{fig:IHMCSEA}.
+The mechanism uses linear die springs and an angular encoder to measure the joint torque output.
 The choice of linear die spring was based on the spring's predictable force to displacement function as well as favorable energy storage to weight characteristics.
@@
 
-Once assembled, the overall actuator had a  {\bf torque limit of 80 Nm}, with a {\bf velocity limit of 6.8 rad/s}.
-As noted, the {\bf bandwidth of the actuator} is its main drawback, which limits the torque control to approximately {\bf 10 Hz at amplitudes greater than 15 Nm} and can track up to {\bf 30Hz for lower magnitude forces}.
+Fully assembled, the actuator has a  {\bf torque limit of 80 Nm}, with a {\bf velocity limit of 6.8 rad/s}.
+As noted, the {\bf bandwidth of the actuator} is its primary drawback, which limits the torque control to approximately {\bf 10 Hz at amplitudes greater than 15 Nm} and up to {\bf 30Hz for lower magnitude forces}.
 
@@
 
- The IHMC MAE is a lower extremity exoskeleton with a total of ten degrees of freedom (DOF), with five each corresponding to the two legs of the system.
-Of the ten DOFs only six total are actively actuated, the other four are passive.
+ The IHMC MAE lower extremity exoskeleton includes a total of ten degrees of freedom (DOF), six of which are active and four are passive.
+Five DOF correspond to the two legs of the system.
 Each of the powered joints is powered using the SEA joint discussed in the previous subsection.
@@
 The hip yaw DOF consists of a curved spring-loaded roller bearing whose center of rotation is approximately at the user's hip joint.
-
 The knee flexion/extension joint is connected to the hip flexion/extension joint using telescoping tubes and is also actuated.
-
- The final DOF in the IHMC design is the dorsal flexion ankle joint.  This joint is spring loaded such that it provides unidirectional torque to provide toe clearance during walking.
+The final DOF is the dorsal flexion at the ankle joint.
+This joint is spring loaded such that it provides unidirectional torque for toe clearance during walking.
 
  In its current configuration, the IHMC exoskeleton in controlled by an off-board computer and is powered using a tether.
-
 The suit {\bf includes position and force sensors in the actuators in addition to two foot switches per foot}.
-The foot switches are used to detect whether the system in in single or double support phase, as well as providing information on how the overall system load is approximately distributed on the stance legs.
+The foot switches are used to detect whether the system is in single or double support phase.
+It also provides information used to estimate the load distribution to the stance legs.
 
@@
 
- The noted control approaches of the IHMC MAE are both relatively straight forward implementations of low-level torque based control strategies.
-The first control scheme takes full advantage of the SEA actuators ability to directly measure output torque and subsequently to perform closed-loop torque control, which is pictured in \ref{fig:IHMCTOR}.
+ Both IHMC MAE control approaches noted are relatively straight forward implementations of low-level torque control.
+The first control scheme takes full advantage of the SEA actuators' ability to directly measure output torque and subsequently to perform closed-loop torque control, (see \ref{fig:IHMCTOR}).
   \begin{figure}[thpb]
@@
  \end{figure}
- The pictured control scheme takes in a desired torque, compares the desired value to the output torque measured at each individual actuator, then uses this value as well as its derivative to compute an actuator command signal, which in this case is a current.
-The largest question in this controller, as presented in the literature, is how to define the desired torque term $\tau_\text{des}$.
+ The controller takes in a desired torque, compares the desired value to the output torque measured at each individual actuator, then uses this value as well as its derivative to compute an actuator command signal, which in this case is a current.
+The largest question in this controller, as presented in literature, is how to define the desired torque term $\tau_\text{des}$.
 
@@
  \end{figure}
- The controller in Figure \ref{fig:IHMCPOS} shows an outer-loop position-based PD controller which takes in desired joint trajectories and compares them to measured joint positions.
-The values are used to generate desired torque values which are fed into an inner closed-loop torque control loop.
-This control scheme is a good low-level choice if high-impedance very accurate motions of the suit are desired.
-Similarly to the torque-only control loop above, the main question in this control scheme is how the desired joint trajectories are defined.
+ The controller in Figure \ref{fig:IHMCPOS} shows an outer-loop position-based PD controller that takes in desired joint trajectories and compares them to measured joint positions.
+These are used to generate desired torque values which are fed into an inner closed-loop torque control loop.
+This control scheme is a good low-level choice if high-impedance.
+accurate suit motion is desired.
+Similarly to the torque-only control loop above, the main question in this control scheme is how to obtain the desired joint trajectories.
 
- One possibility for defining either the desired torques in the first control scheme or the desired joint trajectories in the second is to use clinical gait analysis data.
-In this approach the controllers highlighted in this subsection would act to play back nominal joint trajectory data.
-The foot sensors in the system would be used to detect phase transitions and to dynamically adjust the prerecorded data to the user in realtime.
+ One possibility for defining either the desired torques / joint trajectories in these control schemes is to use clinical gait analysis data.
+With this approach, the highlighted controllers would effectively play back nominal joint trajectory data.
+The foot sensors in the system would detect phase transitions, allowing the controllers to dynamically adjust the prerecorded data to the user in realtime.
 
@@
 
- The actuator design and use of series-elastic actuation is in general a good idea for many applications, but may be a problem for the highly dynamic maneuvers required by operators in the TALOS project.
+ The design and use of series-elastic actuators is, in general, a good idea for many applications, but may be a problem for the highly dynamic maneuvers required by operators in the TALOS project.
 In a large sense, bandwidth is everything in these suits due to the fact that behavior prediction is very complicated.
@@
 
- The other issue with the IHMC work is that their control approach only addresses the lowest-level of control, and does not in general specify exactly how to generalize the process of of generating reference trajectories, whether they be torque or joint trajectories.
-That said, other researchers do explicitly provided general methods for generating desired trajectories in what amounts to a mid-level of a hierarchical control scheme.
-Thus combining other approaches with the low-level highly accurate torque control provided by the novel SEAs presented in this section could offer a potentially good solution at the actuator level.
+ The other issue with the IHMC is that it only addresses low-level control.
+The designers have not specified how to generalize the process of generating torque / joint reference trajectories.
+That said, other researchers do explicitly provide methods for generating desired trajectories in what amounts to a mid-level of a hierarchical control scheme.
+Thus combining other approaches with the low-level, highly accuracy torque control provided by the novel SEAs presented in this section could offer a potentially good solution at the actuator level.
 
@@
 Neuhaus, "Development of the IHMC Mobility Assist Exoskeleton,"  International Conference on Robotics and Automation, 2009.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-28 exos/hal.tex
cga-fixes
Added bibliographies by section to survey.tex. Use biber to build survey.tex

@@
-\subsection{HAL}
+\section{HAL}
 \label{exo:hal}
+\begin{refsection}[exos/hal.bib]
 
@@
 Though HAL-5 is a full body exoskeleton, only hip, knee joints, and ankle are actuated in the sagittal plane.
-The exoskeleton powers DC motors with harmonic drives at hip, knee, ankle joints (in some models the ankle joints act as passive springs).
-HAL weighs 23kg, includes an on-board AC100V battery as the power source, which is designed to support the power requirements of maximum velocity human walking and standing torque requirements.
+The exoskeleton uses DC motors with harmonic drives at hip, knee, ankle joints (in some models the ankle joints act as passive springs).
+HAL weighs 23kg and includes an on-board AC100V battery as the power source, which is designed to support maximum velocity human walking and standing torque requirements.
 The battery allows for 160 min of continuous operation and enables the exoskeleton to lift up to 70kg \cite{HALassist2011}.
 
-Human operators attach to HAL at the waist by a belt, and at the calf and thigh using harness.
+Human operators attach to HAL at the waist with a belt, and at the calf and thigh using harnesses.
 HAL's frame does not transfer load to the ground.
 Instead, HAL adjusts hip, knee and ankle torques to amplify its operator's torque.
-The exoskelton's sensory system includes bioelectric sensing (including sEMG), angular sensors, acceleration sensors, and center of pressure / center of gravity (COP/COG) sensors.
+The exoskelton's sensory system includes bioelectric sensing (including \textbf{sEMG}), \textbf{angular sensors}, \textbf{acceleration sensors}, and \textbf{center of pressure / center of gravity} (COP/COG) sensors.
 COP/COG sensing is provided through shoes with ground reaction force sensors.
@@
 
-HAL uses two types of control systems depending on application domain.
+HAL has two types of control systems designed for different application domains.
 For gait assistance and rehabilitation, HAL uses an autonomous control system that carries the user through predefined gait trajectories by controlling knee and hip joints (the ankles behave as passive springs).
@@
 The authors note the sEMG model requires significant calibration effort.
-%
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=6in]{exos/figs/hal_viscoelastic_control.png}
-\end{figure}
-%
-%
 Viscoelastic torques are computed according to
@@
 
+%
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=6in]{exos/figs/hal_viscoelastic_control.png}
+\end{figure}
+%
+%
+
 While experiments in \cite{HALmuscleImped2005} only consider leg swing-up and swing-down, a similar control approach in \cite{HALvTorqueImp2002} switches the dynamics based controller to compensate for swing and stance phases in walking gait.
-In this case, the gait transitions are detected by thresholding foot sensors and required angular velocity (and acceleration) data is determined by numerically differentiating angular encoders.
+In this case, the gait transitions are detected by thresholding foot sensors and required angular velocity (and acceleration) data are determined by numerically differentiating angular encoders.
 
@@
 Compared to sensitivity amplification methods, HAL's use of EMG data allows it to estimate intention without amplifying external disturbances.
-This is ideal for combat scenarios.
-However, sEMG data is extremely difficult to work with due to high filtering and calibration requirements.
+While direct sensing of user intention is ideal for combat scenarios, sEMG data is extremely difficult to work with due to high filtering and calibration requirements.
 A hybrid strategy which uses sensitivity amplification and sEMG data (possibly thresholded) to filter user intention could prove effective.
@@
 
-%\bibliographystyle{plain}
-%\bibliography{exos/hal}
+\nocite{*}
+\printbibliography[heading=subbibliography]
 
-The figures in this section were obtained from \cite{HALassist2011,HALmuscleImped2005}.  Materials presented are based on the references above.
+The figures in this section were obtained from \cite{HALmuscleImped2005,HALassist2011}.  Materials presented are based on the references above.
+
+\end{refsection}

### 2015-09-28 exos/xor.tex
cga-fixes
Added bibliographies by section to survey.tex. Use biber to build survey.tex

@@
-\subsection{XoR}
+\section{XoR}
 \label{exo:XoR}
+\begin{refsection}[exos/xor.bib]
 
@@
 
-The XoR is a prototype, light-weight lower-body exoskeleton that uses a hybrid pneumatic-electric drive system that \textbf{reduces weight} while providing \textbf{precise torque control}, \textbf{backdrivability}, and a \textbf{desirable force / velocity} profile.
+The XoR is a prototype light-weight, lower-body exoskeleton that uses a hybrid pneumatic-electric drive system that \textbf{reduces weight} while providing \textbf{precise torque control}, \textbf{backdrivability}, and a \textbf{desirable force / velocity} profile.
 The exoskeleton is designed to serve in rehabilitation settings to augment operators' strength and assist with postural control for persons with disabilities.
@@
 The XoR weighs 30kg and includes 10 DOF with 6 active joints (flexion / extension of hip, knee, and ankles) and 6 passive (hip abduction/adduction joints, hip rotation, and ankle adduction/abduction).
-The active joints are powered by hybrid actuators comprised of an \emph{air muscle} and an electric motor.
+The active joints are powered by \textbf{hybrid actuators} comprised of an \textbf{air muscle} and an \textbf{electric motor}.
 The hybrid actuation scheme uses a unilateral air muscle layout to compensate for gravity and bilateral electric motors with relatively small gear ratio (57.5) to serve as the dynamic compensator.
@@
 For instance, at 30\% contraction, the force produced by FESTO rubber air muscles vanishes.
-To address the issue, XoR is strategically places air muscles to generate forces in desired configurations.
+To address the issue, XoR's air muscles are strategically placed to generate forces in desired configurations.
 
@@
 The electric actuator is a geared, 200W brushless DC motor (Maxon EC powermax 30) rated at 4.7 A, which transmits power to joints through a belt and pulley system with a 57.5 gear ratio.
-the motor's 0.12 Nm output torque provides up to 34.5 Nm (for short duration) at every joint and is backdrivable.
-The design yields a relatively high range of motion (up to 120 degrees) with a load capacity 150 Nm.
+The motor is backdrivable with 0.12 Nm output torque and up to 34.5 Nm (for short duration) at every joint.
+The design yields a relatively high range of motion (up to 120 degrees) with a 150 Nm load capacity.
 However, experiments reveal the need for bi-directional pneumatic actuation at hip joints.
 
-For sensing, XoR is equipped with rotary encoders at joints, an IMU in the backpack, load cells in the feet, and is networked to servers capable of providing EMG, NIRS, and EEG data.
+For sensing, XoR is equipped with \textbf{rotary encoders} at joints, an \textbf{IMU} in the backpack, \textbf{load cells} in the feet, and is networked to servers capable of providing \textbf{EMG}, NIRS, and EEG data.
 A control PC running at 1 KHz, pulse counter, amplifiers, Digital IO are external and so not included in the weight of the unit.
-Human operators are outfitted with a goniometer.
+Human operators are outfitted with a \textbf{goniometer}.
 
@@
 The controller rectifies and low pass filters (10 Hz cut-off) data measured in the pilot's quadriceps femoris, tensor fasciae latate, gluteus medius, and tibialis anterior.
-It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta} \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1 \cdot EMG_n)$, using a linear prediction model, 
+It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta}, \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1 \dots EMG_n)$, using a linear prediction model, 
 \[{\bf x}(k+1) = \mA{\bf x}(k)+\mB{\bf u}(k),\].  
@@
 The XoR prototype uses a promising hybrid actuation scheme to achieve both fast peak torque response and higher sustained torque while remaining lightweight.
-They track human motion with feedback from a goniometer and feedforward control provided by EMG data.
+It tracks human motion with feedback from a goniometer and feedforward control provided by EMG data.
 These human intention / sensing modalities are both difficult to implement due to noise and calibration issues.
@@
 
-% \bibliographystyle{plain}
-% \bibliography{exos/xor}
+\nocite{*}
+\printbibliography[heading=subbibliography]
 
 The figures in this section were obtained from \cite{xorDesign2011,XoRkinemExtraction2012}. Materials presented are based on the references above.
+
+\end{refsection}

### 2015-09-29 exos/bleex.tex
more edits

@@
 
-keywords: sensitivity amplification; strength augmentation.\\
+% keywords: sensitivity amplification; strength augmentation.\\
 
@@
 
-The BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as $d$ (assuming no outside disturbances).
+The BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as ${\bf d}$ (assuming no outside disturbances).
 Neglecting gravity, the exokeleton angular velocity is modeled as
-\[v = G r + S d ,\] 
-where $G$ is the transfer function from actuator inputs ($G$ is the exoskeleton's dynamics), $r$ is the actuator input, and $S$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
+\[\vv = \mG \vr + \mS {\bf d} ,\] 
+where $\mG$ is the transfer function from actuator inputs ($\mG$ is the exoskeleton's dynamics), $\vr$ is the actuator input, and $\mS$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
 
@@
 
-The goal is to maximize sensitivity to $d$ \emph{without direct measurement.}  Sensitivity amplification accomplishes this by creating a feedback loop from a controller, $C$, acting only on exoskeleton variables.
+The goal is to maximize sensitivity to ${\bf d}$ \emph{without direct measurement.}  Sensitivity amplification accomplishes this by creating a feedback loop from a controller, $\mC$, acting only on exoskeleton variables.
 A new sensitivity equation,
-\[S_{new} = \frac{v}{d} = \frac{S}{1 + G C} ,\]
+\[\mS_{new} = \frac{\vv}{{\bf d}} = \frac{\mS}{1 + \mG \mC} ,\]
 is maximized by applying positive feedback.
-To achieve a large sensitivity, BLEEX uses $C = (1-\alpha^{-1})G^{-1}$ so that $\alpha$ provides a direct (scalar) amplification factor.
-A low-pass filter is often added to the $C$ term in order to damp out high frequency dynamics of the exoskeleton, which are not captured in these models.
-Note that the controller depends on an inverse dynamic model of the exoskeleton, $G^{-1}$.
+To achieve a large sensitivity, BLEEX uses $\mC = (1-\alpha^{-1})\mG^{-1}$ so that $\alpha$ provides a direct (scalar) amplification factor.
+A low-pass filter is often added to the $\mC$ term in order to damp out high frequency dynamics of the exoskeleton, which are not captured in these models.
+Note that the controller depends on an inverse dynamic model of the exoskeleton, $\mG^{-1}$.
 Since the model is hybrid, these dynamics switch according to gait phases (single support, double support, stance).
@@
 \end{equation}
-where ${\bf T}$ is a vector of actuator torques, the BLEEX control torque would be
+where ${\bf T}$ is a vector of actuator torques and $\mM({\bf\theta})$, $\mC({\bf\theta},\dot{{\bf\theta}})$, and $\mV({\bf\theta})$ are the mass, Coriolis, and gravity terms (respectively), the BLEEX control torque would be
 \begin{equation}
@@
 \end{refsection}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-29 exos/exosText_MIT.tex
more edits

@@
 Finally, in leg retraction a torque is applied to help with foot placement and weight acceptance.
-A diagram that illustrates how the sensors in the shank and hip joint are used to switch between these various phases is show in Figure \ref{fig:hipControl}
+A diagram that illustrates how the sensors in the shank and hip joint are used to switch between these various phases is show in Figure \ref{fig:hipControl}.
  \begin{figure}[thpb]
@@
 The knee therefore needs to be actively demagnetized at full extension during the last phase of stance to allow it to swing freely during the subsequent swing phase.
-The state machine diagram that illustrates how the various phase transitions for the knee joint were implemented using the exoskeleton's on-board sensors is in Figure \ref{fig:kneeControl}
+The state machine diagram that illustrates how the various phase transitions for the knee joint were implemented using the exoskeleton's on-board sensors is in Figure \ref{fig:kneeControl}.
   \begin{figure}[thpb]
@@
 
- \subsection{Opinion}
+ \subsection{Assessment and Recommendations}
 

### 2015-09-29 exos/exosText_ihmc.tex
more edits

@@
  Both IHMC MAE control approaches noted are relatively straight forward implementations of low-level torque control.
-The first control scheme takes full advantage of the SEA actuators' ability to directly measure output torque and subsequently to perform closed-loop torque control, (see \ref{fig:IHMCTOR}).
+The first control scheme takes full advantage of the SEA actuators' ability to directly measure output torque and subsequently to perform closed-loop torque control, (see Fig~\ref{fig:IHMCTOR}).
   \begin{figure}[thpb]

### 2015-09-29 exos/exosText_roboKnee.tex
more edits

@@
 
-The RoboKnee device was built as a prototype system to enhance human strength, endurance and speed.
-The main design objectoves in the RoboKnee system were to enhance human performance using a low-impedance mechanical system that incorporates a natural user interface, long life, and is comfortable to wear.
-The main control objective of the suit is to provide a ``get out of the way" control scheme, wear the system inherently interacts with the user through a low-impedance interface.
-
-The design of a linear series elastic actuator and a novel control method for controlling a single powered joint at the knee of the operator are presented in this section.
+The RoboKnee is a prototype system designed to enhance human strength, endurance and speed.
+To achieve these goals, RoboKnee uses a low-impedance mechanical system that incorporates a natural user interface, long life, and is comfortable to wear.
+The suit's controller implements a ``get out of the way'' control scheme, where the system interacts with the user through a low-impedance interface.
+This section focuses on RoboKnee's linear series elastic actuator design and the novel control method the device employs to coordinate its single powered knee joint.
 
@@
 
-The RoboKnee actuator was designed to specifically provide low impedance and high force feedback fidelity.
-Series elastic actuators are a more robust option for measuring output forces and torques than using notoriously delicate and expensive load cells.
+The RoboKnee actuator is specifically designed to provide low impedance and high force feedback fidelity.
+The device uses a series elastic actuator to obtain robust output force measurements.
+The series elastic actuation scheme allows RoboKnee to avoid load cells, which are both delicate and expensive.
 
-The actuator used in the RoboKnee device is shown in Figure \ref{fig:roboAct}.
-The actuator consists of a drive train subassembly and an output carriage subassembly.
-The two subassemblies are coupled through die compression springs.
-A servo motor directly drives a ball screw assembly to drive a ball nut flange which pushes on a spring retaining plate.
-Secondary spring retaining plate is coupled to an output assembly.
-Measuring the deflection in the die springs makes it possible to measure the force applied to the output of the linear SEA.
+The actuator used in the RoboKnee device (see Figure \ref{fig:roboAct}) consists of a drive train and output carriage.
+These two sub-assemblies are coupled through die compression springs.
+A servo motor directly connected to a ball screw assembly drives a ball nut flange that pushes on a spring retaining plate.
+A secondary spring retaining plate is coupled to an output assembly.
+By measuring the deflection in the die springs, RoboKnee is able to obtain the force applied to the output of the linear SEA.
 \begin{figure}[thpb]
@@
 The {\bf maximum speed of the actuator is 28 cm/s}.
-The actuators {\bf maximum continuous force is 565N, with a peak maximum force of 1330 N}, with {\bf maximum continuous power of 164 W, and maximum peak power of 634 W}.
+The actuator's {\bf maximum continuous force is 565N and its maximum peak force is 1,330 N}, with a {\bf maximum continuous power of output of 164 W, and maximum peak power of 634 W}.
 
-The force control bandwidth of the RoboKnee actuator is dependent on the magnitude of the force, as significant time delays can occur for larger forces due to the fact that there is an inherent time delay due to the time it takes for the spring to compress.
+The force control bandwidth of the RoboKnee actuator is dependent on the magnitude of the force.
+Larger forces can induce significant time delays due to the time it takes for the spring to compress.
 Thus, the {\bf small force control bandwidth of the system is 35 Hz} and the {\bf high force bandwidth is around 7.5 Hz}.
@@
 
- The RoboKnee mechanism consists of a single actuated degree of freedom at the flexion/extension joint of the operator.
-Note that the mechanism itslef does not provide a pathway for transferring weight from a load directly to the ground.
-The mechanism transfers loads directly to the users musculoskeletal structure, and focuses primarily on augmenting the torque supplied to the users knee joint.
+ The RoboKnee mechanism consists of a single actuated degree of freedom powering the flexion/extension of the operator's knee joint.
+Note that the mechanism itself does not provide a pathway for transferring weight from a load directly to the ground.
+The mechanism transfers loads directly to the users musculoskeletal structure, and focuses primarily on augmenting user knee joint torques.
 
- The RoboKnee mechanism itself is composed of an off-the-shelf knee brace modified with structural pieces to extend the brace and provide actuator attachment points.
-The RoboKnee mechanism is shown in \ref{fig:roboKnee}.
+ The RoboKnee device is composed of an off-the-shelf knee brace modified with structural pieces to extend the brace and provide actuator attachment points (see Figure~\ref{fig:roboKnee}).
  \begin{figure}[thpb]
@@
  \end{figure}
- There exact length between attachment points of the actuator is not listed in the literature, but it appears that the attachment points are at minimum the stroke displacement of the actuator  away from each other, so a {\bf minimum moment arm of 30.5 cm} is a reasonable approximation.
+ The exact length between actuator attachment points is not listed in the literature, but it appears as though the attachment points are at a minimum of one actuator stroke displacement away from each other.
+Thus, we assume a {\bf minimum moment arm of 30.5 cm} is a reasonable approximation.
 
- The RoboKnee mechanism uses a linear potentiometer spanning the spring retaining plates to measure displacement in the springs and subsequently to measure force at the actuator output.
-The forces between the users foot and the ground are also measured using single axis load cells.
+ RoboKnee uses a \textbf{linear potentiometer} spanning the spring retaining plates to measure displacement in the springs and subsequently to measure force at the actuator output.
+The forces between the users foot and the ground are also measured using \textbf{single axis load cells}.
 Potentiometers which measure the knee joint angle are also employed in this design.
@@
 
- The control scheme for the RoboKnee device is a hierarchical approach, which uses a straightforward mid-level force generation scheme coupled with a low-level closed-loop force-based control loop.
-The low-level, or rather joint level control is a PD control loop.
+ RoboKnee employs a hierarchical control strategy, which uses a straightforward mid-level force generation scheme coupled with a low-level closed-loop force-based control loop.
+The low-level (joint level) control is a PD control loop.
 
- The mid-level controller of the RoboKnee uses several simplifying assumptions to perform ``force amplification" to offload a percentage of the torque required by the user to actuate the system's knee joint.
-This is achieved through a positive force feedback amplification which is based on measuring the ground reaction force and using this value to calculate the torque that would be required to produce this force in a static situation, namely, \[ \tau = R \times F,\] where $R$ is the vector from the ground reaction force to the users knee and $F$ is the ground reaction force.
-Note that in the RoboKnee implementation this value can only be estimated due to the fact that the ankle joint angle and the ground reaction force is assumed to be purely vertical (which is a result of the single axis force sensors on the users feet).
+ The RoboKnee's mid-level controller uses several simplifying assumptions to perform ``force amplification'' to offload a percentage of the torque required by the user to actuate the system's knee joint.
+This is achieved through positive force feedback amplification.
+The approach uses the measured ground reaction force to calculate the torque that would be required to produce this force in a static situation, \[ {\bf \tau} = \vR \times \vF.\]  The $\vR$ term represents the vector from the ground reaction force to the user's knee and $\vF$ is the ground reaction force.
+Note that in RoboKnee's implementation, $\vF$ can only be estimated due to the fact that the ankle joint angle and the ground reaction force is assumed to be purely vertical (a limitation induced by the single axis force sensors on the user's feet).
 
- Once the estimated joint torque $\tau$ is calculated, an amplification factor is used in the RoboKnee control scheme to decide how much of the required knee torque will be provided by the control system.
-For example, if the amplification factor were chosen to be one the actuator would completely compensate for the sensed ground reaction forces, and if it were set to zero to exoskeleton would provide no force.
+ Once the estimated joint torque, ${\bf \tau}$, is calculated, an amplification factor specifies how much of the required knee torque will be provided by the control system.
+For example, with unity amplification factor, the actuator would completely compensate for the sensed ground reaction forces.
+Similarly, with zero amplification, the exoskeleton provides no force.
 
- \subsection{Opinion}
+ \subsection{Assessment and Recommendations}
 
@@
 
- The overall design of the RoboKnee mechanism is most likely not the greatest, as it significantly increases the effective volume of the operator-system leg, but as a prototype this concept offers interesting benefits.
+ The overall design of the RoboKnee mechanism is most likely not the greatest, as it significantly increases the effective volume of the operator-system leg, but as a prototype, this concept offers interesting benefits.
 
@@
 2430-2435.
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-29 survey.tex
more edits

@@
 \newcommand{\vT}{{\bf T}}
+\newcommand{\vR}{{\bf R}}
 
@@
 
-I copied figure captions and numbers. Matt did not.  Coordinate?
+See if the Hydraulic Lower Extremity Exo is sensitivity amp and if so change opinion so we don't favor the approach.
 
@@
 
-I don't point out what the terms are in the dynamics equations.
+Discussion blurb -- all these exos either measure user intention directly and have sensor noise and calibration challenges, or measure at the suit and perform amplification.
+The amplification strategy is highly susceptible to outside disturbances and therefore not appropriate in combat situations.
+Combine the two with filtering or directly address the sensing challenge.
 
-See if the Hydraulic Lower Extremity Exo is sensitivity amp and if so change opinion so we don't favor the approach.
+Discussion blurb (pertains to IHMC and probably others) -- issues with high impedance torque control in practice.
 
-Discuss IHMC -- issues with high impedance torque control in practice.
-
-Talk a lot about drawbacks of SEAs in IHMC.  Are these included in MIT?
+Talk a lot about drawbacks of SEAs in IHMC.  Are these included in MIT / RoboKnee?  (low control bandwidth is mentioned in RoboKnee).  
+We should have a blurb about the strengths/weakness of SEAs and possible options -- stiffer springs for force measurement and spring emulation for compliant control?

### 2015-09-29 exos/xor.tex
more edits
working on conclusions and recommendations

@@
 
-keywords: model-based control; strength augmentation; hybrid actuation; pneumatic actuation; electric motors;\\
+% keywords: model-based control; strength augmentation; hybrid actuation; pneumatic actuation; electric motors;\\
 
@@
 The controller rectifies and low pass filters (10 Hz cut-off) data measured in the pilot's quadriceps femoris, tensor fasciae latate, gluteus medius, and tibialis anterior.
-It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta}, \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1 \dots EMG_n)$, using a linear prediction model, 
+It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta}, \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1, \dots , EMG_n)$, using a linear prediction model, 
 \[{\bf x}(k+1) = \mA{\bf x}(k)+\mB{\bf u}(k),\].  

### 2015-09-29 survey.tex
working on conclusions and recommendations

@@
 
-Conclusions and Recommendations to be written.
+All the (strength augmentation) exoskeletons we reviewed can be categorized into one of two general control categories.
+The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
+The first strategy is difficult to implement using standard sensing modalities.
+For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
+It would also be extremely difficult to keep these sensors in place to obtain accurate readings in a combat scenario.
+Similarly, many avoid the use of load cells for measurement of the human-exoskeleton interface fores to avoid outfitting the complicated sensor network required to accurate obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
+
+While sensitivity amplification strategies avoid these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
+As such, they end up amplifying noise and the user must react to stabilize the system.
+This is also unacceptable for a combat exoskeleton.
 

### 2015-09-29 exos/bleex.tex
Draft of conclusion.

@@
 Joint angles, torque, and power requirements are determined from human motion analysis based on a 75-kg human walking on flat ground at roughly 1.3 m/s (the average military male's maximum reported joint limits are also used to derive joint range of motion targets).
-During design, joint motion was intended to be slightly less than the maximum human range of motion for safety; however, some joint ranges had to be reduced to avoid singularities.
+During design, joint motion was intended to be slightly less than the maximum human range of motion for safety; however, some joint ranges had to be reduced to avoid (mechanical) singularities.
 % Ideally, joint motion would provide the maximum human range of motion for safety
@@
 However, the approach has several significant disadvantages including heavy reliance on an accurate exoskeleton model and the fact that sensitivity amplification will amplify disturbances.
-Note that in combat situations this latter point is critical as a sensitivity amplification will amplify forces acting on the suit and the operator will have to re-act to compensate.
+Note that in highly dynamic, contact-rich environments this latter point is critical, as a sensitivity amplification will amplify forces acting on the suit and the operator will have to re-act to compensate.
 The only way to compensate for or resist external forces in this setting is to filter them out, which would be extremely challenging unless additional sensory equipment is provided.
-As one possibility, a sensitivity optimization approach may be paired with sEMG data to estimate user intent and provide such a filter.
+As one possibility, a sensitivity optimization approach may be paired with sEMG (surface electromyography) data to estimate user intent and provide such a filter.
 Note this type of implementation would present its own challenges considering limitations in sensing and calibration of current EMG systems.
+Section~\ref{survey:recommend} discusses new flexible sensory arrays and other technologies and strategies to capture / filter based on user intent.
 

### 2015-09-29 survey.tex
Draft of conclusion.

--- 2015-09-29 survey.tex
@@
 \section{Conclusions and Recommendations}
+\label{survey:recommend}
 
-Conclusions and Recommendations to be written.
+All the (strength augmentation) exoskeletons we reviewed can be categorized into two general control categories.
+The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
+The first strategy is difficult to implement using standard sensing modalities.
+For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
+It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or combat scenarios.
+Similarly, many avoid the use of load cells for measurement of human-exoskeleton interface forces and the complicated sensor network required to accurately obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
 
-% \bibliographystyle{plain}
-% \bibliography{exo}
+While sensitivity amplification strategies circumvent these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
+As such, they end up amplifying noise and the user must react to stabilize the system.
+This is also unacceptable for highly dynamic exoskeletons operating in contact-rich environments.
+
+We emphasize that user intent is critical, and current sensing challenges must be addressed in order to develop dynamic exoskeletons capable of serving in combat scenarios.
+As a stop-gap, it may be possible to combine low information density EMG (thresholded on/off muscle activation signals) and human force/torque or angular measurements to filter user intent.
+Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.
+As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
+Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
+Dense sensing arrays add redundancy and can improve measurement reliability.
+We recommend further study and the development of more advanced sensory systems to address the challenges inherent in dynamic exoskeleton control.
+
+In addition to capturing user intent, it is important to develop accurate models of the human/exoskeleton system.
+Feedback strategies, especially those based on force/torque control loops and series elastic actuation, are limited in bandwidth.
+Highly dynamic behaviors require feedforward models to reduce reliance on feedback and to allow lower-impedance feedback, which reduces resistive forces acting on the user.
+
+%% OTHER POINTS:
+% Model-based approaches are heavily reliant on accurate models 
+% Underactuated -- feedback linearization typically assumes fully actuated systems.
 
@@
 
-Coordinate notes about sEMG and user intention -- BLEEX;
-
-Discussion blurb -- all these exos either measure user intention directly and have sensor noise and calibration challenges, or measure at the suit and perform amplification.
-The amplification strategy is highly susceptible to outside disturbances and therefore not appropriate in combat situations.
-Combine the two with filtering or directly address the sensing challenge.
-
-Discussion blurb (pertains to IHMC and probably others) -- issues with high impedance torque control in practice.
-
 Talk a lot about drawbacks of SEAs in IHMC.  Are these included in MIT / RoboKnee?  (low control bandwidth is mentioned in RoboKnee).  

### 2015-09-29 exos/exosText_MIT.tex
revision of abstract, summary, and refs

@@
 \section{MIT Exoskeleton}
+\label{mit}
+\begin{refsection}[exos/mit.bib]
 
@@
 \includegraphics[width=3.in]{exos/figs/MIT/mitSEA}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Linear series elastic actuator.}
+%  \vspace{-0.2in}
  \label{fig:mitSEA}   
@@
 \includegraphics[width=3.in]{exos/figs/MIT/MITsuit}
-  \caption{}
+  \caption{ Sensors placed on the exoskeleton leg \cite{mit_control_2006}.}
 %  \vspace{-0.2in}
@@
  The interface between the suit and operator consists of shoulder straps, a compliant waist belt, thigh cuffs, and a shoe connection.
+The MIT exoskeleton employs a custom-built sensor that measures the interaction force between the exoskeleton and the operator?s thigh.
+The sensor consisted of a spring pack, where the deflection of the passive springs was measured with a spring-loaded linear potentiometer.
 
@@
 \includegraphics[width=4.in]{exos/figs/MIT/walkingCycle}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{ Summary of the actuation control of the exoskeleton leg as a function of gait cycle with an actuator at the hip and a damper at the knee.
+The ankle is completely passive but is included for completeness.}
+%  \vspace{-0.2in}
  \label{fig:walkingCycle} 
@@
 \includegraphics[width=3.in]{exos/figs/MIT/hipControl}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{State-machine diagram for the hip controller.}
+%  \vspace{-0.2in}
  \label{fig:hipControl} 
@@
 \includegraphics[width=3.in]{exos/figs/MIT/kneeControl}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{State-machine diagram for the knee controller.}
+%  \vspace{-0.2in}
  \label{fig:kneeControl} 
@@
 
+ The hardware specifications and associated figures in this section were taken from \cite{mit_design_2006}.
+The discussion on control design and and the associated figures therein are respectively based on and taken from \cite{mit_control_2006}.
+
+ \printbibliography[heading=subbibliography]
+
+\end{refsection}
+
 % \end{document}

### 2015-09-29 exos/exosText_bodyExt.tex
more edits
revision of abstract, summary, and refs

@@
 \section{Body Extender}
+\label{bodyExtender}
+\begin{refsection}[exos/bodyExt.bib]
 
@@
 \includegraphics[width=3.in]{exos/figs/bodyExt/bodyExt}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{The Body Extender.}
+%  \vspace{-0.2in}
  \label{fig:bodyExt}   
@@
 \includegraphics[width=3.in]{exos/figs/bodyExt/motionConv}
-  \caption{}
+  \caption{CAD view of the internal components of the actuation module.}
 %  \vspace{-0.2in}
@@
 
-The authors in \cite{} use a decoupling strategy in their controllers which enables the individual limbs in the Body Extender to be somewhat decoupled from each other.
+The authors in \cite{body_control_2012} use a decoupling strategy in their controllers which enables the individual limbs in the Body Extender to be somewhat decoupled from each other.
 They claim that they accomplish this using the following model:
@@
 where $\tilde{\vq} = \vq_d - \vq$.
-The details are unclear in \cite{}, but it appears that the authors either propose using predefined trajectories to define $\ddot{\vq}_d$, $\dot{\vq}_d$, and $\vq_d$, or use the assumption in \eqref{eq:forceEq} to determine these values.
+The details are unclear in \cite{body_control_2012}, but it appears that the authors either propose using predefined trajectories to define $\ddot{\vq}_d$, $\dot{\vq}_d$, and $\vq_d$, or use the assumption in \eqref{eq:forceEq} to determine these values.
 %\begin{equation}
@@
 Because the point at which the manipulator arm and object are connected is assumed to be well known, and the arm is instrumented with inertial sensors, estimating the interaction forces of the manipulated object is be assumed to be equivalent to estimating its inertial properties.
-As outlined in \cite{}, this is accomplished using a standard regressor which uses the input and output of the feedback linearizing controller as the basis for the regression.
-The details of this process are left to \cite{}.
+As outlined in \cite{body_control_2012}, this is accomplished using a standard regressor which uses the input and output of the feedback linearizing controller as the basis for the regression.
+The details of this process are left to \cite{body_control_2012}.
 
-\subsection{Opinion}
+\subsection{Assessment and Recommendations}
 
@@
 
-In terms of the controllers highlighted in \cite{}, there are some good ideas.
+In terms of the controllers highlighted in \cite{body_control_2012}, there are some good ideas.
 For example, using a regressor to estimate unknown external forces may be an idea which is directly applicable to the TALOS project.
@@
 
+ The mechanical specifications as well as figures related to mechanical design for the Body Extender system discussed in this section are based on and taken from \cite{body_design_2011}.
+The discussion of the control system is based on work presented in \cite{body_control_2012}.
+
+\printbibliography[heading=subbibliography]
+
+\end{refsection}
+
 % \end{document}
 
-% S. Marcheschi, F. Salsedo, M. Fontana, and M. Bergamasco, ?Body ex- tender: Whole body exoskeleton for human power augmentation,? in Proc. IEEE Int. Conf. Robotics Automation, 2011, pp. 611?616.
+% S. Marcheschi, F. Salsedo, M. Fontana, and M. Bergamasco, Body Extender: Whole body exoskeleton for human power augmentation, in Proc. IEEE Int. Conf. Robotics Automation, 2011, pp. 611?616.
 

### 2015-09-29 exos/exosText_ihmc.tex
revision of abstract, summary, and refs

@@
 \section{IHMC Mobility Assist Exoskeleton (MAE)}
+\label{ihmc}
+\begin{refsection}[exos/ihmc.bib]
+
 Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the device is also provides a ``performance enhancement" control mode, which is within the scope of this document.
@@
 \includegraphics[width=3.in]{exos/figs/ihmc/seaAssm}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Schematic of the RSEA elastic mechanism design, shown in the zero force position.
+The encoder reads a scale tape on the curved cylindrical surface.
+Note that the actuator housing and bearing have been omitted for clarity.}
+%  \vspace{-0.2in}
  \label{fig:IHMCSEA}   
@@
 \includegraphics[width=3.in]{exos/figs/ihmc/ihmcSys}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Side view of the IHMC Mobility Assist Exoskeleton showing the powered and passive degrees of freedom.}
+%  \vspace{-0.2in}
  \label{fig:IHMCSYS}   
@@
 \includegraphics[width=3.in]{exos/figs/ihmc/torqueCon}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Torque control loop in the actuator control system.}
+%  \vspace{-0.2in}
  \label{fig:IHMCTOR}   
@@
 \includegraphics[width=3.in]{exos/figs/ihmc/positionCon}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Control system diagram for an actuator. The torque feedback is fed to an inner control loop while the position is fed back to the overall position controller.}
+%  \vspace{-0.2in}
  \label{fig:IHMCPOS}   
@@
 
+All of the figures as well as hardware and control specifications in this section are taken from \cite{ihmc_2009}.
+
+\printbibliography[heading=subbibliography]
+
+\end{refsection}
+
 % \end{document}

### 2015-09-29 exos/exosText_roboKnee.tex
revision of abstract, summary, and refs

@@
 \section{RoboKnee}
+\label{roboKnee}
+\begin{refsection}[exos/roboKnee.bib]
 
@@
 \includegraphics[width=3.in]{exos/figs/roboKnee/roboSEA}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Exploded view of a series-elastic actuator.}
+%  \vspace{-0.2in}
  \label{fig:roboAct}   
@@
 \includegraphics[width=3.in]{exos/figs/roboKnee/roboBrace}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{RoboKnee design.}
+%  \vspace{-0.2in}
  \label{fig:roboKnee}   
@@
 
- The overall design of the RoboKnee mechanism is most likely not the greatest, as it significantly increases the effective volume of the operator-system leg, but as a prototype, this concept offers interesting benefits.
+ The overall design of the RoboKnee mechanism is most likely not optimal, as it significantly increases the effective volume of the operator-system leg, but as a prototype, this concept offers interesting benefits.
+
+ The discussion as well as figures in this section are respectively based on and taken from \cite{robo_knee_2004}.
+
+\printbibliography[heading=subbibliography]
+
+\end{refsection}
 

### 2015-09-29 survey.tex
revision of abstract, summary, and refs

--- 2015-09-29 survey.tex
@@
 \begin{abstract}
-Abstract to be written.
+In the simplest case, where the environment around a robotic system remains relatively fixed, controlling the dynamic interaction between the robot and the environment to achieve a specific physical objective is difficult.
+In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for for the dynamics of multiple fundamentally linked systems and that are for the most part difficult to model.
+The goal of this document is to provide a survey of the literature on the design, modeling, and control of robotic exoskeletons and to highlight what we believe are several of the most promising currently existing approaches to the complicated goal of enabling dynamic yet safe interaction between operators, exoskeletons, and the environment.
+Where information was available, we summarize specific contributions related to actuation, power storage and generation, sensor type and distribution, mechanical architecture, as well as control system design for a number of different exoskeleton hardware systems.
+Included as part of this report are recommendations on what we see as helpful components as well as potentially problematic issues in current systems and control approaches.
 \end{abstract}
@@
 
-Executive Summary to be written.
+This document summarizes what we have found to be several of the most relevant exoskeleton hardware systems and the associated controllers used to enable the hardware to effectively and safely interact with human operators.
+As we further elaborate on in this document, the different systems that we summarize range from the relatively simple and dramatically underactuated single joint mechanisms to extremely complex and fully actuated whole-body solutions.
+The controllers associated with these systems also ranged from solutions that implemented conceptually simple designs to those which made use of very novel design techniques and/or relatively developmental sensing and actuation technologies.
 
-\section{Scope: What is this paper about?}
+While this document does try to provide a relatively wide overview of existing exoskeleton technology, we did focus our efforts in this report to hardware systems meant to provide power augmentation and magnification as opposed the somewhat larger subset of exoskeletons designed primarily for rehabilitation.
+This choice was made relative to both hardware as well as control system design considerations.
+Exoskeletons designed for rehabilitation are not typically designed as stand-alone systems and are generally limited in the scope of operational modes in which they are capable of reliably providing assistance.
+The hardware designed for the TALOS program will need to be high-powered with a power source that is capable of being carried such that the suit is a completely stand alone unit and will need to run a control system that is capable of providing transparent support to the operator in a large variety operational scenarios.
 
-This paper surveys exoskeleton control architectures and
-algorithms.
-The focus is on exoskeleton control that allows a
-highly trained and top percentile athletic 
-operator to carry a payload that weighs approximately the same amount
-as the operator. We envisage these types of exoskeletons to be useful
-in carrying protective and safety equipment for SWAT teams, police,
-firefighters, and soldiers. 
-We also 
-focus this survey on exoskeleton control for lower body tasks (standing, walking,
-running, jumping, kicking, dodging, ...).
+To this end, the hardware systems that we chose to include summaries of in this document where the BLEEX, HAL, XOR, Body Extender, Hydraulic Lower Etremety, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.
+The different systems covered many of the possible choices of actuation strategies, ranging from DC powered electric motors, to hydraulics, to pneumatics, and combinations thereof.
+Several of the systems used standard to more novel transmission strategies as well, focusing on both efficient as well as high-fidelity force control implementations.
+The transmission strategies included included standard hydraulic systems, hydraulic systems with actively controlled bypass valves, ball-screw mechanisms that include series elastic components, as well as pneumatically actuated pulley-tendon and belt-drive mechanisms.
+
+In terms of sensing and control technologies, the different exoskeletons covered in this report employed a wide variety of the spectrum of available options.
+Sensors were used to measure everything from standard information, like angular displacement of the hardware system's joints, to more novel information, like muscle activity of the user using EMG technology.
+Sensor information was used as the basis for a variety of different closed and open-loop control system designs, ranging from straight-forward force control and trajectory following to novel approaches of fully actuated feedback linearization and sensitivity amplification strategies.
+
+Each of these component technologies are reviewed as they relate to the hardware systems discussed in this report.
+Recommendations on what we feel are the best approaches in these reviews are also provided.
+We provide a summary of these recommendations at the end of this document.
+
+%\section{Scope: What is this paper about?}
+%
+%This paper surveys exoskeleton control architectures and
+%algorithms.
+%The focus is on exoskeleton control that allows a
+%highly trained and top percentile athletic 
+%operator to carry a payload that weighs approximately the same amount
+%as the operator. We envisage these types of exoskeletons to be useful
+%in carrying protective and safety equipment for SWAT teams, police,
+%firefighters, and soldiers. 
+%We also 
+%focus this survey on exoskeleton control for lower body tasks (standing, walking,
+%running, jumping, kicking, dodging, ...).
 
@@
 
-% \input{exos/exo-ul7}
-
-\section{Discussion}
-
-Discussion to be written.
-
 \section{Conclusions and Recommendations}
@@
 
+\end{document}
+
+% \input{exos/exo-ul7}
+
+%\section{Discussion}
+%
+%Discussion to be written.
+%
+%\section{Conclusions and Recommendations}
+%
+%All the (strength augmentation) exoskeletons we reviewed can be categorized into one of two general control categories.
+The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
+The first strategy is difficult to implement using standard sensing modalities.
+For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
+It would also be extremely difficult to keep these sensors in place to obtain accurate readings in a combat scenario.
+Similarly, many avoid the use of load cells for measurement of the human-exoskeleton interface fores to avoid outfitting the complicated sensor network required to accurate obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
+%
+%While sensitivity amplification strategies avoid these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
+As such, they end up amplifying noise and the user must react to stabilize the system.
+This is also unacceptable for a combat exoskeleton.
+%
+
 % \bibliographystyle{plain}
@@
 
-\end{document}
+%\end{document}
 

### 2015-09-29 exos/hal.tex
more edits

@@
 
-keywords: model-based control; predefined gait trajectory; strength augmentation; rehabilitation.\\
+% keywords: model-based control; predefined gait trajectory; strength augmentation; rehabilitation.\\
 
@@
 Compared to sensitivity amplification methods, HAL's use of EMG data allows it to estimate intention without amplifying external disturbances.
-While direct sensing of user intention is ideal for combat scenarios, sEMG data is extremely difficult to work with due to high filtering and calibration requirements.
+While direct sensing of user intention is ideal for noisy, contract-rich environments, sEMG data is extremely difficult to work with due to high filtering and calibration requirements.
 A hybrid strategy which uses sensitivity amplification and sEMG data (possibly thresholded) to filter user intention could prove effective.
-As an alternative, research into nano-fabrication and MEMs technologies has produced new ``robotic skin'' sensing arrays.
-This technology may allow direct measurement of interaction forces (i.e.
-estimating pressure / force and exoskeleton contact locations), which has not been feasible in the past.
-Such measurements could (potentially) filter external disturbances from user generated input to estimate intent.
+Additionally, Section~\ref{survey:recommend} mentions new capabilities in nano-fabrication and MEMs technologies that have produced new high-density, wearable sensing arrays.
+These and similarly advanced sensing technology may facilitate direct measurement of interaction forces (i.e.
+estimating pressure / force and exoskeleton contact locations), which could potentially filter external disturbances from user generated input to estimate intent.
 
@@
 \end{refsection}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "../survey"
+%%% End:

### 2015-09-29 exos/xor.tex

@@
 
-In implementation, the XoR uses rubber air muscles (MDSP-40) connected to pulleys via tendons and to EP regulators in the backpack (a constant pressure and cam system is being tested to reduce weight required for backpack air valves).
+In implementation, the XoR uses rubber air muscles (MDSP-40) connected to pulleys via tendons and to electro-pneumatic regulators in the backpack (a constant pressure and cam system is being tested to reduce weight required for backpack air valves).
 The electric actuator is a geared, 200W brushless DC motor (Maxon EC powermax 30) rated at 4.7 A, which transmits power to joints through a belt and pulley system with a 57.5 gear ratio.
@@
 
-For sensing, XoR is equipped with \textbf{rotary encoders} at joints, an \textbf{IMU} in the backpack, \textbf{load cells} in the feet, and is networked to servers capable of providing \textbf{EMG}, NIRS, and EEG data.
+For sensing, XoR is equipped with \textbf{rotary encoders} at joints, an \textbf{IMU} in the backpack, \textbf{load cells} in the feet, and is networked to servers capable of providing \textbf{EMG}, NIRS (near-infrared spectroscopy), and EEG (electroencephalogram) data.
 A control PC running at 1 KHz, pulse counter, amplifiers, Digital IO are external and so not included in the weight of the unit.
@@
 However, capturing the user intent is important, especially in the presence of external disturbances.
-If the sensory system were improved (e.g, wearable ``robot skin'' sensing arrays), this exoskeleton design could be highly effective.
+If the sensory system were improved (see the discussion in Section~\ref{survey:recommend}), this exoskeleton design could be highly effective.
 

### 2015-09-29 survey.tex

@@
 \section{Conclusions and Recommendations}
+\label{survey:recommend}
 
-All the (strength augmentation) exoskeletons we reviewed can be categorized into one of two general control categories.
+All the (strength augmentation) exoskeletons we reviewed can be categorized into two general control categories.
 The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
@@
 For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
-It would also be extremely difficult to keep these sensors in place to obtain accurate readings in a combat scenario.
-Similarly, many avoid the use of load cells for measurement of the human-exoskeleton interface fores to avoid outfitting the complicated sensor network required to accurate obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
+It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or combat scenarios.
+Similarly, many avoid the use of load cells for measurement of human-exoskeleton interface forces and the complicated sensor network required to accurately obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
 
-While sensitivity amplification strategies avoid these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
+While sensitivity amplification strategies circumvent these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
 As such, they end up amplifying noise and the user must react to stabilize the system.
-This is also unacceptable for a combat exoskeleton.
+This is also unacceptable for highly dynamic exoskeletons operating in contact-rich environments.
+
+We emphasize that user intent is critical, and current sensing challenges must be addressed in order to develop dynamic exoskeletons capable of serving in combat scenarios.
+As a stop-gap, it may be possible to combine low information density EMG (thresholded on/off muscle activation signals) and human force/torque or angular measurements to filter user intent.
+Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.
+As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
+Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
+Dense sensing arrays add redundancy and can improve measurement reliability.
+We recommend further study and the development of more advanced sensory systems to address the challenges inherent in dynamic exoskeleton control.
+
+In addition to capturing user intent, it is important to develop accurate models of the human/exoskeleton system.
+Feedback strategies, especially those based on force/torque control loops and series elastic actuation, are limited in bandwidth.
+Highly dynamic behaviors require feedforward models to reduce reliance on feedback and to allow lower-impedance feedback, which reduces resistive forces acting on the user.
+
+%% OTHER POINTS:
+% Model-based approaches are heavily reliant on accurate models 
+% Underactuated -- feedback linearization typically assumes fully actuated systems.
 
 \end{document}
-
-% \input{exos/exo-ul7}
-
-%\section{Discussion}
-%
-%Discussion to be written.
-%
-%\section{Conclusions and Recommendations}
-%
-%All the (strength augmentation) exoskeletons we reviewed can be categorized into one of two general control categories.
-The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
-The first strategy is difficult to implement using standard sensing modalities.
-For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
-It would also be extremely difficult to keep these sensors in place to obtain accurate readings in a combat scenario.
-Similarly, many avoid the use of load cells for measurement of the human-exoskeleton interface fores to avoid outfitting the complicated sensor network required to accurate obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
-%
-%While sensitivity amplification strategies avoid these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
-As such, they end up amplifying noise and the user must react to stabilize the system.
-This is also unacceptable for a combat exoskeleton.
-%
-
-% \bibliographystyle{plain}
-% \bibliography{exo}
-
-%\end{document}
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:
 

### 2015-09-29 exos/exosText_MIT.tex

@@
 
- The details on the controllers were somewhat sparse in the sense that, at least in \cite{}, the exact means through which the added torques were derived was not provided.
+ The details on the controllers were somewhat sparse in the sense that, at least in \cite{mit_control_2006}, the exact means through which the added torques were derived was not provided.
 Overall though, the controllers were intelligently designed in the sense that they explicitly took the biomechanics of walking into account in their design and implementation.

### 2015-09-29 survey.tex
refs correected
about to pull
fixed bibliography

@@
 \addbibresource{exos/xor.bib}
+\addbibresource{exos/bodyExt.bib}
+\addbibresource{exos/dualMod.bib}
+\addbibresource{exos/ihmc.bib}
+\addbibresource{exos/mit.bib}
+\addbibresource{exos/roboKnee.bib}
 
@@
 In the simplest case, where the environment around a robotic system remains relatively fixed, controlling the dynamic interaction between the robot and the environment to achieve a specific physical objective is difficult.
-In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for for the dynamics of multiple fundamentally linked systems and that are for the most part difficult to model.
+In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems and that are for the most part difficult to model.
 The goal of this document is to provide a survey of the literature on the design, modeling, and control of robotic exoskeletons and to highlight what we believe are several of the most promising currently existing approaches to the complicated goal of enabling dynamic yet safe interaction between operators, exoskeletons, and the environment.
@@
 We should have a blurb about the strengths/weakness of SEAs and possible options -- stiffer springs for force measurement and spring emulation for compliant control?
+
+Add references for new sensing tech
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2015-09-29 survey.tex
more updates

@@
 
-\title{A Survey of Exoskeleton Control Architectures and
+\title{A Survey of Current Exoskeletons and Their Control Architectures and
 Algorithms\\
-(Draft 3.0)}
+(Draft 3.1)}
 
@@
 \section{Executive Summary}
+
+This paper surveys the following exoskeletons in detail:
+\begin{itemize}
+\item
+BLEEX: a ``minimal-sensing'' and aggressive model-based control design point.
+\item
+HAL: an EMG controlled exoskeleton.
+\item
+XoR: a hybrid Pneumatic/Electric-motor system.
+\item
+Body Extender: a fully actuated full body exoskeleton.
+\item
+IHMC MAE: an SEA-based torque controlled exoskeleton.
+\item
+MIT Exoskeleton: an attempt to take advantage of the passive dynamics
+of walking.
+\item
+Roboknee: an example of a single-joint exoskeleton.
+\end{itemize}
+
+We then list other systems in a variety of categories.
+
+\section{Introduction}
 
@@
 
+\input{survey-atkeson}
+
 \section{Conclusions and Recommendations}

### 2015-09-29 survey.tex
more updates

--- 2015-09-29 survey.tex
@@
 \section{Executive Summary}
+
+This paper surveys the following exoskeletons in detail:
+\begin{itemize}
+\item
+{\bf BLEEX:} a ``minimal-sensing'' and aggressive model-based control design point.
+\item
+{\bf HAL:} an EMG controlled exoskeleton.
+\item
+{\bf XoR:} a hybrid Pneumatic/Electric-motor system.
+\item
+{\bf Body Extender:} a fully actuated full body exoskeleton.
+\item
+{\bf HLEE:} uses torque control approach.
+\item
+{\bf IHMC MAE:} an SEA-based torque controlled exoskeleton.
+\item
+{\bf MIT Exoskeleton:} an attempt to take advantage of the passive dynamics
+of walking.
+\item
+{\bf Roboknee:} an example of a single-joint exoskeleton.
+\end{itemize}
+
+We then list other systems in a variety of categories.
+
+\section{Introduction}
 

### 2015-09-29 survey.tex
draft of conclusions

--- 2015-09-29 survey.tex
@@
 Feedback strategies, especially those based on force/torque control loops and series elastic actuation, are limited in bandwidth.
-Highly dynamic behaviors require feedforward models to reduce reliance on feedback and to allow lower-impedance feedback, which reduces resistive forces acting on the user.
+%
+It is important that an exoskeleton's control bandwidth stays well above (typically $2-3\times$) the human control bandwidth ($2-5\,$Hz), to track dynamic motion.
+Several of exoskeletons controllers come close to this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.
+%
+Highly dynamic behaviors require predictive, feedforward models to reduce reliance on feedback and to allow a lower-impedance control response.  
+Such feedforward, model-based controllers depend heavily on accurate exoskeleton / human dynamics and so modeling efforts must be emphasized.
+
+% Finally, we note that 
+Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
+These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.
+There are more advanced control schemes that extend to underactuated systems and may apply here (e.g., optimization-based controllers or virtual model control).
+However, we note that there is a balance between feedback and feedforward control.
+The better one is, the less reliant the system is on the other.
+Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control.
 
@@
 % Underactuated -- feedback linearization typically assumes fully actuated systems.
+% human bandwidth vs control bandwidth
 

### 2015-09-29 survey.tex

@@
 \input{exos/exosText_roboKnee}
+
+\input{survey-atkeson}
 

### 2015-09-30 control.tex
more updates

@@
 dynamics.
+
+\subsection{Bias}
+
+If you ask people who research complex computationally expensive
+high end control
+systems for a living how to control a system,
+they will recommend complex computationally expensive high end control systems.
+It is entirely possible that simple local low dimensional controllers
+designed using traditional (frequency domain) techniques will do
+quite well, especially if the appropriate sensors (such as 
+operator-exoskeleton force sensors) are available.
 

### 2015-09-30 survey-atkeson.tex
more updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Appendix: Other exoskeleton systems}

\subsection{Overviews}

\url{http://powerexoskeletons.com/}\\
\url{http://exoskeletonreport.com/}\\
\url{http://robohub.org/tag/exoskeleton/}\\
\url{http://robohub.org/exoskeletons-new-and-older/}\\
\url{http://neurogadget.com/tag/exoskeleton}\\
\url{http://www.engadget.com/tag/exoskeleton/}\\
\url{http://spectrum.ieee.org/robotics/medical-robots/exoskeletons-around-the-world}\\
\url{http://spectrum.ieee.org/biomedical/bionics/the-rise-of-the-body-bots}\\
\url{http://www.extremetech.com/electronics/139633-will-we-ever-have-iron-man-exoskeletons}\\
\url{prezi.com/jygcvfxiesnm/a-brief-introduction-to-biomechanical-exoskeletons/}\\
\url{http://news.discovery.com/tech/robotics/exoskeleton-robots-top-5.htm}\\
\url{http://blog.equipoisinc.com/new-technologies-emerging-aerospace-defense-manufacturing/}\\
\url{http://nextbigfuture.com/2015/03/lower-body-exoskeleton-audi-chairless.html}\\
\url{http://www.army-technology.com/features/featuremilitary-exoskeletons-uncovered-ironman-suits-a-concrete-possibility}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We do not have sufficient information to survey the following exoskeletons.\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Commercial exoskeletons for augmentation}

\subsubsection{Sarcos XOS/WEAR}

\noindent
We do not have sufficient non-proprietary information to survey
this system.\\
\url{http://www.robaid.com/bionics/raytheons-second-generation-exoskeleton-xos-2.htm}\\
\url{http://techcrunch.com/2010/09/27/sarcos-xos2-the-real-life-iron-man/}\\

\subsubsection{Lockheed Martin HULC, FORTIS}

\noindent
We believe this is in collaboration with Ekso Bionics and is similar
to BLEEX.\\
\url{http://www.lockheedmartin.com/us/products/exoskeleton/hulc.html}\\
\url{http://www.lockheedmartin.com/us/products/exoskeleton/FORTIS.html}\\
\url{http://www.lockheedmartin.com/us/news/features/2014/mfc-103114-relief-daily-grind-industrial-exoskeletons-work.html}\\
\url{http://www.dailytech.com/From+HULC+to+FORTIS+the+Evolution+of+Lockheed+Martins+Incredible+Exosuit/article36421.htm}\\
\url{http://www.wired.com/2014/09/navys-exoskeleton-could-make-workers-20-times-more-productive/}\\
\url{http://robrady.com/design-project/lockheed-martin-fortis-human-powered-exoskeleton}\\
\url{http://www.cnn.com/2013/05/22/tech/innovation/exoskeleton-robot-suit/}

\subsubsection{Revision Military}

\noindent
PROWLER Human Augmentation System (HAS) Exoskeleton.\\
\url{https://www.youtube.com/watch?v=RKcqHaPhkkM}\\
\url{https://www.youtube.com/watch?v=Zq4amM9u-6o}\\

\subsubsection{Activelink/Panasonic}

\noindent
A variety of exoskeleton products.\\
\url{http://activelink.co.jp/en}\\

Assist Suit AWN-03:\\
{\it consists of a 13-pound backpack outfitted with leg supports which allows its wearer to lift up to 33 pounds without any noticeable effort.}\\
\url{http://www.engadget.com/2015/08/20/japans-top-oil-company-is-building-an-aliens-power-loader/}\\

Power Loader Suit:\\
\url{http://pinktentacle.com/2009/09/power-loader-exoskeleton-suit/}\\
\url{http://www.extremetech.com/extreme/173463-worlds-first-affordable-powered-exoskeleton-is-almost-here-prepare-for-mech-wars}\\
\url{http://www.dealstreetasia.com/stories/panasonic-to-release-commercial-exoskeleton-to-market-in-4q-2015-9269/}\\

\subsubsection{DAEWOO}

\noindent
\url{https://www.newscientist.com/article/mg22329803.900-robotic-suit-gives-shipyard-workers-super-strength#.U9_FTPldWSp}\\
\url{http://www.geek.com/science/daewoo-dock-workers-now-using-strength-enhancing-exo-skeletons-1601290/}\\

\subsubsection{Innophys: Muscle Suit}

\noindent
Appears to be a power assist suit for lifting. Sip and puff internface?
Tokyo University of Science Prof. Hiroshi Kobayash.\\
\url{https://innophys.jp/}\\
video: \url{https://www.youtube.com/watch?v=Sqr1l909WPw}\\
video: \url{https://www.youtube.com/watch?v=mNdJgh0ZcrM}\\
video: \url{https://www.youtube.com/watch?v=NtX81pf-3G4}\\
\url{http://blogs.wsj.com/japanrealtime/2014/11/12/wearable-power-assist-device-goes-on-sale-in-japan/}

\subsubsection{Noonee (Swiss)}

\noindent
Chair exoskeleton. Passive?\\
\url{http://www.extremetech.com/extreme/188417-rejoice-commuters-and-workers-the-chairless-chair-exoskeleton-lets-you-sit-down-anywhere-anytime}\\

\subsubsection{Sagawa Electronics (Swiss)}

\noindent
Increases reach. Master exoskeleton inside larger slave exoskeleton. Running
demonstrated.\\
\url{https://www.youtube.com/watch?v=tYgHdxmCcAI}\\
\url{http://www.ohgizmo.com/2013/07/11/mk3-exoskeleton-suit-promises-to-make-schoolgirls-better-taller-stronger/}\\
\url{http://www.bitrebels.com/technology/powered-jacket-exoskeleton-japan/}\\

\subsubsection{RB-3D}

\noindent
Hercule\\
\url{http://www.rb3d.com/produits/exosquelettes/}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Academic exoskeletons for augmentation}

\subsubsection{Kanagawa Institute of Technology}

\noindent
Pneumatic exoskeleton developed at the Kanagawa Institute of Technology,
in Atsugi, Japan, allows Akiko Michihisa, a fitness trainer, ...\\
\url{http://www.ubergizmo.com/2007/10/air-pressure-exoskeleton/}\\

\subsubsection{Monash University/Chen}

\noindent
\url{http://www.dailymail.co.uk/sciencetech/article-2639939/Super-fireman-The-hi-tech-exoskeleton-firefighters-superhuman-abilities.html}\\
\url{http://news.discovery.com/tech/robotics/firefighter-exoskeleton-to-the-rescue-140521.htm}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Commercial exoskeletons for assistance and rehabilitation}

\subsubsection{ReWalk}

\noindent
On the sixth generation of this device.\\
\url{http://rewalk.com/}\\

\subsubsection{Cyberdyne}

\noindent
up to HAL 5 (already covered)\\
\url{http://spectrum.ieee.org/robotics/medical-robots/exoskeletons-are-on-the-march}\\
\url{http://www.computerweekly.com/photostory/2240108388/Photos-Cyberdyne-Hal-robotic-exoskeleton-to-help-paralyzed/7/Japan-beats-the-US-to-it-Cyberdyne-Hal-robotic-exoskeleton-to-help-paralyzed}\\
\url{http://www.dailymail.co.uk/sciencetech/article-2384930/Robotic-exoskeleton-help-rehabilitate-disabled-people-passes-safety-tests--paving-way-sale-UK.html}\\

\subsubsection{Rexbionics}

\noindent
REX exoskeleton\\
\url{http://www.rexbionics.com/}\\
\url{http://www.gizmag.com/rex-robotic-exoskeleton/15736/pictures}\\
\url{http://www.stripes.com/news/exoskeleton-could-benefit-troops-with-spinal-cord-injuries-1.113657}\\
\url{http://www.funis2cool.com/cool/rex-the-robotic-exoskeleton-could-change-the-lives-of-disabled-people.html}\\

\subsubsection{Indego/Parker Hannafin}

\noindent
Commercialization of Vanderbilt exoskeleton\\
\url{http://www.indego.com/indego/en/home}\\
\url{https://www.youtube.com/watch?v=-bYYmZNxaZk}\\
\url{http://www.roboticstrends.com/article/parker_hannifin_indego_exoskeleton}\\
\url{http://www.crainscleveland.com/article/20141112/BLOGS03/141119915/parker-hannifin-accepts-the-challenge-of-commercializing-the-indego--41951672-812546791-1442419256=:6352--}\\

\subsubsection{Honda}

\noindent
2 devicesin the walking assist program: 1) Bodyweight Support Assist and
Stride Management Assist (SMA):\\
\url{http://corporate.honda.com/innovation/walk-assist/}\\
\url{http://world.honda.com/Walking-Assist/}\\
\url{http://www.slashgear.com/honda-demo-walking-assist-exoskeleton-for-elderly-disabled-2211295/}\\
\url{http://singularityhub.com/2010/09/13/hondas-exoskeletons-help-you-walk-like-asimo-video/}\\

\subsubsection{Hyundai}

\noindent
Appears to be similar to BLEEX.\\
Article and video: \url{https://forums.xilinx.com/t5/Xcell-Daily-Blog/Hyundai-s-wearable-exoskeleton-restores-personal-mobility-to-the/ba-p/647031}\\
Hyundai's exo project is lead by Dongjin Hyun who did his
PhD with Homayoon Kazerooni at UC Berkeley and a PostDoc with Sangbae
Kim at MIT.

\subsubsection{3D Systems}

\noindent
3D printed exoskeleton. Using Ekso Bionics technology.\\
\url{http://www.dezeen.com/2014/03/05/3d-printed-exoskeleton-helps-paralysed-users-walk/}\\

\subsubsection{ExoAtlet (Russian)}

\noindent
Appears to be similar to BLEEX.\\
\url{http://robohub.org/clinical-trials-begin-for-russias-first-medical-exoskeleton/}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Academic exoskeletons for assistance and rehabilitation}

\subsubsection{Vanderbilt}

\noindent
Commercialized by Indego/Parker Hannafin/\\
\url{https://en.wikipedia.org/wiki/Vanderbilt_exoskeleton}\\
\url{http://research.vuse.vanderbilt.edu/cim/research_orthosis.html}\\
\url{http://www.dailytech.com/Iron+Man+New+Robotic+Exoskeleton+Helps+Paraplegics+Walk/article29079.htm}\\

\subsubsection{Mindwalker}

\noindent
Key feature is actuation in lateral plane.
Each of the powered joints has a series elastic
actuator, which can deliver 100 Nm torque and 1 kW power.\\
\url{https://mindwalker-project.eu/}\\
\url{http://www.3me.tudelft.nl/en/about-the-faculty/departments/biomechanical-engineering/research/dbl-delft-biorobotics-lab/exoskeleton/}\\
\url{http://www.utwente.nl/ctw/bw/research/projects/MINDWALKER/}\\
\url{http://tnsre.embs.org/2015/03/27/design-control-mindwalker-exoskeleton/}\\

\subsubsection{U. Penn}

\noindent
Student project that got some press.\\
\url{http://www.popsci.com/article/technology/invention-awards-2014-powerful-portable-and-affordable-robotic-exoskeleton}\\

\subsubsection{East China University of Science and Technology}

\noindent
Fully portable hydraulic exoskeleton. Unclear if this is for
assistance or augmentation.\\
Video: \url{http://v.youku.com/v_show/id_XMTg3MTQzNjgw.html}\\
Heng Cao: ``most difficult thing was control bandwidth'', so they
are probably not using high quality valves.
\url{http://mech.ecust.edu.cn/s/131/t/147/6a/07/info27143.htm}

\subsubsection{University of Electronic Science and Technology of China}

\noindent
Prof Hong Cheng did his Posdoc in CMU from 2006-2009 in CS.
He is very famous in China for his exoskeleton work.\\
Chinese Video: \url{http://society.kankanews.com/s/2014-12-17/0016103932.shtml}\\
\url{http://uestcrobot.net/~hcheng/}\\
\url{http://www.uestcrobot.net/?q=node/291}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Government Programs}

\subsubsection{TALOS}

\noindent
We assume you all are familiar with this program.

\subsubsection{DARPA Warrior Web}

\noindent
We assume you all are familiar with this program.

\subsubsection{Japan: NEDO}

\noindent
There was a Japanese government (NEDO) oriented project, ``Robotic
Devices for Nursing Care Project'', that intended to boost the
practical nursing care robots development, including exoskeleton.\\
\url{http://robotcare.jp/?lang=en}\\
\url{http://robotcare.jp/?page_id=29&lang=en}\\
They made a center to test the safety for robots including exoskeletons.\\
\url{http://www.rtnet-biz.jp/rtsic/}\\

\subsubsection{EU: BALANCE program}

\noindent
\url{http://balance-fp7.eu/}\\
\url{http://cordis.europa.eu/project/rcn/106854_en.html}\\

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-09-30 survey.tex
more updates

@@
 
-\title{A Survey of Current Exoskeletons and Their Control Architectures and
+\title{A Survey of Current Exoskeletons\\ and Their Control Architectures and
 Algorithms\\
-(Draft 3.1)}
+(Draft 3.2)}
 
@@
 
-We then list other systems in a variety of categories.
+We then list other exoskeletons in a variety of categories.
 
@@
 
-\input{survey-atkeson}
-
 \section{Conclusions and Recommendations}
@@
 
+\newpage
+
+\input{survey-atkeson}
+
 \end{document}

### 2015-09-30 survey-atkeson.tex
more updates

@@
 \url{http://www.extremetech.com/electronics/139633-will-we-ever-have-iron-man-exoskeletons}\\
-\url{prezi.com/jygcvfxiesnm/a-brief-introduction-to-biomechanical-exoskeletons/}\\
+\url{https://prezi.com/jygcvfxiesnm/a-brief-introduction-to-biomechanical-exoskeletons/}\\
 \url{http://news.discovery.com/tech/robotics/exoskeleton-robots-top-5.htm}\\
@@
 \noindent
-Appears to be a power assist suit for lifting. Sip and puff internface?
+Appears to be a power assist suit for lifting. Sip and puff interface?
 Tokyo University of Science Prof. Hiroshi Kobayash.\\
@@
 \noindent
-2 devicesin the walking assist program: 1) Bodyweight Support Assist and
+2 devices in the walking assist program: 1) Bodyweight Support Assist and
 Stride Management Assist (SMA):\\
@@
 \noindent
-Prof Hong Cheng did his Posdoc in CMU from 2006-2009 in CS.
+Prof Hong Cheng did his Postdoc in CMU from 2006-2009 in CS.
 He is very famous in China for his exoskeleton work.\\

### 2015-09-30 exos/exosText_MIT.tex
updated ass. and rec. on travers sections

--- 2015-09-29 exos/exosText_MIT.tex
@@
 
- The MIT hardware and control system present a very well designed and thought-out system that focuses primarily on efficiency in design.
-The series elastic linear ball screw hardware design seems to be a very smart way to implement efficient drives as well as to perform high fidelity force control.
+ The MIT hardware and control system present a well designed system that focuses primarily on efficiency in design.
+We generally support the series-elastic linear ball screw hardware design for their mechanical efficiency as well as ability to perform high fidelity force control.
+Although, these drives will again be limited by bandwidth issues in highly-dynamic maneuvers.
 The use of a strain gauge in the shank was also a novel means of measuring ground reaction forces.
-This design may offer benefits in terms of repeatability over different operators as foot fall patterns for different users may include significant variations.
+This design may offer benefits in terms of repeatability over different operators, as foot fall patterns may include significant variations for different users or even in the same user in different environments.
 
- The details on the controllers were somewhat sparse in the sense that, at least in \cite{}, the exact means through which the added torques were derived was not provided.
-Overall though, the controllers were intelligently designed in the sense that they explicitly took the biomechanics of walking into account in their design and implementation.
-A similar approach for a variety of different behaviors may be very beneficial to the TALOS project.
+ The details on the controllers were somewhat sparse in the sense that, at least in \cite{mit_control_2006}, the exact means through which the added torques discussed above were derived was not provided.
+Overall, we support the design of the controllers in the sense that they explicitly took the biomechanics of walking into account in their design and implementation.
+A similar approach for a variety of different behaviors may be very beneficial.
 

### 2015-09-30 exos/exosText_bodyExt.tex
updated ass. and rec. on travers sections

@@
 
-The Body Extender suit is one of the only full-body exoskeletons available today, and as such is a somewhat remarkable system.
-In terms of how the Body Extender system is designed and how it is controlled, the suit may be limited.
-Clearly, for the task of handling heavy objects a fully actuated suit may be the best option, but, generally speaking, this seems almost like an over engineered solution.
-The Body Extender is a heavy, overly clunky suit.
-This suit may be alright for quasi-static operations, but it is unlikely that it will be useful for more than pick and place type operations.
-Again, seeing as this is basically what it was designed for, that seems alright.
+The Body Extender suit is one of the only fully-actuated upper and lower extremity exoskeletons available today.
+Clearly, for the task of handling extremely heavy objects, a fully actuated suit may be the best option, but, generally speaking, having the addition weight due to actuating every degree of freedom in the suit is not ideal when considering dynamic operation.
 
-In terms of the controllers highlighted in \cite{body_control_2012}, there are some good ideas.
-For example, using a regressor to estimate unknown external forces may be an idea which is directly applicable to the TALOS project.
-In the case of the Body Extender this regression is dependent on knowing at the location of the external force  relative to the suit, but the idea in general is interesting.
-The idea of using the measured forces to generate desired trajectories around which a stabilizing controller is designed also appears to be a relatively good idea.
-Note that, like any of the model-based control approaches outlined in the exoskeleton literature, or robotics literature for that matter, the feedback linearizing controller in this work is dependent on knowing the model of the robot hardware.
+The controllers highlighted in \cite{body_control_2012} do present some novel components.
+For example, using a regressor to estimate unknown external forces may be an idea which is directly applicable to a variety of different projects.
+Note that in the case of the Body Extender, this regression is dependent on knowing the location of the external force relative to the suit, the manipulator end effectors.
+Thus, an extension of this idea, possibly using a lager distribution of sensors to measure external forces, is necessary to extend this approach more generally.
 
-In terms of the control of the overall suit, it was difficult to find any control papers about the Body Extender that covered anything other than upper-body single-arm control of a manipulated object.
-The strategy of decoupling the individual limbs from each other, seems somewhat feasible, but is not the best way to approach coordinated limb control.
-It was difficult to find literature documenting the specifics of lower-limb control for the Body Extender.
-Presumably a similar strategy to that discussed here, with the addition of gravity compensation for the suit, is used.
+The way that the Body Extender control uses operator-robot measured interaction forces to generate desired trajectories around which a stabilizing controller is designed is generally an idea that we support.
+Note that, like any of the model-based control approaches outlined in the exoskeleton literature the feedback linearizing controller in this work is dependent on knowing the model of the robot hardware with a relatively high-level of fidelity.
+
+In terms of the controlling of the overall suit, i.e., locomotion in addition to manipulation, it was difficult to find any papers that discussed more than upper-body single-arm control of a manipulated object for the Body Extender hardware.
+The stated strategy of decoupling the individual limbs from each other seems somewhat feasible, but is limited relative to coordinated limb control.
 

### 2015-09-30 exos/exosText_dualMode.tex
more edits
revision of abstract, summary, and refs
updated ass. and rec. on travers sections

@@
 \section{Hydraulic Lower Extremity Exoskeleton}
+\label{dualMode}
+\begin{refsection}[exos/dualMod.bib]
 
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/exoSuit}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Exoskeleton robot system.}
+%  \vspace{-0.2in}
  \label{fig:exoSuit}   
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/hydraulicSys}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Dual-mode control}
+%  \vspace{-0.2in}
  \label{fig:hydraulicSys}   
@@
 \includegraphics[width=5.in]{exos/figs/hydLowerExrem/dualModeDia}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Locomotion control strategy during gait : Smoothing method ! Active mode ! Pre-transition method ! Passive mode}
+%  \vspace{-0.2in}
  \label{fig:dualModeDia}   
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/suitDia}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Overall control block diagram.}
+%  \vspace{-0.2in}
  \label{fig:suitDia}   
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/torAssist}
-  \caption{}
-  \vspace{-0.2in}
+%  \caption{Simplified virtual joint torque control block diagram}
+%  \vspace{-0.2in}
 % \label{fig:suitDia}   
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/blockDia}
-  \caption{}
-  \vspace{-0.2in}
+  \caption{Simplified virtual joint torque control block diagram.}
+%  \vspace{-0.2in}
  \label{fig:blockDia}   
@@
 \includegraphics[width=3.in]{exos/figs/hydLowerExrem/weightingForm}
-  \caption{}
+%  \caption{Simplified virtual joint torque control block diagram.}
 %  \vspace{-0.2in}
@@
 
- \subsection{Opinion}
+ \subsection{Assessment and Recommendations}
 
-The design of the Hydraulic Lower Extremity Exoskeleton has a lot going for it.
-The best part of this hardware as well as control design is the fact that the creators of it are doing something relatively straightforward.
-The actuators are composed of what appears to be off-the-shelf components, and the addition of bypass valves to take advantage of passive dynamics during walking to produce more energy conscious behaviors is a good idea.
+The mechanical as well as control design of the Hydraulic Lower Extremity Exoskeleton was straight-forward and contained many aspects which we support.
+In particualr, we supported the use of off-the-shelf components in the hardware design and the addition of bypass valves to take advantage of passive dynamics during walking to produce more energy conscious behaviors.
+
+The control system for this hardware effectively created a ``get out of the way design" using the torque sensed at the joints only.
+In the short-term, this type of policy seems to be a good strategy.
+The other highlight of the outlined control policy was the smoothing function which was used during phase transitions in the walking gait cycle.
+A similar type of smoothing will most likely be beneficial in terms of operator experience for a variety of different operational modes in which different phase transitions occur, e.g., running, jumping, etc.
 
 One potential drawback of this hardware is the power system.
-The documentation in \cite{} was sparse with respect to the exact specifications of the motor and energy storage device which powers the hydraulic system.
+The documentation in \cite{dual_mode_2014} was sparse with respect to the exact specifications of the motor and energy storage device which powers the hydraulic system.
+Additionally, in the long-term control design strategy, pure get out of the way control using measurements of the suit only has drawbacks, as mentioned previously.
+This strategy will ultimately lead to a ``sensitivity amplification" -like behavior.
+This strategy will thus be positively sensitive to user intended motion and negatively sensitive to external disturbances.
 
-The control system for this hardware followed a relatively straight-forward design, effectively creating a ``get out of the way design" using the torque sensed at the joints only.
-In the short-term, this type of policy seems to be a good strategy.
-The other highlight of the outlined control policy was the smoothing function which was used during phase transitions in the walking gait cycle.
-A similar type of smoothing will most likely be beneficial in terms of operator experience for a varity of different operational modes in which different phase transitions occur, e.g., running, jumping, etc.
+The figures as well as hardware and control specifications in this section are taken from \cite{dual_mode_2015}.
+
+\printbibliography[heading=subbibliography]
+
+\end{refsection}
 

### 2015-09-30 exos/exosText_ihmc.tex
updated ass. and rec. on travers sections

@@
 
- The design and use of series-elastic actuators is, in general, a good idea for many applications, but may be a problem for the highly dynamic maneuvers required by operators in the TALOS project.
-In a large sense, bandwidth is everything in these suits due to the fact that behavior prediction is very complicated.
-Simply put, 10Hz control bandwidth at the actuator may be too slow to safely react in dynamically changing environments.
+The design of the IHMC MAE is in many ways centered around being able to perform high-fidelity force control.
+In general we support the ability to perform force control using a direct measurement (or near direct) of the force/torque at the output of the transmission system.
+Series-elastic actuators may provide a solution to this problem for a number of operational scenarios, but caution should be taken when considering this solution for highly dynamic behaviors, as the control bandwidth at the actuator may be too slow to safely react in dynamically changing environments.
 
- The other issue with the IHMC is that it only addresses low-level control.
-The designers have not specified how to generalize the process of generating torque / joint reference trajectories.
-That said, other researchers do explicitly provide methods for generating desired trajectories in what amounts to a mid-level of a hierarchical control scheme.
-Thus combining other approaches with the low-level, highly accuracy torque control provided by the novel SEAs presented in this section could offer a potentially good solution at the actuator level.
+One potential other issue with the IHMC control design is that they only appear to address low-level control.
+The designers have not specified how to generalize the process of generating torque / joint reference trajectories, in other words there is a lack of information with respect to the typical ``mid-level" of a hierarchical control scheme.
+Thus combining other approaches with the low-level, highly accuracy torque control provided by the novel SEAs presented in this section could offer a potentially beneficial control solution.
 

### 2015-09-30 exos/exosText_roboKnee.tex
updated ass. and rec. on travers sections

@@
  \end{figure}
- The exact length between actuator attachment points is not listed in the literature, but it appears as though the attachment points are at a minimum of one actuator stroke displacement away from each other.
-Thus, we assume a {\bf minimum moment arm of 30.5 cm} is a reasonable approximation.
+ The exact length between actuator attachment points is not listed in the literature, but it appears that the attachment points are at minimum the
+stroke displacement (30.5cm) of the actuator away from each other.
 
@@
 
- The RoboKnee is a relatively simple system that highlights several intelligent hardware and control design concepts.
-The use of a linear ball screw with a series elastic element that enables high-fidelity force control is a very good idea.
-The idea of using the measured ground reaction forces to produce the desired knee joint torque is also interesting in the sense that it is very simple and relies on a minimal amount of information.
+ The RoboKnee is single-joint system that highlights several hardware and control design concepts that we generally support.
+We again support the use of a linear ball screw with a series elastic element to enable high mechanical efficiency and high-fidelity force control in low to medium speed maneuvers.
+The concept of using measured ground reaction forces to directly produce desired knee joint torque is interesting in the sense that it is simple and relies on minimal information.
 
- The overall design of the RoboKnee mechanism is most likely not optimal, as it significantly increases the effective volume of the operator-system leg, but as a prototype, this concept offers interesting benefits.
+The largest potential issues with the RoboKnee are that the forces of the mechanism and any load carried by the operator are fully transferred through the users musculoskeletal structure.
+Additionally, the large moment arm of the actuator significantly increases the effective volume of the operator-system leg, and as such may reduce overall mobility in densely-packed areas.
 

### 2015-09-30 survey-atkeson.tex
more updates

@@
 \section{Appendix: Other exoskeleton systems}
+
+The URLs below can be clicked on to view, if your PDF viewer/browser supports that.
 
@@
 
-We do not have sufficient information to survey the following exoskeletons.\\
+We do not have sufficient information to survey the following exoskeletons.
 
@@
 \noindent
-2 devices in the walking assist program: 1) Bodyweight Support Assist and
+2 devices in the Walking Assist Program: Bodyweight Support Assist and
 Stride Management Assist (SMA):\\
@@
 \noindent
-Commercialized by Indego/Parker Hannafin/\\
+Commercialized by Indego/Parker Hannafin\\
 \url{https://en.wikipedia.org/wiki/Vanderbilt_exoskeleton}\\
@@
 Heng Cao: ``most difficult thing was control bandwidth'', so they
-are probably not using high quality valves.
+are probably not using high quality valves.\\
 \url{http://mech.ecust.edu.cn/s/131/t/147/6a/07/info27143.htm}
@@
 \noindent
-We assume you all are familiar with this program.
+We assume the reader is familiar with this program.
 

### 2015-09-30 survey.tex
more updates

@@
 In the simplest case, where the environment around a robotic system remains relatively fixed, controlling the dynamic interaction between the robot and the environment to achieve a specific physical objective is difficult.
-In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems and that are for the most part difficult to model.
+In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems that are for the most part difficult to model.
 The goal of this document is to provide a survey of the literature on the design, modeling, and control of robotic exoskeletons and to highlight what we believe are several of the most promising currently existing approaches to the complicated goal of enabling dynamic yet safe interaction between operators, exoskeletons, and the environment.
@@
 \item
-{\bf BLEEX:} a ``minimal-sensing'' and aggressive model-based control design point.
+{\bf BLEEX:} a ``minimal-sensing'' and aggressive model-based control design that has influenced many subsequent exoskeletons.
 \item
-{\bf HAL:} an EMG controlled exoskeleton.
+{\bf HAL:} an EMG controlled exoskeleton, emphasizing recognizing operator intent.
 \item
-{\bf XoR:} a hybrid Pneumatic/Electric-motor system.
+{\bf XoR:} a hybrid Pneumatic/Electric-motor system, emphasizing the
+value of using multiple actuator types.
 \item
-{\bf Body Extender:} a fully actuated full body exoskeleton.
+{\bf Body Extender:} a fully actuated full body exoskeleton. This is the
+Iron Man option.
 \item
-{\bf HLEE:} uses torque control approach.
+{\bf HLEE:} a nice and well presented example of a torque control approach.
 \item
-{\bf IHMC MAE:} an SEA-based torque controlled exoskeleton.
+{\bf IHMC MAE:} an SEA-based torque controlled exoskeleton, emphasizing
+the value of mixing passive mechanisms (local energy storage) and active
+actuation.
 \item
@@
 \item
-{\bf Roboknee:} an example of a single-joint exoskeleton.
+{\bf RoboKnee:} an example of a single-joint exoskeleton. Powerful single-joint
+or minimal exoskeletons are one way to enable operators to move fast
+and naturally, and greatly increase their strength.
 \end{itemize}
@@
 This document summarizes what we have found to be several of the most relevant exoskeleton hardware systems and the associated controllers used to enable the hardware to effectively and safely interact with human operators.
-As we further elaborate on in this document, the different systems that we summarize range from the relatively simple and dramatically underactuated single joint mechanisms to extremely complex and fully actuated whole-body solutions.
-The controllers associated with these systems also ranged from solutions that implemented conceptually simple designs to those which made use of very novel design techniques and/or relatively developmental sensing and actuation technologies.
+As we further elaborate, the different systems that we summarize range from the relatively simple and dramatically underactuated single joint mechanisms to extremely complex and fully actuated whole-body solutions.
+The controllers associated with these systems also ranged from solutions that implemented conceptually simple designs to those which made use of novel and more complex design techniques and/or relatively untried sensing and actuation technologies.
 
-While this document does try to provide a relatively wide overview of existing exoskeleton technology, we did focus our efforts in this report to hardware systems meant to provide power augmentation and magnification as opposed the somewhat larger subset of exoskeletons designed primarily for rehabilitation.
+While this document does try to provide a relatively wide overview of existing exoskeleton technology, we did focus our efforts in this report on hardware systems meant to provide power augmentation and magnification as opposed the somewhat larger subset of exoskeletons designed primarily for rehabilitation.
 This choice was made relative to both hardware as well as control system design considerations.
 Exoskeletons designed for rehabilitation are not typically designed as stand-alone systems and are generally limited in the scope of operational modes in which they are capable of reliably providing assistance.
-The hardware designed for the TALOS program will need to be high-powered with a power source that is capable of being carried such that the suit is a completely stand alone unit and will need to run a control system that is capable of providing transparent support to the operator in a large variety operational scenarios.
+The hardware designed for athletic operators and extreme tasks will need to be high-powered with a power source that is capable of being carried such that the suit is a completely stand alone unit and will need to run a control system that is capable of providing transparent support to the operator in a large variety of operational scenarios.
 
-To this end, the hardware systems that we chose to include summaries of in this document where the BLEEX, HAL, XOR, Body Extender, Hydraulic Lower Etremety, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.
-The different systems covered many of the possible choices of actuation strategies, ranging from DC powered electric motors, to hydraulics, to pneumatics, and combinations thereof.
-Several of the systems used standard to more novel transmission strategies as well, focusing on both efficient as well as high-fidelity force control implementations.
-The transmission strategies included included standard hydraulic systems, hydraulic systems with actively controlled bypass valves, ball-screw mechanisms that include series elastic components, as well as pneumatically actuated pulley-tendon and belt-drive mechanisms.
+To this end, the hardware systems that we chose to include summaries of in this document were the BLEEX, HAL, XoR, Body Extender, Hydraulic Lower Extremety, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.
+The different systems cover many of the possible choices of actuation strategies, ranging from electric motors, to hydraulics, to pneumatics, and combinations thereof.
+Several of the systems used novel transmission strategies as well, focusing on both efficient as well as high-fidelity force control implementations.
+The transmission strategies included standard hydraulic systems, hydraulic systems with actively controlled bypass valves, ball-screw mechanisms that include series elastic components, as well as pneumatically actuated pulley-tendon and belt-drive mechanisms.
 

### 2015-09-30 control.tex
more updates

@@
 
-\title{A Survey of Possible Exoskeleton Control Architectures and
-Algorithms\\
-(Draft 3.1)}
+\title{Possible Exoskeleton Control Architectures and Algorithms\\
+(Draft 4.0)}
 
@@
 
-\begin{abstract}
-Abstract to be written.
-\end{abstract}
-
 \section{Executive Summary}
 
-Executive Summary to be written.
+We recommend a red team that explores frequency domain control system
+design, as well as an emphasis on ``modern'' control in this paper.
+
+\section{Clickable web links}
+
+The web links (URLs) below can be clicked on to view, if your PDF viewer/browser supports that. On our browsers you actually have to click on the black
+part of a letter, not just in the blue box (if present).
 
@@
 A companion paper surveys implemented exoskeleton control architectures
-and algorithms [survey.pdf].
+and algorithms.\\
+\url{http://www.cs.cmu.edu/~cga/exo/survey.pdf}
 
@@
 operator-exoskeleton force sensors) are available.
+It may even be the case that frequency domain techniques will handle
+structural modeling error (jello-like human operators, deformable materials
+(straps and pads, wires and hoses), and structural deformation and joint
+play in lightweight frames) better than the state space oriented techniques
+advocated here.
+
+We recommend a red team that explores frequency domain control system
+design, as well as an emphasis on ``modern'' control here.
 
@@
 in either optimization or feedback control (barrier
-Lyapunov functions (BLF)~\cite{IEEE06911561}).
+Lyapunov functions (BLF).\\
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6911561}
 
@@
 
-We recommend the following hierarchical control architecture~\cite{IEEE06907051}:
+We recommend the following hierarchical control architecture.
+% review? \url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6907051}
 
@@
 for higher control levels
-to access the actuators, sensors, and other robot hardware~\cite{IEEE07271219}.
+to access the actuators, sensors, and other robot hardware.\\
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7271219}\\
 It implements high servo rate processing including feedback control and online

### 2015-09-30 survey.tex
updated conclusion

@@
 
-All the (strength augmentation) exoskeletons we reviewed can be categorized into two general control categories.
-The exoskeleton either directly measures user intent (e.g., from EMG data or human-exoskeleton interaction forces), or it estimates intent from measurements at exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
-The first strategy is difficult to implement using standard sensing modalities.
+The (strength augmentation) exoskeletons we reviewed can be categorized into two general control categories.
+The exoskeleton either directly measures user intent (e.g., from EMG data or direct human-exoskeleton interaction forces), or it estimates intent from measurements at the exoskeleton joints and applies a sensitivity amplification strategy to shadow the wearer.
+
+Directly measuring user intent is difficult to implement using standard sensing modalities.
 For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
 It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or combat scenarios.
-Similarly, many avoid the use of load cells for measurement of human-exoskeleton interface forces and the complicated sensor network required to accurately obtain interaction forces / torques (humans are compliant thus rigid attachment assumptions are typically inadequate).
+For similar reasons, many systems avoid the use of load cells or complicated sensor networks to accurately obtain operator-robot interaction forces / torques.
 
-While sensitivity amplification strategies circumvent these sensing issues by estimating human intent from exoskeleton joint readings, they cannot distinguish between user generated torques and those produced by the environment.
-As such, they end up amplifying noise and the user must react to stabilize the system.
-This is also unacceptable for highly dynamic exoskeletons operating in contact-rich environments.
+Sensitivity amplification strategies circumvent many of the direct operator-robot sensing issues by estimating human intent from measurements of the exoskeleton hardware.
+The issue with this policy is that it cannot generally distinguish between user generated torques and those produced by external sources.
+As such, they end up amplifying not only the user's intent, but also external disturbances, requiring the operator to expend valuable energy to stabilize the system.
+Additionally, even under the assumption that external forces can be minimized or mitigated by the user, enabling the suit to be highly responsive to perceived user intent implies that the hardware needs to be operating near the margin of stability.
+This is not an optimal strategy for highly dynamic behaviors, as the suit may be pushed beyond the point to which the operator can reasonably stabilize it.
 
-We emphasize that user intent is critical, and current sensing challenges must be addressed in order to develop dynamic exoskeletons capable of serving in combat scenarios.
-As a stop-gap, it may be possible to combine low information density EMG (thresholded on/off muscle activation signals) and human force/torque or angular measurements to filter user intent.
-Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.
-As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
+In light of these opinions, our main recommendation is that a control policy which implements a \emph{modified get out of the way} strategy would provide a reasonable strategy for dynamic exoskeleton control.
+The get out of the way controller used in the Hydraulic Lower Extremity Exoskeleton discussed above would provide a starting point for this strategy.
+This controller utilizes IMUs, foot contact sensors, and joint torque sensors to implement a user-reactive gravity compensating strategy.
+Augmenting the sensor suit and control system to include sensors that directly measure the forces between the operator and the robot would enable this system not only avoid the user, but also to reject external disturbances.
+
+To achieve this type of augmented get out of the way control, it may be possible to use low information density EMG (thresholded on/off muscle activation signals) with robot force/torque and angular measurements to filter operator-robot interaction forces.
+As a longer term solution, we highlight research in the area of MEMs, nano-fabrication, soft sensors, etc., which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
 Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
 Dense sensing arrays add redundancy and can improve measurement reliability.
-We recommend further study and the development of more advanced sensory systems to address the challenges inherent in dynamic exoskeleton control.
 
-In addition to capturing user intent, it is important to develop accurate models of the human/exoskeleton system.
-Feedback strategies, especially those based on force/torque control loops and series elastic actuation, are limited in bandwidth.
-%
+We also highly stress the importance of high-bandwidth control loops.
 It is important that an exoskeleton's control bandwidth stays well above (typically $2-3\times$) the human control bandwidth ($2-5\,$Hz), to track dynamic motion.
-Several of exoskeletons controllers come close to this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.
-%
-Highly dynamic behaviors require predictive, feedforward models to reduce reliance on feedback and to allow a lower-impedance control response.  
-Such feedforward, model-based controllers depend heavily on accurate exoskeleton / human dynamics and so modeling efforts must be emphasized.
+Several of exoskeletons controllers are close to lower limit of this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.
+Note that these bandwidth issues are primarily a function of the exoskeleton hardware, and thus special care/analysis needs to be given/conducted to make sure that the hardware is capable of physically producing required behaviors.
+
+Lastly, the importance of producing hardware that is physically capable of achieving dynamic performance is fundamentally linked to how mass is distributed throughout the system.
+In general, this means that minimizing mass, more specifically distal masses on extremities, is desirable, which in turn means that at certain degrees of freedom must remain unactuated.
+Underactuted control is a complicated problem which can either be ignored, in the sense that the control algorithm assumes the user will provide the necessary forces and torques for the unactuated degrees of freedom, dealt with using optimization-based control methods, or can be partially compensated using passive energy storage and dissipation elements.
+Our recommendation is that a combination of biomechanically-inspired passive elements and optimization-based control methods will ultimately produce the best strategy for dealing with underactuation in higly-dynamic exoskeleton systems.
+
+%Such feedforward, model-based controllers depend heavily on accurate exoskeleton / human dynamics and so modeling efforts must be emphasized.
 
 % Finally, we note that 
-Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
-These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.
+%Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
+%These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.
 There are more advanced control schemes that extend to underactuated systems and may apply here (e.g., optimization-based controllers or virtual model control).
@@
 The better one is, the less reliant the system is on the other.
-Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control.
+%
+%Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control.
 

### 2015-09-30 exos/exosText_dualMode.tex
complete version of survey.tex

@@
 
-The Hydraulic Lower Extremity Exoskeleton's actuators were designed based on human walking data obtained at 4km/h with a 45 kg load.
-Based on this data, the hydraulic capacity system was designed to provide 2050psi and 8lpm.
+The Hydraulic Lower Extremity Exoskeleton's actuators were designed based on human walking data obtained at 4 km/h with a 45 kg load.
+Based on this data, the hydraulic capacity system was designed to provide 2050 psi and 8 LPM.
 The individual actuators were designed to be double acting at the hip and knee flexion/extension joints.
@@
 The mechanical as well as control design of the Hydraulic Lower Extremity Exoskeleton was straight-forward and contained many aspects which we support.
-In particualr, we supported the use of off-the-shelf components in the hardware design and the addition of bypass valves to take advantage of passive dynamics during walking to produce more energy conscious behaviors.
+In particular, we supported the use of off-the-shelf components in the hardware design and the addition of bypass valves to take advantage of passive dynamics during walking to produce more energy conscious behaviors.
 
@@
 One potential drawback of this hardware is the power system.
-The documentation in \cite{dual_mode_2014} was sparse with respect to the exact specifications of the motor and energy storage device which powers the hydraulic system.
+The documentation in \cite{dual_mode_2015} was sparse with respect to the exact specifications of the motor and energy storage device which powers the hydraulic system.
 Additionally, in the long-term control design strategy, pure get out of the way control using measurements of the suit only has drawbacks, as mentioned previously.

### 2015-09-30 exos/hal.tex
complete version of survey.tex

@@
 The hybrid assistive limb (HAL) exoskeleton is developed by the University of Tsukuba and Cyberdyne for human strength augmentation and as an assistive gait device in rehabilitation.
-Although several prototypes have been developed, this section focuses on the most recent HAL-5 model (specifically the non-clicial type, HAL-5 Type-B).
+Although several prototypes have been developed, this section focuses on the most recent HAL-5 model (specifically the non-clinical type, HAL-5 Type-B).
 Though HAL-5 is a full body exoskeleton, only hip, knee joints, and ankle are actuated in the sagittal plane.

### 2015-09-30 control.tex

@@
 
-We recommend a red team that explores frequency domain control system
-design, as well as an emphasis on ``modern'' control in this paper.
+\begin{enumerate}
+\item
+We recommend using {\bf online optimization.} This architecture allows
+a wide variety of control approaches to be implemented and work together.
+\item
+We also recommend a {\bf red team} that explores traditional 
+{\bf frequency domain} control system
+design.
+\item
+We recommend applying as {\bf many sensors} as possible, and asking the performers
+to make it easy to add more by making the sensor network available in the design.
+This maximizes the probability of success
+with respect to control architectures and algorithms by enabling multiple control
+approaches to be implemented, refined, and support each other.
+\item
+We recommend a {\bf ``symbiotic'' control system design} should be used as a backup,
+in case more aggressive design philosophies such as ``invisible'' and ``natural''
+control system designs are not achievable.
+The ``symbiotic'' control system design
+expects the operator to adapt to the exoskeleton and its control,
+and has the control customized for the operator.
+\end{enumerate}
 
@@
 %\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
-A later white paper could discuss
+A future white paper could discuss
 possible combinations of response to user and exoskeleton
@@
 
-A later white paper could discuss how to handle exoskeleton
+A future white paper could discuss how to handle exoskeleton
 malfunctions and damage to the
@@
 
-A later white paper could evaluate using simulation
+A future white paper could evaluate using simulation to assess
 the various control approaches' robustness to modeling error and unmodeled
 dynamics.
+
+A future white paper could survey human learning with respect to exoskeletons
+and teleoperation.\\
+\url{https://www.newscientist.com/article/dn28155-exoskeleton-reveals-the-lengths-we-go-to-save-energy-when-walking/}\\
+
+\begin{verbatim}
+K. E. Gordon, C. R. Kinnaird, and D. P. Ferris, “Locomotor adaptation
+to a soleus emg-controlled antagonistic exoskeleton,” J. Neurophysiol.,
+vol. 109, no. 7, pp. 1804–1814, 2013.
+\end{verbatim}
+
+A future white paper could survey adaptive control of exoskeletons.\\
+{\it
+The relationship between physical human-exoskeleton interaction and
+dynamic factors: using a learning approach for control applications}\\
+\url{http://info.scichina.com/sciFe/EN/abstract/abstract516165.shtml}
+
+A future white paper could survey the role of phase estimation, behavioral clocks,  phase reset events such as impacts, central pattern generators,
+and oscillator approaches in robot and exoskeleton control.
+
+\subsection{Related Surveys}
+
+Active Exoskeleton Control Systems: State of the Art (2012)\\
+\url{http://www.sciencedirect.com/science/article/pii/S1877705812026732}\\
+Our copy:\\
+\url{http://www.cs.cmu.edu/~cga/exo/info/Anam2012.pdf}
 
@@
 (straps and pads, wires and hoses), and structural deformation and joint
-play in lightweight frames) better than the state space oriented techniques
-advocated here.
-
-We recommend a red team that explores frequency domain control system
-design, as well as an emphasis on ``modern'' control here.
+play in lightweight frames) better than the state space oriented 
+and other advanced techniques advocated here.
+
+We recommend a red team that explores traditional 
+frequency domain control system
+design, as well as an emphasis on advanced control approaches.
 
@@
 Typically low level control is implemented
-by a high sampling rate global servo~\cite{Atlas-robot}
-or distributed limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
+by a high sampling rate global servo [Boston Dynamics Atlas humanoid]
+or distributed limb or joint-level high sampling rate servos
+[Sarcos Primus humanoid].
 Often, this controller is provided by the manufacturer of the robot
 or exoskeleton and runs on a separate computer system from the
-``user's'' or task/application controller~\cite{Sarcos,Atlas}.
+``user's'' or task/application controller.
 
@@
 redundant sensing is a separate process and is as independent as possible
-from the control method~\cite{certainty-eq,separation-prin}.
+from the control method.
+%~\cite{certainty-eq,separation-prin}.
 This makes evaluating different control methods much easier.
@@
 commonly used. Variants of this architecture were used by almost all teams
-in the DARPA Robotics Challenge~\cite{}.
+in the DARPA Robotics Challenge.
+%~\cite{}.
 
@@
 An alternative approach to state estimation is to design state estimation
-and control laws simultaneously~\cite{LQG-LTR}. Another alternative is
+and control laws simultaneously.
+%~\cite{LQG-LTR}. 
+Another alternative is
 to have a controller with internal state, and design or learn a policy
 that works well, but does not have an explicit state estimation 
-part~\cite{lead-lag}.
+part.
+%~\cite{lead-lag}.
 
@@
 A key question is where does the desired acceleration $\ddot{\vq}_d$ come from?
-This is discussed in a companion paper [intent.pdf].
+This is discussed in a companion paper.
+\url{http://www.cs.cmu.edu/~cga/exo/intent.pdf}
 
@@
 effects on those joints. This is what has been done with the BLEEX exoskeleton,
-for example.~\cite{}.
+for example.
+%~\cite{}.
 
@@
 Another approach to solving this problem is prioritized control, which does
-not include optimization~\cite{}. For reasons we will describe, we believe
+not include optimization.
+%~\cite{}
+For reasons we will describe, we believe
 online optimization is much more flexible and can handle programming exoskeleton
@@
 
-Quadratic programming~\cite{Wikipedia} solves the following problem:
+Quadratic programming
+%~\cite{Wikipedia} 
+solves the following problem:
 Minimize
@@
 Another approach to shared control is allocating time slices or providing assistance
-on an as-needed basis~\cite{IEEE07140030}.
-
-This recent paper discusses shared control \url{http://www.ri.cmu.edu/pub_files/2015/7/Javdani15Hindsight.pdf}
+on an as-needed basis.\\
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7140030}
+
+This recent paper discusses shared control\\
+\url{http://www.ri.cmu.edu/pub_files/2015/7/Javdani15Hindsight.pdf}
 
@@
 
-Point to DRC papers by multiple groups.
+More details on our work and pointers to other work are in the papers at:\\
+\url{http://www.cs.cmu.edu/~cga/drc}.
 
@@
 This work also received
-a ``Best Oral Paper Award'' at Humanoids 2013~\cite{Feng/etal13}. Figure~\ref{fig:terrain_pic} shows a sequence of snapshots of ATLAS 
+a ``Best Oral Paper Award'' at Humanoids 2013. Figure~\ref{fig:terrain_pic} shows a sequence of snapshots of ATLAS 
 walking on rough terrain and plots of measured trajectories of the feet and the center of mass (CoM) for the DRC Trials.
@@
 biped that did not fall down or need to be rescued by 
-humans during the two days of the evaluations~\cite{cga-hum15}.
+humans during the two days of the evaluations.
 Our robot performance was also the most consistent across the two days.
@@
 from the actual evaluations. Videos of our robot's performance during the
-DRC Finals are available~\cite{cga-web-page}.
+DRC Finals are available.\\
+\url{http://www.cs.cmu.edu/~cga/drc}.
 
@@
 We optimize with a time horizon of several seconds
-using Differential Dynamic Programming~\cite{ddp}.
+using Differential Dynamic Programming.
+%~\cite{ddp}.
 
 For walking, our planner is similar in spirit to Kajita's Preview Control for a Linear Inverted
-Pendulum Model (LIPM)~\cite{kajita03}. We are also using a CoM
+Pendulum Model (LIPM)
+%~\cite{kajita03}.
+We are also using a CoM
 model, reason about Zero Moment Point (ZMP), and use future information 
@@
 %We explicitly add the vertical height in our  CoM model to handle height variations on rough terrain. 
-Like capture point methods~\cite{capture_pt}, we take the 
+Like capture point methods,
+%~\cite{capture_pt}, 
+we take the 
 next few steps into consideration but
-do not plan to come to rest at the end. Ogura et al.~\cite{wabian2} 
+do not plan to come to rest at the end. Ogura et al.
+%~\cite{wabian2} 
 investigated generating human-like walking with heel-strike and toe-off by 
@@
 
-For the QP, we use a formulation developed in our group 
-\cite{Stephens-thesis,Whitman10,Whitman-thesis}. Unlike~\cite{mistry10,usc_id1} who 
-use orthogonal decomposition to project the allowable motions into the null 
+For the QP, we use a formulation developed in our group.
+%\cite{Stephens-thesis,Whitman10,Whitman-thesis}. 
+Unlike those
+% ~\cite{mistry10,usc_id1}
+who use orthogonal decomposition to project the allowable motions into the null 
 space of the constraint Jacobian, and minimize costs in the contact constraints and the commands, we directly 
@@
 limits. Although it becomes a bigger QP problem, we are still able to solve it 
-in real time. Hutter el al.~\cite{eth_id} resolved redundancy in inverse 
+in real time. Hutter el al.
+%~\cite{eth_id}
+resolved redundancy in inverse 
 dynamics using a kinematic task prioritization approach that ensures lower 
@@
 appropriate state information to all other modules. The details of this work on state estimation are
-described in~\cite{xinji_com_est,Xinjilefu-thesis}.
+described in.
+\url{http://www.cs.cmu.edu/~cga/drc}.
+%~\cite{xinji_com_est,Xinjilefu-thesis}.
 
@@
 $\mJ$ is an appropriate Jacobian matrix, $\mK_1$ is a gain matrix,
-and $\vf$ is the force vector to be controlled~\cite{IEEE06990981}.
+and $\vf$ is the force vector to be controlled.\\
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=6990981}
+%~\cite{IEEE06990981}.
 
@@
 redundant sensing is a separate process and is as independent as possible
-from the control method~\cite{certainty-eq,separation-prin}.
+from the control method
+%~\cite{certainty-eq,separation-prin}.
 This makes evaluating different control methods much easier.
@@
 are measured. We use the full internal dynamics as the process model,
-assuming the root dynamics are known. See Chapter 3 of~\cite{Xinjelefu-thesis}.
+assuming the root dynamics are known. See Chapter 3 of\\
+\url{http://www.cs.cmu.edu/~xxinjile/pdf/main.pdf}\\
+%~\cite{Xinjelefu-thesis}.
 The internal dynamics are linearized and covariances updated
@@
 For higher level (overall) state estimation,
-we developed a center of mass location and velocity estimator~\cite{}.
+we developed a center of mass location and velocity estimator.
+%~\cite{}.
 The center of mass filter is modeled as a multiple model EKF with contact switching
@@
 push the robot or exoskeleton over.
-We compute a ``corrected capture point'' (CCP)~\cite{capture},
+We compute a ``corrected capture point'' (CCP),
+%~\cite{capture},
 which is an offset to the current capture point.
@@
 current center of mass height $z$. We assume a constant center of mass height
-in estimating the corrected capture point based on the estimated capture point
-as described in~\cite{capture}:
+in estimating the corrected capture point based on the estimated capture point:
+% as described in~\cite{capture}:
 \begin{equation}
@@
 improve both velocity and 
-acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
+acceleration estimation
+%~\cite{Xinjilefu-thesis}, 
+which should improve
 Sensitivity Amplification Control.
 
-Disturbance observers~\cite{IEEE06197032}.
+% Disturbance observers~\cite{IEEE06197032}.
 
@@
 This question needs to be addressed using simulation, and is deferred to a 
-possible future white paper [robust.pdf].
+possible future white paper.
 
@@
 On the other hand, feedback control can lead to steady state or growing
-oscillation, which we colloquially refer to as being unstable or instability.
+oscillations, which we colloquially refer to as being unstable or an
+instability.
 Since walking and running are also oscillations which we refer to as stable
@@
 descent in simulation to improve the control robustness.
-One technique to do this is multiple-model design~\cite{acc12}.
+One technique to do this is multiple-model design.
+%~\cite{acc12}.
+\url{http://www.cs.cmu.edu/~cga/papers/acc2012-tr.pdf}
 
@@
 \subsection{Integral admittance control}
+
+Integral admittance control is a refinement of 
+sensitivity amplification control.
+This section is quotes from:\\
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7139989}
 
@@
 achieved. This framework also ensures that the coupled stability
-and passivity are guaranteed.}~\cite{Nagarajan_etal_2015}
+and passivity are guaranteed.}
+%~\cite{Nagarajan_etal_2015}
 
@@
 is stable and passive. ... 
-}~\cite{Nagarajan_etal_2015}
+}
+%~\cite{Nagarajan_etal_2015}
 
@@
 feedback with a second-order low-pass Butterworth filter. ...
-}~\cite{Nagarajan_etal_2015}
+}
+%~\cite{Nagarajan_etal_2015}
 
@@
 walking by predicting the change from the swing to the stance
-in advance.}~\cite{IEEE07222598}
+in advance.}\\
+%~\cite{IEEE07222598}
+\url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7222598}
 
@@
 
-\section{Discussion}
-
-Discussion to be written.
+%\section{Discussion}
+
+%Discussion to be written.
 
@@
 
-Conclusions and Recommendations to be written.
-
-1) We recommend a particular control 
-overall architecture in Section~\ref{sec:architecture}.
+0) We recommend a red team that explores traditional 
+frequency domain control system
+design, as well as an emphasis on advanced control approaches.
+
+1) We recommend a particular overall control 
+architecture in Section~\ref{sec:architecture}.
 We do not expect this to be controversial.
@@
 and fit the necessary torque sensing.
+We do not expect this to be controversial.
 
@@
 in case more aggressive design philosophies such as ``invisible'' and ``natural''
-control system designs are not achievable.,
+control system designs are not achievable.
 The ``symbiotic'' control system design
@@
 
-7) We recommend making use of virtual model control specialize
+7) We recommend making use of virtual model control specialized
 for the various tasks. This can greatly increase operator/exoskeleton system
@@
 
-\bibliographystyle{plain}
-\bibliography{exo}
+%\bibliographystyle{plain}
+%\bibliography{exo}
 
 \end{document}
-
-I have incorporated your suggestions about robustness and underactuation
-in the white paper draft 3.1 now at:
-www.cs.cmu.edu/~cga/exo/control.pdf
-
-State estimation (to deal with sensor noise and to fuse disparate sensors)
-is discussed in section 21 and mentioned in 5.5, 5.8,
-
-Robustness (sensitivity to modeling error and unmodeled dynamics) is
-discussed in section 22
-
-Underactuation is discussed in section 18 and the following sections.
-
-Shared control is discussed in section 19.5
-
-********************************************
-
-A future white paper could explore using simulation studies to experimentally
-assess robustness of various control approaches.
-
-A future white paper could explore using simulation studies to numerically
-estimate:
-1) how necessary is lateral actuation
-2) what is the penalty (weight, drag of actuator, complexity, maintenance)
-
-A future white paper could explore how to handle failure and damage, again using
-simulation.

### 2015-09-30 survey.tex

@@
 In the simplest case, where the environment around a robotic system remains relatively fixed, controlling the dynamic interaction between the robot and the environment to achieve a specific physical objective is difficult.
-In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems that are for the most part difficult to model.
+In the case where an operator effectively wears the robot, i.e., a robotic exoskeleton, and would like to move through unknown and dynamically evolving environments, the situation becomes very complicated; the robotic system needs to account for the dynamics of multiple fundamentally linked systems that are, for the most part, difficult to model.
 The goal of this document is to provide a survey of the literature on the design, modeling, and control of robotic exoskeletons and to highlight what we believe are several of the most promising currently existing approaches to the complicated goal of enabling dynamic yet safe interaction between operators, exoskeletons, and the environment.
@@
 
-We then list other exoskeletons in a variety of categories.
+\noindent Exoskeletons in a variety of categories are listed in the Appendix.\\
+
+\noindent Based on our review, we have made the following conclusions:
+\begin{itemize}
+\item Most exoskeletons estimate user intent indirectly, from measurements of suit variables.
+\item Due to limitations in current sensory systems, it is difficult to sense user intent directly. 
+\item As direct measurement of user intent is critical in dynamic, noisy environments, we recommend better fusion of low information density sensory data obtained from users with closed-loop control based on suit variables.
+\item In the longer term, we recommend investigation and development of more advanced human-exoskeleton sensing technology to capture user intent.
+\end{itemize}
 
@@
 As such, they end up amplifying not only the user's intent, but also external disturbances, requiring the operator to expend valuable energy to stabilize the system.
-Additionally, even under the assumption that external forces can be minimized or mitigated by the user, enabling the suit to be highly responsive to perceived user intent implies that the hardware needs to be operating near the margin of stability.
+In dynamic situations the operator may not be capable of actively stabilizing the exoskeleton.
+% Additionally, even under the assumption that external forces can be minimized or mitigated by the user, enabling the suit to be highly responsive to perceived user intent implies that the hardware needs to be operating near the margin of stability.
 This is not an optimal strategy for highly dynamic behaviors, as the suit may be pushed beyond the point to which the operator can reasonably stabilize it.
 
-In light of these opinions, our main recommendation is that a control policy which implements a \emph{modified get out of the way} strategy would provide a reasonable strategy for dynamic exoskeleton control.
-The get out of the way controller used in the Hydraulic Lower Extremity Exoskeleton discussed above would provide a starting point for this strategy.
-This controller utilizes IMUs, foot contact sensors, and joint torque sensors to implement a user-reactive gravity compensating strategy.
-Augmenting the sensor suit and control system to include sensors that directly measure the forces between the operator and the robot would enable this system not only avoid the user, but also to reject external disturbances.
-
-To achieve this type of augmented get out of the way control, it may be possible to use low information density EMG (thresholded on/off muscle activation signals) with robot force/torque and angular measurements to filter operator-robot interaction forces.
+% Based on these observations, our main recommendation is that a \emph{modified get out of the way} control strategy provides a reasonable policy for dynamic exoskeletons.
+Based on these observations, we recommend (at least in the short term) fusing the two predominant approaches to exoskeleton control.
+That is, combining sensitivity amplification strategies with direct user measurement to robustly distinguish user intent from external disturbances.
+% For instance, the get out of the way controller used in the Hydraulic Lower Extremity Exoskeleton would provide a starting point for this strategy.
+This controller utilizes IMUs, foot contact sensors, and joint torque sensors to implement a user-reactive gravity compensating strategy that does not directly measure the user.
+Augmenting the sensor suite with sensors that that measure user generated forces could allow this system to identify and reject external disturbances while simultaneously shadowing user motion.
+%
+% We emphasize that user intent is critical, and current sensing challenges must be addressed in order to develop dynamic exoskeletons capable of serving in combat scenarios.
+As a stop-gap, it may be possible to combine low information density EMG (thresholded on/off muscle activation signals) and human force/torque or angular measurements to filter user intent.
+Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.
+% As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
+Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
+Dense sensing arrays add redundancy and can improve measurement reliability.
+We recommend further study and the development of more advanced sensory systems to address the challenges inherent in dynamic exoskeleton control.
+%
+For instance, it may be possible to use low information density EMG (thresholded on/off muscle activation signals) with robot force/torque and angular measurements to filter disturbances from operator-robot interaction forces.
 As a longer term solution, we highlight research in the area of MEMs, nano-fabrication, soft sensors, etc., which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
 Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
-Dense sensing arrays add redundancy and can improve measurement reliability.
+Dense sensing arrays add redundancy and can improve measurement reliability and, thereby, feedback control performance.
 
@@
 It is important that an exoskeleton's control bandwidth stays well above (typically $2-3\times$) the human control bandwidth ($2-5\,$Hz), to track dynamic motion.
-Several of exoskeletons controllers are close to lower limit of this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.
+Several of the reviewed exoskeletons controllers are close to lower limit of this human bandwidth threshold under no-load conditions, and so can interfere in dynamic tasks.
 Note that these bandwidth issues are primarily a function of the exoskeleton hardware, and thus special care/analysis needs to be given/conducted to make sure that the hardware is capable of physically producing required behaviors.
@@
 Lastly, the importance of producing hardware that is physically capable of achieving dynamic performance is fundamentally linked to how mass is distributed throughout the system.
-In general, this means that minimizing mass, more specifically distal masses on extremities, is desirable, which in turn means that at certain degrees of freedom must remain unactuated.
-Underactuted control is a complicated problem which can either be ignored, in the sense that the control algorithm assumes the user will provide the necessary forces and torques for the unactuated degrees of freedom, dealt with using optimization-based control methods, or can be partially compensated using passive energy storage and dissipation elements.
-Our recommendation is that a combination of biomechanically-inspired passive elements and optimization-based control methods will ultimately produce the best strategy for dealing with underactuation in higly-dynamic exoskeleton systems.
+In general, it is desirable to minimize mass, particularly in distal regions.
+The design objective is often achieved by focusing actuation and leaving certain degrees of freedom passive (leading to an underactued system).
+Underactuted control is a complicated problem that is often ignored in the sense that controllers assume the operator will provide necessary forces and torques for the un-actuated degrees of freedom.
+Though it is possible to deal with the underactuated control problem directly (e.g., optimization-based controllers), there is a balance between feedback and feedforward control.
+% The better one is, the less reliant the system is on the other.  
+By developing better sensors and feedback controllers that directly measure and shadow operator intent, exoskeleton controllers will become more robust and less sensitive to errors due to underactuated system assumptions and dynamic model inaccuracies.
+
+%
+%, or can be partially compensated using passive energy storage and dissipation elements.  
+% we recommend a combination of biomechanically-inspired passive elements and optimization-based control methods, as these should ultimately produce the best strategy for dealing with underactuation in highly-dynamic exoskeleton systems.
 
@@
 % Finally, we note that 
-%Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
-%These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.
+% Many of the common model-based control methods in this survey employ feedback linearization and inverse dynamics strategies for dynamic exoskeleton control.  
+% These methods (incorrectly) assume the exoskeleton is fully actuated and require users to provide certain torques at uncontrolled joints.
 There are more advanced control schemes that extend to underactuated systems and may apply here (e.g., optimization-based controllers or virtual model control).
@@
 %
-%Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control.
+% Improved feedback and direct sensing of user intent / human interaction forces therefore also offers to reduce sensitivity to high accuracy models and the fully actuated assumptions in inverse dynamics control.
 

### 2015-09-30 control.tex
more updates

@@
 
-What to do about this to achieve robust online optimalization-based control is
+What to do about this to achieve robust online optimization-based control is
 currently a research issue. In actual implementations, skilled engineers manually
@@
 action to formulate a unified control design framework called
-Integral Admittance [torque to angle] Shaping, which designs exoskeleton con-
-trollers capable of producing the desired dynamic response
+Integral Admittance [torque to angle] Shaping, which designs exoskeleton 
+controllers capable of producing the desired dynamic response
 for the assisted limb. In this framework, a virtual increase

### 2015-10-01 exos/bleex.tex
more updates

@@
 
-The BLEEX team has successfully implemented both sensitivity amplification \cite{sesitivityAmpPaper2005} and a hybrid assitive control scheme \cite{bleex_hybrid_control_2006} that switches (based on gait phase) between sensitivity amplification and a position control regulating desired torque.
-This section focuses on sensitivity amplification, as the hybrid assitive strategy did not perform as well in walking trials.
+The BLEEX team has successfully implemented both sensitivity amplification \cite{sesitivityAmpPaper2005} and a hybrid assistive control scheme \cite{bleex_hybrid_control_2006} that switches (based on gait phase) between sensitivity amplification and a position control regulating desired torque.
+This section focuses on sensitivity amplification, as the hybrid assistive strategy did not perform as well in walking trials.
 
@@
 They highlight difficulties in properly outfitting humans with necessary sensing equipment and challenges in modeling the interaction between the human and exoskeleton (e.g., non-rigid, non-fixed contact points).
-Instead, the team develops controllers based on an inverse dynamic model (in the saggital plane) of only the exoskeleton.
+Instead, the team develops controllers based on an inverse dynamic model (in the sagittal plane) of only the exoskeleton.
 
@@
 The BLEEX sensitivity amplification control strategy models the torque applied by the human pilot on the exoskeleton as ${\bf d}$ (assuming no outside disturbances).
-Neglecting gravity, the exokeleton angular velocity is modeled as
+Neglecting gravity, the exoskeleton angular velocity is modeled as
 \[\vv = \mG \vr + \mS {\bf d} ,\] 
-where $\mG$ is the transfer function from actuator inputs ($\mG$ is the exoskeleton's dynamics), $\vr$ is the actuator input, and $\mS$ is the sensitivity or transfer function from human torque to exokeleton angular velocity.
+where $\mG$ is the transfer function from actuator inputs ($\mG$ is the exoskeleton's dynamics), $\vr$ is the actuator input, and $\mS$ is the sensitivity or transfer function from human torque to exoskeleton angular velocity.
 

### 2015-10-01 exos/exosText_MIT.tex
more updates

@@
  The interface between the suit and operator consists of shoulder straps, a compliant waist belt, thigh cuffs, and a shoe connection.
-The MIT exoskeleton employs a custom-built sensor that measures the interaction force between the exoskeleton and the operator?s thigh.
+The MIT exoskeleton employs a custom-built sensor that measures the interaction force between the exoskeleton and the operator's thigh.
 The sensor consisted of a spring pack, where the deflection of the passive springs was measured with a spring-loaded linear potentiometer.
@@
  The exoskeleton implements a {\bf variable damper knee mechanism} in the flexion/extension degree of freedom using a magnetorheological damper.
-The damper can exert a {\bf maximum braking torque of 60 Nm and consums on average 1W of electrical power}.
+The damper can exert a {\bf maximum braking torque of 60 Nm and consumes on average 1W of electrical power}.
 Additionally, a linear spring at the ankles captures negative energy during dorsilflexion.
@@
 The suit uses a 48V battery pack and is instrumented with \textbf{rotary potentiometers} at the hip and knee.
-Additionally, \textbf{strain gauges} were placed on the structure of the exoskeleton shank to measured sagittal bending moment and compression force.
+Additionally, \textbf{strain gages} were placed on the structure of the exoskeleton shank to measured sagittal bending moment and compression force.
 
@@
 
- The knee controller in the MIT exoskeleton effectively functions independently from he hip controller.
+ The knee controller in the MIT exoskeleton effectively functions independently from the hip controller.
 During knee strike, the variable damper in the knee exerts a torque proportional to the rotational velocity of the knee joint.
@@
 Although, these drives will again be limited by bandwidth issues in highly-dynamic maneuvers.
-The use of a strain gauge in the shank was also a novel means of measuring ground reaction forces.
+The use of a strain gage in the shank was also a novel means of measuring ground reaction forces.
 This design may offer benefits in terms of repeatability over different operators, as foot fall patterns may include significant variations for different users or even in the same user in different environments.

### 2015-10-01 exos/exosText_bodyExt.tex
more updates

@@
  Except for forearm pronosupination, each of the actuators in the Body Extender suit are driven by a linear actuator that is composed of an electrical motor, incremental encoder, ball screw, and angular contact ball bearings.
-The motors are frameless DC torque controlled motors.
+The motors are frame-less DC torque controlled motors.
 The {\bf peak torque of the motors is 7 Nm, with a maximum continuous torque of 6 Nm}.
@@
 
-The feedback linearizing controller in \cite{} operates under two fundamental assumptions and only appears to relate directly to the case of a single manipulator arm.
+The feedback linearizing controller in 
+%\cite{}
+operates under two fundamental assumptions and only appears to relate directly to the case of a single manipulator arm.
 The first assumption of the controller is that the mass and inertial properties of an object being manipulated are estimated online.
@@
 The Body Extender suit is one of the only fully-actuated upper and lower extremity exoskeletons available today.
-Clearly, for the task of handling extremely heavy objects, a fully actuated suit may be the best option, but, generally speaking, having the addition weight due to actuating every degree of freedom in the suit is not ideal when considering dynamic operation.
+Clearly, for the task of handling extremely heavy objects, a fully actuated suit may be the best option, but, generally speaking, having the additional weight due to actuating every degree of freedom in the suit is not ideal when considering dynamic operation.
 
@@
 The stated strategy of decoupling the individual limbs from each other seems somewhat feasible, but is limited relative to coordinated limb control.
+
+The speed of the actuators (approximately 60 deg/s) is far to slow for
+dynamic human movement.
 

### 2015-10-01 exos/exosText_ihmc.tex
more updates

@@
 
-Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the device is also provides a ``performance enhancement" control mode, which is within the scope of this document.
-The main design principle highlighted in the IHMC MAE hardware (a recurring theme throughout exoskeleton literature) is that directly measuring user intent can be difficult for a number of reasons.
+Although the IHMC MAE was primarily designed as a disabled assist exoskeleton, the device also provides a ``performance enhancement" control mode, which is within the scope of this document.
+The main design principle highlighted in the IHMC MAE hardware (a recurring theme throughout the exoskeleton literature) is that directly measuring user intent can be difficult for a number of reasons.
 The IHMC designers resolve the issue using a new control and feedback system, which for the IHMC MAE, is primarily based on a unique actuation scheme.
@@
 \includegraphics[width=3.in]{exos/figs/ihmc/seaAssm}
-  \caption{Schematic of the RSEA elastic mechanism design, shown in the zero force position.
+  \caption{Schematic of the SEA elastic mechanism design, shown in the zero force position.
 The encoder reads a scale tape on the curved cylindrical surface.

### 2015-10-01 exos/hal.tex
more updates

@@
 Instead, HAL adjusts hip, knee and ankle torques to amplify its operator's torque.
-The exoskelton's sensory system includes bioelectric sensing (including \textbf{sEMG}), \textbf{angular sensors}, \textbf{acceleration sensors}, and \textbf{center of pressure / center of gravity} (COP/COG) sensors.
+The exoskeleton's sensory system includes bioelectric sensing (including \textbf{sEMG}), \textbf{angular sensors}, \textbf{acceleration sensors}, and \textbf{center of pressure / center of gravity} (COP/COG) sensors.
 COP/COG sensing is provided through shoes with ground reaction force sensors.
@@
 A relatively autonomous torque assist strategy in \cite{HALmodelControlKnee2010} recognizes user intention to take a step by thresholding sEMG data.
-The approach provides a knee torque response including an assitive torque component, a viscous damping component that reduces high velocity motion for safety, and a gravity compensation torque.
+The approach provides a knee torque response including an assistive torque component, a viscous damping component that reduces high velocity motion for safety, and a gravity compensation torque.
 %In \cite{HALvTorqueImp2002}, HAL uses myoelectric data to estimate muscle torque based on the difference between flexor and extensor muscle activities.

### 2015-10-01 exos/xor.tex
more updates

@@
 The active joints are powered by \textbf{hybrid actuators} comprised of an \textbf{air muscle} and an \textbf{electric motor}.
-The hybrid actuation scheme uses a unilateral air muscle layout to compensate for gravity and bilateral electric motors with relatively small gear ratio (57.5) to serve as the dynamic compensator.
+The hybrid actuation scheme uses a unilateral air muscle layout to compensate for gravity and bilateral electric motors with a relatively small gear ratio (57.5) to serve as the dynamic compensator.
 
@@
 The hybrid drive system sums the two to develop a desired torque profile that achieves the benefits of both and reduces weight and motor size.
-The system offers high torque control with negligable stick-slip and backdrivability.
+The system offers high torque control with negligible stick-slip and backdrivability.
 
@@
 It sets desired joint angles and velocities, ${\bf x} = ({\bf \theta}, \dot{\bf \theta})$, from the EMG data, ${\bf u} = (EMG_1, \dots , EMG_n)$, using a linear prediction model, 
-\[{\bf x}(k+1) = \mA{\bf x}(k)+\mB{\bf u}(k),\].  
+\[{\bf x}(k+1) = \mA{\bf x}(k)+\mB{\bf u}(k),\]  
 The controller uses the angular measurements required to perform CCA and EMG data to derive the model's $\mA$ and $\mB$ terms.
@@
 Without the EMG data, the mean squared error increased to $2.1$.
-Additional experiments confirm the air muscles can effectively perform gravity compensation when used in conjunction with electric actuators (to correct for torque errors induced by inaccuracies in the model mapping position to toque in air muscles).
+Additional experiments confirm the air muscles can effectively perform gravity compensation when used in conjunction with electric actuators (to correct for torque errors induced by inaccuracies in the model mapping position to torque in air muscles).
 

### 2015-10-01 survey.tex
more updates

@@
 Algorithms\\
-(Draft 3.2)}
+(Draft 4.0)}
 
@@
 \item
-{\bf XoR:} a hybrid Pneumatic/Electric-motor system, emphasizing the
-value of using multiple actuator types.
+{\bf XoR:} a hybrid pneumatic/electric-motor system, emphasizing the
+value of using multiple actuator types. It also combines the use
+of brain signals and EMG for estimating operator intent.
 \item
@@
 
-To this end, the hardware systems that we chose to include summaries of in this document were the BLEEX, HAL, XoR, Body Extender, Hydraulic Lower Extremety, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.
+To this end, the hardware systems that we chose to include summaries of in this document were the BLEEX, HAL, XoR, Body Extender, Hydraulic Lower Extremity, IHMC Mobility Assist, MIT, and RoboKnee exoskeletons.
 The different systems cover many of the possible choices of actuation strategies, ranging from electric motors, to hydraulics, to pneumatics, and combinations thereof.
@@
 For instance, EMG data is notoriously noisy and presents both modeling and calibration challenges.
-It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or combat scenarios.
-For similar reasons, many systems avoid the use of load cells or complicated sensor networks to accurately obtain operator-robot interaction forces / torques.
+It would also be extremely difficult to keep these sensors in place to obtain accurate readings in dynamic environments or activities.
+For similar reasons, many systems avoid the use of load cells or complicated sensor networks to accurately obtain operator-robot interaction forces/torques.
 
@@
 Such an approach may prove useful in a hybrid strategy, filtering external disturbance forces during sensitivity amplification control.
-% As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has lead to the development of new, flexible ``robotic skin'' sensing arrays.
-Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure / force and exoskeleton contact locations), which was not feasible in the recent past.
+% As a longer term solution, we highlight research in the area of MEMs and nano-fabrication, which has led to the development of new, flexible ``robotic skin'' sensing arrays.
+Integrating these sensing arrays into operator clothing (or attaching directly to the exoskeleton), it may be possible to directly measure human interaction forces (i.e., estimating pressure/force and exoskeleton contact locations), which was not feasible in the recent past.
 Dense sensing arrays add redundancy and can improve measurement reliability.
@@
 In general, it is desirable to minimize mass, particularly in distal regions.
-The design objective is often achieved by focusing actuation and leaving certain degrees of freedom passive (leading to an underactued system).
-Underactuted control is a complicated problem that is often ignored in the sense that controllers assume the operator will provide necessary forces and torques for the un-actuated degrees of freedom.
+The design objective is often achieved by focusing actuation and leaving certain degrees of freedom passive (leading to an underactuated system).
+Underactuated control is a complicated problem that is often ignored in the sense that controllers assume the operator will provide necessary forces and torques for the un-actuated degrees of freedom.
 Though it is possible to deal with the underactuated control problem directly (e.g., optimization-based controllers), there is a balance between feedback and feedforward control.

### 2015-10-01 control.tex
more updates

@@
 %\usepackage{stfloats}
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
 \usepackage{hyperref}
@@
 \item
-Valve control (for fluidic systems):
+Valve control (for fluidic systems)
 \begin{equation}
@@
 + \mK_{\dot{\vtau}} ( \dot{\vtau} - \dot{\vtau}_d ) + \vu_{\ff}
+\label{eq:valve}
 \end{equation}
@@
 a third order model of actuator plus rigid body dynamics.
+
 If electric motor and power amplifier dynamics are taken into account in electric
@@
 an electrical current servo in an electric system.
+
+Often this type of control is referred to as a ``torque control inner loop''
+with a ``position/velocity control outer loop''. In this approach
+the inner and outer loops are combined in one equation (eq.~\ref{eq:valve}),
+and there is no
+longer ``inner'' and ``outer'' control quantities. All feedback is
+treated the same way.
 \item
@@
 
-\subsection{Behavior Selection}
+\subsection{``Sentence'' Control: Behavior Selection}
 
@@
 To maintain compatibility in our joint representation, we will similarly
-map operator ``muscle commands'' $\myo$ to operator joint torques $\vtau_{\myo}$.
+map operator ``muscle commands'' $\vu_{\myo}$ to operator joint torques $\vtau_{\myo}$.
 \begin{equation}
@@
 sensitivity amplification control.
-This section is quotes from:\\
+This section quotes from:\\
 \url{http://ieeexplore.ieee.org/xpl/login.jsp?tp=&arnumber=7139989}
 
-{\it [Exoskeletons] have the implicit property
+{\it ``[Exoskeletons] have the implicit property
 of causing a virtual modification of the dynamic response of
@@
 
-{\it ... the impedance of the coupled human-exoskeleton
+{\it ``... the impedance of the coupled human-exoskeleton
 system needs to be reduced below that of the unassisted
@@
 {\it
-... Inertia compensation is more complex ...
+``... Inertia compensation is more complex ...
 It can be shown that using only positive acceleration
@@
 This work uses filtered acceleration
-feedback with a second-order low-pass Butterworth filter. ...
+feedback with a second-order low-pass Butterworth filter. ...''
 }
@@
 
-{\it The robot utilized the dual-mode control scheme, which is
+{\it ``The robot utilized the dual-mode control scheme, which is
 comprised of the active control for the stance phase and the
@@
 walking by predicting the change from the swing to the stance
-in advance.}\\
+in advance.''}\\
 %~\cite{IEEE07222598}
@@
 has been shown to be an effective method to generate the
-locomotion for an exoskeleton robot.[3],[4],[21]
+locomotion for an exoskeleton robot.
 \begin{verbatim}
-H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
-the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
-Conf. Robot. Autom., pp. 4364-4371, 2005.
-Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
-for Human Performance Amplification,” University of California,2003.
-582
 Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
@@
 sudden changes(command jump) at the phase transitions, a
-smoothing method [5],[19] is introduced and a pre-transition
+smoothing method is introduced and a pre-transition
 method is used to solve the swing delay due to the internal
@@
 \begin{verbatim}
-H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
-Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
-Robotics Research, pp. 561-573, 2006.
 oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
